---
title: "Canine Skin Microbiome: ITS Microbial Diversity Analysis "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

**Purpose:**
Investigate microbial diversity across skin, air and negative control samples.

## Load R packages
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(knitr)
library(FSA)
library(tidyverse)
library(vegan)
library(phyloseq)
library(ggplot2)
library(ggtree)
library(scales)  
library(RColorBrewer)
library(segmented)
library(stringr)
library(pairwiseAdonis)
library(ggpubr)
library(changepoint)
library(dplyr)
library(lme4)
library(emmeans)
library(nortest)
library(mcp)
library(purrr)
library(kableExtra)
library(forcats)
 
```

## Create phyloseq objects
```{r, message=FALSE, warning=FALSE, echo=TRUE}
load("../01_input_files/ASV_tables/decontam_allseqruns_filtered_canine_skin_microbiome.rda")
ps <- phyloseq(
  otu_table(ASV_feature_table_decontam, taxa_are_rows = FALSE),
  sample_data(sample_metadata_decontam),
  tax_table(taxa_annotations_decontam)
)

new_ids <- paste0("ASV_", seq_len(ntaxa(ps)))

tax <- as(tax_table(ps), "matrix")
rownames(tax) <- new_ids
tax <- cbind(
  tax,
  genus_species = paste(tax[, "Genus"], tax[, "Species"], sep = " ")
)

otu <- as(otu_table(ps), "matrix")
colnames(otu) <- new_ids

ps <- phyloseq(
  otu_table(otu, taxa_are_rows = FALSE),
  sample_data(ps),
  tax_table(tax)
)


# Get reads and metadata per sample
reads_per_sample <- sample_sums(ps)
sample_data <- as(sample_data(ps), "data.frame")
 
sample_names_ps <- sample_names(ps)

# Ensure sample order matches
if (!all(sample_names_ps == rownames(sample_data))) {
  sample_data <- sample_data[sample_names_ps, ]
}

# Create dataframe with sample-level stats
df <- data.frame(
  sample = sample_names_ps,
  total_reads = reads_per_sample,
  Sample_Plate = sample_data$Sample_Plate,
  Individual = sample_data$Individual,
  stringsAsFactors = FALSE
)

mean_reads <- mean(df$total_reads)
min_reads  <- min(df$total_reads)
max_reads  <- max(df$total_reads)

mean_reads

# Calculate total reads and ASVs per sample and add to metadata
otu <- otu_table(ps)
if (taxa_are_rows(otu)) {
  total_asvs <- colSums(otu > 0)
} else {
  total_asvs <- rowSums(otu > 0)
}

total_reads <- sample_sums(ps)
 
sample_data$total_asv <- total_asvs
sample_data$total_reads <- total_reads

sample_data(ps) <- sample_data(sample_data)

sample_data <- sample_data %>%
  arrange(Surgery_Type, Individual) %>%
  mutate(Individual = factor(Individual, levels = unique(Individual)))

my_cols <- colorRampPalette(brewer.pal(8, "Dark2"))(20)

# Aggregate reads and ASVs per Individual 
by_ind <- aggregate(
  cbind(total_reads, total_asv) ~ Individual + Sample_Plate,
  data = sample_data,
  FUN = sum
)

print(by_ind)
```

# Negative Controls Analysis
```{r, message=FALSE, warning=FALSE, echo=TRUE}

ps_negatives <- prune_samples(
  sample_data(ps)$Location_Type %in% c("Air", "Blank", "Chlorhex"),
  ps
)

ps_negatives <- prune_taxa(taxa_sums(ps_negatives) > 0, ps_negatives)

# remove ASVs uncharacterized phyla
ps_negatives_1 <- subset_taxa(
  ps_negatives,
  !is.na(Phylum) & !Phylum %in% c("", "uncharacterized")
)

 
unknown_asvs <- rownames(tax_table(ps_negatives))[
  is.na(tax_table(ps_negatives)[, "Phylum"]) |
    tax_table(ps_negatives)[, "Phylum"] %in% c("", "uncharacterized")
]

unknown_count <- length(unknown_asvs)
total_asvs <- nrow(tax_table(ps_negatives))
unknown_percent <- (unknown_count / total_asvs) * 100
 
cat("ASVs with uncharacterized phyla:", unknown_count, "\n")
cat("% ASVs with uncharacterized phyla:", unknown_percent, "\n")

# remove ASVs uncharacterized phyla
ps_negatives_1 <- subset_taxa(
  ps_negatives,
  !is.na(Phylum) & !Phylum %in% c("", "uncharacterized")
)


 
# ASVs removed
n_asv_before <- ntaxa(ps_negatives)
n_asv_after  <- ntaxa(ps_negatives_1)
n_asv_removed <- n_asv_before - n_asv_after
pct_asv_removed <- 100 * n_asv_removed / n_asv_before

# Reads removed
reads_before <- sum(otu_table(ps_negatives))
reads_after  <- sum(otu_table(ps_negatives_1))
reads_removed <- reads_before - reads_after
pct_reads_removed <- 100 * reads_removed / reads_before


data.frame(
  metric = c("ASVs", "Reads"),
  before = c(n_asv_before, reads_before),
  after = c(n_asv_after, reads_after),
  removed = c(n_asv_removed, reads_removed),
  percent_removed = c(pct_asv_removed, pct_reads_removed)
)

low_depth_samples <- sample_sums(ps_negatives_1)[sample_sums(ps_negatives_1) < 2000]

low_depth_samples

# Remove low-depth samples
ps_negatives_2 <- prune_samples(sample_sums(ps_negatives_1) >= 2000, ps_negatives_1)

summary(sample_sums(ps_negatives_2))

zero_samples <- sample_names(ps_negatives_2)[sample_sums(ps_negatives_2) == 0]

zero_samples

# Remove samples with all 0s (if applicable)
ps_negatives_2 <- prune_samples(sample_sums(ps_negatives_2) > 0, ps_negatives_2)

 

# Normalize read counts by dividing by total read count per sample (relative abundance) 
ps_negatives_rel_2 <- transform_sample_counts(ps_negatives_2, function(x) x / sum(x))


 
```

## Top 10 ASVs in Negative Samples
```{r, message=FALSE, warning=FALSE, echo=TRUE}
otu <- otu_table(ps_negatives_rel_2)
if (!taxa_are_rows(ps_negatives_rel_2)) {
  otu <- t(otu)
}


# average relative abundances ASVs
asv_average_rel_abundance <- rowMeans(otu, na.rm = TRUE)

Sample_Location <- as.data.frame(sample_data(ps_negatives_rel_2))$Sample_Location
 

asv_in_kit <- apply(otu, 1, function(x) sum(x[Sample_Location == "Blank"] > 0, na.rm = TRUE))
asv_in_air <- apply(otu, 1, function(x) sum(x[Sample_Location == "Air"] > 0, na.rm = TRUE))
asv_in_chlorhex <- apply(otu, 1, function(x) sum(x[Sample_Location == "Chlorhex"] > 0, na.rm = TRUE))

ord <- order(asv_average_rel_abundance, decreasing = TRUE)
tax <- as.data.frame(tax)
tax$ASV <- rownames(tax)

asvs_negative_samples <- tibble(
  ASV = names(asv_average_rel_abundance)[ord],
  average_rel_abundance = asv_average_rel_abundance[ord],
  asv_in_kit = asv_in_kit[ord],
  asv_in_air = asv_in_air[ord],
  asv_in_chlorhex = asv_in_chlorhex[ord]
) %>%
  left_join(tax, by = "ASV") %>%
  dplyr::select(
    ASV, average_rel_abundance, asv_in_kit, asv_in_air, asv_in_chlorhex,
    Phylum, Family, Order, Genus, Species, genus_species
  )

asvs_negative_samples

top10_asvs <- asvs_negative_samples %>%
  dplyr::slice_head(n = 10)   

top10_asvs

# total ASVs detected in each negative-control type
total_asvs_kit <- sum(asv_in_kit > 0, na.rm = TRUE)
total_asvs_air <- sum(asv_in_air > 0, na.rm = TRUE)
total_asvs_chlorhex <- sum(asv_in_chlorhex > 0, na.rm = TRUE)
 
cat("Total ASV count in kit samples:", total_asvs_kit, "\n")
cat("Total ASV count in air samples:", total_asvs_air, "\n")
cat("Total ASV count in chlorhex samples:", total_asvs_chlorhex, "\n")

```

## Top 10 Phyla in Negative Samples
```{r, message=FALSE, warning=FALSE, echo=TRUE}
ps_negatives_rel_2_df <- psmelt(ps_negatives_rel_2) 

tax <- as(tax_table(ps_negatives_rel_2), "matrix")
family_names <- unique(na.omit(tax[, "Phylum"]))

cat("Number of unique Phylum names:", length(family_names), "\n")

 top10_phyla <- ps_negatives_rel_2_df %>%
  dplyr::group_by(Sample, Phylum) %>%
  # add up the ASV relative abundances per Phylum in each sample
  dplyr::summarise(total_abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
   dplyr::group_by(Phylum) %>%
   # get the mean relative abundance across all samples per Phylum
  dplyr::summarise(mean_rel_abundance = mean(total_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_rel_abundance)) %>%
  dplyr::slice_head(n = 10) %>%
  dplyr::pull(Phylum)
 
top10_phyla

legend_order <- c(
  "p__Ascomycota",
  "p__Basidiomycota",
  "p__Anthophyta",
  "p__Eukaryota_phy_Incertae_sedis",
  "p__Mortierellomycota",
  "p__Mucoromycota",
  "p__Chytridiomycota"
)

phylum_colors <- c(
  "p__Ascomycota"                   = "#1b9e77",
  "p__Basidiomycota"                = "#a6761d",
  "p__Anthophyta"                   = "#7570b3",
  "p__Eukaryota_phy_Incertae_sedis" = "yellow",
  "p__Mortierellomycota"            = "#d95f02",
  "p__Mucoromycota"                 = "#e7298a",
  "p__Chytridiomycota"              = "#66a61e"
)

 
phy_df_2 <- ps_negatives_rel_2_df %>%
  dplyr::group_by(Sample_Site, Sample, Phylum) %>%
  dplyr::summarise(total_abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
   dplyr::group_by(Sample_Site, Phylum) %>%
  dplyr::summarise(mean_rel_abundance = mean(total_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_rel_abundance)) 

phy_df_2 <- phy_df_2[phy_df_2$Phylum %in% top10_phyla, ]

phy_df_2$Phylum <- factor(phy_df_2$Phylum, levels = legend_order)

ggplot(phy_df_2, aes(x = Sample_Site, y = mean_rel_abundance, fill = Phylum, group = Phylum)) +
  ggtitle("Top 10 Phyla in Negative Samples") +
  geom_col(position = "stack") +
  scale_fill_manual(values = phylum_colors, breaks = legend_order, drop = FALSE) +
  coord_cartesian(ylim = c(0, 1.0)) +
  labs(x = "Sample Type", y = "Average relative abundance across samples", fill = "Phylum") +
  
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```
 
```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
## Family 
phy_df_1_negatives <- psmelt(ps_negatives_rel_1) 

tax <- as(tax_table(ps_negatives_rel_1), "matrix")
family_names <- unique(na.omit(tax[, "Family"]))

cat("Number of unique Family names:", length(family_names), "\n")

 top10_phyla <- phy_df_1_negatives %>%
  dplyr::mutate(Family = dplyr::if_else(is.na(Family), "Unknown", Family)) %>%
  dplyr::group_by(Sample, Family) %>%
  # add up the ASV relative abundances per Family in each sample
  dplyr::summarise(total_abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
   dplyr::group_by(Family) %>%
   # get the mean relative abundance across all samples per Family
  dplyr::summarise(mean_rel_abundance = mean(total_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_rel_abundance)) %>%
  dplyr::slice_head(n = 10) %>%
  dplyr::pull(Family)
 
legend_order <- c(
  "Staphylococcaceae",
  "Streptococcaceae",
  "Corynebacteriaceae",
  "Lactobacillaceae",
  "Moraxellaceae",
  "Neisseriaceae",
  "Pseudomonadaceae",
  "Enterobacteriaceae",
  "Lysobacteraceae",
  "Unknown"
)

phylum_colors <- c(
  "Unknown" = "gray",
  "Staphylococcaceae" = "#e7298a",
  "Streptococcaceae" = "blue",
  "Neisseriaceae" = "lightblue",
  "Moraxellaceae" = "#a6761d",
  "Corynebacteriaceae" = "yellow",
  "Lactobacillaceae" = "darkgreen",
  "Pseudomonadaceae" = "green",
   "Enterobacteriaceae" = "purple",
  "Lysobacteraceae" = "red",
  "Other" = "orange"
)

phy_df_2 <- phy_df_1_negatives %>%
  dplyr::mutate(Family = dplyr::if_else(is.na(Family), "Unknown", Family)) %>%
  dplyr::group_by(Sample_Site, Sample, Family) %>%
  dplyr::summarise(total_abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::group_by(Sample_Site, Family) %>%
  dplyr::summarise(mean_rel_abundance = mean(total_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_rel_abundance))

phy_df_2 <- phy_df_2 %>%
  dplyr::filter(Family %in% top10_phyla)

phy_df_2$Family <- factor(phy_df_2$Family, levels = legend_order)

ggplot(phy_df_2, aes(x = Sample_Site, y = mean_rel_abundance, fill = Family)) +
  ggtitle("Top 10 Families in Negative Samples") +
  geom_col(position = "stack") +
  scale_fill_manual(values = phylum_colors, breaks = legend_order, drop = FALSE) +
  coord_cartesian(ylim = c(0, 1.0)) +
  labs(x = "Sample Type", y = "Average relative abundance across samples", fill = "Family") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Species in Negative Samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE} 
tax <- as(tax_table(ps_negatives_rel_2), "matrix")

no_species <- is.na(tax[, "Species"]) | tax[, "Species"] == "" | tax[, "Species"] == "NA"

count_no_species <- sum(no_species, na.rm = TRUE)
total_asvs <- nrow(tax)
pct_no_species <- 100 * count_no_species / total_asvs

cat("ASVs with no species assignment:", count_no_species, "\n")
cat("Percentage with no species assignment:", round(pct_no_species, 2), "%\n")

```

 
# Bioinformatics Analysis

## Taxa-Based Filtering 
### Remove uncharacterized Phyla
```{r, message=FALSE, warning=FALSE, echo=TRUE}
ps_positives <- prune_samples(
  grepl("Skin", sample_data(ps)$Location_Type),
  ps
)

otu_matrix <- as(otu_table(ps_positives), "matrix")
tax <- as(tax_table(ps_positives), "matrix")

unknown_asvs <- rownames(tax)[
  is.na(tax[, "Phylum"]) | tax[, "Phylum"] %in% c("", "uncharacterized")
]

ps_1 <- prune_taxa(!taxa_names(ps_positives) %in% unknown_asvs, ps_positives)

 
# ASVs removed
n_asv_before <- ntaxa(ps_positives)
n_asv_after  <- ntaxa(ps_1)
n_asv_removed <- n_asv_before - n_asv_after
pct_asv_removed <- 100 * n_asv_removed / n_asv_before

# Reads removed
reads_before <- sum(otu_table(ps_positives))
reads_after  <- sum(otu_table(ps_1))
reads_removed <- reads_before - reads_after
pct_reads_removed <- 100 * reads_removed / reads_before


data.frame(
  metric = c("ASVs", "Reads"),
  before = c(n_asv_before, reads_before),
  after = c(n_asv_after, reads_after),
  removed = c(n_asv_removed, reads_removed),
  percent_removed = c(pct_asv_removed, pct_reads_removed)
)

low_depth_samples <- sample_sums(ps_1)[sample_sums(ps_1) < 2000]

low_depth_samples

# Remove low-depth samples
ps_1 <- prune_samples(sample_sums(ps_1) >= 2000, ps_1)

summary(sample_sums(ps_1))

zero_samples <- sample_names(ps_1)[sample_sums(ps_1) == 0]

zero_samples

length(zero_samples)

# Remove samples with all 0s (if applicable)
ps_1 <- prune_samples(sample_sums(ps_1) > 0, ps_1)
 
 
out_dir <- "../01_input_files/phyloseq"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
saveRDS(ps_1, file = file.path(out_dir, "ps_1.rds"))
 
```

### Check: Low-prevalence taxa
No phyla were removed. 
```{r, fig.width=11, fig.height=8, echo=TRUE}
# Compute prevalence for each ASV 
prevdf = apply(X = otu_table(ps_1),
               MARGIN = ifelse(taxa_are_rows(ps_1), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})
 
# Add taxa and total read counts to prevalence df
prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps_1),
                    tax_table(ps_1))

# Are there phyla that are comprised of mostly low-prevalence features? 
# Compute per Phylum: Average number of ASVs and total number of ASVs across samples
phylum_summary <- plyr::ddply(prevdf, "Phylum", function(df1){
  cbind(mean(df1$Prevalence), sum(df1$Prevalence))
})
colnames(phylum_summary) <- c("Phylum", "Mean_Prevalence", "Sum_Prevalence")

# Phyla by Total ASV Prevalence 
phylum_summary_sorted <- phylum_summary[order(-phylum_summary$Sum_Prevalence), ]
print(phylum_summary_sorted, row.names = FALSE)
ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(ps_1),color=Phylum)) +
  ggtitle("Prevalence of Phyla Across Skin-Swab Samples", subtitle="dashed line = 5% prevalence") +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none") 
  
```

## Check: Sequencing Run Batch Effect

### PERMANOVA Test 

**Notes:**

* Batch effects were estimated using PERMANOVA analysis when both batch and Dog_ID were included
 
 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Normalization using relative abundances
ps_rel_1 <- transform_sample_counts(ps_1, function(x) x / sum(x))

individual_list <- unique(sample_data(ps_rel_1)$Individual)

bray_dist <- phyloseq::distance(ps_rel_1, method = "bray")

set.seed(100)
bray_pcoa <- ordinate(ps_rel_1, method = "PCoA", distance =  bray_dist)
p1 <- plot_ordination(ps_rel_1, bray_pcoa, color = "Sample_Plate")


p1 + ggtitle("PCoA of Bray-Curtis Dissimilarities", subtitle = "Ellipses encircle and color samples belonging to the same sequencing run.") +
  geom_point(size = 1.5) + 
  theme_bw() +
  stat_ellipse(aes(fill = Sample_Plate, group = Sample_Plate), alpha = 0.2, geom = "polygon")


bray_matrix <- as.matrix(bray_dist)

metadata <- as(sample_data(ps_rel_1), "data.frame")
 
# PERMANOVA test
set.seed(100)
adonis2(bray_matrix ~ Sample_Plate + Dog_ID, data = metadata, permutations = 999, by = "terms")  

```

 
 
 
# Skin samples Analysis
```{r, message=FALSE, warning=FALSE, echo=TRUE}
 unknown_asvs <- rownames(tax_table(ps_rel_1))[
  is.na(tax_table(ps_rel_1)[, "Phylum"]) |
    tax_table(ps_rel_1)[, "Phylum"] %in% c("", "uncharacterized")
]

unknown_count <- length(unknown_asvs)
total_asvs <- nrow(tax_table(ps_rel_1))
unknown_percent <- (unknown_count / total_asvs) * 100
 
cat("ASVs with uncharacterized phyla:", unknown_count, "\n")
cat("% ASVs with uncharacterized phyla:", unknown_percent, "\n")

# remove ASVs uncharacterized phyla
ps_rel_1 <- subset_taxa(
  ps_rel_1,
  !is.na(Phylum) & !Phylum %in% c("", "uncharacterized")
)

```

## Top 10 ASVs
```{r, message=FALSE, warning=FALSE, echo=TRUE}
otu <- otu_table(ps_rel_1)
if (!taxa_are_rows(ps_rel_1)) {
  otu <- t(otu)
}


# average relative abundances ASVs
asv_average_rel_abundance <- rowMeans(otu, na.rm = TRUE)

Sample_Location <- as.data.frame(sample_data(ps_rel_1))$Sample_Location
 

asv_in_stifle <- apply(otu, 1, function(x) sum(x[Sample_Location == "Stifle"] > 0, na.rm = TRUE))
asv_in_ab <- apply(otu, 1, function(x) sum(x[Sample_Location == "Abdomen"] > 0, na.rm = TRUE))


ord <- order(asv_average_rel_abundance, decreasing = TRUE)
tax <- as.data.frame(tax)
tax$ASV <- rownames(tax)

asvs_positive_samples <- tibble(
  ASV = names(asv_average_rel_abundance)[ord],
  average_rel_abundance = asv_average_rel_abundance[ord],
  asv_in_stifle = asv_in_stifle[ord],
  asv_in_ab = asv_in_ab[ord]
) %>%
  left_join(tax, by = "ASV") %>%
  dplyr::select(
    ASV, average_rel_abundance, asv_in_stifle, asv_in_ab,
    Phylum, Family, Order, Genus, Species, genus_species
  )

asvs_positive_samples

top10_asvs <- asvs_positive_samples %>%
  dplyr::slice_head(n = 10)  
 
top10_asvs

# total ASVs detected in each negative-control type
total_asv_in_stifle <- sum(asv_in_stifle > 0, na.rm = TRUE)
total_asv_in_ab <- sum(asv_in_ab > 0, na.rm = TRUE)
  
cat("Total ASV count in kit samples:", total_asv_in_stifle, "\n")
cat("Total ASV count in air samples:", total_asv_in_ab, "\n")
 
```

## Top 10 Phyla
```{r, message=FALSE, warning=FALSE, echo=TRUE}
phy_df_1 <- psmelt(ps_rel_1) 

tax <- as(tax_table(ps_rel_1), "matrix")
family_names <- unique(na.omit(tax[, "Phylum"]))

cat("Number of unique Phylum names:", length(family_names), "\n")

 top10_phyla <- phy_df_1 %>%
  dplyr::group_by(Sample, Phylum) %>%
  # add up the ASV relative abundances per Phylum in each sample
  dplyr::summarise(total_abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
   dplyr::group_by(Phylum) %>%
   # get the mean relative abundance across all samples per Phylum
  dplyr::summarise(mean_rel_abundance = mean(total_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_rel_abundance)) %>%
  dplyr::slice_head(n = 10) %>%
  dplyr::pull(Phylum)
 

legend_order <- c(
  "p__Ascomycota",
  "p__Basidiomycota",
  "p__Anthophyta",
  "p__Eukaryota_phy_Incertae_sedis",
  "p__Mortierellomycota",
  "p__Mucoromycota",
  "p__Chytridiomycota",
  "p__Rozellomycota"
)

phylum_colors <- c(
  "p__Ascomycota"                   = "#1b9e77",
  "p__Basidiomycota"                = "#a6761d",
  "p__Anthophyta"                   = "#7570b3",
  "p__Eukaryota_phy_Incertae_sedis" = "yellow",
  "p__Mortierellomycota"            = "#d95f02",
  "p__Mucoromycota"                 = "#e7298a",
  "p__Chytridiomycota"              = "#66a61e",
  "p__Rozellomycota"= "red"
)


 
phy_df_2 <- phy_df_1 %>%
  dplyr::group_by(Sample_Site, Sample, Phylum) %>%
  dplyr::summarise(total_abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
   dplyr::group_by(Sample_Site, Phylum) %>%
  dplyr::summarise(mean_rel_abundance = mean(total_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_rel_abundance)) 

phy_df_2 <- phy_df_2[phy_df_2$Phylum %in% top10_phyla, ]

phy_df_2$Phylum <- factor(phy_df_2$Phylum, levels = legend_order)

ggplot(phy_df_2, aes(x = Sample_Site, y = mean_rel_abundance, fill = Phylum, group = Phylum)) +
  ggtitle("Top 10 Phyla in Skin Samples") +
  geom_col(position = "stack") +
  scale_fill_manual(values = phylum_colors, breaks = legend_order, drop = FALSE) +
  coord_cartesian(ylim = c(0, 1.0)) +
  labs(x = "Sample Type", y = "Average relative abundance across samples", fill = "Phylum") +
  
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

 
```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
 ## Family 
tax <- as(tax_table(ps_rel_1), "matrix")
family_names <- unique(na.omit(tax[, "Family"]))

cat("Number of unique Family names:", length(family_names), "\n")

 top10_phyla <- phy_df_1 %>%
  dplyr::mutate(Family = dplyr::if_else(is.na(Family), "Unknown", Family)) %>%
  dplyr::group_by(Sample, Family) %>%
  # add up the ASV relative abundances per Family in each sample
  dplyr::summarise(total_abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
   dplyr::group_by(Family) %>%
   # get the mean relative abundance across all samples per Family
  dplyr::summarise(mean_rel_abundance = mean(total_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_rel_abundance)) %>%
  dplyr::slice_head(n = 10) %>%
  dplyr::pull(Family)
 
legend_order <- c(
  "Staphylococcaceae",
  "Unknown",
  "Streptococcaceae",
  "Neisseriaceae",
  "Moraxellaceae",
  "Pasteurellaceae",
  "Corynebacteriaceae",
  "Lactobacillaceae",
  "Lachnospiraceae",
  "Micrococcaceae"
)

phylum_colors <- c(
  "Unknown" = "gray",
  "Staphylococcaceae" = "#e7298a",
  "Streptococcaceae" = "blue",
  "Neisseriaceae" = "lightblue",
  "Moraxellaceae" = "#a6761d",
  "Corynebacteriaceae" = "yellow",
  "Lactobacillaceae" = "darkgreen",
  "Pasteurellaceae" = "orange",
   "Lachnospiraceae" = "brown",
  "Micrococcaceae" = "pink"
)

phy_df_2 <- phy_df_1 %>%
  dplyr::mutate(Family = dplyr::if_else(is.na(Family), "Unknown", Family)) %>%
  dplyr::group_by(Sample_Site, Sample, Family) %>%
  dplyr::summarise(total_abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::group_by(Sample_Site, Family) %>%
  dplyr::summarise(mean_rel_abundance = mean(total_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_rel_abundance))

phy_df_2 <- phy_df_2 %>%
  dplyr::filter(Family %in% top10_phyla)

phy_df_2$Family <- factor(phy_df_2$Family, levels = legend_order)

ggplot(phy_df_2, aes(x = Sample_Site, y = mean_rel_abundance, fill = Family)) +
  ggtitle("Top 10 Families in Skin Samples") +
  geom_col(position = "stack") +
  scale_fill_manual(values = phylum_colors, breaks = legend_order, drop = FALSE) +
  coord_cartesian(ylim = c(0, 1.0)) +
  labs(x = "Sample Type", y = "Average relative abundance across samples", fill = "Family") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Species 
```{r, message=FALSE, warning=FALSE, echo=TRUE} 
tax <- as(tax_table(ps_rel_1), "matrix")

no_species <- is.na(tax[, "Species"]) | tax[, "Species"] == "" | tax[, "Species"] == "NA"

count_no_species <- sum(no_species, na.rm = TRUE)
total_asvs <- nrow(tax)
pct_no_species <- 100 * count_no_species / total_asvs

cat("ASVs with no species assignment:", count_no_species, "\n")
cat("Percentage with no species assignment:", round(pct_no_species, 2), "%\n")

```


 
 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```

