---
title: "Canine Skin Microbiome: Curate Read and ASV Samples "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

**Purpose:**
  Analyze read count and ASVs across samples, specifically in comparison to controls (Chlor and Kit) with "zero" microbial signatures. 
  
**Overview**
 1. Curation of reads and ASVs https://benjjneb.github.io/FelineBileManuscript/Analyses/FelineBile_Curate.html  2. Perform decontam diagnostics 
https://benjjneb.github.io/FelineBileManuscript/Analyses/FelineBile_Core.html  

# Load R Packages
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(decontam); packageVersion("decontam")
library(ggplot2); packageVersion("ggplot2")
library(gridExtra)
library(data.table)
library(tidyverse)
library(stringr)
library(dada2)
library(phyloseq); packageVersion("phyloseq")
library(knitr)
library(phangorn)
```
# Select which sequencing library to process 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
prefix <- "CH2"
```

# Input files 
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Chimeras removed dada2 ASV feature table 
ASV_feature_table_og <- readRDS("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/input_csv_files/ASV_tables/04_final_pool_truncR150_maxE22/canine_skin_microbiome_filtered_nochim_asvs.rds")
# Remove rows where row names contain "Air"
ASV_feature_table_og <- ASV_feature_table_og[!grepl("Air", rownames(ASV_feature_table_og)), , drop = FALSE]

# Full metadata 
sample_metadata_og <- read.delim("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/input_csv_files/metadata/CH-All_16S_Metadata_20220124.tsv", header=TRUE)
# rownames = Samples
rownames(sample_metadata_og) <- sample_metadata_og$SampleID
nrow(sample_metadata_og)
sample_metadata_og <- sample_metadata_og[!grepl("Air", sample_metadata_og$SampleID), , drop = FALSE]
nrow(sample_metadata_og)
# ASV sequences annotated 
taxa_annotations_og <- readRDS("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/input_csv_files/ASV_tables/04_final_pool_truncR150_maxE22/canine_skin_microbiome_filtered_nochim_asvs_taxa.rds")

# Check to ensure feature table and taxa have identical sequences 
if(!identical(getSequences(taxa_annotations_og), getSequences(ASV_feature_table_og))) stop("Table mismatch: taxa_annotations, ASV_feature_table")

table(sample_metadata_og$Sample_Site, useNA="ifany")
```

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Input each individual sequencing library DNA concentration for pooling (prior to diluting for sequencing prep)

sample_DNA_library_CH1_clean <- read.csv("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/input_csv_files/metadata/CH1-4_PoolCalcs_CH1.csv") %>% 
  filter(!stringr::str_detect(Sample, "Air")) %>%
  filter(!is.na(Well) & Well != "") %>% 
  filter(!is.na(Primer) & Primer == "16S") %>% 
  select(Sample, Well, nM) %>%
  mutate(
    Sample = str_replace_all(Sample, "[_\\s]+", "-"),
    raw_DNA_conc = str_replace_all(nM, "nmol/L|nM", ""),
    raw_DNA_conc = as.numeric(str_trim(raw_DNA_conc)),
    pooled_DNA_conc = 200
  ) %>% 
  select(Sample, raw_DNA_conc, pooled_DNA_conc) %>% 
  distinct(Sample, .keep_all = TRUE)

sample_DNA_library_CH2_clean <- read.csv("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/input_csv_files/metadata/CH1-4_PoolCalcs_CH2.csv") %>% 
  mutate(Sample = case_match(
    Sample,
    "Chlor1-16S" ~ "Chlor1b-16S",
    "Chlor2-16S" ~ "Chlor2b-16S",
    "Chlor3-16S" ~ "Chlor3b-16S",
    .default = Sample
  )) %>% 
  filter(!is.na(Well) & Well != "") %>% 
  filter(!is.na(Primer) & Primer == "16S") %>% 
  filter(!stringr::str_detect(Sample, "Air")) %>%
  select(Sample, Well, nM) %>%
  mutate(
    Sample = str_replace_all(Sample, "[_\\s]+", "-"),
    raw_DNA_conc = str_replace_all(nM, "nmol/L|nM", ""),
    raw_DNA_conc = as.numeric(str_trim(raw_DNA_conc)),
    pooled_DNA_conc = 100
  ) %>% 
  select(Sample, raw_DNA_conc, pooled_DNA_conc) %>% 
  distinct(Sample, .keep_all = TRUE)

sample_DNA_library_CH3_clean <- read.csv("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/input_csv_files/metadata/CH1-4_PoolCalcs_CH3.csv") %>% 
  filter(!is.na(Well) & Well != "") %>% 
  filter(!is.na(Primer) & Primer == "16S") %>% 
  filter(!stringr::str_detect(Sample, "Air")) %>%
  select(Sample, Well, nM) %>%
  mutate(
    Sample = str_replace_all(Sample, "[_\\s]+", "-"),
    raw_DNA_conc = str_replace_all(nM, "nmol/L|nM", ""),
    raw_DNA_conc = as.numeric(str_trim(raw_DNA_conc)),
    pooled_DNA_conc = 10
  ) %>% 
  select(Sample, raw_DNA_conc, pooled_DNA_conc) %>% 
  distinct(Sample, .keep_all = TRUE)

sample_DNA_library_CH4_clean <- read.csv("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/input_csv_files/metadata/CH1-4_PoolCalcs_CH4.csv") %>% 
  mutate(Sample = case_match(
    Sample,
    "CH5 Kit Ctrl 16S" ~ "CH5b-Kit-Ctrl-16S",
    .default = Sample
  )) %>% 
  filter(!is.na(Well) & Well != "") %>% 
  filter(!is.na(Primer) & Primer == "16S") %>% 
  filter(!stringr::str_detect(Sample, "Air")) %>%
  select(Sample, Well, nM) %>%
  mutate(
    Sample = str_replace_all(Sample, "[_\\s]+", "-"),
    raw_DNA_conc = str_replace_all(nM, "nmol/L| nM", ""),
    raw_DNA_conc = as.numeric(str_trim(raw_DNA_conc)),
    pooled_DNA_conc = 50
  ) %>% 
  select(Sample, raw_DNA_conc, pooled_DNA_conc) %>% 
  distinct(Sample, .keep_all = TRUE)

# Combine all sequencing runs 
all_libraries_clean <- bind_rows(
  CH1 = sample_DNA_library_CH1_clean,
  CH2 = sample_DNA_library_CH2_clean,
  CH3 = sample_DNA_library_CH3_clean,
  CH4 = sample_DNA_library_CH4_clean,
  .id = "run"
)


# Distribution of sample DNA concentrations for each sequencing run 
ggplot(all_libraries_clean, aes(x = raw_DNA_conc, fill = run)) +
  ggtitle(paste0("Post-PCR DNA concentrations"))+
  geom_histogram(bins = 15, color = "black", alpha = 0.7, position = "identity") +
  labs(x = "DNA concentration (nM)", y = "Count of samples", fill = "Run") +
  theme_minimal()

# Subset metadata to samples processed within a given sequencing run 
sample_metadata_subset <- sample_metadata_og %>% 
  select(-any_of("pooled_DNA_conc")) %>% 
  left_join(get(paste0("sample_DNA_library_", prefix, "_clean")), by = c("SampleID" = "Sample")) %>% 
  filter(!is.na(pooled_DNA_conc))

rownames(sample_metadata_subset) <- sample_metadata_subset$SampleID

# Subset ASV feature table to samples processed within a given sequencing run 
shared_samples <- intersect(rownames(ASV_feature_table_og), sample_metadata_subset$SampleID)
ASV_table_subset <- ASV_feature_table_og[shared_samples, , drop = FALSE]
# Remove ASVs with 0 counts across all subsetted samples
ASV_table_subset <- ASV_table_subset[, colSums(ASV_table_subset) > 0, drop = FALSE]

shared_ASVs <- intersect(colnames(ASV_table_subset), rownames(taxa_annotations_og))
taxa_annotations_subset <- taxa_annotations_og[shared_ASVs, , drop = FALSE]

# Number of ASVs per bacterial Phylum
table(taxa_annotations_subset[,"Phylum"], useNA="ifany")

# Percent of each bacterial Phylum present in ASV feature table 
tapply(colSums(ASV_table_subset), taxa_annotations_subset[,"Phylum"], sum)/sum(ASV_table_subset)

```

# Curate Data
Define a generic column with TRUE for all negative controls, and FALSE for real (positive) samples
```{r, message=FALSE, warning=FALSE, echo=TRUE}
sample_metadata_subset$NegativeControl <- grepl("Chlorhex|Kit", sample_metadata_subset$Sample_Site)
table(sample_metadata_subset$Sample_Site, sample_metadata_subset$NegativeControl)

class(sample_metadata_subset$raw_DNA_conc)
summary(sample_metadata_subset$raw_DNA_conc)
```


```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
# Distribution of sample DNA concentrations for each sequencing run 
ggplot(sample_metadata_subset, aes(x = raw_DNA_conc)) +
  ggtitle(paste0("Post-PCR DNA concentrations for Seq. Run: ", prefix)) +
  geom_histogram(bins = 15, color = "black", alpha = 0.7, position = "identity") +
  labs(x = "DNA conc (nM)", y = "Count of samples", fill = "Run") +
  theme_minimal()

```


## Inspect library size distribution (read depths) for samples  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# 1. Verify all samples match exactly
if(!setequal(rownames(sample_metadata_subset), rownames(ASV_table_subset))) {
  stop("Mismatched samples between metadata and ASV table.")
}

# 2. Force ASV table rows into the exact same order as metadata
ASV_table_subset <- ASV_table_subset[rownames(sample_metadata_subset), ]

# 3. Calculate LibrarySize safely
sample_metadata_subset$LibrarySize <- rowSums(ASV_table_subset)

# 4. Plot with correct coloring
ggplot(data=sample_metadata_subset, aes(x=rank(LibrarySize), y=LibrarySize, color=Sample_Site)) + 
  geom_point(size = 1.5) + 
  scale_color_manual(values = c(
    "Chlorhex" = "orange",
    "Kit" = "green",
    "Surgical" = "blue",
    "Control" = "purple"
  ))
sort(sample_metadata_subset$LibrarySize)[1:30]
table(sample_metadata_subset$Sample_Site, sample_metadata_subset$NegativeControl)

cat("Print which category (positive = FALSE or negative = TRUE) are the low depth samples (<2k reads) from")
table(sample_metadata_subset$Sample_Site[sample_metadata_subset$LibrarySize < 2000], 
      sample_metadata_subset$NegativeControl[sample_metadata_subset$LibrarySize < 2000])

ps <- phyloseq(otu_table(ASV_table_subset, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_subset), 
               tax_table(taxa_annotations_subset))

```

## Distribution of ASVs 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Distribution shows the number of unique ASVs found in a given number of samples 

asv_counts <- colSums(ASV_table_subset > 0)
plot_data <- as.data.frame(table(asv_counts))
colnames(plot_data) <- c("Num_Samples", "Num_ASVs")
plot_data$Num_Samples <- as.numeric(as.character(plot_data$Num_Samples))
ggplot(plot_data, aes(x = Num_Samples, y = Num_ASVs)) +
  geom_bar(stat = "identity", fill = "gray", color = "black") +
  scale_x_continuous(limits = c(0, max(plot_data$Num_Samples) + 50), breaks = seq(0, max(plot_data$Num_Samples) + 50, by = 50)) +
  scale_y_log10() + 
  labs(
    title = "How many unique ASVs are present in a given number of samples?",
    x = "Number of Samples",
    y = "Number of Unique ASVs (Log10)"
  ) +
  theme_minimal()

```


## ASV prevalence vs abundance across all samples
```{r, message=FALSE, warning=FALSE, echo=TRUE}
df.asv <- data.frame(Sequence=colnames(ASV_table_subset), 
                     Prevalence=colSums((ASV_table_subset)>0), # number of samples with a given ASV (prevalence)
                     Prevalence.Pos=colSums(((ASV_table_subset)[!sample_metadata_subset$NegativeControl,]>0)),
                     Abundance=colSums((ASV_table_subset)), # number of reads from a given ASV (abundance)
                     Abundance.Pos=colSums(((ASV_table_subset)[!sample_metadata_subset$NegativeControl,])))

ggplot(data=df.asv, aes(x=Prevalence, y=Abundance)) + geom_point(size = 0.5) + scale_y_log10()+ ggtitle("Full Dataset (with negative controls)")


ggplot(data=df.asv, aes(x=Prevalence.Pos, y=Abundance.Pos)) + geom_point(size = 0.5) + scale_y_log10() + ggtitle("Positive Samples")


```


# Initial ASV-filtering   
ASVs found in >= 5% of positive samples OR 1000 total reads
```{r, message=FALSE, warning=FALSE, echo=TRUE}
postive_samples <- nrow((ASV_table_subset)[!sample_metadata_subset$NegativeControl, ])
cat(paste0("Number of positive samples: ", postive_samples, "\n"))
total_reads <- sum(ASV_table_subset)
cat(paste0("Total number of reads:", total_reads, "\n"))

# 5% of positive samples
prevalence_threshold <- round(0.05*postive_samples)
# If 5% is = 1, make the threshold 2
if (prevalence_threshold == 1) {
  prevalence_threshold <- prevalence_threshold + 1
}
cat(paste0("5% prevalence threshold: ", prevalence_threshold, "\n"))

abundance_threshold <- 1000
 
true_asv_names <- df.asv$Sequence[df.asv$Prevalence.Pos >= prevalence_threshold | df.asv$Abundance.Pos >= abundance_threshold]

columns_to_keep <- c(true_asv_names)

# Subset ASV feature table to only those with 1000 total reads across samples OR found in at least 5% of positive samples
ASV_feature_table.filt <- ASV_table_subset[, columns_to_keep]

 
total_positive_ASVs <- sum(df.asv$Prevalence.Pos > 1)

print(paste("Number of ASVs found in samples (negative and positive samples)"))
ncol(ASV_table_subset)
print(paste("Total number of reads found in samples (negative and positive samples)"))
sum(ASV_table_subset)
print(paste("Number of ASVs in positive samples:"))
print(total_positive_ASVs)
print(paste("Number of ASVs found in at least", prevalence_threshold, "positive samples or with total abundances >=", abundance_threshold, "reads"))
ncol(ASV_feature_table.filt)
print(paste("% of ASVs found in at least", prevalence_threshold, "positive samples or with total abundances >=", abundance_threshold, "reads"))
(ncol(ASV_feature_table.filt)/ncol(ASV_table_subset))*100
print(paste("% of Total Reads from ASVs found in at least", prevalence_threshold, "positive samples or with total abundances >=", abundance_threshold, "reads"))
(sum(ASV_feature_table.filt)/sum(ASV_table_subset))*100
```

## Filter metadata and taxa tables according to new ASV table
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# Subset metadata and taxa table 
sample_metadata.filt <- sample_metadata_subset[match(rownames(ASV_feature_table.filt), rownames(sample_metadata_subset)), ]
taxa_annotations.filt <- taxa_annotations_subset[match(colnames(ASV_feature_table.filt), rownames(taxa_annotations_subset)), ]
 
if(!identical(rownames(taxa_annotations.filt), getSequences(ASV_feature_table.filt))) {
  stop("Table mismatch: taxa_annotations.filt and ASV_feature_table.filt are not identical.")
}
```

## Inspect library size distribution (read depths) for samples  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# 1. Verify all samples match exactly
if(!setequal(rownames(sample_metadata.filt), rownames(ASV_feature_table.filt))) {
  stop("Mismatched samples between metadata and ASV table.")
}

# 2. Force ASV table rows into the exact same order as metadata
ASV_feature_table.filt <- ASV_feature_table.filt[rownames(sample_metadata.filt), ]

# 3. Calculate LibrarySize safely
sample_metadata.filt$LibrarySize <- rowSums(ASV_feature_table.filt)

# 4. Plot with correct coloring
ggplot(data=sample_metadata.filt, aes(x=rank(LibrarySize), y=LibrarySize, color=Sample_Site)) + 
  geom_point(size = 1.0) + 
  scale_color_manual(values = c(
    "Chlorhex" = "orange",
    "Kit" = "green",
    "Surgical" = "blue",
    "Control" = "purple"
  ))
sort(sample_metadata.filt$LibrarySize)[1:30]
table(sample_metadata.filt$Sample_Site, sample_metadata.filt$NegativeControl)

cat("Print which category (positive = FALSE or negative = TRUE) are the low depth samples (<2k reads) from")
table(sample_metadata.filt$Sample_Site[sample_metadata.filt$LibrarySize < 2000], 
      sample_metadata.filt$NegativeControl[sample_metadata.filt$LibrarySize < 2000])
```

# Run Decontam Diagnostics: Combined Method
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(100)
contam <- isContaminant(ASV_feature_table.filt, conc=sample_metadata.filt$raw_DNA_conc, neg=sample_metadata.filt$NegativeControl, method="combined", threshold = 0.5)
asv_names <- colnames(ASV_feature_table.filt)
contam$ASV       <- asv_names   
contam$Phylum <- taxa_annotations.filt[,"Phylum"]
contam$Genus <- taxa_annotations.filt[,"Genus"]
contam$Frequency <- cut(contam$freq, c(0, 0.25, 0.5, 0.75, 1.0), labels=c("1-25%", "25-50%", "50-75%", "75-100%"))
contam$Prevalence <- cut(contam$prev, c(0, 20, 40, 60, 500), labels=c("0-20", "20-40", "40-60", "60+"))
contam$Abundance <- colSums(ASV_feature_table.filt)
rownames(contam) <- NULL 


histo <- ggplot(contam, aes(x=p)) + 
  labs(title='Decontam Method: Combined', subtitle='Organized by Prevalence', x = 'decontam Score', y='Number of ASVs') + 
  geom_histogram(binwidth=0.02) +
  ylim(0, 650)

histo
histo + facet_wrap(~Prevalence)

i.top10 <- order(contam$p, decreasing=FALSE)[1:10]
contam[i.top10,]

cat("Number of contaminants (method = combined, threshold = 0.5):", sum(contam$contaminant))



``` 

Remove low-depth samples and then remove contaminant ASVs
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Filter out samples with <2k reads
keep_samples <- rownames(sample_metadata.filt)[sample_metadata.filt$LibrarySize >= 2000]
final_samples <- intersect(keep_samples, rownames(ASV_feature_table.filt))
sample_metadata.filt_2 <- sample_metadata.filt[final_samples, , drop = FALSE]
ASV_feature_table.filt_2 <- ASV_feature_table.filt[final_samples, , drop = FALSE]
taxa_annotations.filt_2 <- taxa_annotations.filt[match(colnames(ASV_feature_table.filt_2), rownames(taxa_annotations.filt)), ]
 
# Remove contaminants
ASV_feature_table_decontam <- ASV_feature_table.filt_2[, !contam$contaminant]
# Subset metadata and taxa table 
sample_metadata_decontam <- sample_metadata.filt[match(rownames(ASV_feature_table_decontam), rownames(sample_metadata.filt)), ]
taxa_annotations_decontam <- taxa_annotations.filt[match(colnames(ASV_feature_table_decontam), rownames(taxa_annotations.filt)), ]


ps_og <- phyloseq(
  otu_table(as.matrix(ASV_table_subset), taxa_are_rows = FALSE), 
  sample_data(sample_metadata_subset), 
  tax_table(as.matrix(taxa_annotations_subset))
)

ps_og_no_neg <- prune_samples(
  grepl("Skin", sample_data(ps_og)$Location_Type),
  ps_og
)


ps_filtered <- phyloseq(
  otu_table(as.matrix(ASV_feature_table.filt_2), taxa_are_rows = FALSE), 
  sample_data(sample_metadata.filt_2), 
  tax_table(as.matrix(taxa_annotations.filt_2))
)

ps_filtered_no_neg <- prune_samples(
  grepl("Skin", sample_data(ps_filtered)$Location_Type),
  ps_filtered
)


ps_decontam <- phyloseq(
  otu_table(as.matrix(ASV_feature_table_decontam), taxa_are_rows = FALSE), 
  sample_data(sample_metadata_decontam), 
  tax_table(as.matrix(taxa_annotations_decontam))
)

ps_decontam_no_neg <- prune_samples(
  grepl("Skin", sample_data(ps_decontam)$Location_Type),
  ps_decontam
)
 
 
# Baseline Positive Samples
asvs_og        <- ntaxa(ps_og_no_neg)
reads_og       <- sum(sample_sums(ps_og_no_neg))

# Filtered Positive Samples
asvs_filtered  <- ntaxa(ps_filtered_no_neg)
reads_filtered <- sum(sample_sums(ps_filtered_no_neg))

# Decontaminated Positive Samples
asvs_decontam  <- ntaxa(ps_decontam_no_neg)
reads_decontam <- sum(sample_sums(ps_decontam_no_neg))


 
pct_asvs_filt   <- (asvs_filtered / asvs_og) * 100
pct_reads_filt  <- (reads_filtered / reads_og) * 100

pct_asvs_dec    <- (asvs_decontam / asvs_og) * 100
pct_reads_dec   <- (reads_decontam / reads_og) * 100


 
pipeline_summary <- data.frame(
  Pipeline_Step = c(
    "Original Positive Samples", 
    "Filtered Positive Samples", 
    "Decontaminated Positive Samples"
  ),
  Total_ASVs = c(
    asvs_og, 
    asvs_filtered, 
    asvs_decontam
  ),
  ASVs_Retained = c(
    "100.00%", 
    paste0(round(pct_asvs_filt, 2), "%"), 
    paste0(round(pct_asvs_dec, 2), "%")
  ),
  Total_Reads = c(
    reads_og, 
    reads_filtered, 
    reads_decontam
  ),
  Reads_Retained = c(
    "100.00%", 
    paste0(round(pct_reads_filt, 2), "%"), 
    paste0(round(pct_reads_dec, 2), "%")
  )
)


 
cat("### Positive Sample Processing Summary\n\n")

kable(
  pipeline_summary, 
  format = "markdown", 
  col.names = c("Pipeline Step", "Total ASVs", "% ASVs Retained", "Total Reads", "% Reads Retained"),
  align = c("l", "r", "r", "r", "r")
)



 
save(ASV_feature_table_decontam, taxa_annotations_decontam, sample_metadata_decontam, 
     file = paste0("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/R_scripts/", prefix, "_filtered_canine_skin_microbiome.rda"))

 
saveRDS(ps_decontam_no_neg, 
        file = paste0("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/R_scripts/", prefix, "_ASV_feature_table_decontam.rds"))

```

## Inspect library size distribution (read depths) for samples  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# 1. Verify all samples match exactly
if(!setequal(rownames(sample_metadata_decontam), rownames(ASV_feature_table_decontam))) {
  stop("Mismatched samples between metadata and ASV table.")
}

# 2. Force ASV table rows into the exact same order as metadata
ASV_feature_table_decontam <- ASV_feature_table_decontam[rownames(sample_metadata_decontam), ]

# 3. Calculate LibrarySize safely
sample_metadata_decontam$LibrarySize <- rowSums(ASV_feature_table_decontam)

# 4. Plot with correct coloring
ggplot(data=sample_metadata_decontam, aes(x=rank(LibrarySize), y=LibrarySize, color=Sample_Site)) + 
  geom_point(size = 1.0) + 
  scale_color_manual(values = c(
    "Chlorhex" = "orange",
    "Kit" = "green",
    "Surgical" = "blue",
    "Control" = "purple"
  ))
sort(sample_metadata_decontam$LibrarySize)[1:30]
table(sample_metadata_decontam$Sample_Site, sample_metadata_decontam$NegativeControl)

cat("Print which category (positive = FALSE or negative = TRUE) are the low depth samples (<2k reads) from")
table(sample_metadata_decontam$Sample_Site[sample_metadata_decontam$LibrarySize < 2000], 
      sample_metadata_decontam$NegativeControl[sample_metadata_decontam$LibrarySize < 2000])
```
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```
 