---
title: "Canine Skin Microbiome: Curate Read and ASV Samples "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

**Purpose:**
  Analyze read count and ASVs across samples, specifically in comparison to controls (Chlor and Kit) with "zero" microbial signatures. 
  
**Overview**
 1. Curation of reads and ASVs https://benjjneb.github.io/FelineBileManuscript/Analyses/FelineBile_Curate.html  
 2. Perform decontam diagnostics  https://benjjneb.github.io/FelineBileManuscript/Analyses/FelineBile_Core.html  

# Load R Packages
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(decontam); packageVersion("decontam")
library(ggplot2); packageVersion("ggplot2")
library(gridExtra)
library(data.table)
library(tidyverse)
library(stringr)
library(dada2)
library(phyloseq); packageVersion("phyloseq")
library(knitr)
library(phangorn)
library(kableExtra)
library(gt)
```
# Input files 
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Chimeras removed dada2 ASV feature table 
ASV_feature_table_og <- readRDS("../01_input_files/dada2_results/canine_skin_microbiome_filtered_nochim_asvs.rds")
# Remove rows where row names contain "Air"
#ASV_feature_table_og <- ASV_feature_table_og[!grepl("Air", rownames(ASV_feature_table_og)), , drop = FALSE]

# Full metadata 
sample_metadata_og <- read.delim("../01_input_files/metadata/CH-All_Metadata_20220124.tsv", header=TRUE)
# rownames = Samples
rownames(sample_metadata_og) <- sample_metadata_og$SampleID
nrow(sample_metadata_og)
#sample_metadata_og <- sample_metadata_og[!grepl("Air", sample_metadata_og$SampleID), , drop = FALSE]
nrow(sample_metadata_og)
# ASV sequences annotated 
taxa_annotations_og <- readRDS("../01_input_files/dada2_results/canine_skin_microbiome_filtered_nochim_asvs_taxa_species.rds")
 

table(sample_metadata_og$Sample_Site, useNA="ifany")

out_dir <- "../02_output_files/02_image_files"
```

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Input each individual sequencing library DNA concentration for pooling (prior to diluting for sequencing prep)

sample_DNA_library_CH1_clean <- read.csv("../01_input_files/metadata/PoolCalcs_CH1.csv") %>% 
# filter(!stringr::str_detect(Sample, "Air")) %>%
  filter(!is.na(Well) & Well != "") %>% 
  filter(!is.na(Primer) & Primer == "16S") %>% 
  dplyr::select(Sample, Well, nM) %>%
  mutate(
    Sample = str_replace_all(Sample, "[_\\s]+", "-"),
    raw_DNA_conc = str_replace_all(nM, "nmol/L|nM", ""),
    raw_DNA_conc = as.numeric(str_trim(raw_DNA_conc)),
    pooled_DNA_conc = 200
  ) %>% 
  dplyr::select(Sample, raw_DNA_conc, pooled_DNA_conc) %>% 
  distinct(Sample, .keep_all = TRUE)

sample_DNA_library_CH2_clean <- read.csv("../01_input_files/metadata/PoolCalcs_CH2.csv") %>% 
  mutate(Sample = case_match(
    Sample,
    "Chlor1-16S" ~ "Chlor1b-16S",
    "Chlor2-16S" ~ "Chlor2b-16S",
    "Chlor3-16S" ~ "Chlor3b-16S",
    .default = Sample
  )) %>% 
  filter(!is.na(Well) & Well != "") %>% 
  filter(!is.na(Primer) & Primer == "16S") %>% 
 #filter(!stringr::str_detect(Sample, "Air")) %>%
  dplyr::select(Sample, Well, nM) %>%
  mutate(
    Sample = str_replace_all(Sample, "[_\\s]+", "-"),
    raw_DNA_conc = str_replace_all(nM, "nmol/L|nM", ""),
    raw_DNA_conc = as.numeric(str_trim(raw_DNA_conc)),
    pooled_DNA_conc = 100
  ) %>% 
  dplyr::select(Sample, raw_DNA_conc, pooled_DNA_conc) %>% 
  distinct(Sample, .keep_all = TRUE)

sample_DNA_library_CH3_clean <- read.csv("../01_input_files/metadata/PoolCalcs_CH3.csv") %>% 
  filter(!is.na(Well) & Well != "") %>% 
  filter(!is.na(Primer) & Primer == "16S") %>% 
  #filter(!stringr::str_detect(Sample, "Air")) %>%
  dplyr::select(Sample, Well, nM) %>%
  mutate(
    Sample = str_replace_all(Sample, "[_\\s]+", "-"),
    raw_DNA_conc = str_replace_all(nM, "nmol/L|nM", ""),
    raw_DNA_conc = as.numeric(str_trim(raw_DNA_conc)),
    pooled_DNA_conc = 10
  ) %>% 
  dplyr::select(Sample, raw_DNA_conc, pooled_DNA_conc) %>% 
  distinct(Sample, .keep_all = TRUE)

sample_DNA_library_CH4_clean <- read.csv("../01_input_files/metadata/PoolCalcs_CH4.csv") %>% 
  mutate(Sample = case_match(
    Sample,
    "CH5 Kit Ctrl 16S" ~ "CH5b-Kit-Ctrl-16S",
    .default = Sample
  )) %>% 
  filter(!is.na(Well) & Well != "") %>% 
  filter(!is.na(Primer) & Primer == "16S") %>% 
#  filter(!stringr::str_detect(Sample, "Air")) %>%
  dplyr::select(Sample, Well, nM) %>%
  mutate(
    Sample = str_replace_all(Sample, "[_\\s]+", "-"),
    raw_DNA_conc = str_replace_all(nM, "nmol/L| nM", ""),
    raw_DNA_conc = as.numeric(str_trim(raw_DNA_conc)),
    pooled_DNA_conc = 50
  ) %>% 
  dplyr::select(Sample, raw_DNA_conc, pooled_DNA_conc) %>% 
  distinct(Sample, .keep_all = TRUE)

# Combine all sequencing runs 
all_libraries_clean <- bind_rows(
  CH1 = sample_DNA_library_CH1_clean,
  CH2 = sample_DNA_library_CH2_clean,
  CH3 = sample_DNA_library_CH3_clean,
  CH4 = sample_DNA_library_CH4_clean,
  .id = "run"
)

 
 sample_metadata_subset <- sample_metadata_og %>% 
  dplyr::select(-any_of("pooled_DNA_conc")) %>% 
 left_join(get(paste0("all_libraries_clean")), by = c("SampleID" = "Sample")) %>% 
  filter(!is.na(pooled_DNA_conc))

rownames(sample_metadata_subset) <- sample_metadata_subset$SampleID
 
# Number of ASVs per bacterial Phylum
table(taxa_annotations_og[,"Phylum"], useNA="ifany")

# Percent of each bacterial Phylum present in ASV feature table 
tapply(colSums(ASV_feature_table_og), taxa_annotations_og[,"Phylum"], sum)/sum(ASV_feature_table_og)


```

# Curate Data
Define a generic column with TRUE for all negative controls, and FALSE for real (positive) samples
```{r, message=FALSE, warning=FALSE, echo=TRUE}
sample_metadata_subset$NegativeControl <- grepl("Chlorhex|Kit|Air", sample_metadata_subset$Sample_Site)
table(sample_metadata_subset$Sample_Site, sample_metadata_subset$NegativeControl)

class(sample_metadata_subset$raw_DNA_conc)
summary(sample_metadata_subset$raw_DNA_conc)



dog_library_samples <- sample_metadata_subset %>%
  count(Sample_Site, Location_Type, Sample_Plate, name = "n") %>%
  pivot_wider(names_from = Sample_Plate, values_from = n, values_fill = 0)

totals <- dog_library_samples %>%
  dplyr::select(-Sample_Site, -Location_Type) %>%
  dplyr::summarise(dplyr::across(everything(), sum)) %>%
  dplyr::mutate(Sample_Site = "Total", Location_Type = NA_character_) %>%
  dplyr::select(Sample_Site, Location_Type, everything())

dog_library_samples <- dplyr::bind_rows(dog_library_samples, totals)

total_columns <- ncol(dog_library_samples)
sample_plate_index <- which(colnames(dog_library_samples) == "Sample_plate")

align_vector <- rep("|c", total_columns)
align_vector[total_columns] <- "|c|" # Close the rightmost edge of the table
align_string <- paste(align_vector, collapse = "")

out_table <- file.path(out_dir, "total_samples_by_type.png")


dog_library_samples %>% 
  kable(caption = "Sample Type Count per Sequencing Run", align = "c") %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"), 
    full_width = FALSE
  ) %>% 
  column_spec(sample_plate_index, border_right = "3px solid #333333 !important;", color = "black") %>% 
  column_spec(1:ncol(dog_library_samples), color = "black") %>%
  row_spec(nrow(dog_library_samples), bold = TRUE, background = "#f2f2f2", color = "black")  

save_kable(
  dog_library_samples,
  file = out_table,
  zoom = 3,
  vwidth = 600,
  vheight = 1200
)
 
 
sample_metadata_subset_dogs <- subset(sample_metadata_subset, !grepl("Chlor|Kit", Individual))

dog_library_table <- with(
  subset(sample_metadata_subset_dogs),
  table(Individual, Sample_Plate)
)
dog_library_table <- rbind(
  dog_library_table,
  Total = colSums(dog_library_table)
)
 
 

out_table <- file.path(out_dir, "total_samples_by_dog.png")
 

dog_library_table_bold <- dog_library_table
dog_library_table_bold[dog_library_table_bold != 0] <- paste0(
  "<b>", dog_library_table_bold[dog_library_table_bold != 0], "</b>"
)

dog_library_table_bold <- kable(
  dog_library_table_bold,
  caption = "Dog Sample Count per Sequencing Run",
  align = "c",
  escape = FALSE,
  format = "html"
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  ) %>%
  column_spec(1:5, color = "black") %>%
  row_spec(
    nrow(dog_library_table_bold),
    bold = TRUE,
    color = "black",
    background = "#f2f2f2"
  )

save_kable(
  dog_library_table_bold,
  file = out_table,
  zoom = 3,
  vwidth = 400,
  vheight = 1200
)

dog_library_table_bold

```


```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
# Distribution of sample DNA concentrations for each sequencing run 
ggplot(sample_metadata_subset, aes(x = raw_DNA_conc, fill = Sample_Plate)) + 
  geom_histogram(bins = 15, alpha = 0.7, position = "identity", color = "black") + 
  ggtitle("Post-PCR DNA concentrations") + 
  labs(x = "DNA conc (nM)", y = "Count of samples", fill = "Seq. Run") + 
  theme_minimal()


```


## Inspect library size distribution (read depths) for samples  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
# Verify all samples match exactly
if(!setequal(rownames(sample_metadata_subset), rownames(ASV_feature_table_og))) {
  stop("Mismatched samples between metadata and ASV table.")
}

# 2. Force ASV table rows into the exact same order as metadata
ASV_feature_table_og <- ASV_feature_table_og[rownames(sample_metadata_subset), ]

# 3. Calculate LibrarySize safely
sample_metadata_subset$LibrarySize <- rowSums(ASV_feature_table_og)

sample_metadata_subset$Sample_Site <- factor(
  sample_metadata_subset$Sample_Site,
  levels = c("Control", "Surgical", "Air", "Chlorhex", "Kit")
)

sample_metadata_subset2 <- sample_metadata_subset |>
  dplyr::mutate(front = Sample_Site %in% c("Air", "Kit", "Chlorhex")) |>
  dplyr::arrange(front)

ggplot(sample_metadata_subset2, aes(x = rank(LibrarySize), y = LibrarySize, color = Sample_Site)) + 
   scale_y_continuous(
    limits = c(0, 175000),
    breaks = seq(0, 175000, by = 50000),
    labels = scales::comma
  ) +
  geom_point(size = 2.0) + 
  scale_color_manual(values = c(
    "Chlorhex" = "orange",
    "Kit" = "green",
    "Surgical" = "blue",
    "Control" = "purple",
    "Air" = "red"
  ),
    labels = c(
      "Chlorhex" = "Chlorhex",
      "Kit" = "Kit",
      "Surgical" = "Surgical (Skin)",
      "Control" = "Control (Skin)",
      "Air" = "Air"
    )
  ) +
  theme(
    legend.position = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.key = element_rect(fill = NA, color = NA)
  ) + guides(color = guide_legend(override.aes = list(size = 3)))

sort(sample_metadata_subset$LibrarySize)[1:30]
table(sample_metadata_subset$Sample_Site, sample_metadata_subset$NegativeControl)

cat("Print which category (positive = FALSE or negative = TRUE) are the low depth samples (<2k reads) from")
table(sample_metadata_subset$Sample_Site[sample_metadata_subset$LibrarySize < 2000], 
      sample_metadata_subset$NegativeControl[sample_metadata_subset$LibrarySize < 2000])

ps <- phyloseq(otu_table(ASV_feature_table_og, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_subset), 
               tax_table(taxa_annotations_og))

```

## Distribution of ASVs 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
# Distribution shows the number of unique ASVs found in a given number of samples 

asv_counts <- colSums(ASV_feature_table_og > 0)
plot_data <- as.data.frame(table(asv_counts))
colnames(plot_data) <- c("Num_Samples", "Num_ASVs")
plot_data$Num_Samples <- as.numeric(as.character(plot_data$Num_Samples))
ggplot(plot_data, aes(x = Num_Samples, y = Num_ASVs)) +
  geom_bar(stat = "identity", fill = "gray", color = "black") +
  scale_x_continuous(limits = c(0, max(plot_data$Num_Samples) + 50), breaks = seq(0, max(plot_data$Num_Samples) + 50, by = 50)) +
  scale_y_log10() + 
  labs(
    title = "How many unique ASVs are present in a given number of samples?",
    x = "Number of Samples",
    y = "Number of Unique ASVs (Log10)"
  ) +
  theme_minimal()

```


## ASV prevalence vs abundance across all samples
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
df.asv <- data.frame(Sequence=colnames(ASV_feature_table_og), 
                     Prevalence=colSums((ASV_feature_table_og)>0), # number of samples with a given ASV (prevalence)
                     Prevalence.Pos=colSums(((ASV_feature_table_og)[!sample_metadata_subset$NegativeControl,]>0)),
                     Abundance=colSums((ASV_feature_table_og)), # number of reads from a given ASV (abundance)
                     Abundance.Pos=colSums(((ASV_feature_table_og)[!sample_metadata_subset$NegativeControl,])))

ggplot(data=df.asv, aes(x=Prevalence, y=Abundance)) + geom_point(size = 0.5) + scale_y_log10()+ ggtitle("Full Dataset (with negative controls)")


ggplot(data=df.asv, aes(x=Prevalence.Pos, y=Abundance.Pos)) + geom_point(size = 0.5) + scale_y_log10() + ggtitle("Positive Samples")


```


# Initial ASV-filtering   
ASVs found in >= 10% of positive samples OR >=0.01% of total reads
```{r, message=FALSE, warning=FALSE, echo=TRUE}
postive_samples <- nrow((ASV_feature_table_og)[!sample_metadata_subset$NegativeControl, ])
cat(paste0("Number of positive samples: ", postive_samples, "\n"))
total_reads <- sum(ASV_feature_table_og)
cat(paste0("Total number of reads:", total_reads, "\n"))

# 5% of positive samples
prevalence_threshold <- round(0.05*postive_samples)
 
cat(paste0("10% prevalence threshold: ", prevalence_threshold, "\n"))

# 0.01% of total reads threshold
abundance_threshold <- 0.0001*(total_reads)
 
true_asv_names <- df.asv$Sequence[df.asv$Prevalence.Pos >= prevalence_threshold | df.asv$Abundance.Pos >= abundance_threshold]

columns_to_keep <- c(true_asv_names)

# Subset ASV feature table to only those with >=0.01% total reads across samples OR found in at least 5% of positive samples
ASV_feature_table.filt <- ASV_feature_table_og[, columns_to_keep]

 
total_positive_ASVs <- sum(df.asv$Prevalence.Pos > 1)

print(paste("Number of ASVs found in samples (negative and positive samples)"))
ncol(ASV_feature_table_og)
print(paste("Total number of reads found in samples (negative and positive samples)"))
sum(ASV_feature_table_og)
print(paste("Number of ASVs in positive samples:"))
print(total_positive_ASVs)
print(paste("Number of ASVs found in at least", prevalence_threshold, "positive samples or with total abundances >=", abundance_threshold, "reads"))
ncol(ASV_feature_table.filt)
print(paste("% of ASVs found in at least", prevalence_threshold, "positive samples or with total abundances >=", abundance_threshold, "reads"))
(ncol(ASV_feature_table.filt)/ncol(ASV_feature_table_og))*100
print(paste("% of Total Reads from ASVs found in at least", prevalence_threshold, "positive samples or with total abundances >=", abundance_threshold, "reads"))
(sum(ASV_feature_table.filt)/sum(ASV_feature_table_og))*100
```

## Filter metadata and taxa tables according to new ASV table
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# Subset metadata and taxa table 
sample_metadata.filt <- sample_metadata_subset[match(rownames(ASV_feature_table.filt), rownames(sample_metadata_subset)), ]
taxa_annotations.filt <- taxa_annotations_og[match(colnames(ASV_feature_table.filt), rownames(taxa_annotations_og)), ]
 
if(!identical(rownames(taxa_annotations.filt), getSequences(ASV_feature_table.filt))) {
  stop("Table mismatch: taxa_annotations.filt and ASV_feature_table.filt are not identical.")
}
```

## Inspect library size distribution (read depths) for samples  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
# Verify all samples match exactly
if(!setequal(rownames(sample_metadata.filt), rownames(ASV_feature_table.filt))) {
  stop("Mismatched samples between metadata and ASV table.")
}

 
ASV_feature_table.filt <- ASV_feature_table.filt[rownames(sample_metadata.filt), ]

 
sample_metadata.filt$LibrarySize <- rowSums(ASV_feature_table.filt)

sample_metadata.filt2 <- sample_metadata.filt |>
  dplyr::mutate(front = Sample_Site %in% c("Air", "Kit", "Chlorhex")) |>
  dplyr::arrange(front)

ggplot(sample_metadata.filt2, aes(x = rank(LibrarySize), y = LibrarySize, color = Sample_Site)) + 
   scale_y_continuous(
    limits = c(0, 175000),
    breaks = seq(0, 175000, by = 50000),
    labels = scales::comma
  ) + 
  geom_point(size = 2.0) + 
  scale_color_manual(values = c(
    "Chlorhex" = "orange",
    "Kit" = "green",
    "Surgical" = "blue",
    "Control" = "purple",
    "Air" = "red"
  ),
    labels = c(
      "Chlorhex" = "Chlorhex",
      "Kit" = "Kit",
      "Surgical" = "Surgical (Skin)",
      "Control" = "Control (Skin)",
      "Air" = "Air"
    )
  ) +
  theme(
    legend.position = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.key = element_rect(fill = NA, color = NA)
  ) + guides(color = guide_legend(override.aes = list(size = 3)))

sort(sample_metadata.filt$LibrarySize)[1:30]
table(sample_metadata.filt$Sample_Site, sample_metadata.filt$NegativeControl)

cat("Print which category (positive = FALSE or negative = TRUE) are the low depth samples (<2k reads) from")
table(sample_metadata.filt$Sample_Site[sample_metadata.filt$LibrarySize < 2000], 
      sample_metadata.filt$NegativeControl[sample_metadata.filt$LibrarySize < 2000])
```

# Run Decontam Diagnostics: Combined method with sequencing run as batches 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(100)
contam <- isContaminant(ASV_feature_table.filt, batch = sample_metadata_subset$Sample_Plate, batch.combine = c("minimum"), conc=sample_metadata.filt$raw_DNA_conc, neg=sample_metadata.filt$NegativeControl, method="combined", threshold = 0.1)
asv_names <- colnames(ASV_feature_table.filt)
contam$ASV       <- asv_names   
contam$Phylum <- taxa_annotations.filt[,"Phylum"]
contam$Genus <- taxa_annotations.filt[,"Genus"]
contam$Frequency <- cut(contam$freq, c(0, 0.25, 0.5, 0.75, 1.0), labels=c("1-25%", "25-50%", "50-75%", "75-100%"))
contam$Prevalence <- cut(contam$prev, c(0, 20, 40, 60, 500), labels=c("0-20", "20-40", "40-60", "60+"))
contam$Abundance <- colSums(ASV_feature_table.filt)
rownames(contam) <- NULL 


histo <- ggplot(contam, aes(x=p)) + 
  labs(title='Decontam Method: Prevalence', subtitle='Organized by Prevalence', x = 'decontam Score', y='Number of ASVs') + 
  geom_histogram(binwidth=0.02) +
  ylim(0, 650)

histo
histo + facet_wrap(~Prevalence)

i.top10 <- order(contam$p, decreasing=FALSE)[1:10]
contam[i.top10,]

cat("Number of contaminants (method = prevalence, threshold = 0.1):", sum(contam$contaminant))



``` 

Remove low-depth samples and then remove contaminant ASVs
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Filter out samples with <2k reads
keep_samples <- rownames(sample_metadata.filt)[sample_metadata.filt$LibrarySize >= 2000]
final_samples <- intersect(keep_samples, rownames(ASV_feature_table.filt))
sample_metadata.filt_2 <- sample_metadata.filt[final_samples, , drop = FALSE]
ASV_feature_table.filt_2 <- ASV_feature_table.filt[final_samples, , drop = FALSE]
taxa_annotations.filt_2 <- taxa_annotations.filt[match(colnames(ASV_feature_table.filt_2), rownames(taxa_annotations.filt)), ]
 
# Remove contaminants
ASV_feature_table_decontam <- ASV_feature_table.filt_2[, !contam$contaminant]
# Subset metadata and taxa table 
sample_metadata_decontam <- sample_metadata.filt[match(rownames(ASV_feature_table_decontam), rownames(sample_metadata.filt)), ]
taxa_annotations_decontam <- taxa_annotations.filt[match(colnames(ASV_feature_table_decontam), rownames(taxa_annotations.filt)), ]


ps_og <- phyloseq(
  otu_table(as.matrix(ASV_feature_table_og), taxa_are_rows = FALSE), 
  sample_data(sample_metadata_subset), 
  tax_table(as.matrix(taxa_annotations_og))
)

ps_og_no_neg <- prune_samples(
  grepl("Skin", sample_data(ps_og)$Location_Type),
  ps_og
)


ps_filtered <- phyloseq(
  otu_table(as.matrix(ASV_feature_table.filt_2), taxa_are_rows = FALSE), 
  sample_data(sample_metadata.filt_2), 
  tax_table(as.matrix(taxa_annotations.filt_2))
)

ps_filtered_no_neg <- prune_samples(
  grepl("Skin", sample_data(ps_filtered)$Location_Type),
  ps_filtered
)


ps_decontam <- phyloseq(
  otu_table(as.matrix(ASV_feature_table_decontam), taxa_are_rows = FALSE), 
  sample_data(sample_metadata_decontam), 
  tax_table(as.matrix(taxa_annotations_decontam))
)

ps_decontam_no_neg <- prune_samples(
  grepl("Skin", sample_data(ps_decontam)$Location_Type),
  ps_decontam
)
 
 
# Baseline Positive Samples
asvs_og        <- ntaxa(ps_og_no_neg)
reads_og       <- sum(sample_sums(ps_og_no_neg))

# Filtered Positive Samples
asvs_filtered  <- ntaxa(ps_filtered_no_neg)
reads_filtered <- sum(sample_sums(ps_filtered_no_neg))

# Decontaminated Positive Samples
asvs_decontam  <- ntaxa(ps_decontam_no_neg)
reads_decontam <- sum(sample_sums(ps_decontam_no_neg))


 
pct_asvs_filt   <- (asvs_filtered / asvs_og) * 100
pct_reads_filt  <- (reads_filtered / reads_og) * 100

pct_asvs_dec    <- (asvs_decontam / asvs_og) * 100
pct_reads_dec   <- (reads_decontam / reads_og) * 100


 
pipeline_summary <- data.frame(
  Pipeline_Step = c(
    "Original Positive Samples", 
    "Filtered Positive Samples", 
    "Decontaminated Positive Samples"
  ),
  Total_ASVs = c(
    asvs_og, 
    asvs_filtered, 
    asvs_decontam
  ),
  ASVs_Retained = c(
    "100.00%", 
    paste0(round(pct_asvs_filt, 2), "%"), 
    paste0(round(pct_asvs_dec, 2), "%")
  ),
  Total_Reads = c(
    reads_og, 
    reads_filtered, 
    reads_decontam
  ),
  Reads_Retained = c(
    "100.00%", 
    paste0(round(pct_reads_filt, 2), "%"), 
    paste0(round(pct_reads_dec, 2), "%")
  )
)


 
cat("### Positive Sample Processing Summary\n\n")

kable(
  pipeline_summary, 
  format = "markdown", 
  col.names = c("Pipeline Step", "Total ASVs", "% ASVs Retained", "Total Reads", "% Reads Retained"),
  align = c("l", "r", "r", "r", "r")
)



 
save(ASV_feature_table_decontam, taxa_annotations_decontam, sample_metadata_decontam, 
     file = paste0("../02_output_files/01_filtered_ASV_tables/allseqruns_filtered_canine_skin_microbiome.rda"))

 
saveRDS(ps_decontam_no_neg, 
        file = paste0("../02_output_files/01_filtered_ASV_tables/allseqruns_ASV_feature_table_decontam.rds"))

```

## Inspect library size distribution (read depths) for samples  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
# 1. Verify all samples match exactly
if(!setequal(rownames(sample_metadata_decontam), rownames(ASV_feature_table_decontam))) {
  stop("Mismatched samples between metadata and ASV table.")
}

# 2. Force ASV table rows into the exact same order as metadata
ASV_feature_table_decontam <- ASV_feature_table_decontam[rownames(sample_metadata_decontam), ]

# 3. Calculate Library Size safely
sample_metadata_decontam$LibrarySize <- rowSums(ASV_feature_table_decontam)

options(scipen = 999)
 
sample_metadata_decontam2 <- sample_metadata_decontam |>
  dplyr::mutate(front = Sample_Site %in% c("Air", "Kit", "Chlorhex")) |>
  dplyr::arrange(front)

ggplot(sample_metadata_decontam2, aes(x = rank(LibrarySize), y = LibrarySize, color = Sample_Site)) + 
   scale_y_continuous(
    limits = c(0, 175000),
    breaks = seq(0, 175000, by = 50000),
    labels = scales::comma
  ) +
  geom_point(size = 2.0) + 
  scale_color_manual(values = c(
    "Chlorhex" = "orange",
    "Kit" = "green",
    "Surgical" = "blue",
    "Control" = "purple",
    "Air" = "red"
  ),
    labels = c(
      "Chlorhex" = "Chlorhex",
      "Kit" = "Kit",
      "Surgical" = "Surgical (Skin)",
      "Control" = "Control (Skin)",
      "Air" = "Air"
    ))+
  theme(
    legend.position = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.key = element_rect(fill = NA, color = NA)
  ) + guides(color = guide_legend(override.aes = list(size = 3)))

sort(sample_metadata_decontam$LibrarySize)[1:30]
table(sample_metadata_decontam$Sample_Site, sample_metadata_decontam$NegativeControl)

```

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```
 