---
title: "Canine Skin Microbiome: Microbial Diversity Analysis "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

**Purpose:**
Analyze microbial diversity within and across timepoints, comparing control and surgical skin sites. 

# Load files

```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(knitr)
library(FSA)
library(tidyverse)
library(vegan)
library(phyloseq)
library(ggplot2)
library(ggtree)
library(scales)  
library(RColorBrewer)
library(segmented)
library(stringr)
library(pairwiseAdonis)
library(ggpubr)
library(changepoint)
library(dplyr)
library(lme4)
```

## Create phyloseq objects
1. Load each output file for each sequencing run processed in decontam 
2. Filter out negative-controls (only Skin-Control and Skin-Surgical sites)
3. Merge all phyloseq objects for downstream analysis
```{r, message=FALSE, warning=FALSE, echo=TRUE}
load("../02_output_files/01_filtered_ASV_tables/CH1_filtered_canine_skin_microbiome.rda")

ps1 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps1 <- prune_samples(
  grepl("Skin", sample_data(ps1)$Location_Type),
  ps1
)
load("../02_output_files/01_filtered_ASV_tables/CH2_filtered_canine_skin_microbiome.rda")

ps2 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps2 <- prune_samples(
  grepl("Skin", sample_data(ps2)$Location_Type),
  ps2
)
load("../02_output_files/01_filtered_ASV_tables/CH3_filtered_canine_skin_microbiome.rda")

ps3 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps3 <- prune_samples(
  grepl("Skin", sample_data(ps3)$Location_Type),
  ps3
)
load("../02_output_files/01_filtered_ASV_tables/CH4_filtered_canine_skin_microbiome.rda")

ps4 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
  ps4 <- prune_samples(
  grepl("Skin", sample_data(ps4)$Location_Type),
  ps4
)
ps <- merge_phyloseq(ps1, ps2, ps3, ps4)
ps <- prune_samples(
  grepl("Skin", sample_data(ps)$Location_Type),
  ps
)
```

# ASV + Sample Curation 

## Remove Unknown ASVs
Remove ASVs with uncharacterized Phylum and output how many were removed across the sequencing runs. 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
otu_matrix <- as(otu_table(ps), "matrix")

# Get total ASV count per Sample_Plate
# Sample_Plate indicates sequencing run 
total_asvs_per_plate <- data.frame(
  Sample_Plate = sample_data(ps)$Sample_Plate,
  as.data.frame(otu_matrix)
) %>%
  group_by(Sample_Plate) %>%
  summarise(
    Total_ASVs = sum(colSums(across(-where(is.character))) > 0),
    .groups = "drop"
  )

unknown_asvs <- rownames(tax_table(ps))[
  is.na(tax_table(ps)[, "Phylum"]) | tax_table(ps)[, "Phylum"] %in% c("", "uncharacterized")
]


otu_unknown <- otu_matrix[, unknown_asvs, drop = FALSE]

# Provide summary for uncharacterized ASV count per seq. run
out <- data.frame(
  Sample_Plate = sample_data(ps)$Sample_Plate,
  as.data.frame(otu_unknown)
) %>%
  group_by(Sample_Plate) %>%
  summarise(
    Unknown_ASVs_Count = sum(colSums(across(-where(is.character))) > 0),
    .groups = "drop"
  ) %>%
  left_join(total_asvs_per_plate, by = "Sample_Plate") %>%
  mutate(
    Percent_Removed = 100 * Unknown_ASVs_Count / Total_ASVs
  )
print(out)
 
# Subset phyloseq object for downstream analysis
ps_1 <- subset_taxa(ps, !is.na(Phylum) & !Phylum %in% c("", "uncharacterized"))
```

### Check for low-prevalence taxa
```{r, fig.width=11, fig.height=8}
# Compute prevalence for each ASV 
prevdf = apply(X = otu_table(ps_1),
               MARGIN = ifelse(taxa_are_rows(ps_1), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})
 
# Add taxa and total read counts to prevalence df
prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps_1),
                    tax_table(ps_1))

# Are there phyla that are comprised of mostly low-prevalence features? 
# Compute the total and average prevalences of the features in each phylum.
phylum_summary <- plyr::ddply(prevdf, "Phylum", function(df1){
  cbind(mean(df1$Prevalence), sum(df1$Prevalence))
})
colnames(phylum_summary) <- c("Phylum", "Mean_Prevalence", "Sum_Prevalence")

# Phyla by Total ASV Prevalence 
phylum_summary_sorted <- phylum_summary[order(-phylum_summary$Sum_Prevalence), ]
print(phylum_summary_sorted, row.names = FALSE)
ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(ps_1),color=Phylum)) +
  ggtitle("Prevalence of Phyla Across Skin-Swab Samples", subtitle="dashed line = 5% prevalence") +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```
### At each timepoint, keep only samples with control and surgical pairs  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Extract sample data
sample_check <- as(sample_data(ps_1), "data.frame")

# Set timepoint order 
day_levels <- c(
  "Admission",
  "Day 0a - Pre Surgery",
  "Day 0c - Post Surgery",
  "Day 1",
  "Day 3",
  "Week 1",
  "Week 2",
  "Week 3",
  "Week 4",
  "Week 5",
  "Week 6"
)
sample_check$Day_Order <- factor(sample_check$Day_Order, levels = day_levels)
sample_check$Sample_Site <- factor(sample_check$Sample_Site, levels = c("Control", "Surgical"))

# 2. Count skin samples across timepoints
count_df <- sample_check %>% 
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>% 
  distinct() %>% 
  group_by(Day_Order, Sample_Site) %>% 
  summarise(n = n(), .groups = "drop")
 

count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)
count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Control", "Surgical"))

p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Control" = "blue", "Surgical" = "purple")) +
  labs(
    x = "Timepoint",
    y = "Number of Samples",
    fill = "Skin Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)

# Check which dogs are missing samples (either control or surgical) at each timepoint
missing_sample_unique <- sample_check %>% 
  group_by(Individual, Day_Order) %>% 
  summarise(
    site_absent = if (all(c("Control", "Surgical") %in% Sample_Site)) NA_character_ 
                  else paste(setdiff(c("Control", "Surgical"), unique(Sample_Site)), collapse = ", "), 
    Sample_Unique = paste(sort(unique(Sample_Unique)), collapse = ", "), 
    n_sites = n_distinct(Sample_Site), 
    .groups = "drop"
  ) %>% 
  filter(n_sites < 2) %>% 
  dplyr::select(Individual, Day_Order, site_absent, Sample_Unique) %>%
  # Arrange by Individual and chronological factor timepoint
  arrange(Individual, Day_Order)

print(missing_sample_unique)

# Keep only samples that have control and surgical at each Day_Order
paired_groups <- sample_check %>%
  group_by(Individual, Day_Order) %>%
  summarize(n_sites = n_distinct(Sample_Site), .groups = "drop") %>%
  filter(n_sites == 2) %>%
  transmute(Individual, Day_Order)
filtered_sample_data <- sample_check %>%
  inner_join(paired_groups, by = c("Individual", "Day_Order"))

 
# Prune phyloseq object so it only contains these remaining samples 
ps_2 <- prune_samples(filtered_sample_data$SampleID, ps_1)
 

# Count unique samples by Day_Order and Sample_Site
count_df <- filtered_sample_data %>%
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>%
  distinct() %>%
  group_by(Day_Order, Sample_Site) %>%
  summarise(n = n(), .groups = "drop")


count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)


count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Control", "Surgical"))

# Updated after removing dog-timepoint samples with only one site
p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Control" = "blue", "Surgical" = "purple")) +
  labs(
    x = "Timepoint",
    y = "Number of Samples",
    fill = "Skin Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)
```
### Generate final stats per seq. run and sample
```{r, message=FALSE, warning=FALSE, echo=TRUE}
out_dir <- "../02_output_files/02_image_files"

# Get reads and metadata per sample
reads_per_sample <- sample_sums(ps_2)
md <- as(sample_data(ps_2), "data.frame")
 
sample_names_ps <- sample_names(ps_2)

# Ensure sample order matches
if (!all(sample_names_ps == rownames(md))) {
  md <- md[sample_names_ps, ]
}

# Create dataframe with sample-level stats
df <- data.frame(
  sample = sample_names_ps,
  total_reads = reads_per_sample,
  Sample_Plate = md$Sample_Plate,
  Individual = md$Individual,
  stringsAsFactors = FALSE
)

# Calculate total reads and ASVs per sample and add to metadata
otu <- otu_table(ps_2)
if (taxa_are_rows(otu)) {
  total_asvs <- colSums(otu > 0)
} else {
  total_asvs <- rowSums(otu > 0)
}

total_reads <- sample_sums(ps_2)
 
md$total_asv <- total_asvs
md$total_reads <- total_reads

sample_data(ps_2) <- sample_data(md)

# Aggregate reads and ASVs per Individual (summing across all samples from same dog)
by_ind <- aggregate(
  cbind(total_reads, total_asv) ~ Individual + Sample_Plate,
  data = md,
  FUN = sum
)

# Boxplot for total reads by Sample_Plate (each point = one Individual)
p_reads <- ggplot(by_ind, aes(x = Sample_Plate, y = total_reads, fill = Sample_Plate)) +
  geom_boxplot() +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "K"),
    name = "Total Reads (per Individual, in thousands)"
  ) +
  labs(
    title = "Total Reads per Individual by Sample Plate",
    x = "Sample Plate"
  ) +
  theme_minimal()

out_reads <- paste0(out_dir, "/total_reads_by_sample_plate.png")
ggplot2::ggsave(out_reads, p_reads, width = 8, height = 6)
p_reads

# Boxplot for total ASVs by Sample_Plate (each point = one Individual)
p_asv <- ggplot(by_ind, aes(x = Sample_Plate, y = total_asv, fill = Sample_Plate)) +
  geom_boxplot() +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "k"),
    name = "Total ASVs (per Individual, in thousands)"
  ) +
  labs(
    title = "Total ASVs per Individual by Sample Plate",
    x = "Sample Plate"
  ) +
  theme_minimal()
p_asv

out_asv <- paste0(out_dir, "/total_asvs_by_sample_plate.png")
ggplot2::ggsave(out_asv, p_asv, width = 8, height = 6)

print(by_ind)

```

# Shannon Alpha-Diversity   
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
# Calculate alpha-diversity using Shannon Diversity Index for each skin sample 
alphadiv_plot_data <- estimate_richness(ps_2, measures = "Shannon") %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(as.data.frame(sample_data(ps_2)), by = "SampleID") %>% 
  group_by(Individual) %>% 
  mutate(Surgical_Location = Sample_Location[Sample_Site == "Surgical"][1]) %>% 
  ungroup() %>% 
  mutate(Surgical_Location = factor(Surgical_Location)) %>% 
  arrange(Surgical_Location, Individual)
ggplot(alphadiv_plot_data, aes(x = factor(Day_Order), y = Shannon, color = Sample_Site, group = Sample_Site)) + 
  geom_point(size = 1, position = position_dodge(width = 0.3)) + 
  geom_line(position = position_dodge(width = 0.3)) + 
  facet_wrap(
    Surgical_Location ~ Individual, 
    labeller = label_both
  ) + 
  scale_color_manual(values = c("Control" = "purple", "Surgical" = "blue")) + 
  theme_bw() + 
  labs(
    title = "Shannon Diversity Across Timepoints per Dog",
    x = "Timepoint",
    y = "Shannon Index",
    color = "Skin Site"
  ) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

# Box Plot of Shannon Index values across timepoints (dog's labeled with which type of surgery)
plot_richness(ps_2, 
              x = "Day_Order", 
              measures = "Shannon") + 
  geom_boxplot(aes(fill = Sample_Site), show.legend = TRUE) +
  facet_grid(. ~ Sample_Site) + 
  scale_fill_manual(values = c("Control" = "purple", "Surgical" = "blue")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(title = "Shannon Diversity Across Timepoints",
       x = "Timepoint",
       fill = "Skin Site")

# Summary of alpha diversity across timepoints 
alphadiv_all <- estimate_richness(ps_2, measures = c("Shannon")) %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(as.data.frame(sample_data(ps_2)), by = "SampleID")  
alphadiv_summary <- alphadiv_all %>%
  summarise(
    mean_shannon = mean(Shannon, na.rm = TRUE),
    sd_shannon   = sd(Shannon, na.rm = TRUE),
    n            = n(),
    .by          = c(Day_Order, Sample_Site)  
  ) 
alphadiv_summary
```

# Relative Abundances
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
# Normalization using relative abundances
ps_rel <- transform_sample_counts(ps_2, function(x) x / sum(x))
individual_list <- unique(sample_data(ps_rel)$Individual)

# create color palette so phyla is the same color across all individual dog plots 
all_phyla <- tax_table(ps_rel)[, "Phylum"] %>% as.vector() %>% unique() %>% na.omit() %>% sort()
n_phyla <- length(all_phyla)
base_cols <- brewer.pal(min(n_phyla, 12), "Paired")
phylum_cols_vec <- colorRampPalette(base_cols)(n_phyla)
phylum_colors <- setNames(phylum_cols_vec, all_phyla)

# Generate per dog plots showing (top 20) phyla diversity across timepoints
for (ind in individual_list) {
  ps_ind <- subset_samples(ps_rel, Individual == ind)
    plot_df <- psmelt(ps_ind) %>% 
    filter(Sample_Site %in% c("Control", "Surgical")) %>% 
    mutate(
      Surgical_Location = unique(Sample_Location[Sample_Site == "Surgical" & !is.na(Sample_Location)])[1]
    ) %>% 
    mutate(Surgical_Location = factor(Surgical_Location))
  
  # Identify top 20 phyla specific to this dog's profile
  top20_ind <- plot_df %>% 
    group_by(Phylum) %>% 
    summarise(Total_Abundance = sum(Abundance), .groups = "drop") %>% 
    arrange(desc(Total_Abundance)) %>% 
    slice_head(n = 20) %>% 
    pull(Phylum)
  
  phylum_levels_ind <- c(top20_ind, "Other")
   
  plot_df <- plot_df %>% mutate(
    Phylum = ifelse(Phylum %in% top20_ind, Phylum, "Other"),
    Phylum = factor(Phylum, levels = phylum_levels_ind),
    Sample_Site = factor(Sample_Site, levels = c("Control", "Surgical")),
    Surgical_Location = factor(Surgical_Location, levels = c("Abdomen", "Stifle")),
    Day_Order = factor(Day_Order)
  )
  
  local_cols <- phylum_colors[top20_ind]
  missing_indices <- is.na(local_cols)
  if(any(missing_indices)) {
    fallback_pool <- brewer.pal(12, "Paired")
    local_cols[missing_indices] <- sample(fallback_pool, sum(missing_indices), replace = TRUE)
  }
  values_ind <- c(local_cols, "Other" = "grey70")
  
  p <- ggplot(plot_df, aes(x = Day_Order, y = Abundance, fill = Phylum)) + 
    geom_col(position = "stack", width = 0.7) +  
    facet_wrap(~Sample_Site, nrow = 1) +          
    scale_fill_manual(
      values = values_ind,
      breaks = phylum_levels_ind,
      na.value = "black"
    ) + 
    theme_minimal() + 
    theme(
      axis.text.x = element_text(angle = 75, hjust = 1),
      panel.spacing = unit(1.5, "lines")        
    ) + 
    labs(
      title = paste0("Relative Abundance Across Timepoints by Phyla"),
      subtitle = paste0("Dog: ", ind, " Surgical Location: ", unique(plot_df$Surgical_Location)),
      x = "Timepoint",
      y = "Relative Abundance"
    )
  print(p)
}

 
```

# Bray-Curtis Beta-Diversity + PCoA
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# Generate bray-curtis distance matrix using ASV relative abundances
metadata <- data.frame(sample_data(ps_rel))
bray_dist <- phyloseq::distance(ps_rel, method = "bray")

set.seed(100)
# generate PCoA using bray-curtis distance matrix
bray_pcoa <- ordinate(physeq = ps_rel, method = "PCoA", distance = bray_dist)
```

# Statistical Analysis 

## Sequencing Run Batch Effect

### PERMANOVA 
While PERMANOVA tests if the centers of sequencing clusters are in different locations, PERMDISP tests if the sizes of the clusters are significantly different, regardless of where they are on the PCoA plot.
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Draw ellipses around samples from same sequencing run to visualize batch effects
p1 <- plot_ordination(ps_rel, bray_pcoa, color = "Sample_Plate")
p1 + ggtitle("PCoA of Beta-Diversity (Bray-Curtis)", subtitle = "Ellipses encircle samples belonging to the same sequencing run.") +
  geom_point(size = 1.5) + 
  theme_bw() +
  stat_ellipse(aes(fill = Sample_Plate, group = Sample_Plate), alpha = 0.2, geom = "polygon")

# PERMANOVA test  
adonis2(bray_dist ~ Sample_Plate, data = metadata, permutations = 999)


# PERMDISP2 test (multivariate analogue of Levene's test)
betadisper_result <- betadisper(bray_dist, group = metadata$Sample_Plate)

average_by_plate <- data.frame(
  Plate = metadata$Sample_Plate,
  Dist_to_Centroid = betadisper_result$distances
)

average_by_plate_table <- average_by_plate %>%
  group_by(Plate) %>%
  summarize(
    Mean_Dist = mean(Dist_to_Centroid, na.rm = TRUE),
    SD_Dist   = sd(Dist_to_Centroid, na.rm = TRUE),
    CV_Percent = (SD_Dist / Mean_Dist *100) 
  )

print(average_by_plate_table)

# Extract the PCoA coordinates and centroids
df_points <- as.data.frame(scores(betadisper_result, display = "sites"))
df_points$Sample_Plate <- metadata$Sample_Plate

df_centroids <- as.data.frame(scores(betadisper_result, display = "centroids"))
df_centroids$Sample_Plate <- rownames(df_centroids)

# Pairwise post-hoc comparisons between sequencing runs
print(TukeyHSD(betadisper_result))
```

### PERMDISP2   
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# PERMDISP2 test (multivariate analogue of Levene's test)
betadisper_result <- betadisper(bray_dist, group = metadata$Sample_Plate)

average_by_plate <- data.frame(
  Plate = metadata$Sample_Plate,
  Dist_to_Centroid = betadisper_result$distances
)

average_by_plate_table <- average_by_plate %>%
  group_by(Plate) %>%
  summarize(
    Mean_Dist = mean(Dist_to_Centroid, na.rm = TRUE),
    SD_Dist   = sd(Dist_to_Centroid, na.rm = TRUE),
    CV_Percent = (SD_Dist / Mean_Dist *100) 
  )

print(average_by_plate_table)

# Extract the PCoA coordinates and centroids
df_points <- as.data.frame(scores(betadisper_result, display = "sites"))
df_points$Sample_Plate <- metadata$Sample_Plate

df_centroids <- as.data.frame(scores(betadisper_result, display = "centroids"))
df_centroids$Sample_Plate <- rownames(df_centroids)

set.seed(100)
print(permutest(betadisper_result, perm=999))
# Pairwise post-hoc comparisons between sequencing runs
print(TukeyHSD(betadisper_result))
```

## Timepoints after Surgery

### PERMANOVA
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(100)
# Run PERMANOVA constraining permutations within plates to account for sequencing batch effects
global_permanova_strata <- adonis2(
  bray_dist ~ Day_Order, 
  data = metadata, 
  strata = metadata$Sample_Plate,
  permutations = 999
)

print(global_permanova_strata)
```

### Pairwise PERMANOVA: Control-Surgical Bray-Curtis Distances 
```{r,fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# Extract pairwise differences between control annd surgical sites for each dog-timepoint
bray_matrix <- as.matrix(bray_dist)
paired_distances <- metadata %>%
  group_by(Dog_ID, Day_Order) %>%
  filter(all(c("Control", "Surgical") %in% Sample_Site)) %>%
  ungroup() %>%
  group_by(Dog_ID, Day_Order) %>%
  summarise(
    Control_Sample = SampleID[Sample_Site == "Control"][1],
    Surgical_Sample = SampleID[Sample_Site == "Surgical"][1],
    Bray_Distance = bray_matrix[Control_Sample, Surgical_Sample],
    Sample_Plate = Sample_Plate[1],
    Day_Order = Day_Order[1],   
    .groups = "drop"
  )


print("Group means by Day_Order:")
print(paired_distances %>%
  group_by(Day_Order) %>%
  summarise(
    Mean_Bray = mean(Bray_Distance),
    Median_Bray = median(Bray_Distance),
    SD_Bray = sd(Bray_Distance),
    N = n(),
    .groups = "drop"
  ))

# Plot Distribution of Bray-Curtis Distances
histogram_distances <- ggplot(paired_distances, aes(x = Bray_Distance)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of Bray-Curtis Distances",
    x = "Bray-Curtis Distance",
    y = "Number of Control-Surgical Pairwise Comparisons (n = 208)"
  )
print(histogram_distances)

set.seed(100)
pairwise_results <- pairwise.adonis2(
  bray_dist ~ Day_Order,
  strata = "Sample_Plate",
  data = metadata,
  nperm = 999)

p_values <- sapply(names(pairwise_results)[-1], function(x) {
  pairwise_results[[x]][["Pr(>F)"]][1]
})

# Benjamini-Hochberg (BH) adjustment since we are testing multiple comparisons
p_adjusted <- p.adjust(p_values, method = "BH")

pairwise_df <- data.frame(
  Comparison = names(p_values),
  p_value    = p_values,
  p_adj      = p_adjusted,
  row.names  = NULL
)

groups <- str_split_fixed(pairwise_df$Comparison, "_vs_", 2)

pairwise_df$Group1 <- groups[, 1]
pairwise_df$Group2 <- groups[, 2]

pairwise_df <- pairwise_df[, c("Group1", "Group2", "p_value", "p_adj")]

alpha <- 0.05
pairwise_tidy <- pairwise_df %>%
  filter(!is.na(p_adj)) %>% 
  mutate(
    Significant = ifelse(p_adj <= alpha, "Yes", "No"),
    Stars = case_when(
      p_adj <= 0.001 ~ "***",
      p_adj <= 0.01  ~ "**",
      p_adj <= 0.05  ~ "*",
      p_adj <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p_adj = round(p_adj, 4) ) %>%
  arrange(p_adj) 

# Filter out insignificant pairwise results
sig_df <- pairwise_tidy %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,    
    label = Stars
  )
x_levels <- levels(factor(paired_distances$Day_Order))

sig_df <- sig_df %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p_value) %>%
  mutate(
    y.position = max(paired_distances$Bray_Distance, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )
print(sig_df)
median_df <- paired_distances %>%
  group_by(Day_Order) %>%
  summarise(median_dist = median(Bray_Distance, na.rm = TRUE), .groups = "drop")

ggplot(paired_distances, aes(x = factor(Day_Order), y = Bray_Distance)) +
  geom_violin(fill = "lightblue", trim = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +
  geom_point(data = median_df, aes(x = Day_Order, y = median_dist),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = median_df, aes(x = Day_Order, y = median_dist, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +
  stat_pvalue_manual(
    sig_df,
    label = "label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.5,
    size = 5
  ) +
  labs(
    title = "Control vs. Surgical Site Bray-Curtis Distances Across Timepoints",
    subtitle = "Trend line indicating median bray-curtis distances",
    x = "Timepoint",
    y = "Bray-Curtis Distance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )


```


```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}

### Fit Point-Change Model to Pairwise Bray-Curtis Dissimilarities 
# Change timepoints to number of days since surgery
paired_distances_numeric <- paired_distances %>%
  mutate(
    Day_Order = case_when(
      Day_Order == "Admission" ~ -1,            
      Day_Order == "Day 0a - Pre Surgery" ~ -0.5,
      Day_Order == "Day 0c - Post Surgery" ~ 0.5,
      Day_Order == "Day 1" ~ 1,
      Day_Order == "Day 3" ~ 3,
      Day_Order == "Week 1" ~ 7,
      Day_Order == "Week 2" ~ 14,
      Day_Order == "Week 3" ~ 21,
      Day_Order == "Week 4" ~ 28,
      Day_Order == "Week 5" ~ 35,
      Day_Order == "Week 6" ~ 42,
      TRUE ~ NA_real_  
    )
  )

paired_distances_numeric$Dog_ID <- factor(paired_distances_numeric$Dog_ID)
paired_distances_numeric$Sample_Plate <- factor(paired_distances_numeric$Sample_Plate)

tp_med <- paired_distances %>%
  group_by(Day_Order) %>%
  summarise(med_bray = median(Bray_Distance, na.rm = TRUE), .groups = "drop") %>%
  arrange(Day_Order)

x <- tp_med$med_bray

fit_mean <- cpt.mean(x, method = "PELT", penalty = "MBIC")
fit_var  <- cpt.var(x,  method = "PELT", penalty = "MBIC")
fit_both <- cpt.meanvar(x, method = "PELT", penalty = "MBIC")

cpts(fit_mean)
cpts(fit_var)
cpts(fit_both)

plot(fit_both, main = "Changepoint on median Bray-Curtis by timepoint")


# Check for specific changes between before and after surgery 
bp <- -0.5   # -0.5 = Day 0c - Pre Surgery
paired_distances_numeric <- paired_distances_numeric %>%
  mutate(
    Day_c = Day_Order - bp,
    post_bp = ifelse(Day_Order > bp, 1, 0),
    Day_after_bp = Day_c * post_bp
  )

lm_changepoint_pre_surgery <- lmer(
  Bray_Distance ~ Day_c + Day_after_bp + Sample_Plate + (1 | Dog_ID),
  data = paired_distances_numeric
)


bp <- 0.5   # 0.5 = Day 0c - Post Surgery
paired_distances_numeric <- paired_distances_numeric %>%
  mutate(
    Day_c = Day_Order - bp,
    post_bp = ifelse(Day_Order > bp, 1, 0),
    Day_after_bp = Day_c * post_bp
  )

lm_changepoint_post_surgery <- lmer(
  Bray_Distance ~ Day_c + Day_after_bp + Sample_Plate + (1 | Dog_ID),
  data = paired_distances_numeric
)

print("Changepoint: Pre-Surgery")
summary(lm_changepoint_pre_surgery)

print("Changepoint: Post-Surgery")
summary(lm_changepoint_post_surgery)

anova(lm_changepoint_pre_surgery, lm_changepoint_post_surgery)



newdat <- expand.grid(
  Day_Order = seq(min(paired_distances_numeric$Day_Order, na.rm = TRUE),
                  max(paired_distances_numeric$Day_Order, na.rm = TRUE),
                  length.out = 100),
  Sample_Plate = levels(factor(paired_distances_numeric$Sample_Plate))
)

bp <- 0.5
newdat <- newdat %>%
  mutate(
    Day_c = Day_Order - bp,
    post_bp = ifelse(Day_Order > bp, 1, 0),
    Day_after_bp = Day_c * post_bp
  )

X <- model.matrix(~ Day_c + Day_after_bp + Sample_Plate, data = newdat)
beta <- fixef(lm_changepoint_post_surgery)
V <- vcov(lm_changepoint_post_surgery)

newdat$fit <- as.numeric(X %*% beta)
newdat$se <- sqrt(diag(X %*% V %*% t(X)))
newdat$lwr <- newdat$fit - 1.96 * newdat$se
newdat$upr <- newdat$fit + 1.96 * newdat$se

ggplot(newdat, aes(x = Day_Order, y = fit, color = Sample_Plate, fill = Sample_Plate)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, color = NA) +
  theme_classic() +
  labs(x = "Day Order", y = "Predicted Bray Distance")


ggplot() +
  geom_point(data = paired_distances_numeric,
             aes(x = Day_Order, y = Bray_Distance, color = Sample_Plate),
             alpha = 0.5) +
  geom_line(data = newdat,
            aes(x = Day_Order, y = fit, color = Sample_Plate),
            linewidth = 1) +
  geom_ribbon(data = newdat,
              aes(x = Day_Order, ymin = lwr, ymax = upr, fill = Sample_Plate),
              alpha = 0.2, color = NA) +
  theme_classic() +
  labs(x = "Day Order", y = "Bray Distance")


bp <- 0.5

pred_grid <- data.frame(
  Day_Order = seq(min(paired_distances_numeric$Day_Order, na.rm = TRUE),
                  max(paired_distances_numeric$Day_Order, na.rm = TRUE),
                  by = 0.5),
  Sample_Plate = "CHLib1",
  Dog_ID = paired_distances_numeric$Dog_ID[1]
)

pred_grid$Day_c <- pred_grid$Day_Order - bp
pred_grid$post_bp <- ifelse(pred_grid$Day_Order > bp, 1, 0)
pred_grid$Day_after_bp <- pred_grid$Day_c * pred_grid$post_bp

pred_grid$fit <- predict(lm_changepoint_post_surgery, newdata = pred_grid, re.form = NA)

target_day <- -1
target_fit <- pred_grid$fit[which.min(abs(pred_grid$Day_Order - target_day))]

tol <- 42   # adjust as needed
matches <- pred_grid[abs(pred_grid$fit - target_fit) <= tol, ]

matches
```

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```
