---
title: "Canine Skin Microbiome: Microbial Diversity Analysis "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

**Purpose:**
Analyze microbial diversity within and across timepoints, comparing control and surgical skin sites. 

# Load files

```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(knitr)
library(FSA)
library(tidyverse)
library(vegan)
library(phyloseq)
library(ggplot2)
library(ggtree)
library(scales)  
library(RColorBrewer)
library(segmented)
library(stringr)
library(pairwiseAdonis)
```

## Create phyloseq objects
1. Load each output file for each sequencing run processed in decontam 
2. Filter out negative-controls (only Skin-Control and Skin-Surgical sites)
3. Merge all phyloseq objects for downstream analysis
```{r, message=FALSE, warning=FALSE, echo=TRUE}
load("../02_output_files/01_filtered_ASV_tables/CH1_filtered_canine_skin_microbiome.rda")

ps1 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps1 <- prune_samples(
  grepl("Skin", sample_data(ps1)$Location_Type),
  ps1
)
load("../02_output_files/01_filtered_ASV_tables/CH2_filtered_canine_skin_microbiome.rda")

ps2 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps2 <- prune_samples(
  grepl("Skin", sample_data(ps2)$Location_Type),
  ps2
)
load("../02_output_files/01_filtered_ASV_tables/CH3_filtered_canine_skin_microbiome.rda")

ps3 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps3 <- prune_samples(
  grepl("Skin", sample_data(ps3)$Location_Type),
  ps3
)
load("../02_output_files/01_filtered_ASV_tables/CH4_filtered_canine_skin_microbiome.rda")

ps4 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
  ps4 <- prune_samples(
  grepl("Skin", sample_data(ps4)$Location_Type),
  ps4
)
ps <- merge_phyloseq(ps1, ps2, ps3, ps4)
ps <- prune_samples(
  grepl("Skin", sample_data(ps)$Location_Type),
  ps
)
```

# ASV + Sample Curation 

## Remove Unknown ASVs
Remove ASVs with uncharacterized Phylum and output how many were removed across the sequencing runs. 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
otu_matrix <- as(otu_table(ps), "matrix")

# Get total ASV count per Sample_Plate
# Sample_Plate indicates sequencing run 
total_asvs_per_plate <- data.frame(
  Sample_Plate = sample_data(ps)$Sample_Plate,
  as.data.frame(otu_matrix)
) %>%
  group_by(Sample_Plate) %>%
  summarise(
    Total_ASVs = sum(colSums(across(-where(is.character))) > 0),
    .groups = "drop"
  )

unknown_asvs <- rownames(tax_table(ps))[
  is.na(tax_table(ps)[, "Phylum"]) | tax_table(ps)[, "Phylum"] %in% c("", "uncharacterized")
]


otu_unknown <- otu_matrix[, unknown_asvs, drop = FALSE]

# Provide summary for uncharacterized ASV count per seq. run
out <- data.frame(
  Sample_Plate = sample_data(ps)$Sample_Plate,
  as.data.frame(otu_unknown)
) %>%
  group_by(Sample_Plate) %>%
  summarise(
    Unknown_ASVs_Count = sum(colSums(across(-where(is.character))) > 0),
    .groups = "drop"
  ) %>%
  left_join(total_asvs_per_plate, by = "Sample_Plate") %>%
  mutate(
    Percent_Removed = 100 * Unknown_ASVs_Count / Total_ASVs
  )
print(out)
 
# Subset phyloseq object for downstream analysis
ps_1 <- subset_taxa(ps, !is.na(Phylum) & !Phylum %in% c("", "uncharacterized"))
```

### Check for low-prevalent taxa
```{r, fig.width=11, fig.height=8}
# Compute prevalence for each ASV 
prevdf = apply(X = otu_table(ps_1),
               MARGIN = ifelse(taxa_are_rows(ps_1), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})
 
# Add taxa and total read counts to prevalence df
prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps_1),
                    tax_table(ps_1))

# Are there phyla that are comprised of mostly low-prevalence features? 
# Compute the total and average prevalences of the features in each phylum.
phylum_summary <- plyr::ddply(prevdf, "Phylum", function(df1){
  cbind(mean(df1$Prevalence), sum(df1$Prevalence))
})
colnames(phylum_summary) <- c("Phylum", "Mean_Prevalence", "Sum_Prevalence")

# Phyla by Total ASV Prevalence 
phylum_summary_sorted <- phylum_summary[order(-phylum_summary$Sum_Prevalence), ]
print(phylum_summary_sorted, row.names = FALSE)
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(ps_1),color=Phylum)) +
  ggtitle("Prevalence of Phyla Across Skin-Swab Samples") +
  ggsubtitle("y-intercept = 5% prevalence"
  )
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```
### At each timepoint, keep only samples with control and surgical pairs  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Extract sample data
sample_check <- as(sample_data(ps_1), "data.frame")

# Set timepoint order 
day_levels <- c(
  "Admission",
  "Day 0a - Pre Surgery",
  "Day 0c - Post Surgery",
  "Day 1",
  "Day 3",
  "Week 1",
  "Week 2",
  "Week 3",
  "Week 4",
  "Week 5",
  "Week 6"
)
sample_check$Day_Order <- factor(sample_check$Day_Order, levels = day_levels)
sample_check$Sample_Site <- factor(sample_check$Sample_Site, levels = c("Control", "Surgical"))

# 2. Count skin samples across timepoints
count_df <- sample_check %>% 
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>% 
  distinct() %>% 
  group_by(Day_Order, Sample_Site) %>% 
  summarise(n = n(), .groups = "drop")
 

count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)
count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Control", "Surgical"))

p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Control" = "blue", "Surgical" = "purple")) +
  labs(
    x = "Timepoint",
    y = "Number of Samples",
    fill = "Skin Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)

# Check which dogs are missing samples (either control or surgical) at each timepoint
missing_sample_unique <- sample_check %>% 
  group_by(Individual, Day_Order) %>% 
  summarise(
    site_absent = if (all(c("Control", "Surgical") %in% Sample_Site)) NA_character_ 
                  else paste(setdiff(c("Control", "Surgical"), unique(Sample_Site)), collapse = ", "), 
    Sample_Unique = paste(sort(unique(Sample_Unique)), collapse = ", "), 
    n_sites = n_distinct(Sample_Site), 
    .groups = "drop"
  ) %>% 
  filter(n_sites < 2) %>% 
  dplyr::select(Individual, Day_Order, site_absent, Sample_Unique) %>%
  # Arrange by Individual and chronological factor timepoint
  arrange(Individual, Day_Order)

print(missing_sample_unique)

# Keep only samples that have control and surgical at each Day_Order
paired_groups <- sample_data %>%
  group_by(Individual, Day_Order) %>%
  summarize(n_sites = n_distinct(Sample_Site), .groups = "drop") %>%
  filter(n_sites == 2) %>%
  transmute(Individual, Day_Order)
filtered_sample_data <- sample_data %>%
  inner_join(paired_groups, by = c("Individual", "Day_Order"))

sample_data(ps_1)
nrow(sample_data(ps_1))

# Prune phyloseq object so it only contains these remaining samples 
ps_1 <- prune_samples(filtered_sample_data$SampleID, ps_1)
nrow(sample_data(ps_1))

# Count unique samples by Day_Order and Sample_Site
count_df <- filtered_sample_data %>%
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>%
  distinct() %>%
  group_by(Day_Order, Sample_Site) %>%
  summarise(n = n(), .groups = "drop")


count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)


count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Control", "Surgical"))

# Updated after removing dog-timepoint samples with missing pair
p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Control" = "blue", "Surgical" = "purple")) +
  labs(
    x = "Timepoint",
    y = "Number of Samples",
    fill = "Skin Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)
```
### Generate final stats per seq. run and sample
```{r, message=FALSE, warning=FALSE, echo=TRUE}
out_dir <- "../02_output_files/02_image_files"

# Get reads and metadata per sample
reads_per_sample <- sample_sums(ps_1)
md <- as(sample_data(ps_1), "data.frame")
 
sample_names_ps <- sample_names(ps_1)

# Ensure sample order matches
if (!all(sample_names_ps == rownames(md))) {
  md <- md[sample_names_ps, ]
}

# Create dataframe with sample-level stats
df <- data.frame(
  sample = sample_names_ps,
  total_reads = reads_per_sample,
  Sample_Plate = md$Sample_Plate,
  Individual = md$Individual,
  stringsAsFactors = FALSE
)

# Calculate total ASVs per sample
# 1. Calculate total ASVs and total reads
otu <- otu_table(ps_1)
if (taxa_are_rows(otu)) {
  total_asvs <- colSums(otu > 0)
} else {
  total_asvs <- rowSums(otu > 0)
}

total_reads <- sample_sums(ps_1)

# 2. Extract metadata, append columns, and reassign
 
md$total_asv <- total_asvs
md$total_reads <- total_reads

sample_data(ps_1) <- sample_data(md)

# Aggregate reads and ASVs per Individual (summing across all samples from that Individual)
by_ind <- aggregate(
  cbind(total_reads, total_asv) ~ Individual + Sample_Plate,
  data = md,
  FUN = sum
)

# Boxplot for total reads by Sample_Plate (each point = one Individual)
p_reads <- ggplot(by_ind, aes(x = Sample_Plate, y = total_reads, fill = Sample_Plate)) +
  geom_boxplot() +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "K"),
    name = "Total Reads (per Individual, in thousands)"
  ) +
  labs(
    title = "Total Reads per Individual by Sample Plate",
    x = "Sample Plate"
  ) +
  theme_minimal()

out_reads <- paste0(out_dir, "/total_reads_by_sample_plate.png")
ggplot2::ggsave(out_reads, p_reads, width = 8, height = 6)
p_reads
# Boxplot for total ASVs by Sample_Plate (each point = one Individual)
p_asv <- ggplot(by_ind, aes(x = Sample_Plate, y = total_asv, fill = Sample_Plate)) +
  geom_boxplot() +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "K"),
    name = "Total ASVs (per Individual, in thousands)"
  ) +
  labs(
    title = "Total ASVs per Individual by Sample Plate",
    x = "Sample Plate"
  ) +
  theme_minimal()
p_asv
out_asv <- paste0(out_dir, "/total_asvs_by_sample_plate.png")
ggplot2::ggsave(out_asv, p_asv, width = 8, height = 6)

# Print summary to verify - each row is one Individual
print(by_ind)

```

# Shannon Alpha-Diversity 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
# 1. Filter and prepare the alpha diversity data frame
alphadiv_plot_data <- estimate_richness(ps_1, measures = "Shannon") %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(as.data.frame(sample_data(ps_1)), by = "SampleID")

# 2. Generate the faceted dot plot for each individual dog
ggplot(alphadiv_plot_data, aes(x = factor(Day_Order), y = Shannon, color = Sample_Site, group = Individual)) +
  geom_point(size = 1) +
  geom_line() + # Optional: connects the dots chronologically for each dog
  facet_wrap(~ Individual) + 
  scale_color_manual(values = c("Control" = "purple", "Surgical" = "blue")) + 
  theme_bw() + 
  labs(
    title = "Shannon Diversity Across Timepoints per Dog", 
    x = "Timepoint (Day Order)", 
    y = "Shannon Index",
    color = "Sample Site"
  ) +
  theme(axis.text.x = element_text(angle = 90))

 
plot_richness(ps_1, 
              x = "Day_Order", 
              measures = "Shannon") + 
  geom_boxplot(aes(fill = Sample_Site), show.legend = TRUE) +
  facet_grid(. ~ Sample_Site) + # Adjusted facet since there is only 1 metric now
  scale_fill_manual(values = c("Control" = "purple", "Surgical" = "blue")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(title = "Shannon Diversity",
       x = "Timepoint",
       fill = "Sample Site")

# Alpha diversity across timepoints for the Control Groups 
alphadiv <- estimate_richness(ps_1, measures = c("Shannon")) %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(as.data.frame(sample_data(ps_1)), by = "SampleID")  %>% 
  filter(Sample_Site == "Control")  

kruskal.test(Shannon ~ Day_Order, data = alphadiv)
 
dunnS <- dunnTest(Shannon ~ Day_Order,
                  data=alphadiv,
                  method="bh")
 dunnS
 
# Alpha diversity across timepoints for the Surgery Groups 
alphadiv <- estimate_richness(ps_1, measures = c("Shannon")) %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(as.data.frame(sample_data(ps_1)), by = "SampleID")  %>% 
  filter(Sample_Site == "Surgical") 

kruskal.test(Shannon ~ Day_Order, data = alphadiv)
 
dunnS <- dunnTest(Shannon ~ Day_Order,
                  data=alphadiv,
                  method="bh")
 dunnS
 
```

# Relative Abundances
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
# Normalization using relative abundances
ps_rel <- transform_sample_counts(ps_1, function(x) x / sum(x))

individual_list <- unique(sample_data(ps_rel)$Individual)

# create color pallette 
all_phyla <- tax_table(ps_rel)[, "Phylum"] %>% as.vector() %>% unique() %>% na.omit() %>% sort()
n_phyla <- length(all_phyla)
base_cols <- brewer.pal(min(n_phyla, 12), "Paired")
phylum_cols_vec <- colorRampPalette(base_cols)(n_phyla)
phylum_colors <- setNames(phylum_cols_vec, all_phyla)

# 4. Loop through each individual to generate exactly ONE plot per dog
for (ind in individual_list) {
  
  # Subset the main relative abundance object for this specific dog
  ps_ind <- subset_samples(ps_rel, Individual == ind)
  
  # Skip to the next dog if this individual has no samples
  if(nsamples(ps_ind) == 0) next
  
  # Melt and clean data for both conditions simultaneously
  plot_df <- psmelt(ps_ind) %>% 
    filter(Sample_Site %in% c("Control", "Surgical")) %>% 
    group_by(Day_Order, Phylum, Sample_Site) %>% 
    summarise(Abundance = sum(Abundance), .groups = "drop") %>% 
    rename(Condition = Sample_Site)
  
  # Identify top 20 phyla specific to this dog's profile
  top20_ind <- plot_df %>% 
    group_by(Phylum) %>% 
    summarise(Total_Abundance = sum(Abundance), .groups = "drop") %>% 
    arrange(desc(Total_Abundance)) %>% 
    slice_head(n = 20) %>% 
    pull(Phylum)
  
  phylum_levels_ind <- c(top20_ind, "Other")
  
  # Recode lower abundance phyla into "Other"
  plot_df <- plot_df %>% mutate(
    Phylum = ifelse(Phylum %in% top20_ind, Phylum, "Other"),
    Phylum = factor(Phylum, levels = phylum_levels_ind),
    Condition = factor(Condition, levels = c("Control", "Surgical")),
    Day_Order = factor(Day_Order)
  )
  
  # Map colors safely from master palette; fallback to gray grid if missing
  local_cols <- phylum_colors[top20_ind]
  missing_indices <- is.na(local_cols)
  if(any(missing_indices)) {
    fallback_pool <- brewer.pal(12, "Paired")
    local_cols[missing_indices] <- sample(fallback_pool, sum(missing_indices), replace = TRUE)
  }
  values_ind <- c(local_cols, "Other" = "grey70")
  
  # Generate the single combined plot for this dog
  p <- ggplot(plot_df, aes(x = Day_Order, y = Abundance, fill = Phylum)) + 
    geom_col(position = "stack", width = 0.7) + # Changed to standard stacked bars
    facet_wrap(~Condition, nrow = 1) +          # Side-by-side comparison panels
    scale_fill_manual(
      values = values_ind,
      breaks = phylum_levels_ind,
      na.value = "black"
    ) + 
    theme_minimal() + 
    theme(
      axis.text.x = element_text(angle = 75, hjust = 1),
      panel.spacing = unit(1.5, "lines")        # Clear visual separation between panels
    ) + 
    labs(
      title = paste0("Relative Abundance Across Timepoints by Phyla"),
      subtitle = paste0("Dog: ", ind),
      x = "Timepoint (Day Order)",
      y = "Relative Abundance"
    )
  
  # Output the plot to your R session
  print(p)
}

 
```

# Bray-Curtis Beta-Diversity 
## Seq. Runs (batch-effect)
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(100)
# Calculate bray-curtis distance matrix on samples using ASV relative abundances
metadata <- data.frame(sample_data(ps_rel))
bray_dist <- phyloseq::distance(ps_rel, method = "bray")
# generate PCoA using the sample distance matrix
bray_pcoa <- ordinate(physeq = ps_rel, method = "PCoA", distance = bray_dist)

# Color samples by Sequencing Run
p1 <- plot_ordination(ps_rel, bray_pcoa, color = "Sample_Plate")
p1 + ggtitle("PCoA of Beta-Diversity (Bray-Curtis)") +
  geom_point(size = 1.5) + 
  theme_bw() +
  stat_ellipse(aes(fill = Sample_Plate, group = Sample_Plate), alpha = 0.2, geom = "polygon")
# PERMANOVA test for significant differences amongst sequencing runs 
adonis2(bray_dist ~ Sample_Plate, data = metadata, permutations = 999)
```
```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
print("Within a cluster what is the average distance to the centroid?")
# Distance of each 16S sample to its sequencing batch centroid 
# Method = PERMDISP2 (multivariate analogue of Levene's test)
betadisper_result <- betadisper(bray_dist, group = metadata$Sample_Plate)

print("What are the differences between the libraries?")
average_by_plate <- data.frame(
  Plate = metadata$Sample_Plate,
  Dist_to_Centroid = betadisper_result$distances
)

average_by_plate_table <- average_by_plate %>%
  group_by(Plate) %>%
  summarize(
    Mean_Dist = mean(Dist_to_Centroid, na.rm = TRUE),
    SD_Dist   = sd(Dist_to_Centroid, na.rm = TRUE),
    CV_Percent = (SD_Dist / Mean_Dist *100) 
  )

print(average_by_plate_table)

# Extract the PCoA coordinates and centroids
df_points <- as.data.frame(scores(betadisper_result, display = "sites"))
df_points$Sample_Plate <- metadata$Sample_Plate

df_centroids <- as.data.frame(scores(betadisper_result, display = "centroids"))
df_centroids$Sample_Plate <- rownames(df_centroids)

ggplot(df_points, aes(x = PCoA1, y = PCoA2, color = Sample_Plate)) +
  geom_segment(data = merge(df_points, df_centroids, by = "Sample_Plate", suffixes = c("", ".cen")),
               aes(xend = PCoA1.cen, yend = PCoA2.cen), alpha = 0.5) +
  geom_point(size = 2) +
  geom_point(data = df_centroids, size = 5, shape = 17) + 
  stat_ellipse(aes(fill = Sample_Plate, group = Sample_Plate), alpha = 0.1, geom = "polygon", color = NA) +
  ggtitle("PCoA of Beta-Diversity (Bray-Curtis) with Centroids") +
  theme_bw()

# Permutation test for significance
permutest(betadisper_result, permutations = 999)

# SEq Run Pairwise Post-hoc comparisons
TukeyHSD(betadisper_result)


# Within each sequencing run, test if there is a signficant difference amongst the centroid  
split_idx <- split(seq_len(nrow(metadata)), metadata$Sample_Plate)

results <- lapply(split_idx, function(i) {
  d_sub <- as.dist(as.matrix(bray_dist)[i, i])
  md_sub <- metadata[i, ]
  bd <- betadisper(d_sub, group = md_sub$Day_Order)
  list(
    anova = anova(bd),
    perm = permutest(bd, permutations = 999)
  )
})

print(results)
```
## Timepoints 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Run PERMANOVA constraining permutations within plates
global_permanova_strata <- adonis2(
  bray_dist ~ Day_Order, 
  data = md, 
  strata = md$Sample_Plate,
  permutations = 999
)

print(global_permanova_strata)
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
 
## Merge rooted tree to phyloseq object  
# Unifrac distance matrix 
# Create random tree matching the taxa names
random_tree <- rtree(ntaxa(ps_rel), rooted = TRUE, tip.label = taxa_names(ps))
plot(random_tree)

# Merge tree into the phyloseq object
ps_2 <- merge_phyloseq(ps_rel, random_tree)
ps_2

# Calculate weighted UniFrac distance matrix 
dist_unifrac_2 <- phyloseq::distance(ps_2, method = "wunifrac")

# Run PCoA using the distance matrix
uni_2 <- ordinate(physeq = ps_2, method = "PCoA", distance = dist_unifrac_2)


# Run PERMANOVA constraining permutations within plates
global_permanova_strata <- adonis2(
  dist_unifrac_2 ~ Day_Order, 
  data = md, 
  strata = md$Sample_Plate,
  permutations = 999
)

print(global_permanova_strata)
```

# Analyze Control-Surgical Bray-Curtis Distances Across Timepoints 
```{r,fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
paired_distances <- md %>%
  group_by(Dog_ID, Day_Order) %>%
  filter(all(c("Control", "Surgical") %in% Sample_Site)) %>%
  ungroup() %>%
  group_by(Dog_ID, Day_Order) %>%
  summarise(
    Control_Sample = SampleID[Sample_Site == "Control"][1],
    Surgical_Sample = SampleID[Sample_Site == "Surgical"][1],
    Bray_Distance = bray_mat[Control_Sample, Surgical_Sample],
    Sample_Plate = Sample_Plate[1],
    Day_Order = Day_Order[1],  # Add the character label
    .groups = "drop"
  )


# Show group means
print("Group means by Day_Order:")
print(paired_distances %>%
  group_by(Day_Order) %>%
  summarise(
    Mean_Bray = mean(Bray_Distance),
    Median_Bray = median(Bray_Distance),
    SD_Bray = sd(Bray_Distance),
    N = n(),
    .groups = "drop"
  ))


# Print the results
print(paired_distances)

# Generate the histogram
histogram_distances <- ggplot(paired_distances, aes(x = Bray_Distance)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of Average Bray-Curtis Distances per Dog",
    x = "Average Bray-Curtis Distance",
    y = "Count of Dogs"
  )
print(histogram_distances)

 
pairwise_results <- pairwise.adonis2(
  bray_dist ~ Day_Order,
  strata = "Sample_Plate",
  data = md)
 
print(pairwise_results)

#  Pull the raw p-values 
p_values <- sapply(names(pairwise_results)[-1], function(x) {
  pairwise_results[[x]][["Pr(>F)"]][1]
})

# Benjamini-Hochberg (BH) adjustment   
p_adjusted <- p.adjust(p_values, method = "BH")

# 3. Compile them into a clean, flat data frame
pairwise_df <- data.frame(
  Comparison = names(p_values),
  p_value    = p_values,
  p_adj      = p_adjusted,
  row.names  = NULL
)


# 1. Split the Comparison column by the "_vs_" delimiter
groups <- str_split_fixed(pairwise_df$Comparison, "_vs_", 2)

# 2. Add the new clean columns back to your dataframe
pairwise_df$Group1 <- groups[, 1]
pairwise_df$Group2 <- groups[, 2]

# 3. Clean up: Reorder columns and drop the original messy header
pairwise_df <- pairwise_df[, c("Group1", "Group2", "p_value", "p_adj")]

# 4. Print the final result
print(pairwise_df)

pairwise_tidy <- pairwise_df %>%
# Drop empty matrix intersections
  filter(!is.na(p_adj)) %>% 
  # Add significance labels
  mutate(
    Significant = ifelse(p_adj <= alpha, "Yes", "No"),
    Stars = case_when(
      p_adj <= 0.001 ~ "***",
      p_adj <= 0.01  ~ "**",
      p_adj <= 0.05  ~ "*",
      p_adj <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p_adj = round(p_adj, 4) ) %>%
  arrange(p_adj) 

print(pairwise_tidy)

# Keep only significant distances
sig_df <- pairwise_tidy %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,   # fill below
    label = Stars
  )
x_levels <- levels(factor(paired_distances$Day_Order))

sig_df <- sig_df %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p_value) %>%
  mutate(
    y.position = max(paired_distances$Bray_Distance, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )

median_df <- paired_distances %>%
  group_by(Day_Order) %>%
  summarise(median_dist = median(Bray_Distance, na.rm = TRUE), .groups = "drop")

ggplot(paired_distances, aes(x = factor(Day_Order), y = Bray_Distance)) +
  geom_violin(fill = "lightblue", trim = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +
  geom_point(data = median_df, aes(x = Day_Order, y = median_dist),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = median_df, aes(x = Day_Order, y = median_dist, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +
  stat_pvalue_manual(
    sig_df,
    label = "label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.5,
    size = 5
  ) +
  labs(
    title = "Control vs. Surgical Site Bray-Curtis Distances Across Timepoints",
    subtitle = "Trend line indicating average bray-curtis distances; * adj.p-value =< 0.05",
    x = "Timepoint",
    y = "Bray-Curtis Distance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )
```

# Fit Model to Pairwise Bray-Curtis Dissimilarities between Control and Surgical Skin Site Across Timepoints 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
paired_distances <- paired_distances %>%
  mutate(
    Day_Order = case_when(
      Day_Order == "Admission" ~ -2,            
      Day_Order == "Day 0a - Pre Surgery" ~ -1,
      Day_Order == "Day 0c - Post Surgery" ~ 0,
      Day_Order == "Day 1" ~ 1,
      Day_Order == "Day 3" ~ 3,
      Day_Order == "Week 1" ~ 7,
      Day_Order == "Week 2" ~ 14,
      Day_Order == "Week 3" ~ 21,
      Day_Order == "Week 4" ~ 28,
      Day_Order == "Week 5" ~ 35,
      Day_Order == "Week 6" ~ 42,
      TRUE ~ NA_real_  # Handle any unmatched values
    )
  )



lm <- lm(Bray_Distance ~ Day_Order + Sample_Plate,
            data = paired_distances)
summary(lm)
os <- segmented(lm, seg.Z = ~Day_Order)
summary(os)
confint(os)
 
library(ggplot2)
library(segmented)

# 1. Create a sequence grid for Day_Order (X-axis)
# We choose a single reference plate (e.g., "Plate_1" or the first unique plate) 
# so R can calculate predictions for a single line profile.
reference_plate <- unique(paired_distances$Sample_Plate)[1] 

grid_data <- data.frame(
  Day_Order = seq(min(paired_distances$Day_Order), 
                  max(paired_distances$Day_Order), 
                  length.out = 100),
  Sample_Plate = reference_plate
)

# 2. Generate predictions with confidence intervals
# Note: For 'segmented' objects, standard predict interval="confidence" may lack 
# full breakpoint variance, but it computes the segment errors.
pred_vals <- predict(os, newdata = grid_data, interval = "confidence")

# 3. Combine grid data and predictions
plot_data <- cbind(grid_data, pred_vals)

# 4. Plot using geom_ribbon (X = Day_Order, Y = Bray_Distance)
ggplot(plot_data, aes(x = Day_Order, y = fit)) +
  geom_line(color = "blue", size = 1) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "blue", alpha = 0.2) +
  # Add the raw experimental scatter data
  geom_point(data = paired_distances, aes(x = Day_Order, y = Bray_Distance), alpha = 0.5) +
  labs(x = "Day Order", y = "Avg Bray Distance", title = "Segmented Regression with CI Band")

```

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```
