---
title: "Canine Skin Microbiome: Microbial Diversity Analysis "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

**Purpose:**
Analyze microbial diversity within and across timepoints, comparing control and surgical skin sites. 
    
# Load R Packages
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library("knitr")
library("FSA")
library("tidyverse")
library("vegan")
library("phyloseq")
library("ggplot2")
library("ggtree")
library("scales") # Crucial for label_comma() and log_breaks() to work
library(RColorBrewer)
library(segmented)
set.seed(100)
```

# Create phyloseq objects
```{r, message=FALSE, warning=FALSE, echo=TRUE}
load("../02_output_files/01_filtered_ASV_tables/CH1_filtered_canine_skin_microbiome.rda")

ps1 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps1 <- prune_samples(
  grepl("Skin", sample_data(ps1)$Location_Type),
  ps1
)
load("../02_output_files/01_filtered_ASV_tables/CH2_filtered_canine_skin_microbiome.rda")

ps2 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps2 <- prune_samples(
  grepl("Skin", sample_data(ps2)$Location_Type),
  ps2
)
load("../02_output_files/01_filtered_ASV_tables/CH3_filtered_canine_skin_microbiome.rda")

ps3 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps3 <- prune_samples(
  grepl("Skin", sample_data(ps3)$Location_Type),
  ps3
)
load("../02_output_files/01_filtered_ASV_tables/CH4_filtered_canine_skin_microbiome.rda")

ps4 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
  ps4 <- prune_samples(
  grepl("Skin", sample_data(ps4)$Location_Type),
  ps4
)
ps <- merge_phyloseq(ps1, ps2, ps3, ps4)
ps <- prune_samples(
  grepl("Skin", sample_data(ps)$Location_Type),
  ps
)
```

# ASVs with Unknown Phylum
```{r, message=FALSE, warning=FALSE, echo=TRUE}
otu_matrix <- as(otu_table(ps), "matrix")
if (taxa_are_rows(ps)) {
  otu_matrix <- t(otu_matrix)
}

# First get total ASVs per plate
total_asvs_per_plate <- data.frame(
  Sample_Plate = sample_data(ps)$Sample_Plate,
  as.data.frame(otu_matrix)
) %>%
  group_by(Sample_Plate) %>%
  summarise(
    Total_ASVs = sum(colSums(across(-where(is.character))) > 0),
    .groups = "drop"
  )


unknown_asvs <- rownames(tax_table(ps))[
  is.na(tax_table(ps)[, "Phylum"]) | tax_table(ps)[, "Phylum"] %in% c("", "uncharacterized")
]

otu_matrix <- as(otu_table(ps), "matrix")
if (taxa_are_rows(ps)) {
  otu_matrix <- t(otu_matrix)
}

otu_unknown <- otu_matrix[, unknown_asvs, drop = FALSE]

# Then get unknown ASVs per plate and calculate percentage
out <- data.frame(
  Sample_Plate = sample_data(ps)$Sample_Plate,
  as.data.frame(otu_unknown)
) %>%
  group_by(Sample_Plate) %>%
  summarise(
    Unknown_ASVs_Count = sum(colSums(across(-where(is.character))) > 0),
    .groups = "drop"
  ) %>%
  # Join with total ASVs
  left_join(total_asvs_per_plate, by = "Sample_Plate") %>%
  mutate(
    Percent_Removed = 100 * Unknown_ASVs_Count / Total_ASVs
  )

print(out)
 

# Remove ASVs with unknown Phyla
ps_1 <- subset_taxa(ps, !is.na(Phylum) & !Phylum %in% c("", "uncharacterized"))
 

# Extract sample metadata
meta <- as(sample_data(ps_1), "data.frame")

# Count unique samples by Day_Order and Sample_Site
count_df <- meta %>%
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>%
  distinct() %>%
  group_by(Day_Order, Sample_Site) %>%
  summarise(n = n(), .groups = "drop")

# Set the desired order for Day_Order
day_levels <- c(
  "Admission",
  "Day 0a - Pre Surgery",
  "Day 0c - Post Surgery",
  "Day 1",
  "Day 3",
  "Week 1",
  "Week 2",
  "Week 3",
  "Week 4",
  "Week 5",
  "Week 6"
)

count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)

# Optional: set Sample_Site order
count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Control", "Surgical"))

# Plot
p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Control" = "blue", "Surgical" = "purple")) +
  labs(
    x = "Day Order",
    y = "Number of Samples",
    fill = "Sample Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)

# Check which dogs are missing samples (either control or surgical) at each timepoint
missing_sample_unique <- sample_check %>%
  group_by(Individual, Day_Order) %>%
  summarise(
    site_absent = if (all(c("Control", "Surgical") %in% Sample_Site)) NA_character_
                  else paste(setdiff(c("Control", "Surgical"), unique(Sample_Site)), collapse = ", "),
    Sample_Unique = paste(sort(unique(Sample_Unique)), collapse = ", "),
    n_sites = n_distinct(Sample_Site),
    .groups = "drop"
  ) %>%
  filter(n_sites < 2) %>%
  select(Individual, Day_Order, site_absent, Sample_Unique)

print(missing_sample_unique)


# extract sample data as data.frame
meta <- as(sample_data(ps_1), "data.frame")

# keep only samples that have control and surgical at each Day_Order
paired_groups <- meta %>%
  group_by(Individual, Day_Order) %>%
  summarize(n_sites = n_distinct(Sample_Site), .groups = "drop") %>%
  filter(n_sites == 2) %>%
  transmute(Individual, Day_Order)
filtered_meta <- meta %>%
  inner_join(paired_groups, by = c("Individual", "Day_Order"))

sample_data(ps_1)
nrow(sample_data(ps_1))
#prune phyloseq onject so it only contains these remaining samples 
ps_1 <- prune_samples(filtered_meta$SampleID, ps_1)
nrow(sample_data(ps_1))

# Count unique samples by Day_Order and Sample_Site
count_df <- filtered_meta %>%
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>%
  distinct() %>%
  group_by(Day_Order, Sample_Site) %>%
  summarise(n = n(), .groups = "drop")


count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)

# Optional: set Sample_Site order
count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Control", "Surgical"))

# Plot
p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Control" = "blue", "Surgical" = "purple")) +
  labs(
    x = "Day Order",
    y = "Number of Samples",
    fill = "Sample Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)
```

```{r, fig.width=11, fig.height=8}

# Compute prevalence of each feature, store as data.frame
prevdf = apply(X = otu_table(ps_1),
               MARGIN = ifelse(taxa_are_rows(ps_1), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})
 
# Add taxonomy and total read counts to this data.frame
prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps_1),
                    tax_table(ps_1))

# Are there phyla that are comprised of mostly low-prevalence features? Compute the total and average prevalences of the features in each phylum.
plyr::ddply(prevdf, "Phylum", function(df1){cbind(mean(df1$Prevalence),sum(df1$Prevalence))})
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(ps_1),color=Phylum)) +
  ggtitle("Prevalence of Phyla Across Skin-Swab Samples") +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```
 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
out_dir <- "../02_output_files/02_image_files"

# Get reads and metadata per sample
reads_per_sample <- sample_sums(ps_1)
md <- as(sample_data(ps_1), "data.frame")
 
sample_names_ps <- sample_names(ps_1)

# Ensure sample order matches
if (!all(sample_names_ps == rownames(md))) {
  md <- md[sample_names_ps, ]
}

# Create dataframe with sample-level stats
df <- data.frame(
  sample = sample_names_ps,
  total_reads = reads_per_sample,
  Sample_Plate = md$Sample_Plate,
  Individual = md$Individual,
  stringsAsFactors = FALSE
)

# Calculate total ASVs per sample
# 1. Calculate total ASVs and total reads
otu <- otu_table(ps_1)
if (taxa_are_rows(otu)) {
  total_asvs <- colSums(otu > 0)
} else {
  total_asvs <- rowSums(otu > 0)
}

total_reads <- sample_sums(ps_1)

# 2. Extract metadata, append columns, and reassign
 
md$total_asv <- total_asvs
md$total_reads <- total_reads

sample_data(ps_1) <- sample_data(md)

# Aggregate reads and ASVs per Individual (summing across all samples from that Individual)
by_ind <- aggregate(
  cbind(total_reads, total_asv) ~ Individual + Sample_Plate,
  data = md,
  FUN = sum
)

# Boxplot for total reads by Sample_Plate (each point = one Individual)
p_reads <- ggplot(by_ind, aes(x = Sample_Plate, y = total_reads, fill = Sample_Plate)) +
  geom_boxplot() +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "K"),
    name = "Total Reads (per Individual, in thousands)"
  ) +
  labs(
    title = "Total Reads per Individual by Sample Plate",
    x = "Sample Plate"
  ) +
  theme_minimal()

out_reads <- paste0(out_dir, "/total_reads_by_sample_plate.png")
ggplot2::ggsave(out_reads, p_reads, width = 8, height = 6)
p_reads
# Boxplot for total ASVs by Sample_Plate (each point = one Individual)
p_asv <- ggplot(by_ind, aes(x = Sample_Plate, y = total_asv, fill = Sample_Plate)) +
  geom_boxplot() +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "K"),
    name = "Total ASVs (per Individual, in thousands)"
  ) +
  labs(
    title = "Total ASVs per Individual by Sample Plate",
    x = "Sample Plate"
  ) +
  theme_minimal()
p_asv
out_asv <- paste0(out_dir, "/total_asvs_by_sample_plate.png")
ggplot2::ggsave(out_asv, p_asv, width = 8, height = 6)

# Print summary to verify - each row is one Individual
print(by_ind)




```
# Beta-diversity: Bray-Curtis Distance Matrix, PCoA, PERMANOVA

## Identifying sequencing run batch-effects
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(100)
ps_rel <- transform_sample_counts(ps_1, function(x) x / sum(x)) 

# Calculate bray-curtis distance matrix on samples using ASV relative abundances
metadata <- data.frame(sample_data(ps_rel))
bray_dist <- phyloseq::distance(ps_rel, method = "bray")
# generate PCoA using the sample distance matrix
bray_pcoa <- ordinate(physeq = ps_rel, method = "PCoA", distance = bray_dist)

print("How different are the centroids of the clusters formed by samples within the same sequencing run?")
# Color samples by Sequencing Run
p1 <- plot_ordination(ps_rel, bray_pcoa, color = "Sample_Plate")
p1 + ggtitle("PCoA of Beta-Diversity (Bray-Curtis)") +
  geom_point(size = 1.5) + 
  theme_bw() +
  stat_ellipse(aes(fill = Sample_Plate, group = Sample_Plate), alpha = 0.2, geom = "polygon")
# PERMANOVA test for significant differences amongst sequencing runs 
adonis2(bray_dist ~ Sample_Plate, data = metadata, permutations = 999)


print("Within a cluster what is the average distance to the centroid?")
# Distance of each 16S sample to its sequencing batch centroid 
# Method = PERMDISP2 (multivariate analogue of Levene's test)
betadisper_result <- betadisper(bray_dist, group = metadata$Sample_Plate)

# Extract the PCoA coordinates and centroids
df_points <- as.data.frame(scores(betadisper_result, display = "sites"))
df_points$Sample_Plate <- metadata$Sample_Plate

df_centroids <- as.data.frame(scores(betadisper_result, display = "centroids"))
df_centroids$Sample_Plate <- rownames(df_centroids)

ggplot(df_points, aes(x = PCoA1, y = PCoA2, color = Sample_Plate)) +
  geom_segment(data = merge(df_points, df_centroids, by = "Sample_Plate", suffixes = c("", ".cen")),
               aes(xend = PCoA1.cen, yend = PCoA2.cen), alpha = 0.5) +
  geom_point(size = 2) +
  geom_point(data = df_centroids, size = 5, shape = 17) + 
  stat_ellipse(aes(fill = Sample_Plate, group = Sample_Plate), alpha = 0.1, geom = "polygon", color = NA) +
  ggtitle("PCoA of Beta-Diversity (Bray-Curtis) with Centroids") +
  theme_bw()

print("What are the pairwise differences between the libraries?")
average_by_plate <- data.frame(
  Plate = metadata$Sample_Plate,
  Dist_to_Centroid = betadisper_result$distances
)

average_by_plate_table <- average_by_plate %>%
  group_by(Plate) %>%
  summarize(
    Mean_Dist = mean(Dist_to_Centroid, na.rm = TRUE),
    SD_Dist   = sd(Dist_to_Centroid, na.rm = TRUE),
    CV_Percent = (SD_Dist / Mean_Dist *100) 
  )

print(average_by_plate_table)


# Permutation test for significance
permutest(betadisper_result, permutations = 999)

# SEq Run Pairwise Post-hoc comparisons
TukeyHSD(betadisper_result)
```

## Relative Abundance (surgical site)
```{r, message=FALSE, warning=FALSE, echo=TRUE}
ps_rel_surgical <- subset_samples(ps_rel, Sample_Site == "Surgical")
ps_1_surgical <- subset_samples(ps_1, Sample_Site == "Surgical")
# 2. Get a unique list of all individuals in your dataset
individual_list <- unique(sample_data(ps_rel_surgical)$Individual)

# Get all phyla in your object (no Other here)
all_phyla <- tax_table(ps_rel_surgical)[, "Phylum"] %>%
  as.vector() %>% 
  unique() %>%
  sort()

# Choose colors (expand palette if needed)
n_phyla <- length(all_phyla)  
base_cols <- brewer.pal(min(n_phyla, 12), "Paired")
phylum_cols_vec <- colorRampPalette(base_cols)(n_phyla)

# Named palette for real phyla (no Other)
phylum_levels <- all_phyla
phylum_colors <- setNames(phylum_cols_vec, phylum_levels)

for (ind in individual_list) {
  
  ps_sub <- subset_samples(ps_rel_surgical, Individual == ind)
  
  plot_df <- psmelt(ps_sub) %>%
    group_by(Day_Order, Phylum) %>%
    summarise(Abundance = sum(Abundance), .groups = "drop")
  
  # top 20 Phylum within this individual
  top20_ind <- plot_df %>%
    group_by(Phylum) %>%
    summarise(Total_Abundance = sum(Abundance), .groups = "drop") %>%
    arrange(desc(Total_Abundance)) %>%
    slice_head(n = 20) %>%
    pull(Phylum)
  
  # collapse others into "Other"
  phylum_levels_ind <- c(top20_ind, "Other")
  
  plot_df <- plot_df %>%
    mutate(
      Phylum = ifelse(Phylum %in% top20_ind, Phylum, "Other"),
      Phylum = factor(Phylum, levels = phylum_levels_ind)
    )
  
  # colors for this individual: top20 from global palette + "Other" as gray
  values_ind <- c(
    phylum_colors[top20_ind],    # named vector for the top20
    "Other" = "grey70"           # force Other to gray
  )
  
  p <- ggplot(plot_df, aes(x = Day_Order, y = Abundance, fill = Phylum)) +
    geom_col() +
    scale_fill_manual(
      values = values_ind,
      breaks = phylum_levels_ind,   # legend order = top20 then Other
      na.value = "black"            # if any NA in Phylum, show as black
    ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = paste0("Top 20 Phyla Across Timepoints for Dog: ", ind))
  
  print(p)
}
 
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
 
## Merge rooted tree to phyloseq object  
# Unifrac distance matrix 
# Create random tree matching the taxa names
random_tree <- rtree(ntaxa(ps_rel), rooted = TRUE, tip.label = taxa_names(ps))
plot(random_tree)

# Merge tree into the phyloseq object
ps_2 <- merge_phyloseq(ps_rel, random_tree)
ps_2

# Calculate weighted UniFrac distance matrix 
dist_unifrac_2 <- phyloseq::distance(ps_2, method = "wunifrac")

# Run PCoA using the distance matrix
uni_2 <- ordinate(physeq = ps_2, method = "PCoA", distance = dist_unifrac_2)
```

# Alpha-diversity 

```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
# 1. Filter and prepare the alpha diversity data frame
alphadiv_plot_data <- estimate_richness(ps_1, measures = "Shannon") %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(as.data.frame(sample_data(ps_1)), by = "SampleID")

# 2. Generate the faceted dot plot for each individual dog
ggplot(alphadiv_plot_data, aes(x = factor(Day_Order), y = Shannon, color = Sample_Site, group = Individual)) +
  geom_point(size = 1) +
  geom_line() + # Optional: connects the dots chronologically for each dog
  facet_wrap(~ Individual) + 
  scale_color_manual(values = c("Control" = "purple", "Surgical" = "blue")) + 
  theme_bw() + 
  labs(
    title = "Shannon Diversity Across Timepoints per Dog", 
    x = "Timepoint (Day Order)", 
    y = "Shannon Index",
    color = "Sample Site"
  ) +
  theme(axis.text.x = element_text(angle = 90))

 
plot_richness(ps_1, 
              x = "Day_Order", 
              measures = "Shannon") + 
  geom_boxplot(aes(fill = Sample_Site), show.legend = TRUE) +
  facet_grid(. ~ Sample_Site) + # Adjusted facet since there is only 1 metric now
  scale_fill_manual(values = c("Control" = "purple", "Surgical" = "blue")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(title = "Shannon Diversity",
       x = "Timepoint",
       fill = "Sample Site")

# Alpha diversity across timepoints for the Control Groups 
alphadiv <- estimate_richness(ps_1, measures = c("Shannon")) %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(as.data.frame(sample_data(ps_1)), by = "SampleID")  %>% 
  filter(Sample_Site == "Control")  

kruskal.test(Shannon ~ Day_Order, data = alphadiv)
 
dunnS <- dunnTest(Shannon ~ Day_Order,
                  data=alphadiv,
                  method="bh")
 dunnS
 
# Alpha diversity across timepoints for the Surgery Groups 
alphadiv <- estimate_richness(ps_1, measures = c("Shannon")) %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(as.data.frame(sample_data(ps_1)), by = "SampleID")  %>% 
  filter(Sample_Site == "Surgical") 

kruskal.test(Shannon ~ Day_Order, data = alphadiv)
 
dunnS <- dunnTest(Shannon ~ Day_Order,
                  data=alphadiv,
                  method="bh")
 dunnS
 
```

# Model Changes in Pairwise Bray-Curtis Dissimilarities between Control and Surgical Skin Site Across Timepoints  
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
md <- as(sample_data(ps_1), "data.frame")

bray_mat <- as.matrix(bray_dist)
paired_distances <- md %>%
  group_by(Dog_ID, Day_Order) %>%
  filter(all(c("Control", "Surgical") %in% Sample_Site)) %>%
  ungroup() %>%
  group_by(Dog_ID, Day_Order) %>%
  summarise(
    Control_Sample = SampleID[Sample_Site == "Control"][1],
    Surgical_Sample = SampleID[Sample_Site == "Surgical"][1],
    Bray_Distance = bray_mat[Control_Sample, Surgical_Sample],
    Sample_Plate = Sample_Plate[1],
    .groups = "drop"
  )

print(paired_distances)

# Calculate average distance per Dog_ID for each Day_Order
avg_distances_by_dog <- paired_distances %>%
  group_by(Dog_ID, Day_Order) %>%
  summarise(
    Avg_Bray_Distance = mean(Bray_Distance),
    Sample_Plate = Sample_Plate,
   # total_reads = sum(total_reads),
    .groups = "drop"
  )

# Print the results
print(avg_distances_by_dog)

# Create box-and-whisker plot across Day_Orders
ggplot(data = avg_distances_by_dog, 
       aes(x = Day_Order, y = Avg_Bray_Distance)) +
  geom_boxplot(fill = "darkgreen") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "darkgreen") +
  labs(
    title = "Average Bray-Curtis Distance by Day Order",
    x = "Day Order",
    y = "Average Bray Distance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )

avg_distances_by_dog <- avg_distances_by_dog %>%
  mutate(
    Day_Order = case_when(
      Day_Order == "Admission" ~ -2,           # No quotes!
      Day_Order == "Day 0a - Pre Surgery" ~ -1,
      Day_Order == "Day 0c - Post Surgery" ~ 0,
      Day_Order == "Day 1" ~ 1,
      Day_Order == "Day 3" ~ 3,
      Day_Order == "Week 1" ~ 7,
      Day_Order == "Week 2" ~ 14,
      Day_Order == "Week 3" ~ 21,
      Day_Order == "Week 4" ~ 28,
      Day_Order == "Week 5" ~ 35,
      Day_Order == "Week 6" ~ 42,
      TRUE ~ NA_real_  # Handle any unmatched values
    )
  )

library(lme4)

lmm <- lmer(Avg_Bray_Distance ~ Day_Order + Sample_Plate + (1 | Dog_ID),
            data = avg_distances_by_dog)
summary(lmm)
os <- segmented(lm_model, seg.Z = ~Day_Order)
summary(os)
confint(os)

plot(avg_distances_by_dog$Day_Order, avg_distances_by_dog$Avg_Bray_Distance,
     xlab = "Day Number (relative to surgery)", 
     ylab = "Average Bray Distance",
     main = "Segmented Regression: Bray Distance Over Time")
points(avg_distances_by_dog$Day_Order, avg_distances_by_dog$Avg_Bray_Distance, pch = 19, col = "blue")

pred_vals <- predict(os)
lines(avg_distances_by_dog$Day_Order, pred_vals, col = "red", lwd = 2)



library(dplyr)
library(ggplot2)
md <- as(sample_data(ps_1), "data.frame")
# Step 1: Create a NEW column for the character labels (don't overwrite Day_Order)
md <- md %>%
  mutate(
    Day_Order = case_when(
      Day_Order == -2 ~ "Admission",
      Day_Order == -1 ~ "Pre_Surgery",
      Day_Order == 0 ~ "Post_Surgery",
      Day_Order == 1 ~ "Day_1",
      Day_Order == 3 ~ "Day_3",
      Day_Order == 7 ~ "Week_1",
      Day_Order == 14 ~ "Week_2",
      Day_Order == 21 ~ "Week_3",
      Day_Order == 28 ~ "Week_4",
      Day_Order == 35 ~ "Week_5",
      Day_Order == 42 ~ "Week_6",
      TRUE ~ as.character(Day_Order)
    )
  )

# Step 2: Create paired_distances using the numeric Day_Order (for regression)
paired_distances <- md %>%
  group_by(Dog_ID, Day_Order) %>%
  filter(all(c("Control", "Surgical") %in% Sample_Site)) %>%
  ungroup() %>%
  group_by(Dog_ID, Day_Order) %>%
  summarise(
    Control_Sample = SampleID[Sample_Site == "Control"][1],
    Surgical_Sample = SampleID[Sample_Site == "Surgical"][1],
    Bray_Distance = bray_mat[Control_Sample, Surgical_Sample],
    Sample_Plate = Sample_Plate[1],
    Day_Order = Day_Order[1],  # Add the character label
    .groups = "drop"
  )

# Step 3: Calculate average distance per Dog_ID for each Day_Order (numeric)
avg_distances_by_dog <- paired_distances %>%
  group_by(Dog_ID, Day_Order) %>%
  summarise(
    Avg_Bray_Distance = mean(Bray_Distance),
    Sample_Plate = Sample_Plate,
    Day_Order = Day_Order[1],
    .groups = "drop"
  )

# Step 4: Perform pairwise t-tests using the character Day_Order
alpha <- 0.10

pairwise_results <- pairwise.t.test(
  Avg_Bray_Distance, 
  Day_Order,  # Use character labels for t-test
  data = avg_distances_by_dog,
  p.adjust.method = "none",
  alternative = "two.sided"
)

# Print results
print(pairwise_results)

# Convert to tidy format for easy reading
pairwise_df <- pairwise_results$p.value

pairwise_tidy <- pairwise_df %>%
  as.data.frame() %>%
  mutate(Group1 = rownames(pairwise_df)) %>%
  tidyr::pivot_longer(
    cols = -Group1,
    names_to = "Group2",
    values_to = "p_value"
  ) %>%
  filter(Group1 < Group2) %>%
  mutate(
    Significant = ifelse(p_value <= alpha, "Yes", "No"),
    p_value = round(p_value, 4),
    Stars = case_when(
      p_value <= 0.01 ~ "**",
      p_value <= 0.05 ~ "*",
      p_value <= alpha ~ ".",
      TRUE ~ ""
    )
  )

print(pairwise_tidy)

# Show group means
print("Group means by Day_Order:")
print(avg_distances_by_dog %>%
  group_by(Day_Order) %>%
  summarise(
    Mean_Bray = mean(Avg_Bray_Distance),
    SD_Bray = sd(Avg_Bray_Distance),
    N = n(),
    .groups = "drop"
  ) %>%
  arrange(match(Day_Order, c("Admission", "Pre_Surgery", "Post_Surgery", "Day_1", "Day_3", "Week_1", "Week_2", "Week_3", "Week_4", "Week_5", "Week_6"))))

# Get detailed results as a table
pairwise_df <- pairwise_results$p.value

# Convert to tidy format with comparisons
pairwise_tidy <- pairwise_df %>%
  as.data.frame() %>%
  mutate(Group1 = rownames(pairwise_df)) %>%
  tidyr::pivot_longer(
    cols = -Group1,
    names_to = "Group2",
    values_to = "p_value"
  ) %>%
  filter(Group1 < Group2) %>%  # Remove duplicates and self-comparisons
  mutate(
    Significant = ifelse(p_value <= alpha, "Yes", "No"),
    p_value = round(p_value, 4)
  )

print(pairwise_tidy)

# Add significance stars
pairwise_tidy <- pairwise_tidy %>%
  mutate(
    Stars = case_when(
      p_value <= 0.01 ~ "**",
      p_value <= 0.05 ~ "*",
      p_value <= alpha ~ ".",
      TRUE ~ ""
    )
  )

print(pairwise_tidy)
# Create comparison table with effect sizes
comparison_table <- avg_distances_by_dog %>%
  group_by(Day_Order) %>%
  summarise(
    Mean_Bray = mean(Avg_Bray_Distance),
    SD_Bray = sd(Avg_Bray_Distance),
    N = n(),
    .groups = "drop"
  )

print(comparison_table)

# Plot with pairwise significance
ggplot(avg_distances_by_dog, aes(x = Day_Order, y = Avg_Bray_Distance)) +
  geom_boxplot(fill = "lightblue", outlier.alpha = 0.3) +
  geom_jitter(width = 0.15, alpha = 0.4, color = "darkblue") +
  labs(
    title = "Bray Distance by Day Order",
    x = "Day Order",
    y = "Average Bray Distance"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```
