---
title: "Canine Skin Microbiome: Microbial Diversity Analysis "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

**Purpose:**
Analyze microbial diversity within and across timepoints, comparing control and surgical skin sites. 
    
# Load R Packages
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library("knitr")
library("FSA")
library("tidyverse")
library("vegan")
library("phyloseq")
library("ggplot2")
library("ggtree")
library("scales") # Crucial for label_comma() and log_breaks() to work
library(RColorBrewer)
set.seed(100)
```

# Create phyloseq objects
```{r, message=FALSE, warning=FALSE, echo=TRUE}
load("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/R_scripts/CH1_filtered_canine_skin_microbiome.rda")

ps1 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps1 <- prune_samples(
  grepl("Skin", sample_data(ps1)$Location_Type),
  ps1
)
load("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/R_scripts/CH2_filtered_canine_skin_microbiome.rda")

ps2 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps2 <- prune_samples(
  grepl("Skin", sample_data(ps2)$Location_Type),
  ps2
)
load("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/R_scripts/CH3_filtered_canine_skin_microbiome.rda")

ps3 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps3 <- prune_samples(
  grepl("Skin", sample_data(ps3)$Location_Type),
  ps3
)
load("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/R_scripts/CH4_filtered_canine_skin_microbiome.rda")

ps4 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
  ps4 <- prune_samples(
  grepl("Skin", sample_data(ps4)$Location_Type),
  ps4
)
ps <- merge_phyloseq(ps1, ps2, ps3, ps4)
ps <- prune_samples(
  grepl("Skin", sample_data(ps)$Location_Type),
  ps
)
#ps <- subset_taxa(ps, !is.na(Phylum) & !Phylum %in% c("", "uncharacterized"))

ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
ps_list <- list(ps1, ps2, ps3, ps4)
run_names <- paste0("CH", 1:4)

out_dir <- "/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/R_scripts"

# Aggregate reads and ASVs per Individual within each run
ind_stats <- do.call(rbind, lapply(seq_along(ps_list), function(i) {
  ps <- ps_list[[i]]
  
  reads_per_sample <- sample_sums(ps)
  md <- as(sample_data(ps), "data.frame")
  sample_names_ps <- sample_names(ps)
  
  if (!all(sample_names_ps == rownames(md))) {
    md <- md[sample_names_ps, ]
  }
  
  df <- data.frame(
    sample = sample_names_ps,
    total_reads = reads_per_sample,
    Individual = md$Individual,
    stringsAsFactors = FALSE
  )
  
  otu <- otu_table(ps)
  if (taxa_are_rows(otu)) {
    df$total_asv <- colSums(otu > 0)
  } else {
    df$total_asv <- rowSums(otu > 0)
  }
  
  # Sum reads and ASVs per Individual
  by_ind <- aggregate(
    cbind(total_reads, total_asv) ~ Individual,
    data = df,
    FUN = sum
  )
  
  data.frame(
    sequencing_run = run_names[i],
    Individual = by_ind$Individual,
    total_reads = by_ind$total_reads,
    total_asv = by_ind$total_asv,
    stringsAsFactors = FALSE
  )
}))

# Boxplot for total reads by sequencing run
p_reads <- ggplot(ind_stats, aes(x = sequencing_run, y = total_reads, fill=sequencing_run)) +
  geom_boxplot() +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "K"),
    name = "Total Reads (per Individual, in thousands)"
  ) +
  labs(
    title = "Total Reads per Individual by Sequencing Run",
    x = "Sequencing Run"
  ) +
  theme_minimal()

out_reads <- paste0(out_dir, "/total_reads_by_sequencing_run.png")
ggplot2::ggsave(out_reads, p_reads, width = 8, height = 6)

# Boxplot for total ASVs by sequencing run (y-axis in thousands)
p_asv <- ggplot(ind_stats, aes(x = sequencing_run, y = total_asv,  fill=sequencing_run)) +
  geom_boxplot() +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "K"),
    name = "Total ASVs (per Individual, in thousands)"
  ) +
  labs(
    title = "Total ASVs per Individual by Sequencing Run",
    x = "Sequencing Run"
  ) +
  theme_minimal()

out_asv <- paste0(out_dir, "/total_asvs_by_sequencing_run.png")
ggplot2::ggsave(out_asv, p_asv, width = 8, height = 6)
 

```
# Beta-diversity: Bray-Curtis Distance Matrix, PCoA, PERMANOVA

## Check for sequencing run batch-effects
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Calculate bray-curtis distance matrix on samples using ASV relative abundances
metadata <- data.frame(sample_data(ps_rel))
bray_dist <- phyloseq::distance(ps_rel, method = "bray")
# generate PCoA using the sample distance matrix
bray_pcoa <- ordinate(physeq = ps_rel, method = "PCoA", distance = bray_dist)

print("How different are the centroids of the clusters formed by samples within the same sequencing run?")
# Color samples by Sequencing Run
p1 <- plot_ordination(ps_rel, bray_pcoa, color = "Sample_Plate")
p1 + ggtitle("PCoA of Beta-Diversity (Bray-Curtis)") +
  geom_point(size = 1.5) + 
  theme_bw() +
  stat_ellipse(aes(fill = Sample_Plate, group = Sample_Plate), alpha = 0.2, geom = "polygon")
# PERMANOVA test for significant differences amongst sequencing runs 
adonis2(bray_distance ~ Sample_Plate, data = metadata, permutations = 999)


print("Within a cluster what is the average distance to the centroid?")
# Distance of each 16S sample to its sequencing batch centroid 
# Method = PERMDISP2 (multivariate analogue of Levene's test)
betadisper_result <- betadisper(bray_dist, group = metadata$Sample_Plate)

# Extract the PCoA coordinates and centroids
df_points <- as.data.frame(scores(betadisper_result, display = "sites"))
df_points$Sample_Plate <- metadata$Sample_Plate

df_centroids <- as.data.frame(scores(betadisper_result, display = "centroids"))
df_centroids$Sample_Plate <- rownames(df_centroids)

ggplot(df_points, aes(x = PCoA1, y = PCoA2, color = Sample_Plate)) +
  geom_segment(data = merge(df_points, df_centroids, by = "Sample_Plate", suffixes = c("", ".cen")),
               aes(xend = PCoA1.cen, yend = PCoA2.cen), alpha = 0.5) +
  geom_point(size = 2) +
  geom_point(data = df_centroids, size = 5, shape = 17) + 
  stat_ellipse(aes(fill = Sample_Plate, group = Sample_Plate), alpha = 0.1, geom = "polygon", color = NA) +
  ggtitle("PCoA of Beta-Diversity (Bray-Curtis) with Centroids") +
  theme_bw()


average_by_plate <- data.frame(
  Plate = metadata$Sample_Plate,
  Dist_to_Centroid = betadisper_result$distances
)

average_by_plate_table <- average_by_plate %>%
  group_by(Plate) %>%
  summarize(
    Mean_Dist = mean(Dist_to_Centroid, na.rm = TRUE),
    SD_Dist   = sd(Dist_to_Centroid, na.rm = TRUE),
    CV_Percent = (SD_Dist / Mean_Dist *100) 
  )

print(average_by_plate_table)

# Permutation test for significance
permutest(betadisper_result, permutations = 999)

# SEq Run Pairwise Post-hoc comparisons
TukeyHSD(betadisper_result)
```

### Merge rooted tree to phyloseq object  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Unifrac distance matrix 
# Create random tree matching the taxa names
random_tree <- rtree(ntaxa(ps_rel), rooted = TRUE, tip.label = taxa_names(ps))
plot(random_tree)

# Merge tree into the phyloseq object
ps_2 <- merge_phyloseq(ps_rel, random_tree)
ps_2

# Calculate weighted UniFrac distance matrix 
dist_unifrac_2 <- phyloseq::distance(ps_2, method = "wunifrac")
distance_bray_curtis <- phyloseq::distance(ps_rel, method = "bray")

# Run PCoA using the distance matrix
uni_2 <- ordinate(physeq = ps_2, method = "PCoA", distance = dist_unifrac_2)
bray_1 <- ordinate(physeq = ps_rel, method = "PCoA", distance = distance_bray_curtis)
```


## Second Check for differences between Control and Surgical Sites 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
p1 <- plot_ordination(ps_2, uni_2, color = "Sample_Site")
p1 + 
  geom_point(size = 1.5) + 
  theme_bw() + 
  scale_color_manual(values = c("Control" = "purple", "Surgical" = "blue"))
sam_2 <- data.frame(sample_data(ps_2))
adonis2(dist_unifrac_2 ~ Day_Order + Individual + Day_Order:Individual, data = sam_2, permutations = 999)
```

# Alpha-diversity 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# 1. Filter and prepare the alpha diversity data frame
alphadiv_plot_data <- estimate_richness(ps_rel, measures = "Shannon") %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(as.data.frame(sample_data(ps_rel)), by = "SampleID")

# 2. Generate the faceted dot plot for each individual dog
ggplot(alphadiv_plot_data, aes(x = factor(Day_Order), y = Shannon, color = Sample_Site, group = Individual)) +
  geom_point(size = 1) +
  geom_line() + # Optional: connects the dots chronologically for each dog
  facet_wrap(~ Individual) + 
  scale_color_manual(values = c("Control" = "purple", "Surgical" = "blue")) + 
  theme_bw() + 
  labs(
    title = "Shannon Diversity Across Timepoints per Dog", 
    x = "Timepoint (Day Order)", 
    y = "Shannon Index",
    color = "Sample Site"
  ) +
  theme(axis.text.x = element_text(angle = 90))

# 2. Plot Shannon index ONLY (Observed richness will not work on relative abundance)
plot_richness(ps_rel, 
              x = "Day_Order", 
              measures = "Shannon") + 
  geom_boxplot(aes(fill = Sample_Site), show.legend = TRUE) +
  facet_grid(. ~ Sample_Site) + # Adjusted facet since there is only 1 metric now
  scale_fill_manual(values = c("Control" = "purple", "Surgical" = "blue")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(title = "Shannon Diversity (Relative Abundance)",
       x = "Timepoint",
       fill = "Sample Site")

# Alpha diversity across timepoints for the Control Groups 
alphadiv <- estimate_richness(ps_rel, measures = c("Shannon")) %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(as.data.frame(sample_data(ps_rel)), by = "SampleID")  %>% 
  filter(Sample_Site == "Control")  

kruskal.test(Shannon ~ Day_Order, data = alphadiv)
 
dunnS <- dunnTest(Shannon ~ Day_Order,
                  data=alphadiv,
                  method="bh")
 dunnS
 
# Alpha diversity across timepoints for the Surgery Groups 
alphadiv <- estimate_richness(ps_rel, measures = c("Shannon")) %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(as.data.frame(sample_data(ps_rel)), by = "SampleID")  %>% 
  filter(Sample_Site == "Surgical") 

kruskal.test(Shannon ~ Day_Order, data = alphadiv)
 
dunnS <- dunnTest(Shannon ~ Day_Order,
                  data=alphadiv,
                  method="bh")
 dunnS
 
```

## Relative Abundance (surgical site)
```{r, message=FALSE, warning=FALSE, echo=TRUE}
ps_rel_surgical <- subset_samples(ps_rel, Sample_Site == "Surgical")

# 2. Get a unique list of all individuals in your dataset
individual_list <- unique(sample_data(ps_rel_surgical)$Individual)

# Get all phyla in your object (no Other here)
all_phyla <- tax_table(ps_rel_surgical)[, "Phylum"] %>%
  as.vector() %>% 
  unique() %>%
  sort()

# Choose colors (expand palette if needed)
n_phyla <- length(all_phyla)  
base_cols <- brewer.pal(min(n_phyla, 12), "Paired")
phylum_cols_vec <- colorRampPalette(base_cols)(n_phyla)

# Named palette for real phyla (no Other)
phylum_levels <- all_phyla
phylum_colors <- setNames(phylum_cols_vec, phylum_levels)

for (ind in individual_list) {
  
  ps_sub <- subset_samples(ps_rel_surgical, Individual == ind)
  
  plot_df <- psmelt(ps_sub) %>%
    group_by(Day_Order, Phylum) %>%
    summarise(Abundance = sum(Abundance), .groups = "drop")
  
  # top 20 Phylum within this individual
  top20_ind <- plot_df %>%
    group_by(Phylum) %>%
    summarise(Total_Abundance = sum(Abundance), .groups = "drop") %>%
    arrange(desc(Total_Abundance)) %>%
    slice_head(n = 20) %>%
    pull(Phylum)
  
  # collapse others into "Other"
  phylum_levels_ind <- c(top20_ind, "Other")
  
  plot_df <- plot_df %>%
    mutate(
      Phylum = ifelse(Phylum %in% top20_ind, Phylum, "Other"),
      Phylum = factor(Phylum, levels = phylum_levels_ind)
    )
  
  # colors for this individual: top20 from global palette + "Other" as gray
  values_ind <- c(
    phylum_colors[top20_ind],    # named vector for the top20
    "Other" = "grey70"           # force Other to gray
  )
  
  p <- ggplot(plot_df, aes(x = Day_Order, y = Abundance, fill = Phylum)) +
    geom_col() +
    scale_fill_manual(
      values = values_ind,
      breaks = phylum_levels_ind,   # legend order = top20 then Other
      na.value = "black"            # if any NA in Phylum, show as black
    ) +
    labs(title = paste0("Top 20 Phyla Across Timepoints for Dog: ", ind))
  
  print(p)
}

```

## Examine Differences between: Individuals, Surgery-Type, and Timepoints
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Remove Control samples  
ps_3 <- subset_samples(ps_2, !grepl("Control", Sample_Site))
ps_3 <- prune_taxa(taxa_sums(ps_3) > 0, ps_3)

# Calculate new UniFrac distance for subsetted data
dist_unifrac_3 <- phyloseq::distance(ps_3, method = "unifrac")
uni_3 <- ordinate(physeq = ps_3, method = "PCoA", distance = dist_unifrac_3)

# Plot by Individual
p3 <- plot_ordination(ps_3, uni_3, color = "Individual")
p3 + geom_point(size = 1.5) + theme_bw()
sam_2 <- data.frame(sample_data(ps_2))
adonis2(dist_unifrac_2 ~ Day_Order + Individual + Day_Order:Individual, data = sam_2, permutations = 999)


# Plot by Surgery_Type
p2 <- plot_ordination(ps_3, uni_3, color = "Surgery_Type")
p2 + 
  geom_point(size = 1.5) + 
  theme_bw() + 
  scale_color_manual(values = c("Abdomen" = "orange", "Stifle" = "pink"))
sam_2 <- data.frame(sample_data(ps_2))
adonis2(dist_unifrac_2 ~ Day_Order + Individual + Day_Order:Individual, data = sam_2, permutations = 999)

# Plot by Day_Order
p4 <- plot_ordination(ps_3, uni_3, color = "Day_Order")
p4 + geom_point(size = 1.5) + theme_bw()
sam_2 <- data.frame(sample_data(ps_2))
adonis2(dist_unifrac_2 ~ Day_Order + Individual + Day_Order:Individual, data = sam_2, permutations = 999)
```
 

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```
