---
title: "Canine Skin Microbiome: 16S Change-Point Analysis "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

**Purpose:**
Investigate how canine skin microbial composition changes in result of disturbance (surgery).

**Null Hypothesis:**
No change-point in bray-curtis dissimilarity between paired control and surgical sites of dogs before surgery. 
 
**Description:**
Each dog has its own control prior to surgery as well as a control site that did not undergo surgery. This allowed us to model how within-dog control–surgical dissimilarity changed with number of days since surgery.

**References:**

* Longitudinal profiling of the microbiome at four body sites reveals core stability and individualized dynamics during health and disease (Cell Host Microbe - April 2025) <https://pmc.ncbi.nlm.nih.gov/articles/PMC11022754/>

* Metagenomic characterisation of canine skin reveals a core healthy skin microbiome (Nature - August 2024)
<https://www.nature.com/articles/s41598-024-63999-5>

* Cutaneous Microbiome Profiles Following Chlorhexidine Treatment in a 72-Hour Daily Follow-Up Paired Design: a Pilot Study (ASM - June 2022) <https://journals.asm.org/doi/full/10.1128/spectrum.01753-21>

# Load R packages
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(knitr)
library(FSA)
library(tidyverse)
library(vegan)
library(phyloseq)
library(ggplot2)
library(ggtree)
library(scales)  
library(RColorBrewer)
library(segmented)
library(stringr)
library(pairwiseAdonis)
library(ggpubr)
library(changepoint)
library(dplyr)
library(lme4)
library(emmeans)
library(nortest)
library(mcp)
library(purrr)
library(kableExtra)
library(forcats)
 
```


# Generate Paired-Only  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
ps_1 <- readRDS("../01_input_files/phyloseq/ps_1.rds")
metadata <- as(sample_data(ps_1), "data.frame")
 
# Set timepoint order 
day_levels <- c(
  "Admission",
  "Day 0a - Pre Surgery",
  "Day 0c - Post Surgery",
  "Day 1",
  "Day 3",
  "Week 1",
  "Week 2",
  "Week 3",
  "Week 4",
  "Week 5",
  "Week 6"
)
metadata$Day_Order <- factor(metadata$Day_Order, levels = day_levels)
metadata$Sample_Site <- factor(metadata$Sample_Site, levels = c("Control", "Surgical"))

#Count skin samples across timepoints
count_df <- metadata %>% 
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>% 
  distinct() %>% 
  group_by(Day_Order, Sample_Site) %>% 
  summarise(n = n(), .groups = "drop")
 

count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)
count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Control", "Surgical"))

p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Control" = "purple", "Surgical" = "blue")) +
  labs(
    x = "Timepoint",
    y = "Number of Samples",
    fill = "Skin Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)

# Check which dogs are missing samples (either control or surgical) at each timepoint
missing_sample_unique <- metadata %>% 
  group_by(Individual, Day_Order) %>% 
  summarise(
    site_absent = if (all(c("Control", "Surgical") %in% Sample_Site)) NA_character_ 
                  else paste(setdiff(c("Control", "Surgical"), unique(Sample_Site)), collapse = ", "), 
    Sample_Unique = paste(sort(unique(Sample_Unique)), collapse = ", "), 
    n_sites = n_distinct(Sample_Site), 
    .groups = "drop"
  ) %>% 
  filter(n_sites < 2) %>% 
  dplyr::select(Individual, Day_Order, site_absent, Sample_Unique) %>%
  arrange(Individual, Day_Order)

print(missing_sample_unique)

# Keep only samples that have control and surgical at each Day_Order
paired_groups <- metadata %>%
  group_by(Individual, Day_Order) %>%
  summarize(n_sites = n_distinct(Sample_Site), .groups = "drop") %>%
  filter(n_sites == 2) %>%
  transmute(Individual, Day_Order)
paired_sample_data <- metadata %>%
  inner_join(paired_groups, by = c("Individual", "Day_Order"))

 
# Prune phyloseq object so it only contains these remaining samples 
ps_2 <- prune_samples(paired_sample_data$SampleID, ps_1)

 
 
out_dir <- "../01_input_files/phyloseq"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
saveRDS(ps_2, file = file.path(out_dir, "ps_2.rds"))
 

# Normalize read counts by dividing by total read count per sample (relative abundance) 
ps_rel_2 <- transform_sample_counts(ps_2, function(x) x / sum(x))
rel_abundance_vector <- as.numeric(
  as(phyloseq::otu_table(ps_rel_2), "matrix")
)

x <- rel_abundance_vector
# Plot Distribution of Transformed Bray-Curtis Distances
histogram_distances <- ggplot(data = data.frame(x), aes(x = x)) +
  geom_histogram(binwidth = 0.05, fill = "gray", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of ASVs",
    x = "ASV relative abundance",
    y = "Count"
  )

print(histogram_distances)


# Count unique samples by Day_Order and Sample_Site
count_df <- paired_sample_data %>%
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>%
  distinct() %>%
  group_by(Day_Order, Sample_Site) %>%
  summarise(n = n(), .groups = "drop")


count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)


count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Control", "Surgical"))

# Updated after removing dog-timepoint samples with only one site
p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Control" = "purple", "Surgical" = "blue")) +
  labs(
    x = "Timepoint",
    y = "Number of Samples",
    fill = "Skin Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)

paired_sample_data$Sample_Site <- factor(paired_sample_data$Sample_Site)
paired_sample_data$Dog_ID <- factor(paired_sample_data$Dog_ID)
paired_sample_data$Sample_Plate <- factor(paired_sample_data$Sample_Plate)
paired_sample_data$Collection_Day <- as.numeric(paired_sample_data$Collection_Day)


dog_library_table <- with(
  paired_sample_data,
  table(Individual, Sample_Plate)
)
dog_library_table <- rbind(
  dog_library_table,
  Total = colSums(dog_library_table)
)
dog_library_table
  

out_table <- paste0(out_dir, "/total_paired_samples_by_individual.png")
 
dog_library_table_bold <- dog_library_table
dog_library_table_bold[dog_library_table_bold != 0] <- paste0(
  "<b>", dog_library_table_bold[dog_library_table_bold != 0], "</b>"
)

 
paste0("Number of dogs: ", length(unique(paired_sample_data$Dog_ID))) 
paste0("Number of samples: ", length(unique(paired_sample_data$Sample)))

```

```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
kbl_obj <- kable(
  dog_library_table_bold,
  caption = "Dog Paired-Sample Count per Sequencing Run",
  align = "c",
  escape = FALSE,
  format = "html"
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  ) %>%
  column_spec(1:10, color = "black") %>%
  row_spec(
    nrow(dog_library_table_bold),
    bold = TRUE,
    color = "black",
    background = "#f2f2f2"
  )
 
kbl_obj


save_kable(
  kbl_obj,
  file = out_table,
  zoom = 3,
  vwidth = 400,
  vheight = 1200
)
```


## Total Reads + ASVs per Individual 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
out_dir <- "../02_output_files/images"

# Get reads and metadata per sample
reads_per_sample <- sample_sums(ps_2)
paired_sample_data <- as(sample_data(ps_2), "data.frame")
 
sample_names_ps <- sample_names(ps_2)

# Ensure sample order matches
if (!all(sample_names_ps == rownames(paired_sample_data))) {
  paired_sample_data <- paired_sample_data[sample_names_ps, ]
}

# Create dataframe with sample-level stats
df <- data.frame(
  sample = sample_names_ps,
  total_reads = reads_per_sample,
  Sample_Plate = paired_sample_data$Sample_Plate,
  Individual = paired_sample_data$Individual,
  stringsAsFactors = FALSE
)

mean_reads <- mean(df$total_reads)
min_reads  <- min(df$total_reads)
max_reads  <- max(df$total_reads)

mean_reads
min_reads
max_reads

# Calculate total reads and ASVs per sample and add to metadata
otu <- otu_table(ps_2)
if (taxa_are_rows(otu)) {
  total_asvs <- colSums(otu > 0)
} else {
  total_asvs <- rowSums(otu > 0)
}

total_reads <- sample_sums(ps_2)
 
paired_sample_data$total_asv <- total_asvs
paired_sample_data$total_reads <- total_reads

sample_data(ps_2) <- sample_data(paired_sample_data)

paired_sample_data <- paired_sample_data %>%
  arrange(Surgery_Type, Individual) %>%
  mutate(Individual = factor(Individual, levels = unique(Individual)))

my_cols <- colorRampPalette(brewer.pal(8, "Dark2"))(20)

# Aggregate reads and ASVs per Individual 
by_ind <- aggregate(
  cbind(total_reads, total_asv) ~ Individual + Sample_Plate,
  data = paired_sample_data,
  FUN = sum
)

print(by_ind)

# Boxplot for total reads by Individual (each point = one sample)
p_reads <- ggplot(paired_sample_data, aes(x = Individual, y = total_reads, fill = Surgery_Type)) +
  geom_boxplot()  +
  scale_fill_manual(values = c(
    "Abdomen" = "orange",
    "Stifle" = "pink"
  )) +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "K"),
    name = "Total Reads (per Sample, in thousands)"
  ) +
  labs(
    title = "Total Reads per Sample by Dog",
    x = "Dog_ID"
  ) +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    plot.title = element_text(size = 18, face = "bold"),
    legend.position = "none"
  )

p_reads

out_reads <- paste0(out_dir, "/total_reads_by_individual_paired.png")
ggplot2::ggsave(out_reads, p_reads, width = 8, height = 6)
 

# Boxplot for total ASVs by Individual (each point = one sample)
p_asv <- ggplot(paired_sample_data, aes(x = Individual, y = total_asv, fill = Surgery_Type)) +
  geom_boxplot() +
  scale_fill_manual(values = c(
    "Abdomen" = "orange",
    "Stifle" = "pink"
  )) +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "k"),
    name = "Total ASVs (per Individual, in thousands)"
  ) +
  labs(
    title = "Total ASVs per Sample by Dog",
    x = "Dog_ID"
  ) +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    plot.title = element_text(size = 18, face = "bold"),
    legend.position = "none"
  )
p_asv

out_asv <- paste0(out_dir, "/total_asvs_by_individual_paired.png")
ggplot2::ggsave(out_asv, p_asv, width = 8, height = 6)

```
 
# Shannon Alpha-Diversity 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
alphadiv_plot_data <- estimate_richness(ps_2, measures = "Shannon") %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(as.data.frame(sample_data(ps_2)), by = "SampleID") %>% 
  group_by(Individual) %>% 
  mutate(Surgical_Location = Sample_Location[Sample_Site == "Surgical"][1]) %>% 
  ungroup() %>% 
  mutate(Surgical_Location = factor(Surgical_Location)) %>% 
  arrange(Surgical_Location, Individual)
ggplot(alphadiv_plot_data, aes(x = factor(Day_Order), y = Shannon, color = Sample_Site, group = Sample_Site)) + 
  geom_point(size = 1, position = position_dodge(width = 0.3)) + 
  geom_line(position = position_dodge(width = 0.3)) + 
  facet_wrap(
    Surgical_Location ~ Individual, 
    labeller = label_both
  ) + 
  scale_color_manual(values = c("Control" = "purple", "Surgical" = "blue")) + 
  theme_bw() + 
  labs(
    title = "Shannon Diversity Across Timepoints per Dog",
    x = "Timepoint",
    y = "Shannon Index",
    color = "Skin Site"
  ) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

alphadiv_plot_data_surg <- estimate_richness(ps_2, measures = "Shannon") %>% 
  filter(sample_data(ps_2)$Sample_Site == "Surgical")  %>%
  rownames_to_column(var = "SampleID") %>% 
  left_join(as.data.frame(sample_data(ps_2)), by = "SampleID") %>% 
  group_by(Individual) %>% 
  mutate(Surgical_Location = Sample_Location[Sample_Site == "Surgical"][1]) %>% 
  ungroup() %>% 
  mutate(Surgical_Location = factor(Surgical_Location)) %>% 
  arrange(Surgical_Location, Individual)
ggplot(alphadiv_plot_data_surg, aes(x = factor(Day_Order), y = Shannon, color = Sample_Site, group = Sample_Site)) + 
  geom_point(size = 1, position = position_dodge(width = 0.3)) + 
  geom_line(position = position_dodge(width = 0.3)) + 
  facet_wrap(
    Surgical_Location ~ Individual, 
    labeller = label_both
  ) + 
  scale_color_manual(values = c("Surgical" = "blue")) + 
  theme_bw() + 
  labs(
    title = "Shannon Diversity Across Timepoints per Dog",
    x = "Timepoint",
    y = "Shannon Index",
    color = "Skin Site"
  ) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

 

alphadiv_plot_surgical <- estimate_richness(ps_2, measures = "Shannon") %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(as.data.frame(sample_data(ps_2)), by = "SampleID") %>% 
  filter(Sample_Site == "Surgical") %>% 
  mutate(Sample_Location = factor(Sample_Location)) %>% 
  arrange(Sample_Location, Individual)

alphadiv_plot_surgical_median <- alphadiv_plot_surgical %>%
  group_by(Day_Order) %>%
  summarise(median_Shannon = median(Shannon, na.rm = TRUE), .groups = "drop")
 

ggplot(alphadiv_plot_surgical, aes(x = Day_Order, y = Shannon)) +
  geom_boxplot(fill = "#3399FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = alphadiv_plot_surgical_median$median_Shannon[1], linetype = "dashed", color="red") +
  geom_point(data = alphadiv_plot_surgical_median, aes(x = Day_Order, y = median_Shannon),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = alphadiv_plot_surgical_median, aes(x = Day_Order, y = median_Shannon, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +
  labs(
    title = "Surgical-Sites Shannon Diversity Across Timepoints",
    subtitle = "Trend line indicating median Shannon Diversity per timepoint, horizontal red line indicates median of Admission",
    x = "Timepoint",
    y = "Shannon Diversity"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )


alphadiv_plot_control <- estimate_richness(ps_2, measures = "Shannon") %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(as.data.frame(sample_data(ps_2)), by = "SampleID") %>% 
  filter(Sample_Site == "Control") %>% 
  mutate(Sample_Location = factor(Sample_Location)) %>% 
  arrange(Sample_Location, Individual)

alphadiv_plot_control_median <- alphadiv_plot_control %>%
  group_by(Day_Order) %>%
  summarise(median_Shannon = median(Shannon, na.rm = TRUE), .groups = "drop")
 

ggplot(alphadiv_plot_control, aes(x = Day_Order, y = Shannon)) +
  geom_boxplot(fill = "#CC99FF", alpha = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = alphadiv_plot_control_median$median_Shannon[1], linetype = "dashed", color="red") +
  geom_point(data = alphadiv_plot_control_median, aes(x = Day_Order, y = median_Shannon),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = alphadiv_plot_control_median, aes(x = Day_Order, y = median_Shannon, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +
  labs(
    title = "Control-Site Shannon Diversity Across Timepoints",
    subtitle = "Trend line indicating median Shannon Diversity per timepoint, horizontal red line indicates median of Admission",
    x = "Timepoint",
    y = "Shannon Diversity"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )



 


```
  
# Bray-Curtis (BC) Dissimilarity  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# Generate bray-curtis distance matrix using ASV relative abundances
bray_dist <- phyloseq::distance(ps_rel_2, method = "bray")
bray_matrix <- as.matrix(bray_dist)

sd <- as(sample_data(ps_rel_2), "data.frame")
 
 
sd <- sd %>%
  mutate(
    Time_Group = case_when(
      Day_Order %in% c("Admission", "Day 0a - Pre Surgery") ~ "Before_Surgery",
      Day_Order %in% c("Admission", "Day 0c - Post Surgery") ~ "Post_Surgery",
      Day_Order %in% c("Day 1", "Day 3")                   ~ "Inflammation",
      Day_Order %in% c("Week 1", "Week 2")                   ~ "Proliferation",
      Day_Order %in% c("Week 3", "Week 4", "Week 5", "Week 6")       ~ "Maturation",
      TRUE                                                 ~ "NA"
    )
  )


sd$Time_Group <- factor(sd$Time_Group, levels = c("Before_Surgery", "Post_Surgery", "Inflammation", "Proliferation", "Maturation"))
 
sample_data(ps_rel_2) <- sample_data(sd)

 
```


# Check: Skin Site Effect (Abdomen vs Stifle)

### PERMANOVA Test 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Across all days is there a difference between stifle and abdomen? 
set.seed(100)
bray_pcoa <- ordinate(ps_rel_2, method = "PCoA", distance = as.dist(bray_dist))

p1 <- plot_ordination(ps_rel_2, bray_pcoa, color = "Sample_Location") +
  geom_point(size = 1.5) +
  scale_color_manual(values = c("Stifle" = "pink", "Abdomen" = "orange")) +
  theme_bw() +
  ggtitle(
    "PCoA of Bray-Curtis Dissimilarities",
    subtitle = "Ellipses encircle and color samples belonging to the same site location."
  )  

p1
 

# PERMANOVA test on  
set.seed(100)
adonis2(bray_matrix ~ Sample_Location, data = paired_sample_data, permutations = 999, by = "margin")  




# All days: Control vs Surgical differences in Abdomen Groups (aka stifle control vs abdomen sites)
metadata_abdomen <- paired_sample_data %>%
  filter(Surgery_Type == "Abdomen")
abdomen_bray_dist <- bray_matrix[metadata_abdomen$SampleID, metadata_abdomen$SampleID]
abdomen_bray_dist_obj <- as.dist(abdomen_bray_dist)

set.seed(100)
adonis2(abdomen_bray_dist_obj ~ Sample_Site, data = metadata_abdomen, permutations = 999, by = "margin")  


# All days: Control vs Surgical differences in Stifle Groups
metadata_stifle <- paired_sample_data %>%
  filter(Surgery_Type == "Stifle")
Stifle_bray_dist <- bray_matrix[metadata_stifle$SampleID, metadata_stifle$SampleID]
Stifle_bray_dist_obj <- as.dist(Stifle_bray_dist)

set.seed(100)
adonis2(Stifle_bray_dist_obj ~ Sample_Site, data = metadata_stifle, permutations = 999, by = "margin")  




# Before Surgery? 
paired_sample_data_before <- paired_sample_data %>%
  filter(Day_Order %in% c("Admission", "Day 0a - Pre-Surgery"))

# subset phyloseq object
ps_rel_2_before_surgery <- subset_samples(
  ps_rel_2,
  Day_Order %in% c("Admission", "Day 0a - Pre-Surgery")
)

# subset BC matrix 
paired_sample_data_before <- paired_sample_data_before %>%
  filter(SampleID %in% rownames(bray_matrix)) %>%
  arrange(match(SampleID, rownames(bray_matrix)))

before_surgery_bray_dist <- bray_matrix[
  paired_sample_data_before$SampleID,
  paired_sample_data_before$SampleID
]


# PERMANOVA test on  
set.seed(100)
adonis2(before_surgery_bray_dist ~ Sample_Location, data = paired_sample_data_before, permutations = 999, by = "margin")  


# Before surgery: Control vs Surgical differences in Abdomen Groups (aka stifle control vs abdomen sites)
metadata_abdomen <- paired_sample_data_before %>%
  filter(Surgery_Type == "Abdomen")
abdomen_bray_dist <- before_surgery_bray_dist[metadata_abdomen$SampleID, metadata_abdomen$SampleID]
abdomen_bray_dist_obj <- as.dist(abdomen_bray_dist)

set.seed(100)
adonis2(abdomen_bray_dist_obj ~ Sample_Site, data = metadata_abdomen, permutations = 999, by = "margin")  


# Before surgery: Control vs Surgical differences in Stifle Groups
metadata_stifle <- paired_sample_data_before %>%
  filter(Surgery_Type == "Stifle")
Stifle_bray_dist <- before_surgery_bray_dist[metadata_stifle$SampleID, metadata_stifle$SampleID]
Stifle_bray_dist_obj <- as.dist(Stifle_bray_dist)

set.seed(100)
adonis2(Stifle_bray_dist_obj ~ Sample_Site, data = metadata_stifle, permutations = 999, by = "margin")  

```

### LMM
```{r, message=FALSE, warning=FALSE, echo=TRUE}
bray_matrix <- as.matrix(before_surgery_bray_dist)
# How does the surgical site (abdomen or stifle) influence the dissimilarity to control site (stifle for all) before disturbance (ie surgery)? 
bray_arcsin_s <- asin(sqrt(bray_matrix))
paired_distances <- paired_sample_data_before %>%
 group_by(Dog_ID, Day_Order, Surgery_Type) %>%
  summarise(
    Control_Sample = SampleID[Sample_Site == "Control"][1],
    Surgical_Sample = SampleID[Sample_Site == "Surgical"][1],
    Bray_Distance = bray_arcsin_s[Control_Sample, Surgical_Sample],
    Sample_Plate = Sample_Plate[1],
    Day_Order = Day_Order[1],   
    Collection_Day = Collection_Day[1],
   .groups = "drop"
  )

lmer_bray <- lm(
  Bray_Distance ~ Surgery_Type,
  data = paired_distances
)

anova(lmer_bray)
summary(lmer_bray)
```


# Change-point analysis 

## PERMANOVA  
```{r,fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# Generate bray-curtis distance matrix using ASV relative abundances
bray_dist <- phyloseq::distance(ps_rel_2, method = "bray")
bray_matrix <- as.matrix(bray_dist)

# ordinate PCoA from full bray-curtis matrix 
set.seed(100)

bray_pcoa <- ordinate(ps_rel_2, method = "PCoA", distance =  bray_dist)

p1 <- plot_ordination(ps_rel_2, bray_pcoa, color = "Time_Group")


p1 + ggtitle("PCoA of Bray-Curtis Dissimilarities", subtitle = "Ellipses encircle and color samples belonging to the same healing stage.") +
  geom_point(size = 1.5) + 
  theme_bw() +
  stat_ellipse(aes(fill = Time_Group, group = Time_Group), alpha = 0.2, geom = "polygon")

scores <- bray_pcoa$vectors
pc1 <- scores[, 1]

pc1_df <- as.data.frame(pc1) %>%
  rownames_to_column("SampleID")  
 
paired_sample_data_pc1 <- left_join(paired_sample_data, pc1_df, by = "SampleID")

# extract PCoA coordinates (axis 1) from control and surgical samples 
metadata_ctrl <- paired_sample_data_pc1 %>%
  filter(Sample_Site == "Control")
metadata_ctrl_median <- metadata_ctrl %>%
  group_by(Day_Order) %>%
  summarise(median_PCOA = median(pc1, na.rm = TRUE), .groups = "drop")
 
metadata_surg <- paired_sample_data_pc1 %>%
  filter(Sample_Site == "Surgical")
metadata_surg_median <- metadata_surg %>%
  group_by(Day_Order) %>%
  summarise(median_PCOA = median(pc1, na.rm = TRUE), .groups = "drop")


 
# Run PERMANOVA on control-subset bray curtis matrix 
set.seed(100)
ctrl_bray_dist <- bray_matrix[metadata_ctrl$SampleID, metadata_ctrl$SampleID]
ctrl_bray_dist_obj <- as.dist(ctrl_bray_dist)
ps_rel_2_ctrl <- ps_rel_2 %>% subset_samples(Sample_Site == "Control")

ctrl_permanova <- adonis2(
  ctrl_bray_dist_obj ~ Day_Order, 
  data = metadata_ctrl, 
  strata = metadata_ctrl$Dog_ID,
  permutations = 999
)

print(ctrl_permanova)


# Run pairwise - PERMANOVA on control-subset bray curtis matrix 
set.seed(100)
pairwise_results_ctrl <- pairwise.adonis2(
  ctrl_bray_dist_obj ~ Day_Order,
  strata = "Dog_ID",
  data = metadata_ctrl,
  nperm = 999)

p_values <- sapply(names(pairwise_results_ctrl)[-1], function(x) {
  pairwise_results_ctrl[[x]][["Pr(>F)"]][1]
})

# Benjamini-Hochberg (BH) adjustment since we are testing multiple comparisons
p_adjusted <- p.adjust(p_values, method = "BH")

pairwise_df <- data.frame(
  Comparison = names(p_values),
  p_value    = p_values,
  p_adj      = p_adjusted,
  row.names  = NULL
)

groups <- str_split_fixed(pairwise_df$Comparison, "_vs_", 2)

pairwise_df$Group1 <- groups[, 1]
pairwise_df$Group2 <- groups[, 2]

pairwise_df <- pairwise_df[, c("Group1", "Group2", "p_value", "p_adj")]

alpha <- 0.05
pairwise_tidy_ctrl <- pairwise_df %>%
  filter(!is.na(p_adj)) %>% 
  mutate(
    Significant = ifelse(p_adj <= alpha, "Yes", "No"),
    Stars = case_when(
      p_adj <= 0.001 ~ "***",
      p_adj <= 0.01  ~ "**",
      p_adj <= 0.05  ~ "*",
      p_adj <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p_adj = round(p_adj, 4) ) %>%
  arrange(p_adj) 
 


# Filter out insignificant pairwise results

 sig_df_ctrl <- pairwise_tidy_ctrl %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,    
    label = Stars
  )
x_levels <- levels(factor(metadata_ctrl$Day_Order))

sig_df_ctrl <- sig_df_ctrl %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p_value) %>%
  mutate(
    y.position = max(metadata_ctrl$pc1, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )
print(sig_df_ctrl)
 
 
ggplot(metadata_ctrl, aes(x = Day_Order, y = pc1)) +
  geom_boxplot(fill = "#CC99FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = metadata_ctrl_median$median_PCOA[1], linetype = "dashed", color="red") +
  geom_point(data = metadata_ctrl_median, aes(x = Day_Order, y = median_PCOA),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = metadata_ctrl_median, aes(x = Day_Order, y = median_PCOA, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Control-Site: PCoA Axis 1 Coordinates Across Days",
    subtitle = "Trend line indicating median PCOA coordinates per timepoint, horizontal red line indicates median_PCOA of Admission",
    x = "Timepoint",
    y = "PCoA 1 Coordinates"
  ) +
  theme_minimal() +
  
  stat_pvalue_manual(
    sig_df_ctrl,
    label = "label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.5,
    size = 5
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  ) 

 
 
 
 

# Surgical groups
set.seed(100)

surg_bray_dist <- bray_matrix[metadata_surg$SampleID, metadata_surg$SampleID]
surg_bray_dist_obj <- as.dist(surg_bray_dist)
ps_rel_2_surg <- ps_rel_2 %>% subset_samples(Sample_Site == "Surgical")

surg_permanova <- adonis2(
  surg_bray_dist_obj ~ Day_Order, 
  data = metadata_surg, 
  strata = metadata_surg$Dog_ID,
  permutations = 999
)

print(surg_permanova)



### Pairwise PERMANOVA  
set.seed(100)
pairwise_results_surg <- pairwise.adonis2(
  surg_bray_dist_obj ~ Day_Order,
  strata = "Dog_ID",
  data = metadata_surg,
  nperm = 999)

p_values <- sapply(names(pairwise_results_surg)[-1], function(x) {
  pairwise_results_surg[[x]][["Pr(>F)"]][1]
})

# Benjamini-Hochberg (BH) adjustment since we are testing multiple comparisons
p_adjusted <- p.adjust(p_values, method = "BH")

pairwise_df <- data.frame(
  Comparison = names(p_values),
  p_value    = p_values,
  p_adj      = p_adjusted,
  row.names  = NULL
)

groups <- str_split_fixed(pairwise_df$Comparison, "_vs_", 2)

pairwise_df$Group1 <- groups[, 1]
pairwise_df$Group2 <- groups[, 2]

pairwise_df <- pairwise_df[, c("Group1", "Group2", "p_value", "p_adj")]

alpha <- 0.05
pairwise_tidy_surg <- pairwise_df %>%
  filter(!is.na(p_adj)) %>% 
  mutate(
    Significant = ifelse(p_adj <= alpha, "Yes", "No"),
    Stars = case_when(
      p_adj <= 0.001 ~ "***",
      p_adj <= 0.01  ~ "**",
      p_adj <= 0.05  ~ "*",
      p_adj <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p_adj = round(p_adj, 4) ) %>%
  arrange(p_adj) 

 


 
 sig_df_surg <- pairwise_tidy_surg %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,    
    label = Stars
  )
x_levels <- levels(factor(metadata_surg$Day_Order))

sig_df_surg <- sig_df_surg %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p_value) %>%
  mutate(
    y.position = max(metadata_surg$pc1, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )
print(sig_df_surg)

 
 
ggplot(metadata_surg, aes(x = Day_Order, y = pc1)) +
  geom_boxplot(fill = "#3399FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = metadata_surg_median$median_PCOA[1], linetype = "dashed", color="red") +
  geom_point(data = metadata_surg_median, aes(x = Day_Order, y = median_PCOA),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = metadata_surg_median, aes(x = Day_Order, y = median_PCOA, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Surgical-Site: PCoA Axis 1 Coordinates Across Days",
    subtitle = "Trend line indicating median PCOA coordinates per timepoint, horizontal red line indicates median_PCOA of Admission",
    x = "Timepoint",
    y = "PCoA 1 Coordinates"
  ) +
  theme_minimal() +
  
  stat_pvalue_manual(
    sig_df_surg,
    label = "label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.5,
    size = 5
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )
  

 

```

 
 
## LMM  

The pairwise Bray-Curtis distances were transformed to improve normality before LMM analysis. An arcsine transformation was ultimately selected, after evaluating eight potential transformations, including log, square-root and arcsine of the original distance values. Normality of the transformed data was assessed using the Anderson–Darling test, and the transformation with the lowest Anderson–Darling statistic was chosen. 

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

### Transformation BC
eps <- 1e-8

x <- as.vector(bray_matrix)
# Plot Distribution of Transformed Bray-Curtis Distances
histogram_distances <- ggplot(data = data.frame(x), aes(x = x)) +
  geom_histogram(binwidth = 0.05, fill = "gray", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of Bray-Curtis Distances",
    x = "Bray-Curtis Distance",
    y = "Count"
  )

print(histogram_distances)

transforms <- list(
  log1p     = log1p(x),
  log_eps   = log(x + eps),
  sqrt      = sqrt(x),
  sqrt1p    = sqrt(x + 1),
  arcsin_s  = asin(sqrt(x / max(x))),
  arcsin_p  = asin(sqrt(pmin(pmax(x, 0), 1)))
)

res <- data.frame(
  transformation = names(transforms),
  A = sapply(transforms, function(v) ad.test(v)$statistic),
  p = sapply(transforms, function(v) ad.test(v)$p.value)
)

res <- res[order(res$A), ]
res
best_transform <- res$transformation[1]
best_transform


bray_arcsin_s <- asin(sqrt(bray_matrix))

# get vector for histogram
bray_vec <- as.vector(bray_arcsin_s)

# Plot Distribution of Transformed Bray-Curtis Distances
histogram_distances <- ggplot(data = data.frame(bray_vec), aes(x = bray_vec)) +
  geom_histogram(binwidth = 0.05, fill = "gray", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of Transformed Bray-Curtis Distances",
    x = "(arcsine-transformed) Bray-Curtis Distance",
    y = "Count"
  )

print(histogram_distances)



```
 
### BC: Control-Surgical Paired    
 
Model = lmer(Bray_Distance ~ Day_Order + (1 | Dog_ID))
* Contrasts between estimated marginal means of timepoints from model, using Tukey-adjusting p-values:

**Results:**

  * Pre Surgery vs Post Surgery: estimate = -0.1787, adj-p = 0.0295
  * Post Surgery vs Week 3: estimate = 0.1699, adj-p = 0.0431
  * Post Surgery vs Week 5: estimate = 0.1695, adj-p = 0.0441
  
**Notes:**
Estimated marginal means from the mixed model showed a significant increase in BC dissimilarity between control and surgical sites immediately after surgery compared with Pre-Surgery, followed by a significant decline by Weeks 3 and 5. 

```{r,fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# before transformation
paired_distances <- paired_sample_data %>%
 group_by(Dog_ID, Day_Order) %>%
  summarise(
    Control_Sample = SampleID[Sample_Site == "Control"][1],
    Surgical_Sample = SampleID[Sample_Site == "Surgical"][1],
    Bray_Distance = bray_matrix[Control_Sample, Surgical_Sample],
    Sample_Plate = Sample_Plate[1],
    Day_Order = Day_Order[1],   
    Collection_Day = Collection_Day[1],
   .groups = "drop"
  )
lmer_bray <- lmer(
  Bray_Distance ~ Day_Order + (1 | Dog_ID),
  data = paired_distances
)
paired_distances$fitted <- fitted(lmer_bray)
paired_distances$resid <- resid(lmer_bray)
ggplot(paired_distances, aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  facet_wrap(~ Dog_ID) +
  theme_classic() +
  labs(title = "Residual Q-Q plots by Dog_ID")

bray_arcsin_s <- as.matrix(bray_arcsin_s)
# generate paired-BC matrix where control and surgical samples within dogs are measured
paired_distances <- paired_sample_data %>%
 group_by(Dog_ID, Day_Order) %>%
  summarise(
    Control_Sample = SampleID[Sample_Site == "Control"][1],
    Surgical_Sample = SampleID[Sample_Site == "Surgical"][1],
    Bray_Distance = bray_arcsin_s[Control_Sample, Surgical_Sample],
    Sample_Plate = Sample_Plate[1],
    Day_Order = Day_Order[1],   
    Collection_Day = Collection_Day[1],
   .groups = "drop"
  )
 

print("Group means by Day_Order:")
print(paired_distances %>%
  group_by(Day_Order) %>%
  summarise(
    Mean_Bray = mean(Bray_Distance),
    Median_Bray = median(Bray_Distance),
    SD_Bray = sd(Bray_Distance),
    N = n(),
    .groups = "drop"  ) %>% 
    arrange(Day_Order)
)

# Already grouped by dog in paired_distances
# Random effects to account for per dog replicates
lmer_bray <- lmer(
  Bray_Distance ~ Day_Order + (1 | Dog_ID),
  data = paired_distances
)

anova(lmer_bray)
summary(lmer_bray)
 

paired_distances$fitted <- fitted(lmer_bray)
paired_distances$resid <- resid(lmer_bray)

# Fitted vs residuals
ggplot(paired_distances, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(x = "Fitted values", y = "Residuals")

# Residuals vs each independent variable
ggplot(paired_distances, aes(x = Day_Order, y = resid)) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(x = "Day_Order", y = "Residuals")

ggplot(paired_distances, aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  facet_wrap(~ Day_Order) +
  theme_classic() +
  labs(title = "Residual Q-Q plots by Day_Order")
 

# estimated marginal means 
emm <- emmeans(lmer_bray, ~ Day_Order)

emm_df <- as.data.frame(emm)[, c("Day_Order", "emmean", "SE", "df", "lower.CL", "upper.CL")]

alpha <- 0.05
 
contrasts <- data.frame(pairs(emm, Day_Order = "tukey"))

contrasts <- contrasts %>%
  mutate(
    Group1 = case_when(
      str_detect(contrast, "^\\(") ~ str_match(contrast, "^\\((.*?)\\)")[,2],
      TRUE ~ str_match(contrast, "^(.*?) - ")[,2]
    ),
    Group2 = case_when(
      str_detect(contrast, "\\) - \\(") ~ str_match(contrast, "- \\((.*?)\\)$")[,2],
      str_detect(contrast, "^\\(") ~ str_replace(contrast, "^\\(.*?\\) - ", ""),
      TRUE ~ str_match(contrast, "- (.*)$")[,2]
    ),
    Group1 = str_remove_all(str_trim(Group1), "[()]"),
    Group2 = str_remove_all(str_trim(Group2), "[()]")
  )

pairwise_tidy <- contrasts %>%
  filter(!is.na(p.value)) %>% 
  mutate(
    Significant = ifelse(p.value <= alpha, "Yes", "No"),
    Stars = case_when(
      p.value <= 0.001 ~ "***",
      p.value <= 0.01  ~ "**",
      p.value <= 0.05  ~ "*",
      p.value <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p.value = round(p.value, 4) ) %>%
  arrange(p.value) 

# Filter out insignificant pairwise results
sig_df <- pairwise_tidy %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,    
    label = Stars
  )
x_levels <- levels(factor(paired_distances$Day_Order))

sig_df <- sig_df %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p.value) %>%
  mutate(
    y.position = max(paired_distances$Bray_Distance, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )
print(sig_df)

 

ggplot(paired_distances, aes(x = Day_Order, y = Bray_Distance)) +
  geom_boxplot(fill = "lightblue", alpha = 0.8) +
  geom_jitter(width = 0.15, alpha = 0.5, color = "black") +

  geom_hline(yintercept = emm_df$emmean[1], linetype = "dashed", color="red") +
  geom_point(data = emm_df, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "navyblue", size = 3) +
  geom_line(data = emm_df, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "navyblue", linewidth = 1) +
  
  stat_pvalue_manual(
    sig_df,
    label = "label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.5,
    size = 5
  ) +
  labs(
    title = "Paired Control-vs-Surgical Bray-Curtis Distances Across Timepoints",
    subtitle = "Trend line indicating estimated marginal means per timepoint, horizontal red line indicates emmean of Admission",
    x = "Timepoint",
    y = "Bray-Curtis Distance (transformed)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )
```

### BC: Admission vs. Timepoints per Sample_Site

Model =  lmer(Bray_Distance ~ Day_Order*Sample_Site + (1 | Dog_ID) 
* Contrasts between estimated marginal means of timepoints from model, using Tukey-adjusting p-values 

**Results:**

Surgical Samples: 
  * Pre Surgery vs Post Surgery: estimate = -0.1857, adj-p = 0.0098
  * Pre Surgery vs Week 5: estimate = -0.1582, adj-p = 0.0432

Control samples:
* There is no significant differences between timepoints in how dissimilar control samples are to the baseline Admission 

**Notes:**
The post-surgery surgical samples and the week 5 surgical samples, are significantly more dissimilar to the baseline admission samples compared to the level of dissimilarity the pre-surgery samples are to baseline. 

```{r,fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
 # generate BC matrix where admission compared to the rest of the samples within Sample_Site groups (control and surgical) and within dogs are measured
admission_vs_all_same_site <- paired_sample_data %>%
  group_by(Dog_ID, Sample_Site) %>%
  mutate(
    Admission_Sample = SampleID[Day_Order == "Admission"][1]
  ) %>%
  ungroup() %>%
  filter(!is.na(Admission_Sample)) %>%
  filter(SampleID != Admission_Sample) %>%
  mutate(
    Bray_Distance = mapply(function(ref, other) {
      bray_arcsin_s[ref, other]
    }, Admission_Sample, SampleID)
  ) %>%
  dplyr::select(Dog_ID, Sample_Site, Admission_Sample, SampleID, Day_Order,
         Sample_Plate, Collection_Day, Bray_Distance)
  
# Already grouped by dog 
# Random effects to account for per dog replicates
lmer_bray <- lmer(
  Bray_Distance ~ Day_Order*Sample_Site + (1 | Dog_ID),
  data = admission_vs_all_same_site
)

anova(lmer_bray)
summary(lmer_bray)
 
admission_vs_all_same_site$fitted <- fitted(lmer_bray)
admission_vs_all_same_site$resid <- resid(lmer_bray)

# Fitted vs residuals
ggplot(admission_vs_all_same_site, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(x = "Fitted values", y = "Residuals")

# Residuals vs each independent variable
ggplot(admission_vs_all_same_site, aes(x = Day_Order, y = resid)) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(x = "Day_Order", y = "Residuals")

ggplot(admission_vs_all_same_site, aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  facet_wrap(~ Dog_ID) +
  theme_classic() +
  labs(title = "Residual Q-Q plots by Dog_ID")


 

# estimated marginal means of Day-Order within each site (control and surgical)
emm <- emmeans(lmer_bray, ~ Day_Order|Sample_Site)

alpha <- 0.05
 
contrasts <- as.data.frame(pairs(emm, adjust = "tukey"))

contrasts <- contrasts %>%
  mutate(
    Group1 = case_when(
      str_detect(contrast, "^\\(") ~ str_match(contrast, "^\\((.*?)\\)")[,2],
      TRUE ~ str_match(contrast, "^(.*?) - ")[,2]
    ),
    Group2 = case_when(
      str_detect(contrast, "\\) - \\(") ~ str_match(contrast, "- \\((.*?)\\)$")[,2],
      str_detect(contrast, "^\\(") ~ str_replace(contrast, "^\\(.*?\\) - ", ""),
      TRUE ~ str_match(contrast, "- (.*)$")[,2]
    ),
    Group1 = str_remove_all(str_trim(Group1), "[()]"),
    Group2 = str_remove_all(str_trim(Group2), "[()]")
  )

pairwise_tidy <- contrasts %>%
  filter(!is.na(p.value)) %>% 
  mutate(
    Significant = ifelse(p.value <= alpha, "Yes", "No"),
    Stars = case_when(
      p.value <= 0.001 ~ "***",
      p.value <= 0.01  ~ "**",
      p.value <= 0.05  ~ "*",
      p.value <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p.value = round(p.value, 4) ) %>%
  arrange(p.value) 

# Filter out insignificant pairwise results
sig_df <- pairwise_tidy %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,    
    label = Stars
  )
x_levels <- levels(factor(admission_vs_all_same_site$Day_Order))

sig_df <- sig_df %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p.value) %>%
  mutate(
    y.position = max(admission_vs_all_same_site$Bray_Distance, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )
 

 
emm_df <- as.data.frame(emm)[, c("Day_Order", "Sample_Site", "emmean", "SE", "df", "lower.CL", "upper.CL")]

emm_df_surg <- emm_df %>% 
  filter(Sample_Site == "Surgical")
 sig_df_surg <- sig_df %>% 
  filter(Sample_Site == "Surgical")
 
admission_vs_all_same_site_surg <-  admission_vs_all_same_site %>% 
  filter(Sample_Site == "Surgical")

print(sig_df_surg)

ggplot(admission_vs_all_same_site_surg, aes(x = Day_Order, y = Bray_Distance)) +
  geom_boxplot(fill = "#3399FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = emm_df_surg$emmean[1], linetype = "dashed", color="red") +
  geom_point(data = emm_df_surg, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = emm_df_surg, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Surgical-Site: BC Dissimilarity to Admission Across Timepoints",
    subtitle = "Trend line indicating estimated marginal means per timepoint, horizontal red line indicates emmean of Pre-Surgery",
    x = "Timepoint",
    y = "Bray-Curtis Dissimilarity to Admission Sample"
  ) +
  theme_minimal() +
  stat_pvalue_manual(
    sig_df_surg,
    label = "label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.5,
    size = 5
  ) +
    theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  ) 
  

 emm_df_ctrl <- emm_df %>% 
  filter(Sample_Site == "Control")
sig_df_ctrl <- sig_df %>% 
  filter(Sample_Site == "Control")

print(sig_df_ctrl)

admission_vs_all_same_site_ctrl <-  admission_vs_all_same_site %>% 
  filter(Sample_Site == "Control")

ggplot(admission_vs_all_same_site_ctrl, aes(x = Day_Order, y = Bray_Distance)) +
  geom_boxplot(fill = "#CC99FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = emm_df_ctrl$emmean[1], linetype = "dashed", color="red") +
  geom_point(data = emm_df_ctrl, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = emm_df_ctrl, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Control-Site: BC Dissimilarity to Admission Across Timepoints",
    subtitle = "Trend line indicating estimated marginal means per timepoint, horizontal red line indicates emmean of Admission",
    x = "Timepoint",
    y = "Bray-Curtis Dissimilarity to Admission Sample"
  ) +
  theme_minimal()  + 
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )  
```

### PCoA Axis 1: Control vs. Surgical 

Axis 1 explains about 8.99% of the variation 

Model =  lmer(pc1 ~ Day_Order*Sample_Site + (1 | Dog_ID))
* Contrasts between estimated marginal means of timepoints from model, using Tukey-adjusting p-values 

**Results:**

Surgical Samples: 
  * Admission vs Post Surgery: estimate = 0.1127, adj-p = 0.0488
  * Post Surgery vs Week 2: estimate = -0.1806, adj-p = 0.0003
  * Post Surgery vs Week 3: estimate = -0.1664, adj-p = 0.0028
  * Post Surgery vs Week 4: estimate = -0.1832, adj-p = 0.0001
  * Post Surgery vs Week 5: estimate = -0.1834, adj-p = 0.0047
  * Post Surgery vs Week 6: estimate = -0.2049, adj-p = 0.0002
  * Day 1 vs Week 4: estimate = -0.1113, adj-p = 0.0351
  * Day 1 vs Week 2: estimate = -0.1087, adj-p = 0.0474
 
Control samples:
  * Day 1 vs Week 5: estimate = -0.1476, adj-p =  0.0221 
  * Admission vs Week 5: estimate = -0.1123, adj-p = 0.0278

```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
 # PCoA
set.seed(100)

bray_pcoa <- ordinate(ps_rel_2, method = "PCoA", distance =  bray_dist)
 

scores <- bray_pcoa$vectors
pc1 <- scores[, 1]
pc2 <- scores[, 2]

pc1_scores_df <- as.data.frame(pc1) %>%
  rownames_to_column("SampleID")  

pc2_scores_df <- as.data.frame(pc2) %>%
  rownames_to_column("SampleID")  

 
paired_sample_data_1 <- left_join(paired_sample_data, pc1_scores_df, by = "SampleID")
paired_sample_data_2 <- left_join(paired_sample_data_1, pc2_scores_df, by = "SampleID")


paired_sample_data_surgical <- paired_sample_data_2 %>%
    filter(Sample_Site == "Surgical")
  

paired_sample_data_control <- paired_sample_data_2 %>%
    filter(Sample_Site == "Control")


coord <- as.vector(paired_sample_data_2$pc1)

# Plot Distribution of PCoA 1 coordinates
histogram_distances <- ggplot(data = data.frame(coord), aes(x = coord)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of PCoA 1 coordinates",
    x = "PCoA coordinates",
    y = "Count"
  )

print(histogram_distances)

  

lmer_pcoa_1 <- lmer(data = paired_sample_data_2, pc1 ~ Day_Order*Sample_Site + (1 | Dog_ID))

anova(lmer_pcoa_1)
summary(lmer_pcoa_1)

paired_sample_data_2$fitted_pc1 <- fitted(lmer_pcoa_1)
paired_sample_data_2$resid_pc1 <- resid(lmer_pcoa_1)

# Fitted vs residuals
ggplot(paired_sample_data_2, aes(x = fitted_pc1, y = resid_pc1)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(x = "Fitted values", y = "Residuals")

# Residuals vs each independent variable
ggplot(paired_sample_data_2, aes(x = Day_Order, y = resid_pc1)) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(x = "Day_Order", y = "Residuals")

ggplot(paired_sample_data_2, aes(sample = resid_pc1)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  facet_wrap(~ Dog_ID) +
  theme_classic() +
  labs(title = "Residual Q-Q plots by Dog_ID")


 

# estimated marginal means of Day-Order within each site (control and surgical)
emm <- emmeans(lmer_pcoa_1, ~ Day_Order|Sample_Site)

alpha <- 0.05
 
contrasts <- as.data.frame(pairs(emm, adjust = "tukey"))

contrasts <- contrasts %>%
  mutate(
    Group1 = case_when(
      str_detect(contrast, "^\\(") ~ str_match(contrast, "^\\((.*?)\\)")[,2],
      TRUE ~ str_match(contrast, "^(.*?) - ")[,2]
    ),
    Group2 = case_when(
      str_detect(contrast, "\\) - \\(") ~ str_match(contrast, "- \\((.*?)\\)$")[,2],
      str_detect(contrast, "^\\(") ~ str_replace(contrast, "^\\(.*?\\) - ", ""),
      TRUE ~ str_match(contrast, "- (.*)$")[,2]
    ),
    Group1 = str_remove_all(str_trim(Group1), "[()]"),
    Group2 = str_remove_all(str_trim(Group2), "[()]")
  )

pairwise_tidy <- contrasts %>%
  filter(!is.na(p.value)) %>% 
  mutate(
    Significant = ifelse(p.value <= alpha, "Yes", "No"),
    Stars = case_when(
      p.value <= 0.001 ~ "***",
      p.value <= 0.01  ~ "**",
      p.value <= 0.05  ~ "*",
      p.value <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p.value = round(p.value, 6) ) %>%
  arrange(p.value) 

# Filter out insignificant pairwise results
sig_df <- pairwise_tidy %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,    
    label = Stars
  )
x_levels <- levels(factor(paired_sample_data_2$Day_Order))

sig_df <- sig_df %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p.value) %>%
  mutate(
    y.position = max(paired_sample_data_2$pc1, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )
print(sig_df)

 
emm_df <- as.data.frame(emm)[, c("Day_Order", "Sample_Site", "emmean", "SE", "df", "lower.CL", "upper.CL")]

emm_df_surg <- emm_df %>% 
  filter(Sample_Site == "Surgical")

sig_df_surg <- sig_df %>% 
  filter(Sample_Site == "Surgical")


ggplot(paired_sample_data_surgical, aes(x = Day_Order, y = pc1)) +
  geom_boxplot(fill = "#3399FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = emm_df_surg$emmean[1], linetype = "dashed", color="red") +
  geom_point(data = emm_df_surg, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = emm_df_surg, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Surgical-Site: PCoA 1 Across Timepoints",
    subtitle = "Trend line indicating estimated marginal means per timepoint, horizontal red line indicates emmean of Admission",
    x = "Timepoint",
    y = "PCoA 1"
  ) +
  theme_minimal() +
  
  stat_pvalue_manual(
    sig_df_surg,
    label = "label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.5,
    size = 5
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )
  

emm_df_ctrl <- emm_df %>% 
  filter(Sample_Site == "Control")

sig_df_ctrl <- sig_df %>% 
  filter(Sample_Site == "Control")

ggplot(paired_sample_data_control, aes(x = Day_Order, y = pc1)) +
  geom_boxplot(fill = "#CC99FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = emm_df_ctrl$emmean[1], linetype = "dashed", color="red") +
  geom_point(data = emm_df_ctrl, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = emm_df_ctrl, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Control-Site: PCoA 1 Across Timepoints",
    subtitle = "Trend line indicating estimated marginal means per timepoint, horizontal red line indicates emmean of Pre-Surgery",
    x = "Timepoint",
    y = "PCoA 1"
  ) +
  theme_minimal() +
  
  stat_pvalue_manual(
    sig_df_ctrl,
    label = "label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.5,
    size = 5
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  ) 
```

### PCoA Axis 2: Control vs. Surgical 

Axis 2 explains about 6.13% of the variation

Model =  lmer(pc2 ~ Day_Order*Sample_Site + (1 | Dog_ID))
* Contrasts between estimated marginal means of timepoints from model, using Tukey-adjusting p-values 

**Results:**

Surgical Samples: 
There are no significant differences between PCoA Axis 2 coordinates amongst timepoints

Control Samples: 
There are no significant differences between PCoA Axis 2 coordinates amongst timepoints

```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 

coord <- as.vector(paired_sample_data_2$pc2)
# Plot Distribution of PCoA coordinates
histogram_distances <- ggplot(data = data.frame(coord), aes(x = coord)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of PCoA 2 coordinates",
    x = "PCoA coordinates",
    y = "Count"
  )

print(histogram_distances)


 

lmer_pcoa_2 <- lmer(data = paired_sample_data_2, pc2 ~ Day_Order*Sample_Site + (1 | Dog_ID))

anova(lmer_pcoa_2)
summary(lmer_pcoa_2)

paired_sample_data_2$fitted_pc2 <- fitted(lmer_pcoa_2)
paired_sample_data_2$resid_pc2 <- resid(lmer_pcoa_2)

# Fitted vs residuals
ggplot(paired_sample_data_2, aes(x = fitted_pc2, y = resid_pc2)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(x = "Fitted values", y = "Residuals")

# Residuals vs each independent variable
ggplot(paired_sample_data_2, aes(x = Day_Order, y = resid_pc2)) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(x = "Day_Order", y = "Residuals")

ggplot(paired_sample_data_2, aes(sample = resid_pc2)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  facet_wrap(~ Dog_ID) +
  theme_classic() +
  labs(title = "Residual Q-Q plots by Dog_ID")


 

# estimated marginal means of Day-Order within each site (control and surgical)
emm <- emmeans(lmer_pcoa_2, ~ Day_Order|Sample_Site)

alpha <- 0.05
 
contrasts <- as.data.frame(pairs(emm, adjust = "tukey"))

contrasts <- contrasts %>%
  mutate(
    Group1 = case_when(
      str_detect(contrast, "^\\(") ~ str_match(contrast, "^\\((.*?)\\)")[,2],
      TRUE ~ str_match(contrast, "^(.*?) - ")[,2]
    ),
    Group2 = case_when(
      str_detect(contrast, "\\) - \\(") ~ str_match(contrast, "- \\((.*?)\\)$")[,2],
      str_detect(contrast, "^\\(") ~ str_replace(contrast, "^\\(.*?\\) - ", ""),
      TRUE ~ str_match(contrast, "- (.*)$")[,2]
    ),
    Group1 = str_remove_all(str_trim(Group1), "[()]"),
    Group2 = str_remove_all(str_trim(Group2), "[()]")
  )

pairwise_tidy <- contrasts %>%
  filter(!is.na(p.value)) %>% 
  mutate(
    Significant = ifelse(p.value <= alpha, "Yes", "No"),
    Stars = case_when(
      p.value <= 0.001 ~ "***",
      p.value <= 0.01  ~ "**",
      p.value <= 0.05  ~ "*",
      p.value <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p.value = round(p.value, 4) ) %>%
  arrange(p.value) 

# Filter out insignificant pairwise results
sig_df <- pairwise_tidy %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,    
    label = Stars
  )
x_levels <- levels(factor(paired_sample_data_2$Day_Order))

sig_df <- sig_df %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p.value) %>%
  mutate(
    y.position = max(paired_sample_data_2$pc2, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )
 

 
emm_df <- as.data.frame(emm)[, c("Day_Order", "Sample_Site", "emmean", "SE", "df", "lower.CL", "upper.CL")]

emm_df_surg <- emm_df %>% 
  filter(Sample_Site == "Surgical")
 sig_df_surg <- sig_df %>% 
  filter(Sample_Site == "Surgical")
 
print(sig_df_surg)
ggplot(paired_sample_data_surgical, aes(x = Day_Order, y = pc2)) +
  geom_boxplot(fill = "#3399FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = emm_df_surg$emmean[1], linetype = "dashed", color="red") +
  geom_point(data = emm_df_surg, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = emm_df_surg, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Surgical-Site: PCoA 2 Across Timepoints",
    subtitle = "Trend line indicating estimated marginal means per timepoint, horizontal red line indicates emmean of Admission",
    x = "Timepoint",
    y = "PCoA 2"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )
  

 emm_df_ctrl <- emm_df %>% 
  filter(Sample_Site == "Control")
sig_df_ctrl <- sig_df %>% 
  filter(Sample_Site == "Control")

print(sig_df_ctrl)
ggplot(paired_sample_data_control, aes(x = Day_Order, y = pc2)) +
  geom_boxplot(fill = "#CC99FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = emm_df_ctrl$emmean[1], linetype = "dashed", color="red") +
  geom_point(data = emm_df_ctrl, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = emm_df_ctrl, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Control-Site: PCoA 2 Across Timepoints",
    subtitle = "Trend line indicating estimated marginal means per timepoint, horizontal red line indicates emmean of Admission",
    x = "Timepoint",
    y = "PCoA 2"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )  
```
 
### Shannon Index: Control vs. Surgical 

Model =  lmer(Shannon ~ Day_Order*Sample_Site + (1 | Dog_ID))
* Contrasts between estimated marginal means of timepoints from model, using Tukey-adjusting p-values 

**Results:**

Surgical Samples: 
There are no significant differences amongst timepoints

Control Samples: 
  * Week 2 vs Week 4: estimate = 0.8053, adj-p = 0.0218

```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
 
# get vector for histogram
shannon_vec <- as.vector(alphadiv_plot_data$Shannon)

# Plot Distribution of Shannon Diversity
histogram_distances <- ggplot(data = data.frame(shannon_vec), aes(x = shannon_vec)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of Shannon Diversity",
    x = "Shannon Diversity",
    y = "Count"
  )

print(histogram_distances)
 
lmer_shannon  <- lmer(data = alphadiv_plot_data, Shannon ~  Day_Order*Sample_Site + (1 | Dog_ID))
 
summary(lmer_shannon)
anova(lmer_shannon)

alphadiv_plot_data$fitted_shan <- fitted(lmer_shannon)
alphadiv_plot_data$resid_shan <- resid(lmer_shannon)

# Fitted vs residuals
ggplot(alphadiv_plot_data, aes(x = fitted_shan, y = resid_shan)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(x = "Fitted values", y = "Residuals")

# Residuals vs each independent variable
ggplot(alphadiv_plot_data, aes(x = Day_Order, y = resid_shan)) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(x = "Day_Order", y = "Residuals")

ggplot(alphadiv_plot_data, aes(sample = resid_shan)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  facet_wrap(~ Dog_ID) +
  theme_classic() +
  labs(title = "Residual Q-Q plots by Dog_ID")


# estimated marginal means of Day-Order within each site (control and surgical)
emm <- emmeans(lmer_shannon, ~ Day_Order|Sample_Site)

alpha <- 0.05
 
contrasts <- as.data.frame(pairs(emm, adjust = "tukey"))

contrasts <- contrasts %>%
  mutate(
    Group1 = case_when(
      str_detect(contrast, "^\\(") ~ str_match(contrast, "^\\((.*?)\\)")[,2],
      TRUE ~ str_match(contrast, "^(.*?) - ")[,2]
    ),
    Group2 = case_when(
      str_detect(contrast, "\\) - \\(") ~ str_match(contrast, "- \\((.*?)\\)$")[,2],
      str_detect(contrast, "^\\(") ~ str_replace(contrast, "^\\(.*?\\) - ", ""),
      TRUE ~ str_match(contrast, "- (.*)$")[,2]
    ),
    Group1 = str_remove_all(str_trim(Group1), "[()]"),
    Group2 = str_remove_all(str_trim(Group2), "[()]")
  )

pairwise_tidy <- contrasts %>%
  filter(!is.na(p.value)) %>% 
  mutate(
    Significant = ifelse(p.value <= alpha, "Yes", "No"),
    Stars = case_when(
      p.value <= 0.001 ~ "***",
      p.value <= 0.01  ~ "**",
      p.value <= 0.05  ~ "*",
      p.value <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p.value = round(p.value, 4) ) %>%
  arrange(p.value) 

# Filter out insignificant pairwise results
sig_df <- pairwise_tidy %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,    
    label = Stars
  )
x_levels <- levels(factor(alphadiv_plot_data$Day_Order))

sig_df <- sig_df %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p.value) %>%
  mutate(
    y.position = max(alphadiv_plot_data$Shannon, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )
 

 
emm_df <- as.data.frame(emm)[, c("Day_Order", "Sample_Site", "emmean", "SE", "df", "lower.CL", "upper.CL")]

alphadiv_plot_data_surg <- alphadiv_plot_data %>% 
  filter(Sample_Site == "Surgical")

emm_df_surg <- emm_df %>% 
  filter(Sample_Site == "Surgical")
 sig_df_surg <- sig_df %>% 
  filter(Sample_Site == "Surgical")
 
print(sig_df_surg)

ggplot(alphadiv_plot_data_surg, aes(x = Day_Order, y = Shannon)) +
  geom_boxplot(fill = "#3399FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = emm_df_surg$emmean[1], linetype = "dashed", color="red") +
  geom_point(data = emm_df_surg, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = emm_df_surg, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Surgical-Site: Shannon Diversity Across Timepoints",
    subtitle = "Trend line indicating estimated marginal means per timepoint, horizontal red line indicates emmean of Admission",
    x = "Timepoint",
    y = "Shannon Diversity"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )

 
 

alphadiv_plot_data_ctrl <- alphadiv_plot_data %>% 
  filter(Sample_Site == "Control")
emm_df_ctrl <- emm_df %>% 
  filter(Sample_Site == "Control")
 sig_df_ctrl <- sig_df %>% 
  filter(Sample_Site == "Control")
 

print(sig_df_ctrl)
ggplot(alphadiv_plot_data_ctrl, aes(x = Day_Order, y = Shannon)) +
  geom_boxplot(fill = "#CC99FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = emm_df_ctrl$emmean[1], linetype = "dashed", color="red") +
  geom_point(data = emm_df_ctrl, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = emm_df_ctrl, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Control-Site: Shannon Diversity Across Timepoints",
    subtitle = "Trend line indicating estimated marginal means per timepoint, horizontal red line indicates emmean of Admission",
    x = "Timepoint",
    y = "Shannon Diversity"
  ) +
  theme_minimal() +
  
  stat_pvalue_manual(
    sig_df_ctrl,
    label = "label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.5,
    size = 5
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )
```



 

```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE} 
# Surgical Samples
tax_df <- as.data.frame(tax_table(ps_rel_2))

# rename genus_species to either Unknown or Genus_sp. 
tax_df <- tax_df %>%
  dplyr::mutate(
    genus_species = dplyr::if_else(
      is.na(Genus) | Genus == "",
      "Unknown",
      dplyr::if_else(
        !is.na(Species) & Species != "",
        paste0(Genus, "_", Species),
        paste0(Genus, " sp.")
      )
  ))

tax_table(ps_rel_2) <- tax_table(as.matrix(tax_df))

# Subset to just surgical samples  
ps_rel_2_surg <- ps_rel_2 %>%
  subset_samples(
    Sample_Site == "Surgical")  

species_df <- psmelt(ps_rel_2_surg) %>%
  # Sum genus_species ASV counts for each sample (Dog-Day) 
  group_by(Dog_ID, genus_species, Day_Order) %>%
  summarise(
    total_rel_abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  )  

top5_by_day <- species_df %>%
  group_by(Day_Order, genus_species) %>%
  summarise(
    mean_rel_abun = mean(total_rel_abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Day_Order) %>%
  slice_max(order_by = mean_rel_abun, n = 5, with_ties = FALSE) %>%
  ungroup()
 
top_species <- unique(top5_by_day$genus_species)

species_subset <- species_df %>% 
  dplyr::mutate(
    genus_species = dplyr::if_else(
      genus_species %in% top_species,
      genus_species,
      "Other"
    )
  )%>% 
  # for each day get the relative abundance of each species
  group_by(genus_species, Day_Order) %>%
  summarise(mean_rel_abundance = mean(total_rel_abundance, na.rm = TRUE),
    .groups = "drop"
  )
  

species_colors <- c(
  "Unknown" = "gray",
  "Other" = "black",
  "Staphylococcus sp." = "#e41a1c",
  "Streptococcus sp." = "#377eb8",
  "Escherichia-Shigella sp." = "#a6761d",
  "Corynebacterium sp." = "#4daf4a",
  "Pseudomonas sp." = "#984ea3",
  "Lactococcus sp." = "#ffff33",
  "Acinetobacter sp." = "#a65628",
  "Bacteroides sp." = "#66c2a5",
  "Mycoplasmopsis_cynos" = "#b15928",
  "Enterococcus sp." = "#8da0cb",
  "Xanthomonas sp." = "#80b1d3",
  "Conchiformibius sp." = "#ff7f00",
  "[Ruminococcus] gnavus group sp." = "#fdb462",
  "Bifidobacterium sp." = "#ffd92f"
)


#  Compute ordering of genus_species based on mean abundance in the Before group
before_order <- species_subset %>%
  filter(Day_Order == "Admission") %>%
  group_by(genus_species) %>%
  summarise(mean_before = mean(mean_rel_abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_before)) %>%
  pull(genus_species)

before_order <- c(setdiff(before_order, "Other"), "Other")

#  Apply this order to the full data (all Time_Groups)
species_df_2 <- species_subset %>%
  mutate(genus_species = factor(genus_species, levels = before_order))

# 3. Plot with that fixed fill order
plot <- species_df_2 %>% 
  ggplot(aes(x = Day_Order, y = mean_rel_abundance, fill = genus_species)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = species_colors, drop = FALSE) +
  scale_y_continuous(
    limits = c(0, .65),
    breaks = seq(0, .65, by = 0.1)
  ) +
  labs(
    title = "Top Species Across Groups",
    x = "Timepoint ",
    y = "Average Relative Abundance",
    fill = "Species"
  )+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot
```
 
 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE} 
Control Samples
# Subset to just surgical samples  
ps_rel_2_ctrl <- ps_rel_2 %>%
  subset_samples(
    Sample_Site == "Control")  
 

# rename genus_species to either Unknown or Genus_sp. 
species_df <- psmelt(ps_rel_2_ctrl) %>%
  dplyr::mutate(
    genus_species = dplyr::if_else(
      is.na(Genus) | Genus == "",
      "Unknown",
      dplyr::if_else(
        !is.na(Species) & Species != "",
        paste0(Genus, "_", Species),
        paste0(Genus, " sp.")
      )
    )
  ) %>%
  # Sum genus_species ASV counts for each sample (Dog-Day) 
  group_by(Dog_ID, genus_species, Day_Order) %>%
  summarise(
    total_rel_abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  )  %>% 
  dplyr::mutate(
    genus_species = dplyr::if_else(
      genus_species %in% top_species,
      genus_species,
      "Other"
    )
  )%>% 
  # for each day get the relative abundance of each species
  group_by(genus_species, Day_Order) %>%
  summarise(mean_rel_abundance = mean(total_rel_abundance, na.rm = TRUE),
    .groups = "drop"
  )
 


species_colors <- c(
  "Bacteroides_vulgatus" = "#1b9e77",
  "Conchiformibius_steedae" = "#d95f02",
  "Corynebacterium_kroppenstedtii" = "#7570b3",
  "Corynebacterium_matruchotii" = "#e7298a",
  "Corynebacterium_otitidis" = "#66a61e",
  "Erysipelotrichaceae UCG-003_bacterium" = "red",
  "Escherichia-Shigella sp." = "#a6761d",
  "Gemella_palaticanis" = "#666666",
  "Geoalkalibacter sp." = "#1f78b4",
  "Herbaspirillum sp." = "#b2df8a",
  "Malacoplasma sp." = "#fb9a99",
  "Megamonas_funiformis" = "#cab2d6",
  "Mycoplasmopsis_canis" = "#fdbf6f",
  "Mycoplasmopsis_cynos" = "#b15928",
  "Other" = "gray",
  "Porphyromonas_cangingivalis" = "#6a3d9a",
  "Porphyromonas_gulae" = "#ff7f00",
  "Pseudarthrobacter sp." = "#8dd3c7",
  "Ureaplasma sp." = "#ffffb3",
  "Weissella_koreensis" = "#bebada",
  "Xanthomonas sp." = "#80b1d3",
  "[Ruminococcus] gnavus group sp." = "#fdb462"
)

#  Compute ordering of genus_species based on mean abundance in the Before group
before_order <- species_df %>%
  filter(Day_Order == "Admission") %>%
  group_by(genus_species) %>%
  summarise(mean_before = mean(mean_rel_abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_before)) %>%
  pull(genus_species)

before_order <- c(setdiff(before_order, "Other"), "Other")

#  Apply this order to the full data (all Time_Groups)
species_df_2 <- species_df %>%
  mutate(genus_species = factor(genus_species, levels = before_order))

# 3. Plot with that fixed fill order
plot <- species_df_2 %>% 
  ggplot(aes(x = Day_Order, y = mean_rel_abundance, fill = genus_species)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = species_colors, drop = FALSE) +
  scale_y_continuous(
    limits = c(0, 0.13),
    breaks = seq(0, 0.13, by = 0.01)
  ) +
  labs(
    title = "Top Species Across Groups",
    x = "Timepoint ",
    y = "Average Relative Abundance",
    fill = "Species"
  )+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot
```
 
 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```

