---
title: "Canine Skin Microbiome: 16S Differential Abundance Analysis "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

**Purpose:**
Identify bacterial taxa at the genus and ASV levels that differ between surgical and control groups from baseline (Admission and Pre-Surgery Day 0) to the immediate post-surgery timepoint.

**Overview:**
Taxa present in at least 25% of surgical samples across the pre- and post-surgery timepoints were retained for analysis. For each of the 20 dogs, paired relative abundances before and immediately after surgery were compared using Wilcoxon signed-rank tests.

For taxa with q-values ≤0.10, effect sizes were calculated as the paired Wilcoxon location-shift estimate (i.e., the estimated median difference in relative abundance between timepoints). These results were visualized in a heatmap using the signed significance score: sign(estimate) * -log10(p.value). The same set of significant taxa was subsequently tested using paired Wilcoxon signed-rank tests to compare baseline with later healing-stage timepoints.


**Results:**

* Total taxa present in at least 25% of surgical samples across the before- and post-surgery groups:

  - Genus: 314 taxa tested; 32 had q-values \(\leq 0.1\)
  - ASVs: 722 taxa tested; 2 had q-values \(\leq 0.1\)

Immediately after surgery, the most differentially abundant genera generally showed decreased relative abundance compared with baseline. In contrast, later healing stages showed fewer differences relative to baseline. Control samples also tended to show lower relative abundances at the post-surgery timepoint relative to baseline; however, these changes showed limited statistical significance and were less pronounced than those observed at surgical sites.



# Load files 
## R packages
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(knitr)
library(FSA)
library(tidyverse)
library(vegan)
library(phyloseq)
library(dendextend)
library(ggplot2)
library(ggtree)
library(scales)  
library(RColorBrewer)
library(segmented)
library(stringr)
library(pairwiseAdonis)
library(ggpubr)
library(changepoint)
library(dplyr)
library(lme4)
library(emmeans)
library(nortest)
library(mcp)
library(purrr)
library(kableExtra)
library(forcats)
library(microbiome)
library(lmerTest)       
library(broom.mixed)
library(ComplexHeatmap)
library(circlize)
```
 
## Import paired-dataset
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Paired-only ASV feature table
ps <- readRDS("../01_input_files/phyloseq/ps_2.rds")
 
sd <- as(sample_data(ps), "data.frame")

# define healing stages
sd <- sd %>%
  mutate(
    Time_Group = case_when(
      Day_Order %in% c("Admission", "Day 0a - Pre Surgery") ~ "Before_Surgery",
      Day_Order %in% c("Day 0c - Post Surgery")           ~ "Post_Surgery",
      Day_Order %in% c("Day 1", "Day 3")                  ~ "Inflammation",
      Day_Order %in% c("Week 1", "Week 2")                ~ "Proliferation",
      Day_Order %in% c("Week 3", "Week 4", "Week 5", "Week 6")  ~ "Maturation", 
      TRUE                                                 ~ NA_character_
    ),
    Time_Category = case_when(
      Day_Order %in% c("Admission", "Day 0a - Pre Surgery") ~ "Before",
      TRUE                                                  ~ "After"
    )
  )


sd$Time_Group <- factor(sd$Time_Group, levels = c("Before_Surgery", "Post_Surgery", "Inflammation", "Proliferation", "Maturation"))
  
sample_data(ps) <- sample_data(sd)

range(sample_sums(ps)) 
mean(sample_sums(ps))
median(sample_sums(ps))
```

## TSS Normalization (relative abundances) of ASV + Genus
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Normalization of ASV counts 
ps_ASV_rel <- transform_sample_counts(ps, function(x) x / sum(x))
 
# generate df
ASV_rel_df <- psmelt(ps_ASV_rel) %>%
arrange(Dog_ID, Genus, Day_Order)

ASV_rel_df_surg <- ASV_rel_df%>%
   filter(Sample_Site == "Surgical") 
    
  

ASV_rel_df_ctrl <- ASV_rel_df%>%
   filter(Sample_Site == "Control")

# Unique ASVs
ntaxa(ps)
tax <- tax_table(ps)

# % ASVs removed due to unknown Genus 
no_genus <- is.na(tax[, "Genus"]) | tax[, "Genus"] == ""
n_no_genus <- sum(no_genus, na.rm = TRUE)
n_total_asv <- ntaxa(ps)
pct_no_genus <- 100 * n_no_genus / n_total_asv

n_no_genus
pct_no_genus

cat("Excluded", n_no_genus, "ASVs without genus assignment (", round(pct_no_genus, 1), "% of total ASVs).\n")


# Aggregate ASV counts for Genus-Level Analysis + remove ASVs with unknown Genus
ps_genus <- tax_glom(ps, taxrank = "Genus", NArm = TRUE)
#Unique genus
ntaxa(ps_genus)

ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))
 
# ntaxa should be smaller since we aggregated at Genus level 
ntaxa(ps_genus_rel)
 

# turn phyloseq object into a dataframe 
genus_rel_df <- psmelt(ps_genus_rel) %>%
arrange(Dog_ID, Genus, Day_Order)

# Check for genus-level (each sample should have one row for each Genus, if multiple appear then it is still at ASV-level)
check <- genus_rel_df %>%
  count(Sample, Genus) %>%
  filter(n > 1)
 check
 
genus_rel_df_surg <- genus_rel_df%>%
   filter(Sample_Site == "Surgical")

 
genus_rel_df_ctrl <- genus_rel_df%>%
   filter(Sample_Site == "Control")
```


# Genus-level

## Surgical Samples  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
run_paired_wilcoxon_by_genus <- function(
    data,
    time_group_1,
    time_group_2
) {
  
  required_columns <- c("Genus", time_group_1, time_group_2)
  
  missing_columns <- setdiff(required_columns, names(data))
  
  if (length(missing_columns) > 0) {
    stop(
      "The following columns are missing: ",
      paste(missing_columns, collapse = ", ")
    )
  }
  
  results <- data %>%
  group_by(Genus) %>%
  group_modify(~ {

    dat <- .x %>%
      filter(
        !is.na(.data[[time_group_1]]),
        !is.na(.data[[time_group_2]])
      )

    group_1_values <- dat[[time_group_1]]
    group_2_values <- dat[[time_group_2]]

    differences <- group_1_values - group_2_values

    n_pairs <- length(differences)
    n_zero_diff <- sum(differences == 0)
    n_nonzero_diff <- sum(differences != 0)

    test <- wilcox.test(
      group_1_values,
      group_2_values,
      alternative = "two.sided",
      paired = TRUE,
      exact = FALSE,
      conf.int = TRUE
    )

    tibble(
      n_pairs = n_pairs,
      n_zero_diff = n_zero_diff,
      n_nonzero_diff = n_nonzero_diff,

      median_group_1 = median(group_1_values, na.rm = TRUE),
      median_group_2 = median(group_2_values, na.rm = TRUE),
      median_difference = median(differences, na.rm = TRUE),

      stat = unname(test$statistic),
      p = test$p.value,
      estimate = unname(test$estimate),
      conf_low = test$conf.int[1],
      conf_high = test$conf.int[2]
    )
  }) %>%
  ungroup() %>%
  mutate(
    p_adj = p.adjust(p, method = "BH")
  ) %>%
  arrange(p_adj, p, Genus)
  
  return(results)
}

```
 
Distribution of taxa at different prevalence thresholds
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups)
 
prevalence_thresholds <- function(data, threshold) {
  total_samples <- n_distinct(data$Sample)

  data %>%
    group_by(Genus) %>%
    summarise(
      prevalence = n_distinct(Sample[Abundance > 0]) / total_samples,
      .groups = "drop"
    ) %>%
    filter(prevalence >= threshold) %>%
    summarise(n_genera = n()) %>%
    pull(n_genera)
}

total_samples <- n_distinct(genus_rel_df_surg_2$Sample)

threshold_results <- tibble(
  prevalence_threshold = seq(0.05, 0.50, by = 0.05)
) %>%
  mutate(
    min_samples_required = ceiling(total_samples * prevalence_threshold),
    n_genera = map_int(
      prevalence_threshold,
      ~ prevalence_thresholds(genus_rel_df_surg_2, .x)
    )
  )
 
 threshold_results
 
 ggplot(
  threshold_results,
  aes(x = prevalence_threshold, y = n_genera)
) +
   geom_line(linewidth = 0.9, color = "#2C7FB8") +
  geom_point(size = 2.5, color = "#2C7FB8") +
  scale_x_continuous(
    breaks = seq(0.05, 0.50, by = 0.05),
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    x = "Minimum prevalence threshold",
    y = "Number of genera retained",
    title = "Genera retained across prevalence thresholds"
  ) +
  theme_classic(base_size = 13)+ 
  geom_vline(
    xintercept = 0.25,
    linetype = "dashed",
    color = "firebrick"
  )
```

### Before_Surgery (Admission + Pre-Surgery (Day 0)) vs Post-Surgery (Day 0)
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify genera present in at least 25% of samples
taxa_present_10pct <- genus_rel_df_surg_2 %>%
  group_by(Genus) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.25) %>%
  dplyr::select(Genus)

# Number of retained genera
filtered_genus_count <- n_distinct(taxa_present_10pct$Genus)

paste(
  "Total genera present in at least 10% of samples:",
  filtered_genus_count
)
# Keep genera satisfying the prevalence criterion
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "Genus")


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_postsurg <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_postsurg

results_wt_annotated_sig <- results_wt_annotated_postsurg %>%
  filter(results_wt_annotated_postsurg$p_adj <= 0.1)%>%
  arrange(p)

results_wt_annotated_sig

nrow(results_wt_annotated_postsurg)

significant_genera_1 <- unique(results_wt_annotated_sig$Genus) 

 
```
 
### Inflammation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Inflammation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_inflam <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_inflam
  
```
 
### Proliferation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Proliferation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
   dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_prolif <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_prolif
  
```

### Maturation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Maturation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
   dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_matur <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_matur
  
```

### Heatmap
```{r,   message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 

results_wt_annotated <- bind_rows(results_wt_annotated_postsurg, results_wt_annotated_inflam, results_wt_annotated_prolif, results_wt_annotated_matur)

 # get signed log(p-value) for heatmap
 results <- results_wt_annotated %>%
  mutate(
    signed_log10_p = -(sign(estimate) * -log10(p))
  )
 
# generate per-taxa matrix of signed-log(pvalues)
 score_matrix <- results %>%
  dplyr::select(Genus, Time_Group_2, signed_log10_p) %>%
  pivot_wider(
    names_from = Time_Group_2,
    values_from = signed_log10_p
  ) %>%
  column_to_rownames("Genus") %>%
  as.matrix()
  
heatmap_matrix <- score_matrix[
  complete.cases(score_matrix),
  ,
  drop = FALSE
]
 
 

signed_max <- max(abs(heatmap_matrix), na.rm = TRUE)

col_fun <- circlize::colorRamp2(
  c(-4, 0, 4),
  c("#2166AC", "white", "#B2182B")
)
 


 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation"
)
 
 
 
col_dend <- heatmap_matrix |>
  t() |>
  dist(method = "euclidean") |>
  hclust(method = "complete") |>
  as.dendrogram()

# Rotate branches  
col_dend <- dendextend::rotate(
  col_dend,
  order = timepoint_order
)

ht <- ComplexHeatmap::Heatmap(
  heatmap_matrix,
  name = "Signed\n-log10(p)",
  col = col_fun,

  cluster_rows = TRUE,
  cluster_columns = col_dend,
  column_dend_reorder = FALSE,

  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",

  na_col = "grey90",
  column_title = "Differentially Significant Genus in Surgical Groups",
  row_title = "Genus",
  row_names_side = "right",
  row_names_gp = grid::gpar(fontsize = 7)
)

ComplexHeatmap::draw(ht)
 

row_idx <- ComplexHeatmap::row_order(ht)
row_names_in_heatmap_order <- rownames(heatmap_matrix)[row_idx]

row_names_in_heatmap_order
```
 
## Control Samples  
 
### Before_Surgery (Admission + Pre-Surgery (Day 0)) vs Post-Surgery (Day 0)
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
  dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_postsurg <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_postsurg
  
```
 
### Inflammation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Inflammation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
  dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_inflam <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_inflam
  
```
 
### Proliferation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Proliferation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
   dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_prolif <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_prolif
  
```

### Maturation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Maturation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
   dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_matur <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_matur
  
```

### Heatmap
```{r,  message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 

results_wt_annotated <- bind_rows(results_wt_annotated_postsurg, results_wt_annotated_inflam, results_wt_annotated_prolif, results_wt_annotated_matur)
 
 # get signed log(p-value) for heatmap
 results <- results_wt_annotated %>%
  mutate(
    signed_log10_p = -(sign(estimate) * -log10(p))
  )
 
# generate per-taxa matrix of signed-log(pvalues)
 score_matrix <- results %>%
  dplyr::select(Genus, Time_Group_2, signed_log10_p) %>%
  pivot_wider(
    names_from = Time_Group_2,
    values_from = signed_log10_p
  ) %>%
  column_to_rownames("Genus") %>%
  as.matrix()
  
heatmap_matrix <- score_matrix[
  complete.cases(score_matrix),
  ,
  drop = FALSE
]
 
 

signed_max <- max(abs(heatmap_matrix), na.rm = TRUE)
 
col_fun <- circlize::colorRamp2(
  c(-4, 0, 4),
  c("#2166AC", "white", "#B2182B")
)
 


 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation"
)
 
 
 
col_dend <- heatmap_matrix |>
  t() |>
  dist(method = "euclidean") |>
  hclust(method = "complete") |>
  as.dendrogram()

# Rotate branches  
col_dend <- dendextend::rotate(
  col_dend,
  order = timepoint_order
)

# Reorder matrix rows explicitly
heatmap_matrix_ordered <- heatmap_matrix[row_names_in_heatmap_order, , drop = FALSE]

ht <- ComplexHeatmap::Heatmap(
  heatmap_matrix_ordered,
  name = "Signed\n-log10(p)",
  col = col_fun,

  cluster_rows = FALSE,
  cluster_columns = col_dend,
  column_dend_reorder = FALSE,

  na_col = "grey90",
  column_title = "Selected Genera in Control Groups",
  row_title = "Genus",
  row_names_side = "right",
  row_names_gp = grid::gpar(fontsize = 7)
)

ComplexHeatmap::draw(ht)

```


### Plot taxa 
```{r, fig.width=16, fig.height=11, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
genus_rel_df_sig <- genus_rel_df %>% 
  filter(Sample_Site %in% c("Control", "Surgical"))%>%
  filter(Time_Group %in% c("Before_Surgery", "Post_Surgery", "Inflammation"))%>%
  filter(Genus %in% significant_genera_1) %>%
  group_by(Genus, Time_Group, Sample_Site, Dog_ID) %>%
  summarise(
    avg_rel_abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )
 

ggplot(
  genus_rel_df_sig,
  aes(
    x = Time_Group,
    y = avg_rel_abundance,
    fill = Sample_Site,
    group = interaction(Time_Group, Sample_Site)
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.65,
    outlier.alpha = 0.6
  ) +
  geom_jitter(
    aes(color = Sample_Site),
    position = position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.75
    ),
    size = 1.8,
    alpha = 0.7
  ) +
  facet_wrap(~ Genus, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Control" = "#7B2CBF",
      "Surgical" = "#0072B2"
    )
  ) +
  scale_color_manual(
    values = c(
      "Control" = "#7B2CBF",
      "Surgical" = "#0072B2"
    ),
    guide = "none"
  ) +
  labs(
    x = "Time group",
    y = "Relative abundance per dog",
    fill = "Sample site",
    title = "Genus abundance by time group and sample site"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )+ stat_summary(
    fun = median,
    geom = "point",
    aes(group = Sample_Site),
    position = position_dodge(width = 0.75),
    shape = 21,
    size = 2,
    stroke = 0.8,
    fill = "white",
    color = "black"
  )
  
```

 
  

 
 
# ASV-level
 
 Distribution of taxa at different prevalence thresholds
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)
 
prevalence_thresholds <- function(data, threshold) {
  total_samples <- n_distinct(data$Sample)

  data %>%
    group_by(OTU) %>%
    summarise(
      prevalence = n_distinct(Sample[Abundance > 0]) / total_samples,
      .groups = "drop"
    ) %>%
    filter(prevalence >= threshold) %>%
    summarise(n_genera = n()) %>%
    pull(n_genera)
}

total_samples <- n_distinct(genus_rel_df_surg_2$Sample)

threshold_results <- tibble(
  prevalence_threshold = seq(0.05, 0.50, by = 0.05)
) %>%
  mutate(
    min_samples_required = ceiling(total_samples * prevalence_threshold),
    n_genera = map_int(
      prevalence_threshold,
      ~ prevalence_thresholds(genus_rel_df_surg_2, .x)
    )
  )
 
 threshold_results
 
 ggplot(
  threshold_results,
  aes(x = prevalence_threshold, y = n_genera)
) +
   geom_line(linewidth = 0.9, color = "#2C7FB8") +
  geom_point(size = 2.5, color = "#2C7FB8") +
  scale_x_continuous(
    breaks = seq(0.05, 0.50, by = 0.05),
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    x = "Minimum prevalence threshold",
    y = "Number of genera retained",
    title = "Genera retained across prevalence thresholds"
  ) +
  theme_classic(base_size = 13)+ 
  geom_vline(
    xintercept = 0.25,
    linetype = "dashed",
    color = "firebrick"
  )
```

## Surgical Samples  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
run_paired_wilcoxon_by_ASV <- function(
    data,
    time_group_1,
    time_group_2,
    exact = TRUE
) {
  
  required_columns <- c("OTU", time_group_1, time_group_2)
  
  missing_columns <- setdiff(required_columns, names(data))
  
  if (length(missing_columns) > 0) {
    stop(
      "The following columns are missing: ",
      paste(missing_columns, collapse = ", ")
    )
  }
  
  results <- data %>%
  group_by(OTU) %>%
  group_modify(~ {

    dat <- .x %>%
      filter(
        !is.na(.data[[time_group_1]]),
        !is.na(.data[[time_group_2]])
      )

    group_1_values <- dat[[time_group_1]]
    group_2_values <- dat[[time_group_2]]

    differences <- group_1_values - group_2_values

    n_pairs <- length(differences)
    n_zero_diff <- sum(differences == 0)
    n_nonzero_diff <- sum(differences != 0)

    test <- wilcox.test(
      group_1_values,
      group_2_values,
      paired = TRUE,
      exact = exact,
      conf.int = TRUE
    )

    tibble(
      n_pairs = n_pairs,
      n_zero_diff = n_zero_diff,
      n_nonzero_diff = n_nonzero_diff,

      median_group_1 = median(group_1_values, na.rm = TRUE),
      median_group_2 = median(group_2_values, na.rm = TRUE),
      median_difference = median(differences, na.rm = TRUE),

      stat = unname(test$statistic),
      p = test$p.value,

      wilcox_estimate = unname(test$estimate),
      conf_low = test$conf.int[1],
      conf_high = test$conf.int[2]
    )
  }) %>%
  ungroup() %>%
  mutate(
    p_adj = p.adjust(p, method = "BH")
  ) %>%
  arrange(p_adj, p, OTU)
  
  
  return(results)
}

```

### Before_Surgery (Admission + Pre-Surgery) vs Post-Surgery 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
ASV_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify ASVs present in at least 10% of samples
taxa_present_10pct <- ASV_rel_df_surg_2 %>%
  group_by(OTU) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.25) %>%
  dplyr::select(OTU)

# Number of retained genera
filtered_ASV_count <- n_distinct(taxa_present_10pct$OTU)

cat(
  "Total ASV present in at least 10% of samples:",
  filtered_ASV_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
ASV_rel_df_surg_2_filt <- ASV_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "OTU")
 

# Summarise abundance per dog, genus, and time group
ASV_rel_df_median <- ASV_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_rel_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
ASV_rel_df_median_formatted <- ASV_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_rel_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = ASV_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )

 

tax_df <- tax_table(ps) %>%
  as("matrix") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>%
  tibble::as_tibble()

results_wt_annotated_postsurg <- results_wt %>%
  mutate(OTU = as.character(OTU)) %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_postsurg

results_wt_annotated_sig <- results_wt_annotated_postsurg %>%
  filter(results_wt_annotated_postsurg$p_adj <= 0.1)

results_wt_annotated_sig

nrow(results_wt_annotated_postsurg)

significant_genera_1 <- unique(results_wt_annotated_sig$OTU) 
```
 
### Inflammation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Inflammation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_inflam <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_inflam
  
```
 
### Proliferation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Proliferation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_prolif <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_prolif
  
```
 
### Maturation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Maturation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_matur <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_matur
  
```
 
### Heatmap
```{r,   message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 

results_wt_annotated <- bind_rows(results_wt_annotated_postsurg, results_wt_annotated_inflam, results_wt_annotated_prolif, results_wt_annotated_matur)

 # get signed log(p-value) for heatmap
 results <- results_wt_annotated %>%
  mutate(
    signed_log10_p = -(sign(wilcox_estimate) * -log10(p))
  )
 
# generate per-taxa matrix of signed-log(pvalues)
 score_matrix <- results %>%
  dplyr::select(OTU, Time_Group_2, signed_log10_p) %>%
  pivot_wider(
    names_from = Time_Group_2,
    values_from = signed_log10_p
  ) %>%
  column_to_rownames("OTU") %>%
  as.matrix()
  
heatmap_matrix <- score_matrix[
  complete.cases(score_matrix),
  ,
  drop = FALSE
]
 
 

signed_max <- max(abs(heatmap_matrix), na.rm = TRUE)
 
col_fun <- circlize::colorRamp2(
  c(-4, 0, 4),
  c("#2166AC", "white", "#B2182B")
)
 


 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation"
)
 
 
 
col_dend <- heatmap_matrix |>
  t() |>
  dist(method = "euclidean") |>
  hclust(method = "complete") |>
  as.dendrogram()

# Rotate branches  
col_dend <- dendextend::rotate(
  col_dend,
  order = timepoint_order
)

# Reorder matrix rows explicitly
#heatmap_matrix_ordered <- heatmap_matrix[row_names_in_heatmap_order, , drop = FALSE]

ht <- ComplexHeatmap::Heatmap(
  heatmap_matrix,
  name = "Signed\n-log10(p)",
  col = col_fun,

  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_dend_reorder = FALSE,

  na_col = "grey90",
  column_title = "Selected OTU in Surgical Groups",
  row_title = "OTU",
  row_names_side = "right",
  row_names_gp = grid::gpar(fontsize = 7)
)

ComplexHeatmap::draw(ht)

```


## Control Samples  
### Before_Surgery (Admission + Pre-Surgery) vs Post-Surgery 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_postsurg <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_postsurg
  
```
 
 
### Inflammation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Inflammation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_inflam <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_inflam
  
```
 
### Proliferation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Proliferation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_prolif <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_prolif
  
```
 
### Maturation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Maturation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_matur <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_matur
  
```
 
### Heatmap
```{r,  message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 

results_wt_annotated <- bind_rows(results_wt_annotated_postsurg, results_wt_annotated_inflam, results_wt_annotated_prolif, results_wt_annotated_matur)
 
 # get signed log(p-value) for heatmap
 results <- results_wt_annotated %>%
  mutate(
    signed_log10_p = -(sign(wilcox_estimate) * -log10(p))
  )
 
# generate per-taxa matrix of signed-log(pvalues)
 score_matrix <- results %>%
  dplyr::select(OTU, Time_Group_2, signed_log10_p) %>%
  pivot_wider(
    names_from = Time_Group_2,
    values_from = signed_log10_p
  ) %>%
  column_to_rownames("OTU") %>%
  as.matrix()
  
heatmap_matrix <- score_matrix[
  complete.cases(score_matrix),
  ,
  drop = FALSE
]
 
 

signed_max <- max(abs(heatmap_matrix), na.rm = TRUE)
 
col_fun <- circlize::colorRamp2(
  c(-4, 0, 4),
  c("#2166AC", "white", "#B2182B")
)
 


 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation"
)
 
 
 
col_dend <- heatmap_matrix |>
  t() |>
  dist(method = "euclidean") |>
  hclust(method = "complete") |>
  as.dendrogram()

# Rotate branches  
col_dend <- dendextend::rotate(
  col_dend,
  order = timepoint_order
)

# Reorder matrix rows explicitly
#heatmap_matrix_ordered <- heatmap_matrix[row_names_in_heatmap_order, , drop = FALSE]

ht <- ComplexHeatmap::Heatmap(
  heatmap_matrix,
  name = "Signed\n-log10(p)",
  col = col_fun,

  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_dend_reorder = FALSE,

  na_col = "grey90",
  column_title = "Selected OTU in Control Groups",
  row_title = "OTU",
  row_names_side = "right",
  row_names_gp = grid::gpar(fontsize = 7)
)

ComplexHeatmap::draw(ht)

```

### Plot taxa 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
genus_rel_df_sig <- ASV_rel_df %>% 
  filter(Sample_Site %in% c("Control", "Surgical"))%>%
 filter(Time_Group %in% c("Before_Surgery", "Post_Surgery", "Inflammation"))%>%
  filter(OTU %in% significant_genera_1) %>%
  group_by(OTU, Time_Group, Sample_Site, Dog_ID) %>%
  summarise(
    avg_rel_abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )
 
genus_rel_df_sig

ggplot(
  genus_rel_df_sig,
  aes(
    x = Time_Group,
    y = avg_rel_abundance,
    fill = Sample_Site,
    group = interaction(Time_Group, Sample_Site)
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.65,
    outlier.alpha = 0.6
  ) +
  geom_jitter(
    aes(color = Sample_Site),
    position = position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.75
    ),
    size = 1.8,
    alpha = 0.7
  ) +
  facet_wrap(~ OTU, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Control" = "#7B2CBF",
      "Surgical" = "#0072B2"
    )
  ) +
  scale_color_manual(
    values = c(
      "Control" = "#7B2CBF",
      "Surgical" = "#0072B2"
    ),
    guide = "none"
  ) +
  labs(
    x = "Time group",
    y = "Relative abundance per dog",
    fill = "Sample site",
    title = "OTU abundance by time group and sample site"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )+ stat_summary(
    fun = median,
    geom = "point",
    aes(group = Sample_Site),
    position = position_dodge(width = 0.75),
    shape = 21,
    size = 2,
    stroke = 0.8,
    fill = "white",
    color = "black"
  )
  
```

 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
  
 
 
 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE} 

# Differential Abundance
# get mean relative abundance for each phylum
phy_phyla_df <- psmelt(ps_rel_2) %>%
  filter(Sample_Location == "Stifle") %>%
  filter(!is.na(Phylum)) %>%
  group_by(Phylum) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rel_abundance))

top_phyla <- phy_phyla_df$Phylum[1:min(10, nrow(phy_phyla_df))]

# Recode non-top phyla as Other
phy_phyla_df <- ps_rel_2 %>%
  subset_samples(Surgery_Type == "Stifle") %>%
  psmelt() %>%
  mutate(
    Phylum = if_else(Phylum %in% top_phyla, as.character(Phylum), "Other"),
    Sample_Site = factor(Sample_Site, levels = c("Control", "Surgical")),
    Day_Order = factor(Day_Order, levels = sort(unique(Day_Order)))
  ) %>%
  group_by(Sample_Site, Day_Order, Phylum) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")%>%
  arrange(desc(mean_rel_abundance))
 

# Colors for top 10 + Other
all_phyla <- c(top_phyla, "Other")
n_phyla <- length(all_phyla)
base_cols <- brewer.pal(min(n_phyla, 12), "Paired")
phylum_cols_vec <- colorRampPalette(base_cols)(n_phyla)
phylum_colors <- setNames(phylum_cols_vec, all_phyla)

plot_site <- function(site_name) {
  ggplot(
    phy_phyla_df %>% filter(Sample_Site == site_name),
    aes(x = Day_Order, y = mean_rel_abundance, fill = factor(Phylum, levels = all_phyla))
  ) +
    geom_col(position = "stack") +
    scale_fill_manual(values = phylum_colors, drop = FALSE) +
    labs(
      title = paste("Sample-Site:", site_name),
      x = "Timepoint",
      y = "Average Relative abundance",
      fill = "Phylum"
    ) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
 

p_control  <- plot_site("Control")
p_surgical <- plot_site("Surgical")
  
print(p_control)
print(p_surgical)

 

# get mean relative abundance for each family
family_df <- psmelt(ps_rel_2) %>%
  filter(Sample_Location == "Stifle") %>%
  filter(!is.na(Family)) %>%
  group_by(Phylum, Family) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rel_abundance)) 

top_10_family <- family_df$Family[1:min(10, nrow(family_df))]

# Recode non-top genus as Other
family_df <- ps_rel_2 %>%
  subset_samples(Surgery_Type == "Stifle") %>%
  psmelt() %>%
  mutate(
    Family = if_else(Family %in% top_10_family, as.character(Family), "Other"),
    Sample_Site = factor(Sample_Site, levels = c("Control", "Surgical")),
    Day_Order = factor(Day_Order, levels = sort(unique(Day_Order)))
  ) %>%
  group_by(Dog_ID, Sample_Site, Day_Order, Family) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")%>%
  arrange(desc(mean_rel_abundance))
 

# Colors for top 10 + Other
all_family <- c(top_10_family, "Other")
n_family <- length(all_family)
base_cols <- brewer.pal(min(n_family, 12), "Paired")
cols_vec <- colorRampPalette(base_cols)(n_family)
colors <- setNames(cols_vec, all_family)

plot_site <- function(site_name) {
  ggplot(
    family_df %>% filter(Sample_Site == site_name),
    aes(x = Day_Order, y = mean_rel_abundance, fill = factor(Family, levels = all_family))
  ) +
    geom_col(position = "stack") +
    scale_fill_manual(values = colors, drop = FALSE) +
    labs(
      title = paste("Sample-Site:", site_name),
      x = "Timepoint",
      y = "Average Relative abundance",
      fill = "Family"
    ) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
 

p_control  <- plot_site("Control")
p_surgical <- plot_site("Surgical")
 
p_control
p_surgical


# get mean relative abundance for each genus
phy_genus_df <- psmelt(ps_rel_2) %>%
  filter(Sample_Location == "Stifle") %>%
  filter(!is.na(Genus)) %>%
  group_by(Phylum, Genus) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rel_abundance)) 

top_genus <- phy_genus_df$Genus[1:min(10, nrow(phy_genus_df))]

# Recode non-top genus as Other
phy_genus_df <- ps_rel_2 %>%
  subset_samples(Surgery_Type == "Stifle") %>%
  psmelt() %>%
  mutate(
    Genus = if_else(Genus %in% top_genus, as.character(Genus), "Other"),
    Sample_Site = factor(Sample_Site, levels = c("Control", "Surgical")),
    Day_Order = factor(Day_Order, levels = sort(unique(Day_Order)))
  ) %>%
  group_by(Sample_Site, Day_Order, Genus) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")%>%
  arrange(desc(mean_rel_abundance))
 

# Colors for top 10 + Other
all_genus <- c(top_genus, "Other")
n_genus <- length(all_genus)
base_cols <- brewer.pal(min(n_genus, 12), "Paired")
genus_cols_vec <- colorRampPalette(base_cols)(n_genus)
genus_colors <- setNames(genus_cols_vec, all_genus)

plot_site <- function(site_name) {
  ggplot(
    phy_genus_df %>% filter(Sample_Site == site_name),
    aes(x = Day_Order, y = mean_rel_abundance, fill = factor(Genus, levels = all_genus))
  ) +
    geom_col(position = "stack") +
    scale_fill_manual(values = genus_colors, drop = FALSE) +
    labs(
      title = paste("Sample-Site:", site_name),
      x = "Timepoint",
      y = "Average Relative abundance",
      fill = "Genus"
    ) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
 

p_control  <- plot_site("Control")
p_surgical <- plot_site("Surgical")
 
p_control
p_surgical



species_df <- psmelt(ps_rel_2) %>%
  filter(Sample_Location == "Stifle") %>%
  filter(!is.na(Species)) %>%
  group_by(Species) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rel_abundance))

top_species <- species_df$Species[1:min(10, nrow(species_df))]

# Recode non-top phyla as Other
species_df <- ps_rel_2 %>%
  subset_samples(Surgery_Type == "Stifle") %>%
  psmelt() %>%
  mutate(
    Species = if_else(Species %in% top_phyla, as.character(Species), "Other"),
    Sample_Site = factor(Sample_Site, levels = c("Control", "Surgical")),
    Day_Order = factor(Day_Order, levels = sort(unique(Day_Order)))
  ) %>%
  group_by(Sample_Site, Day_Order, Species) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")%>%
  arrange(desc(mean_rel_abundance))
 

# Colors for top 10 + Other
all_phyla <- c(top_phyla, "Other")
n_phyla <- length(all_phyla)
base_cols <- brewer.pal(min(n_phyla, 12), "Paired")
phylum_cols_vec <- colorRampPalette(base_cols)(n_phyla)
phylum_colors <- setNames(phylum_cols_vec, all_phyla)

plot_site <- function(site_name) {
  ggplot(
    species_df %>% filter(Sample_Site == site_name),
    aes(x = Day_Order, y = mean_rel_abundance, fill = factor(Species, levels = all_phyla))
  ) +
    geom_col(position = "stack") +
    scale_fill_manual(values = phylum_colors, drop = FALSE) +
    labs(
      title = paste("Sample-Site:", site_name),
      x = "Timepoint",
      y = "Average Relative abundance",
      fill = "Species"
    ) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
 

p_control  <- plot_site("Control")
p_surgical <- plot_site("Surgical")
  
print(p_control)
print(p_surgical)

```

 
```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
sd <- as(sample_data(ps_rel_2), "data.frame")

sd$Time_Group <- ifelse(
  sd$Day_Order %in% c("Admission", "Day 0a - Pre Surgery"),
  "Before",
  "After"
)

sd$Time_Group <- factor(sd$Time_Group, levels = c("Before", "After"))
sample_data(ps_rel_2) <- sample_data(sd)

# get mean relative abundance for each phylum
phy_phyla_df <- psmelt(ps_rel_2) %>%
  filter(Sample_Location == "Stifle") %>%
  filter(!is.na(Phylum)) %>%
  group_by(Phylum) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rel_abundance))
 

surg_df_phyla <- psmelt(ps_rel_2)%>%
  filter(
    Sample_Location == "Stifle",
    #Sample_Site == "Surgical",
    Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery", "Day 1"),
    !is.na(Phylum)
  ) %>%
  group_by(Dog_ID, Time_Group, Sample_Site, Phylum) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

results_phyla <- surg_df_phyla %>%
  tidyr::pivot_wider(
    id_cols = c(Dog_ID, Phylum, Sample_Site),
    names_from = Time_Group,
    values_from = mean_rel_abundance
  ) %>%
  group_by(Sample_Site) %>%
  filter((Before > 0) | (After > 0))  %>%
  summarise(
    n_pairs = n(),
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    estimate = test[[1]]$estimate,
    conf_low = test[[1]]$conf.int[1],
    conf_high = test[[1]]$conf.int[2],
    .groups = "drop"
  ) %>%
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)

print(results_phyla)

# Post-Hoc 
results_phyla <- surg_df_phyla %>%
  filter(mean_rel_abundance > 0.0001) %>% 
  tidyr::pivot_wider(
    id_cols = c(Dog_ID, Phylum, Sample_Site),
    names_from = Time_Group,
    values_from = mean_rel_abundance
  ) %>%
  group_by(Sample_Site) %>%
  filter(!is.na(Before), !is.na(After)) %>%
  summarise(
    n_pairs = n(),
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)

print(results_phyla)
 
 
## Genus 
 
surg_df_phyla <- psmelt(ps_rel_2)%>%
  filter(
    Sample_Location == "Stifle",
    #Sample_Site == "Surgical",
    Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery", "Day 1"),
    !is.na(Genus)
  ) %>%
  group_by(Dog_ID, Time_Group, Sample_Site, Genus) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")


results_phyla <- surg_df_phyla %>%
  tidyr::pivot_wider(
    id_cols = c(Dog_ID, Genus, Sample_Site),
    names_from = Time_Group,
    values_from = mean_rel_abundance
  ) %>%
  group_by(Sample_Site) %>%
  filter((Before > 0.001) | (After > 0.001))  %>%
  summarise(
    n_pairs = n(),
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)

print(results_phyla)

results_phyla <- surg_df_phyla %>%
  tidyr::pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = mean_rel_abundance
  ) %>%
  group_by(Genus) %>%
  filter((Before > 0), (After > 0)) %>% 
  summarise(
    n_pairs = n(),
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)

print(results_phyla)


## Species 

surg_df_species <- psmelt(ps_rel_2) %>% 
  filter(
 #  Sample_Location == "Stifle", 
    Sample_Site == "Surgical", 
    Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery", "Day 1"), 
    !is.na(Species) 
  ) %>%
  group_by(Dog_ID, Time_Group, Genus, Species) %>% 
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

results_phyla <- surg_df_species %>% 
  pivot_wider(
    id_cols = c(Dog_ID, Genus, Species), 
    names_from = Time_Group, 
    values_from = mean_rel_abundance
  ) %>% 
  filter(!is.na(Before), !is.na(After)) %>%
  filter((Before > 0.001) | (After > 0.001))  %>%
  # 4. Group strictly by the taxonomic level you are testing
  group_by(Genus, Species) %>% 
  
  # 5. Run the statistical analysis safely
  summarise(
    n_pairs = n(),
    # Explicitly calculate the test statistics
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>% 
  
  # 6. Apply Benjamini-Hochberg false discovery rate correction
  mutate(p_adj = p.adjust(p, method = "BH")) %>% 
  arrange(p)

print(results_phyla)

```

 
```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
library(lefser)
library(SummarizedExperiment) 
ps_rel_2_surg <- ps_rel_2 %>%
  subset_samples(
    Sample_Site == "Surgical")%>% 
  subset_samples(Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery")) %>% 
  subset_taxa(!is.na(Species)) 

# Total ASVs across surgical samples from pre- and post-surgery timepoints 
phyloseq::ntaxa(ps_rel_2_surg)
 


sd <- as(sample_data(ps_rel_2_surg), "data.frame")

sd$Time_Group <- ifelse(
  sd$Day_Order %in% c("Admission", "Day 0a - Pre Surgery"),
  "Before",
  "After"
)
sd$Dog_ID <- factor(sd$Dog_ID)
sd$Time_Group <- factor(sd$Time_Group, levels = c("Before", "After"))
sample_data(ps_rel_2_surg) <- sample_data(sd)



keep <- phyloseq::filter_taxa(
  ps_rel_2_surg,
  function(x) {
    sd <- phyloseq::sample_data(ps_rel_2_surg_filt)
    adm  <- sd$Day_Order == "Admission"
    pre  <- sd$Day_Order == "Day 0a - Pre Surgery"
    post <- sd$Day_Order == "Day 0c - Post Surgery"

    # Require presence in all three timepoints
    present_all <- any(x[adm] > 0) && any(x[pre] > 0) && any(x[post] > 0)

    prevalence_all <- mean(x > 0) >= 0.05

    present_all && prevalence_all
  },
  prune = FALSE
)

ps_rel_2_surg_filt <- phyloseq::prune_taxa(keep, ps_rel_2_surg)
phyloseq::ntaxa(ps_rel_2_surg_filt)


ps_lefse <- ps_rel_2_surg_filt  
 

ps_lefse <- tax_glom(ps_lefse, taxrank = "Species", NArm = TRUE)
 

ps_surgical <- subset_samples(ps_lefse, Sample_Site == "Surgical")
ps_surgical <- prune_taxa(taxa_sums(ps_surgical) > 0, ps_surgical)
 
keep_dogs <- sample_data(ps_surgical) |>
  data.frame() |>
  dplyr::group_by(Dog_ID) |>
  dplyr::filter(all(c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery") %in% Day_Order)) |>
  dplyr::distinct(Dog_ID) |>
  dplyr::pull(Dog_ID)

ps_surgical <- prune_samples(sample_data(ps_surgical)$Dog_ID %in% keep_dogs, ps_surgical)

## Split data tables
counts <- as(otu_table(ps_surgical), "matrix")
if (!taxa_are_rows(otu_table(ps_surgical))) {
  counts <- t(counts)
}
coldata <- as(sample_data(ps_surgical), "data.frame")


## Create a SummarizedExperiment object
se <- SummarizedExperiment(assays = list(counts = counts), colData = coldata)

 
table(colData(se)$Time_Group, colData(se)$Dog_ID)

set.seed(1234)
res <- lefser(
  se,
  kruskal.threshold = .1, wilcox.threshold = .1,
  classCol = "Time_Group" 
) 

lefserPlot(res)
```
 
 
 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE} 

# Differential Abundance
# get mean relative abundance for each phylum
  

species_df_summary <- species_df %>% 
  mutate(genus_species = paste(Genus, Species, sep = " ")) %>%
  group_by(Day_Order, Genus, Species, genus_species) %>% 
  summarise(
    mean_rel_abundance = mean(mean_rel_abundance, na.rm = TRUE),
    .groups = "drop"
  )

species_list <- unique(species_df_summary$genus_species)

# Colors for top 10 + Other
all_species <- c(species_list)
n_taxa <- length(all_species)
base_cols <- brewer.pal(min(n_taxa, 12), "Paired")
taxa_cols_vec <- colorRampPalette(base_cols)(n_taxa)
taxa_colors <- setNames(taxa_cols_vec, all_species)

species_df_summary %>%
  mutate(genus_species = factor(genus_species, levels = all_species)) %>%
  ggplot(aes(x = Day_Order, y = mean_rel_abundance, fill = genus_species)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = taxa_colors, drop = FALSE) +
  labs(
    title = paste("Taxa in Pre-Survery vs. Post-Surgery Samples"),
    x = "Timepoint",
    y = "Average Relative abundance",
    fill = "Species"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
 
 
 

```
 

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
 ## Species 

ctrl_df_species <- psmelt(ps_rel_2) %>% 
  filter(
 #  Sample_Location == "Stifle", 
    Sample_Site == "Control", 
    Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery", "Day 1"), 
    !is.na(Species) 
  ) %>%
  group_by(Dog_ID, Time_Group, Genus, Species) %>% 
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

 

results_phyla <- ctrl_df_species %>% 
  pivot_wider(
    id_cols = c(Dog_ID, Genus, Species), 
    names_from = Time_Group, 
    values_from = mean_rel_abundance
  ) %>% 
  filter(!is.na(Before), !is.na(After)) %>%
  filter((Before > 0.001) | (After > 0.001))  %>%
  # 4. Group strictly by the taxonomic level you are testing
  group_by(Genus, Species) %>% 
  
  # 5. Run the statistical analysis safely
  summarise(
    n_pairs = n(),
    # Explicitly calculate the test statistics
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>% 
  
  # 6. Apply Benjamini-Hochberg false discovery rate correction
  mutate(p_adj = p.adjust(p, method = "BH")) %>% 
  arrange(p)

print(results_phyla)
```
 
  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE} 
results_species_annotated_notsig <- results_species_annotated %>%
  filter(results_species_annotated$p > 0.05)

results_species_annotated_notsig
shared_seqs <- intersect(results_species_annotated_sig$OTU, asvs_negative_samples$ASV)

subset_not_in_negatives <- results_species_annotated_sig[
  !results_species_annotated_sig$OTU %in% asvs_negative_samples$ASV,
]
subset_not_in_negatives


 

species_df_formatted_sub <- species_df_formatted %>%
  filter(OTU == "ASV_114")
species_long <- species_df_formatted_sub %>%
  pivot_longer(
    cols = c(Post_Surgery, Maturation II),
    names_to = "Timepoint",
    values_to = "RelAbundance"
  )
ggplot(species_long,
       aes(x = RelAbundance + 1e-6, fill = Timepoint)) +
  ggtitle("Rel Abundance of Signficant ASV_114 (p.adj <0.1)") +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  coord_cartesian(ylim = c(0, 2.5)) +
  theme_bw() +
  labs(x = "Relative abundance (log10 scale)",
       y = "Density")


species_df_formatted_sub <- species_df_formatted %>%
  filter(OTU == "ASV_1488")
species_long <- species_df_formatted_sub %>%
  pivot_longer(
    cols = c(Post_Surgery, Maturation),
    names_to = "Timepoint",
    values_to = "RelAbundance"
  )
ggplot(species_long,
       aes(x = RelAbundance + 1e-6, fill = Timepoint)) +
  ggtitle("Rel Abundance of Non-signficant ASV_1488 (p.adj > 0.1)") +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  coord_cartesian(ylim = c(0, 2.5)) +
  theme_bw() +
  labs(x = "Relative abundance (log10 scale)",
       y = "Density")



 

phylum_colors <- c(
  "Bacillota" = "orange",
  "Pseudomonadota" = "#e7298a",
  "Actinomycetota" = "#1f78b4",
  "Bacteroidota" = "#1b9e77",
  "Cyanobacteriota" = "#a6761d",
  "Thermodesulfobacteriota" = "red",
  "Fusobacteriota" = "yellow",
  "Campylobacterota" = "#66a61e",
  "Acidobacteriota" = "green",
  "Chloroflexota" = "pink",
  "Other" = "gray"
)

phylum_counts <- results_species_annotated_sig %>%
  mutate(Phylum = ifelse(is.na(Phylum) | Phylum == "", "Other", Phylum),
         Phylum = ifelse(Phylum %in% names(phylum_colors), Phylum, "Other")) %>%
  count(Phylum, name = "Count") %>%
  mutate(Phylum = factor(Phylum, levels = Phylum[order(Count, decreasing = TRUE)]))

ggplot(phylum_counts, aes(x = Phylum, y = Count, fill = Phylum)) +
  ggtitle("Phyla of the 45 ASVs found to have significantly higher relative abundances (p_adj =<= 0.1)") +
  geom_col() +
  scale_fill_manual(values = phylum_colors, drop = FALSE) +
  labs(x = "Phylum", y = "Count") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```
 



 