---
title: "Canine Skin Microbiome: 16S Differential Abundance Analysis "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

**Purpose:**
Identify bacterial taxa at the genus and ASV levels that differ between surgical and control groups from baseline (Admission and Pre-Surgery Day 0) to the immediate post-surgery timepoint.

**Overview:**
Taxa present in at least 35% of surgical samples across the pre- and post-surgery timepoints were retained for analysis. For each of the dogs, paired relative abundances before and immediately after surgery were compared using Wilcoxon signed-rank tests.

For taxa with q-values ≤0.10, effect sizes were calculated as the paired Wilcoxon location-shift estimate (i.e., the estimated median difference in relative abundance between timepoints). These results were visualized in a heatmap using the signed significance score: sign(estimate) * -log10(p.value). The same set of significant taxa was subsequently tested using paired Wilcoxon signed-rank tests to compare baseline with later healing-stage timepoints.


**Results:**

* Total taxa present in at least 25% of surgical samples across the before- and post-surgery groups:

  - Genus: 314 taxa tested; 32 had q-values \(\leq 0.1\)
  - ASVs: 722 taxa tested; 2 had q-values \(\leq 0.1\)

Immediately after surgery, the most differentially abundant genera generally showed decreased relative abundance compared with baseline. In contrast, later healing stages showed fewer differences relative to baseline. Control samples also tended to show lower relative abundances at the post-surgery timepoint relative to baseline; however, these changes showed limited statistical significance and were less pronounced than those observed at surgical sites.



# Load files 
## R packages
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(knitr)
library(FSA)
library(tidyverse)
library(vegan)
library(phyloseq)
library(dendextend)
library(ggplot2)
library(ggtree)
library(scales)  
library(RColorBrewer)
library(segmented)
library(stringr)
library(pairwiseAdonis)
library(ggpubr)
library(changepoint)
library(dplyr)
library(lme4)
library(emmeans)
library(nortest)
library(mcp)
library(purrr)
library(kableExtra)
library(forcats)
library(microbiome)
library(lmerTest)       
library(broom.mixed)
library(ComplexHeatmap)
library(circlize)
```
 
## Import paired-dataset
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Paired-only ASV feature table
ps <- readRDS("../01_input_files/phyloseq/ps_2.rds")
 
sd <- as(sample_data(ps), "data.frame")

# define healing stages
sd <- sd %>%
  mutate(
    Time_Group = case_when(
      Day_Order %in% c("Admission", "Day 0a - Pre Surgery") ~ "Before_Surgery",
      Day_Order %in% c("Day 0c - Post Surgery")           ~ "Post_Surgery",
      Day_Order %in% c("Day 1", "Day 3")                  ~ "Inflammation",
      Day_Order %in% c("Week 1", "Week 2")                ~ "Proliferation",
      Day_Order %in% c("Week 3", "Week 4", "Week 5", "Week 6")  ~ "Maturation", 
      TRUE                                                 ~ NA_character_
    ),
    Time_Category = case_when(
      Day_Order %in% c("Admission", "Day 0a - Pre Surgery") ~ "Before",
      TRUE                                                  ~ "After"
    )
  )


sd$Time_Group <- factor(sd$Time_Group, levels = c("Before_Surgery", "Post_Surgery", "Inflammation", "Proliferation", "Maturation"))
  
sample_data(ps) <- sample_data(sd)

range(sample_sums(ps)) 
mean(sample_sums(ps))
median(sample_sums(ps))
```

## TSS Normalization (relative abundances) of ASV + Genus
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Normalization of ASV counts 
ps_ASV_rel <- transform_sample_counts(ps, function(x) x / sum(x))
 
# Import saved ASV ID-to-sequence lookup
asv_sequence_lookup <- read_csv(
  "../01_input_files/ASV_tables/asv_sequence_lookup.csv",
  show_col_types = FALSE
) %>%
  transmute(
    OTU = as.character(OTU),
    ASV_sequence = as.character(ASV_sequence)
  )

# generate df
ASV_rel_df <- psmelt(ps_ASV_rel) %>%
arrange(Dog_ID, OTU, Day_Order)

ASV_rel_df_surg <- ASV_rel_df%>%
   filter(Sample_Site == "Surgical") 
    
  

ASV_rel_df_ctrl <- ASV_rel_df%>%
   filter(Sample_Site == "Control")

# Unique ASVs
ntaxa(ps)
tax <- tax_table(ps)

# % ASVs removed due to unknown Genus 
no_genus <- is.na(tax[, "Genus"]) | tax[, "Genus"] == ""
n_no_genus <- sum(no_genus, na.rm = TRUE)
n_total_asv <- ntaxa(ps)
pct_no_genus <- 100 * n_no_genus / n_total_asv

n_no_genus
pct_no_genus

cat("Excluded", n_no_genus, "ASVs without genus assignment (", round(pct_no_genus, 1), "% of total ASVs).\n")


# Aggregate ASV counts for Genus-Level Analysis + remove ASVs with unknown Genus
ps_genus <- tax_glom(ps, taxrank = "Genus", NArm = TRUE)
#Unique genus
ntaxa(ps_genus)

# turn phyloseq object into a dataframe 
genus_df <- psmelt(ps_genus) %>%
arrange(Dog_ID, Genus, Day_Order)

# Check for genus-level (each sample should have one row for each Genus, if multiple appear then it is still at ASV-level)
check <- genus_df %>%
  count(Sample, Genus) %>%
  filter(n > 1)
 check
 
ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))
 
# ntaxa should be smaller since we aggregated at Genus level 
ntaxa(ps_genus_rel)
 

# turn phyloseq object into a dataframe 
genus_rel_df <- psmelt(ps_genus_rel) %>%
arrange(Dog_ID, Genus, Day_Order)

# Check for genus-level (each sample should have one row for each Genus, if multiple appear then it is still at ASV-level)
check <- genus_rel_df %>%
  count(Sample, Genus) %>%
  filter(n > 1)
 check
 
genus_rel_df_surg <- genus_rel_df%>%
   filter(Sample_Site == "Surgical")

 
genus_rel_df_ctrl <- genus_rel_df%>%
   filter(Sample_Site == "Control")

genus_df_surg <- genus_df%>%
   filter(Sample_Site == "Surgical")
```


# Genus-level

## Surgical Samples  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
run_paired_wilcoxon_by_genus <- function(
    data,
    time_group_1,
    time_group_2
) {
  
  required_columns <- c("Genus", time_group_1, time_group_2)
  
  missing_columns <- setdiff(required_columns, names(data))
  
  if (length(missing_columns) > 0) {
    stop(
      "The following columns are missing: ",
      paste(missing_columns, collapse = ", ")
    )
  }
  
  results <- data %>%
  group_by(Genus) %>%
  group_modify(~ {

    dat <- .x %>%
      filter(
        !is.na(.data[[time_group_1]]),
        !is.na(.data[[time_group_2]])
      )

    group_1_values <- dat[[time_group_1]]
    group_2_values <- dat[[time_group_2]]

    differences <- group_1_values - group_2_values

    n_pairs <- length(differences)
    n_zero_diff <- sum(differences == 0)
    n_nonzero_diff <- sum(differences != 0)

    test <- wilcox.test(
      group_1_values,
      group_2_values,
      alternative = "two.sided",
      paired = TRUE,
      # By default (if exact is not specified), an exact p-value is computed if the samples contain less than 50 finite values and there are no ties. Otherwise, a normal approximation is used.
      exact = NULL,
      # a nonparametric confidence interval and an estimator for the pseudomedian (one-sample case)
      conf.int = TRUE
    )

    tibble(
      n_pairs = n_pairs,
      n_zero_diff = n_zero_diff,
      n_nonzero_diff = n_nonzero_diff,

      median_group_1 = median(group_1_values, na.rm = TRUE),
      median_group_2 = median(group_2_values, na.rm = TRUE),
      median_difference = median(differences, na.rm = TRUE),

      stat = unname(test$statistic),
      p = test$p.value,
      # Hodges–Lehmann pseudomedian 
      estimate = unname(test$estimate),
      conf_low = test$conf.int[1],
      conf_high = test$conf.int[2]
    )
  }) %>%
  ungroup() %>%
  mutate(
    p_adj = p.adjust(p, method = "BH")
  ) %>%
  arrange(p_adj, p, Genus)
  
  return(results)
}

```

Distribution of taxa at different prevalence thresholds
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_2 <- "Before_Surgery"
Time_Group_1 <- "Maturation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups) %>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()

 
prevalence_thresholds <- function(data, threshold) {
  total_samples <- n_distinct(data$Sample)

  data %>%
    group_by(Genus) %>%
    summarise(
      prevalence = n_distinct(Sample[Abundance > 0]) / total_samples,
      .groups = "drop"
    ) %>%
    filter(prevalence >= threshold) %>%
    summarise(n_genera = n()) %>%
    pull(n_genera)
}

total_samples <- n_distinct(genus_rel_df_surg_2$Sample)

threshold_results <- tibble(
  prevalence_threshold = seq(0.05, 0.50, by = 0.05)
) %>%
  mutate(
    min_samples_required = ceiling(total_samples * prevalence_threshold),
    n_genera = map_int(
      prevalence_threshold,
      ~ prevalence_thresholds(genus_rel_df_surg_2, .x)
    )
  )
 
 threshold_results
 
 ggplot(
  threshold_results,
  aes(x = prevalence_threshold, y = n_genera)
) +
   geom_line(linewidth = 0.9, color = "#2C7FB8") +
  geom_point(size = 2.5, color = "#2C7FB8") +
  scale_x_continuous(
    breaks = seq(0.05, 0.50, by = 0.05),
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    x = "Minimum prevalence threshold",
    y = "Number of genera retained",
    title = "Genera retained across prevalence thresholds"
  ) +
  theme_classic(base_size = 13)+ 
  geom_vline(
    xintercept = 0.1,
    linetype = "dashed",
    color = "firebrick"
  )
```

### Before_Surgery (Admission + Pre-Surgery (Day 0)) vs Post-Surgery (Day 0)
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Maturation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups) %>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()

# Identify genera present in at least 35% of samples
taxa_present_10pct <- genus_rel_df_surg_2 %>%
  group_by(Genus) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.35) %>%
  arrange(desc(prevalence)) %>%
  dplyr::select(Genus) 

top10_genus_by_prevalence <- taxa_present_10pct%>%
  slice_head(n = 10)

paste(
  "Top 10 genera by prevlance:",
  top10_genus_by_prevalence
)

# Number of retained genera
filtered_genus_count <- n_distinct(taxa_present_10pct$Genus)

paste(
  "Total genera present in at least 35% of samples:",
  filtered_genus_count
)
# Keep genera satisfying the prevalence criterion
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "Genus")


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_postsurg <- results_wt %>%
  left_join(tax_df, by = "Genus")

nrow(results_wt_annotated_postsurg)

results_wt_annotated_sig <- results_wt_annotated_postsurg %>%
  filter(results_wt_annotated_postsurg$p_adj <= 0.1)%>%
  arrange(p)

results_wt_annotated_sig

results_wt_annotated_postsurg <- results_wt_annotated_postsurg %>%
  filter(results_wt_annotated_postsurg$p <= 0.05)%>%
  arrange(p)

results_wt_annotated_postsurg


significant_genera_2 <- unique(results_wt_annotated_sig$Genus) 

significant_genera_1 <- unique(results_wt_annotated_postsurg$Genus) 

 
```
 
### Inflammation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Inflammation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups) %>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_inflam <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_inflam
  
```
 
### Proliferation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Proliferation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups) %>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
   dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_prolif <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_prolif
  
```

### Maturation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups) %>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
   dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_matur <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_matur
  
```

### Heatmap
```{r,   message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 

results_wt_annotated <- bind_rows(results_wt_annotated_postsurg, results_wt_annotated_inflam, results_wt_annotated_prolif, results_wt_annotated_matur)

 
 # get signed log(p-value) for heatmap
 results <- results_wt_annotated %>%
  mutate(
    signed_log10_p = -(sign(estimate) * -log10(p))
  )
 
# generate per-taxa matrix of signed-log(pvalues)
 score_matrix <- results %>%
  dplyr::select(Genus, Time_Group_2, signed_log10_p) %>%
  pivot_wider(
    names_from = Time_Group_2,
    values_from = signed_log10_p
  ) %>%
  column_to_rownames("Genus") %>%
  as.matrix()
  
heatmap_matrix <- score_matrix[
  complete.cases(score_matrix),
  ,
  drop = FALSE
]
 
 

signed_max <- max(abs(heatmap_matrix), na.rm = TRUE)

col_fun <- circlize::colorRamp2(
  c(-4, 0, 4),
  c("#2166AC", "white", "#B2182B")
)
 


 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation"
)
 
 
 
col_dend <- heatmap_matrix |>
  t() |>
  dist(method = "euclidean") |>
  hclust(method = "complete") |>
  as.dendrogram()

# Rotate branches  
col_dend <- dendextend::rotate(
 col_dend,
  order = timepoint_order
)

heatmap_matrix <- heatmap_matrix[, timepoint_order]

ht <- ComplexHeatmap::Heatmap(
  heatmap_matrix,
  name = "Signed\n-log10(p)",
  col = col_fun,

  cluster_rows = TRUE,
 # cluster_columns = col_dend,
  cluster_columns = FALSE,
  column_dend_reorder = FALSE,

  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",

  na_col = "grey90",
  column_title = "Differentially Significant Genera in Surgical Groups (Relative to Maturation)",
  row_title = "Genus",
  row_names_side = "right",
  row_names_gp = grid::gpar(fontsize = 7)
)

ComplexHeatmap::draw(ht)
 

row_idx <- ComplexHeatmap::row_order(ht)
row_names_in_heatmap_order <- rownames(heatmap_matrix)[row_idx]

row_names_in_heatmap_order
```
 
## Control Samples  
 
### Before_Surgery (Admission + Pre-Surgery (Day 0)) vs Post-Surgery (Day 0)
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups) %>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
  dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_postsurg <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_postsurg
  
```
 
### Inflammation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Inflammation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups) %>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
  dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_inflam <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_inflam
  
```
 
### Proliferation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Proliferation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups) %>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
   dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_prolif <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_prolif
  
```

### Maturation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Maturation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups) %>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
 
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
   dplyr::filter(Genus %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated_matur <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_matur
  
```

### Heatmap
```{r,  message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 

results_wt_annotated <- bind_rows(results_wt_annotated_postsurg, results_wt_annotated_inflam, results_wt_annotated_prolif, results_wt_annotated_matur)
 
 

 # get signed log(p-value) for heatmap
 results <- results_wt_annotated %>%
  mutate(
    signed_log10_p = -(sign(estimate) * -log10(p))
  )
 
# generate per-taxa matrix of signed-log(pvalues)
 score_matrix <- results %>%
  dplyr::select(Genus, Time_Group_2, signed_log10_p) %>%
  pivot_wider(
    names_from = Time_Group_2,
    values_from = signed_log10_p
  ) %>%
  column_to_rownames("Genus") %>%
  as.matrix()
  
heatmap_matrix <- score_matrix[
  complete.cases(score_matrix),
  ,
  drop = FALSE
]
 
 

signed_max <- max(abs(heatmap_matrix), na.rm = TRUE)
 
col_fun <- circlize::colorRamp2(
  c(-4, 0, 4),
  c("#2166AC", "white", "#B2182B")
)
 


 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation"
)
 
 
 
col_dend <- heatmap_matrix |>
  t() |>
  dist(method = "euclidean") |>
  hclust(method = "complete") |>
  as.dendrogram()

# Rotate branches  
col_dend <- dendextend::rotate(
  col_dend,
  order = timepoint_order
)

heatmap_matrix <- heatmap_matrix[, timepoint_order]

# Reorder matrix rows explicitly
heatmap_matrix_ordered <- heatmap_matrix[row_names_in_heatmap_order, , drop = FALSE]

heatmap_matrix <- heatmap_matrix[, timepoint_order]
ht <- ComplexHeatmap::Heatmap(
  heatmap_matrix_ordered,
  name = "Signed\n-log10(p)",
  col = col_fun,

  cluster_rows = FALSE,
 cluster_columns = FALSE,
  column_dend_reorder = FALSE,

  na_col = "grey90",
  column_title = "Selected Genera in Control Groups",
  row_title = "Genus",
  row_names_side = "right",
  row_names_gp = grid::gpar(fontsize = 7)
)

ComplexHeatmap::draw(ht)

```


### Plot taxa 
```{r, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
genus_rel_df_sig <- genus_rel_df %>% 
  filter(Sample_Site %in% c("Control", "Surgical"))%>%
 # filter(Time_Group %in% c("Before_Surgery", "Post_Surgery"))%>%
  filter(Genus %in% significant_genera_2) %>%
  group_by(Genus, Time_Group, Sample_Site, Dog_ID) %>%
  summarise(
    avg_rel_abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )
 

ggplot(
  genus_rel_df_sig,
  aes(
    x = Time_Group,
    y = avg_rel_abundance,
    fill = Sample_Site,
    group = interaction(Time_Group, Sample_Site)
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.65,
    outlier.alpha = 0.6
  ) +
  geom_jitter(
    aes(color = Sample_Site),
    position = position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.75
    ),
    size = 1.8,
    alpha = 0.7
  ) +
  facet_wrap(~ Genus, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Control" = "purple",
      "Surgical" = "blue"
    )
  ) +
  scale_color_manual(
    values = c(
      "Control" = "purple",
      "Surgical" = "blue"
    ),
    guide = "none"
  ) +
  labs(
    x = "Time group",
    y = "Relative abundance per dog",
    fill = "Sample site",
    title = "Genus abundance by time group and sample site"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )+ stat_summary(
    fun = median,
    geom = "point",
    aes(group = Sample_Site),
    position = position_dodge(width = 0.75),
    shape = 21,
    size = 2,
    stroke = 0.8,
    fill = "white",
    color = "black"
  )
  
```

 

## Log-transformed normalized counts
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# count total number of reads for each sample then get the average of all those samples  
avg_depth <- mean(sample_sums(ps_genus))
 
# model the taxa identified in Wilcoxen test as significant
genus_df_sub <- genus_df %>%
  filter(Genus %in% significant_genera_2)%>%
  filter(
    Time_Group %in% c("Before_Surgery", "Maturation")
  )
 # Check for genus-level (each sample should have one row for each Genus, if multiple appear then it is still at ASV-level)
check <- genus_df_sub %>%
  count(Sample, Genus) %>%
  filter(n > 1)
 check
 
 
abund_vector <- as.vector(genus_df_sub$Abundance)

# Plot Distribution  
histogram_distances <- ggplot(data = data.frame(abund_vector), aes(x = abund_vector)) +
  geom_histogram(binwidth = 100, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  #coord_cartesian(xlim = c(0, 1.52)) +
  labs(
    title = "Distribution of abundances",
    x = "abundance",
    y = "Count"
  ) +
  scale_y_log10()
print(histogram_distances)

# multiple relative abundance by the average depth then add 1 before log10()
ps_genus_tss_log <- transform_sample_counts(
  ps_genus,
  function(x) log10((x / sum(x)) * avg_depth + 1)
)

ps_genus_tss_log_df <- psmelt(ps_genus_tss_log)
 
# model the taxa identified in Wilcoxen test as significant
genus_clr_df_sub <- ps_genus_tss_log_df %>%
  filter(Genus %in% significant_genera_2)%>%
  filter(
    Time_Group %in% c("Before_Surgery", "Maturation")
  )
 

 
 
log_abund <- as.vector(genus_clr_df_sub$Abundance)
# Plot Distribution 
histogram_distances <- ggplot(data = data.frame(log_abund), aes(x = log_abund)) +
  geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  #coord_cartesian(xlim = c(0, 1.52)) +
  labs(
    title = "Distribution of Log normalized-abundances",
    x = "Log10-abundance",
    y = "Count"
  )

print(histogram_distances)
```

LMM  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
genus_clr_df_sub <- genus_clr_df_sub %>%
  mutate(
    Sample_Site = forcats::fct_relevel(Sample_Site, "Control"),
    Time_Group = factor(Time_Group)  
  )
 

genus_models <- genus_clr_df_sub %>%
  group_by(Genus) %>%
  group_nest() %>%
  mutate(
    model = map(
      data,
      ~ lmerTest::lmer(Abundance ~ Time_Group*Sample_Site + (1 | Dog_ID), data = .x)
    ),
    sum_tbl = map(model, ~ summary(.x))
  )
 

# Extract coefficient tables with p-values AND the fixed effect (Time_Group)
results <- genus_models %>%
  mutate(
    coefs = map(
      model,
      ~ as.data.frame(summary(.x)$coefficients) %>%
        tibble::rownames_to_column("Term")
    )
  ) %>%
  dplyr::select(Genus, coefs) %>%
  unnest(coefs) %>%
  filter(
    grepl(
      "^Time_Group.*:Sample_SiteSurgical$|^Sample_SiteSurgical:Time_Group",
      Term
    )
  ) %>%
  mutate(
    Time_Group = sub(
      "^Time_Group(.*):Sample_SiteSurgical$",
      "\\1",
      Term
    )
  ) %>%
  arrange(Genus, Time_Group)


# Rename columns  
results <- results %>%
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    df = df,
    statistic = `t value`,
    p.value = `Pr(>|t|)`
  ) 

results
 
 
 
```


 
 
# ASV-level
 
 Distribution of taxa at different prevalence thresholds
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
ASV_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)%>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
 
 
prevalence_thresholds <- function(data, threshold) {
  total_samples <- n_distinct(data$Sample)

  data %>%
    group_by(OTU) %>%
    summarise(
      prevalence = n_distinct(Sample[Abundance > 0]) / total_samples,
      .groups = "drop"
    ) %>%
    filter(prevalence >= threshold) %>%
    summarise(n_genera = n()) %>%
    pull(n_genera)
}

total_samples <- n_distinct(ASV_rel_df_surg_2$Sample)

threshold_results <- tibble(
  prevalence_threshold = seq(0.05, 0.50, by = 0.05)
) %>%
  mutate(
    min_samples_required = ceiling(total_samples * prevalence_threshold),
    n_genera = map_int(
      prevalence_threshold,
      ~ prevalence_thresholds(ASV_rel_df_surg_2, .x)
    )
  )
 
 threshold_results
 
 ggplot(
  threshold_results,
  aes(x = prevalence_threshold, y = n_genera)
) +
   geom_line(linewidth = 0.9, color = "#2C7FB8") +
  geom_point(size = 2.5, color = "#2C7FB8") +
  scale_x_continuous(
    breaks = seq(0.05, 0.50, by = 0.05),
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    x = "Minimum prevalence threshold",
    y = "Number of ASVs retained",
    title = "ASVs retained across prevalence thresholds"
  ) +
  theme_classic(base_size = 13)+ 
  geom_vline(
    xintercept = 0.1,
    linetype = "dashed",
    color = "firebrick"
  )
```

## Surgical Samples  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
run_paired_wilcoxon_by_ASV <- function(
    data,
    time_group_1,
    time_group_2,
    exact = TRUE
) {
  
  required_columns <- c("OTU", time_group_1, time_group_2)
  
  missing_columns <- setdiff(required_columns, names(data))
  
  if (length(missing_columns) > 0) {
    stop(
      "The following columns are missing: ",
      paste(missing_columns, collapse = ", ")
    )
  }
  
  results <- data %>%
  group_by(OTU) %>%
  group_modify(~ {

    dat <- .x %>%
      filter(
        !is.na(.data[[time_group_1]]),
        !is.na(.data[[time_group_2]])
      )

    group_1_values <- dat[[time_group_1]]
    group_2_values <- dat[[time_group_2]]

    differences <- group_1_values - group_2_values

    n_pairs <- length(differences)
    n_zero_diff <- sum(differences == 0)
    n_nonzero_diff <- sum(differences != 0)

    test <- wilcox.test(
      group_1_values,
      group_2_values,
      paired = TRUE,
      exact = exact,
      conf.int = TRUE
    )

    tibble(
      n_pairs = n_pairs,
      n_zero_diff = n_zero_diff,
      n_nonzero_diff = n_nonzero_diff,

      median_group_1 = median(group_1_values, na.rm = TRUE),
      median_group_2 = median(group_2_values, na.rm = TRUE),
      median_difference = median(differences, na.rm = TRUE),

      stat = unname(test$statistic),
      p = test$p.value,

      wilcox_estimate = unname(test$estimate),
      conf_low = test$conf.int[1],
      conf_high = test$conf.int[2]
    )
  }) %>%
  ungroup() %>%
  mutate(
    p_adj = p.adjust(p, method = "BH")
  ) %>%
  left_join(
    asv_sequence_lookup %>%
      dplyr::select(OTU, ASV_sequence),
    by = "OTU"
  ) %>%
  arrange(p_adj, p, OTU)
  
  
  return(results)
}

```

### Before_Surgery (Admission + Pre-Surgery) vs Post-Surgery 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
ASV_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)%>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
# Identify ASVs present in at least 10% of samples
taxa_present_10pct <- ASV_rel_df_surg_2 %>%
  group_by(OTU) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.35) %>%
  arrange(desc(prevalence)) %>%
  dplyr::select(OTU)

top10_ASVs_by_prevalence <- taxa_present_10pct%>%
  slice_head(n = 10)

paste(
  "Top 10 ASVs by prevlance:",
  top10_ASVs_by_prevalence
)


# Number of retained genera
filtered_ASV_count <- n_distinct(taxa_present_10pct$OTU)

cat(
  "Total ASV present in at least 35% of samples:",
  filtered_ASV_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
ASV_rel_df_surg_2_filt <- ASV_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "OTU")
 

# Summarise abundance per dog, genus, and time group
ASV_rel_df_median <- ASV_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_rel_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
ASV_rel_df_median_formatted <- ASV_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_rel_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = ASV_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )

 

tax_df <- tax_table(ps) %>%
  as("matrix") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>%
  tibble::as_tibble()

results_wt_annotated_postsurg <- results_wt %>%
  mutate(OTU = as.character(OTU)) %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_postsurg

results_wt_annotated_sig <- results_wt_annotated_postsurg %>%
  filter(results_wt_annotated_postsurg$p_adj <= 0.1)%>%
  arrange(p)

results_wt_annotated_sig

results_wt_annotated_postsurg <- results_wt_annotated_postsurg %>%
  filter(results_wt_annotated_postsurg$p <= 0.05)%>%
  arrange(p)

results_wt_annotated_postsurg


significant_genera_2 <- unique(results_wt_annotated_sig$OTU) 

significant_genera_1 <- unique(results_wt_annotated_postsurg$OTU) 

```
 
### Inflammation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Inflammation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)%>%
   group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>% 
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_inflam <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_inflam
  
```
 
### Proliferation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Proliferation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)%>%
   group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_prolif <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_prolif
  
```
 
### Maturation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Maturation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)%>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_matur <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_matur
  
```
 
### Heatmap
```{r,   message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 

results_wt_annotated <- bind_rows(results_wt_annotated_postsurg, results_wt_annotated_inflam, results_wt_annotated_prolif, results_wt_annotated_matur)

 # get signed log(p-value) for heatmap
 results <- results_wt_annotated %>%
  mutate(
    signed_log10_p = -(sign(wilcox_estimate) * -log10(p))
  )
 
# generate per-taxa matrix of signed-log(pvalues)
 score_matrix <- results %>%
  dplyr::select(OTU, Time_Group_2, signed_log10_p) %>%
  pivot_wider(
    names_from = Time_Group_2,
    values_from = signed_log10_p
  ) %>%
  column_to_rownames("OTU") %>%
  as.matrix()
  
heatmap_matrix <- score_matrix[
  complete.cases(score_matrix),
  ,
  drop = FALSE
]
 
 

signed_max <- max(abs(heatmap_matrix), na.rm = TRUE)
 
col_fun <- circlize::colorRamp2(
  c(-4, 0, 4),
  c("#2166AC", "white", "#B2182B")
)
 


 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation"
)
 
 
 
col_dend <- heatmap_matrix |>
  t() |>
  dist(method = "euclidean") |>
  hclust(method = "complete") |>
  as.dendrogram()

# Rotate branches  
col_dend <- dendextend::rotate(
  col_dend,
 order = timepoint_order
)

 
ht <- ComplexHeatmap::Heatmap(
  heatmap_matrix,
  name = "Signed\n-log10(p)",
  col = col_fun,

  cluster_rows = TRUE,
  cluster_columns = col_dend,
  column_dend_reorder = FALSE,

  na_col = "grey90",
  column_title = "Selected OTU in Surgical Groups",
  row_title = "OTU",
  row_names_side = "right",
  row_names_gp = grid::gpar(fontsize = 7)
)

ComplexHeatmap::draw(ht)


row_idx <- ComplexHeatmap::row_order(ht)
row_names_in_heatmap_order <- rownames(heatmap_matrix)[row_idx]

row_names_in_heatmap_order

```


## Log-transformed normalized counts
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# count total number of reads for each sample then get the average of all those samples  
avg_depth <- mean(sample_sums(ps))
ASV_df <- psmelt(ps)
# model the taxa identified in Wilcoxen test as significant
genus_df_sub <- ASV_df %>%
  filter(Genus %in% significant_genera_2)%>%
  filter(
    Time_Group %in% c("Before_Surgery", "Post_Surgery")
  )
 # Check for ASV-level
check <- genus_df_sub %>%
  count(Sample, Genus) %>%
  filter(n > 1)
 check
 
 
abund_vector <- as.vector(genus_df_sub$Abundance)

# Plot Distribution  
histogram_distances <- ggplot(data = data.frame(abund_vector), aes(x = abund_vector)) +
  geom_histogram(binwidth = 100, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  #coord_cartesian(xlim = c(0, 1.52)) +
  labs(
    title = "Distribution of abundances",
    x = "abundance",
    y = "Count"
  ) +
  scale_y_log10()
print(histogram_distances)

# multiple relative abundance by the average depth then add 1 before log10()
ps_tss_log <- transform_sample_counts(
  ps,
  function(x) log10((x / sum(x)) * avg_depth + 1)
)

ps_tss_log_df <- psmelt(ps_tss_log)
 
# model the taxa identified in Wilcoxen test as significant
ps_tss_log_df_sub <- ps_tss_log_df %>%
  filter(OTU %in% significant_genera_2)%>%
  filter(
    Time_Group %in% c("Before_Surgery", "Post_Surgery")
  )
 

 
 
log_abund <- as.vector(ps_tss_log_df_sub$Abundance)
# Plot Distribution 
histogram_distances <- ggplot(data = data.frame(log_abund), aes(x = log_abund)) +
  geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  #coord_cartesian(xlim = c(0, 1.52)) +
  labs(
    title = "Distribution of Log normalized-abundances",
    x = "Log10-abundance",
    y = "Count"
  )

print(histogram_distances)
```

LMM  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
ps_tss_log_df_sub <- ps_tss_log_df_sub %>%
  mutate(
    Sample_Site = forcats::fct_relevel(Sample_Site, "Control"),
    Time_Group = factor(Time_Group)  
  )
 

ASV_models <- ps_tss_log_df_sub %>%
  group_by(OTU) %>%
  group_nest() %>%
  mutate(
    model = map(
      data,
      ~ lmerTest::lmer(Abundance ~ Time_Group*Sample_Site + (1 | Dog_ID), data = .x)
    ),
    sum_tbl = map(model, ~ summary(.x))
  )
 

# Extract coefficient tables with p-values AND the fixed effect (Time_Group)
results <- ASV_models %>%
  mutate(
    coefs = map(
      model,
      ~ as.data.frame(summary(.x)$coefficients) %>%
        tibble::rownames_to_column("Term")
    )
  ) %>%
  dplyr::select(OTU, coefs) %>%
  unnest(coefs) %>%
  filter(
    grepl(
      "^Time_Group.*:Sample_SiteSurgical$|^Sample_SiteSurgical:Time_Group",
      Term
    )
  ) %>%
  mutate(
    Time_Group = sub(
      "^Time_Group(.*):Sample_SiteSurgical$",
      "\\1",
      Term
    )
  ) %>%
  arrange(OTU, Time_Group)


# Rename columns  
results <- results %>%
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    df = df,
    statistic = `t value`,
    p.value = `Pr(>|t|)`
  ) 

results
 
 
 
```



## Control Samples  
### Before_Surgery (Admission + Pre-Surgery) vs Post-Surgery 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)%>%
  group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_postsurg <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_postsurg


  
```
 
### Inflammation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Inflammation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)%>%
   group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_inflam <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_inflam
  
```
 
### Proliferation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Proliferation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)%>%
   group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_prolif <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_prolif
  
```
 
### Maturation samples 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Maturation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- ASV_rel_df_ctrl %>%
  filter(Time_Group %in% time_groups)%>%
   group_by(Dog_ID) %>%
  filter(n_distinct(Time_Group) == length(time_groups)) %>%
  ungroup()
# Keep genera signficant in surgical groups
genus_rel_df_surg_2_filt <- genus_rel_df_surg_2 %>%
  dplyr::filter(OTU %in% significant_genera_1)


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_abun = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_abun,
    # when abundance is 0 for the dogID-genus at time_group instead of NA (default)
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2
) %>%
  mutate(
    Time_Group_1 = Time_Group_1,
    Time_Group_2 = Time_Group_2
  )


tax_df <- as.data.frame(tax_table(ps_ASV_rel))

tax_df <- tax_df %>%
  tibble::rownames_to_column(var = "OTU")

results_wt_annotated_matur <- results_wt %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_matur
  
```
 
### Heatmap
```{r,  message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 

results_wt_annotated <- bind_rows(results_wt_annotated_postsurg, results_wt_annotated_inflam, results_wt_annotated_prolif, results_wt_annotated_matur)
 
 # get signed log(p-value) for heatmap
 results <- results_wt_annotated %>%
  mutate(
    signed_log10_p = -(sign(wilcox_estimate) * -log10(p))
  )
 
# generate per-taxa matrix of signed-log(pvalues)
 score_matrix <- results %>%
  dplyr::select(OTU, Time_Group_2, signed_log10_p) %>%
  pivot_wider(
    names_from = Time_Group_2,
    values_from = signed_log10_p
  ) %>%
  column_to_rownames("OTU") %>%
  as.matrix()
  
heatmap_matrix <- score_matrix[
  complete.cases(score_matrix),
  ,
  drop = FALSE
]
 
 

signed_max <- max(abs(heatmap_matrix), na.rm = TRUE)
 
col_fun <- circlize::colorRamp2(
  c(-4, 0, 4),
  c("#2166AC", "white", "#B2182B")
)
 


 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation"
)
 
 
 
col_dend <- heatmap_matrix |>
  t() |>
  dist(method = "euclidean") |>
  hclust(method = "complete") |>
  as.dendrogram()

# Rotate branches  
col_dend <- dendextend::rotate(
  col_dend,
  order = timepoint_order
)

# Reorder matrix rows explicitly
heatmap_matrix_ordered <- heatmap_matrix[row_names_in_heatmap_order, , drop = FALSE]

ht <- ComplexHeatmap::Heatmap(
  heatmap_matrix_ordered,
  name = "Signed\n-log10(p)",
  col = col_fun,

  cluster_rows = FALSE,
  cluster_columns = col_dend,
  column_dend_reorder = FALSE,

  na_col = "grey90",
  column_title = "Selected OTU in Control Groups",
  row_title = "OTU",
  row_names_side = "right",
  row_names_gp = grid::gpar(fontsize = 7)
)

ComplexHeatmap::draw(ht)

```

### Plot taxa 
```{r, fig.width=8, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
genus_rel_df_sig <- ASV_rel_df %>% 
  filter(Sample_Site %in% c("Control", "Surgical"))%>%
# filter(Time_Group %in% c("Before_Surgery", "Post_Surgery"))%>%
  filter(OTU %in% significant_genera_2) %>%
  group_by(OTU, Time_Group, Sample_Site, Dog_ID) %>%
  summarise(
    avg_rel_abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )
 
genus_rel_df_sig

ggplot(
  genus_rel_df_sig,
  aes(
    x = Time_Group,
    y = avg_rel_abundance,
    fill = Sample_Site,
    group = interaction(Time_Group, Sample_Site)
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.65,
    outlier.alpha = 0.6
  ) +
  geom_jitter(
    aes(color = Sample_Site),
    position = position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.75
    ),
    size = 1.8,
    alpha = 0.7
  ) +
  facet_wrap(~ OTU, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Control" = "purple",
      "Surgical" = "blue"
    )
  ) +
  scale_color_manual(
    values = c(
      "Control" = "purple",
      "Surgical" = "blue"
    ),
    guide = "none"
  ) +
  labs(
    x = "Time group",
    y = "Relative abundance per dog",
    fill = "Sample site",
    title = "OTU abundance by time group and sample site"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )+ stat_summary(
    fun = median,
    geom = "point",
    aes(group = Sample_Site),
    position = position_dodge(width = 0.75),
    shape = 21,
    size = 2,
    stroke = 0.8,
    fill = "white",
    color = "black"
  )
  
```

 
# Chlorhex vs. Skin Samples BC

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# phyloseq object with all skin and chlorhex samples
ps_1 <- readRDS("../01_input_files/phyloseq/ps_skin_negatives.rds")
metadata <- as(sample_data(ps_1), "data.frame")
 
# Normalize ASV read counts 
ps_rel_1 <- transform_sample_counts(ps_1, function(x) x / sum(x))
rel_abundance_vector <- as.numeric(
  as(phyloseq::otu_table(ps_rel_1), "matrix")
)

# Generate bray-curtis distance matrix using ASV relative abundances
bray_dist <- phyloseq::distance(ps_rel_1, method = "bray")
bray_matrix <- as.matrix(bray_dist)

# ordinate PCoA from full bray-curtis matrix 
set.seed(100)

bray_pcoa <- ordinate(ps_rel_1, method = "PCoA", distance =  bray_dist)

p1 <- plot_ordination(ps_rel_1, bray_pcoa, color = "Sample_Site")


p1 + ggtitle("PCoA of Bray-Curtis Dissimilarities", subtitle = "Ellipses encircle and color samples belonging to the same sample type.") +
  geom_point(size = 1.5) + 
  theme_bw() +
  stat_ellipse(aes(fill = Sample_Site, group = Sample_Site), alpha = 0.2, geom = "polygon")
  
metadata <- as(sample_data(ps_rel_1), "data.frame")
 

# PERMANOVA test 
set.seed(100)
adonis2(bray_matrix ~ Sample_Site, data = metadata, permutations = 999, by = "margin")  
 


# Set timepoint order 
day_levels <- c(
  "Admission",
  "Day 0a - Pre Surgery",
  "Day 0b - Surgery",
  "Day 0c - Post Surgery",
  "Day 1",
  "Day 3",
  "Week 1",
  "Week 2",
  "Week 3",
  "Week 4",
  "Week 5",
  "Week 6"
)
metadata$Day_Order <- factor(metadata$Day_Order, levels = day_levels)

#Count skin samples across timepoints
count_df <- metadata %>% 
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>% 
  distinct() %>% 
  group_by(Day_Order, Sample_Site) %>% 
  summarise(n = n(), .groups = "drop")
 

count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)
count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Surgical", "Control", "Chlorhex", "Air", "Kit"))

p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Surgical" = "blue", "Control" = "purple", "Chlorhex" = "orange", "Kit" = "green", "Air" = "red")) +
  labs(
    x = "Timepoint",
    y = "Number of Samples",
    fill = "Skin Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)
 

# All Chlorhex samples 
chlorhex_samples <- metadata %>%
  filter(Sample_Site == "Chlorhex")
 
# Dogs with associated chlorhex sample
chlorhex_dogs <- chlorhex_samples %>%
  distinct(Dog_ID) 


paired_groups <- metadata %>%
  filter(Sample_Site %in% c("Surgical", "Control")) %>%
  filter(Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0b - Surgery", "Day 0c - Post Surgery", "Day 1")) %>%
  filter(Dog_ID %in% chlorhex_dogs$Dog_ID) 
  #group_by(Individual) %>%
  #summarize(n_sites = n_distinct(Day_Order), .groups = "drop") %>%
 # filter(n_sites == 2)  

# Dogs with associated chlorhex sample
dogs_with_chlorhex <- paired_groups %>%
  distinct(Dog_ID) 

chlorhex_samples <- metadata %>%
  filter(Sample_Site == "Chlorhex")%>%
  filter(Dog_ID %in% dogs_with_chlorhex$Dog_ID) 

# Combine
metadata_final <- bind_rows(chlorhex_samples, paired_groups) %>%
  arrange(Individual, Day_Order) 
 
 
#Count skin samples across timepoints
count_df <- metadata_final %>% 
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>% 
  distinct() %>% 
  group_by(Day_Order, Sample_Site) %>% 
  summarise(n = n(), .groups = "drop")
 

count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)
count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Surgical", "Control", "Chlorhex"))

p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Surgical" = "blue", "Control" = "purple", "Chlorhex" = "orange")) +
  labs(
    x = "Timepoint",
    y = "Number of Samples",
    fill = "Skin Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)
```

## Surgical Samples vs Chlorhex
```{r,fig.width=11, fig.height=8,message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

 metadata_final <- bind_rows(chlorhex_samples, paired_groups) %>%
  arrange(Individual, Day_Order) %>%
 filter(Sample_Site %in% c("Surgical", "Chlorhex"))  
  
 
# Prune phyloseq object so it only contains these remaining samples 
ps_rel_2 <- prune_samples(metadata_final$SampleID, ps_rel_1)
 
# Generate bray-curtis distance matrix using ASV relative abundances
bray_dist <- phyloseq::distance(ps_rel_2, method = "bray")
bray_matrix <- as.matrix(bray_dist)


bray_arcsin_s <- asin(sqrt(bray_matrix))

 
 # generate BC matrix where chlorhex samples compared to surgical samples across days within dogs  
chlorhex_vs_all_same_site <- metadata_final %>%
  group_by(Dog_ID) %>%
  mutate(
    Chlorhex_sample = SampleID[Day_Order == "Day 0b - Surgery"][1]
  ) %>%
  ungroup() %>%
  filter(!is.na(Chlorhex_sample)) %>%
  filter(SampleID != Chlorhex_sample) %>%
  mutate(
    Bray_Distance = mapply(function(ref, other) {
      bray_arcsin_s[ref, other]
    }, Chlorhex_sample, SampleID)
  ) %>%
  dplyr::select(Dog_ID, Sample_Site, Chlorhex_sample, SampleID, Day_Order,
         Sample_Plate, Collection_Day, Bray_Distance)  
 


# Random effects to account for per dog replicates
lmer_bray <- lmer(
  Bray_Distance ~ Day_Order + (1 | Dog_ID),
  data = chlorhex_vs_all_same_site
)

anova(lmer_bray)
summary(lmer_bray)
 
 
# estimated marginal means of Day-Order within each site (control and surgical)
emm <- emmeans(lmer_bray, ~ Day_Order)

emm_df <- as.data.frame(emm)[, c("Day_Order",  "emmean", "SE", "df", "lower.CL", "upper.CL")]


alpha <- 0.05
 
contrasts <- as.data.frame(pairs(emm, adjust = "tukey"))

contrasts <- contrasts %>%
  mutate(
    Group1 = case_when(
      str_detect(contrast, "^\\(") ~ str_match(contrast, "^\\((.*?)\\)")[,2],
      TRUE ~ str_match(contrast, "^(.*?) - ")[,2]
    ),
    Group2 = case_when(
      str_detect(contrast, "\\) - \\(") ~ str_match(contrast, "- \\((.*?)\\)$")[,2],
      str_detect(contrast, "^\\(") ~ str_replace(contrast, "^\\(.*?\\) - ", ""),
      TRUE ~ str_match(contrast, "- (.*)$")[,2]
    ),
    Group1 = str_remove_all(str_trim(Group1), "[()]"),
    Group2 = str_remove_all(str_trim(Group2), "[()]")
  )

pairwise_tidy <- contrasts %>%
  filter(!is.na(p.value)) %>% 
  mutate(
    Significant = ifelse(p.value <= alpha, "Yes", "No"),
    Stars = case_when(
      p.value <= 0.001 ~ "***",
      p.value <= 0.01  ~ "**",
      p.value <= 0.05  ~ "*",
      p.value <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p.value = round(p.value, 4) ) %>%
  arrange(p.value) 

# Filter out insignificant pairwise results
sig_df <- pairwise_tidy %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,    
    label = Stars
  )
x_levels <- levels(factor(chlorhex_vs_all_same_site$Day_Order))

sig_df <- sig_df %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p.value) %>%
  mutate(
    y.position = max(chlorhex_vs_all_same_site$Bray_Distance, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )
 
 

p_surg <- ggplot(chlorhex_vs_all_same_site, aes(x = Day_Order, y = Bray_Distance)) +
  coord_cartesian(ylim = c(0, 1.7)) +
  geom_boxplot(fill = "#3399FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = emm_df$emmean[1], linetype = "dashed", color="red") +
  geom_point(data = emm_df, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = emm_df, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Surgical-Site: BC Dissimilarity to Chlorhex",
    subtitle = "Trend line indicating estimated marginal means per timepoint, horizontal red line indicates emmean of Admission",
    x = "Timepoint",
    y = "Bray-Curtis Dissimilarity to Chlorhex Sample"
  ) +
  theme_minimal() +
 # stat_pvalue_manual(
   # sig_df_surg,
    #label = "label",
    #xmin = "xmin",
  #  xmax = "xmax",
   # y.position = "y.position",
   # tip.length = 0.01,
  #  bracket.size = 0.5,
   # size = 5
  #) +
    theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  ) +
  coord_cartesian(ylim = c(0.3, 1.6)) 

p_surg
```

## Control Samples vs Chlorhex
```{r,fig.width=11, fig.height=8,message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

metadata_final <- bind_rows(chlorhex_samples, paired_groups) %>%
  arrange(Individual, Day_Order) %>%
 filter(Sample_Site %in% c("Control", "Chlorhex"))  
  
 
# Prune phyloseq object so it only contains these remaining samples 
ps_rel_2 <- prune_samples(metadata_final$SampleID, ps_rel_1)
 
# Generate bray-curtis distance matrix using ASV relative abundances
bray_dist <- phyloseq::distance(ps_rel_2, method = "bray")
bray_matrix <- as.matrix(bray_dist)


bray_arcsin_s <- asin(sqrt(bray_matrix))

 
 # generate BC matrix where chlorhex samples compared to surgical samples across days within dogs  
chlorhex_vs_all_same_site <- metadata_final %>%
  group_by(Dog_ID) %>%
  mutate(
    Chlorhex_sample = SampleID[Day_Order == "Day 0b - Surgery"][1]
  ) %>%
  ungroup() %>%
  filter(!is.na(Chlorhex_sample)) %>%
  filter(SampleID != Chlorhex_sample) %>%
  mutate(
    Bray_Distance = mapply(function(ref, other) {
      bray_arcsin_s[ref, other]
    }, Chlorhex_sample, SampleID)
  ) %>%
  dplyr::select(Dog_ID, Sample_Site, Chlorhex_sample, SampleID, Day_Order,
         Sample_Plate, Collection_Day, Bray_Distance)  
 


# Random effects to account for per dog replicates
lmer_bray <- lmer(
  Bray_Distance ~ Day_Order + (1 | Dog_ID),
  data = chlorhex_vs_all_same_site
)

anova(lmer_bray)
summary(lmer_bray)
 
 
# estimated marginal means of Day-Order within each site (control and surgical)
emm <- emmeans(lmer_bray, ~ Day_Order)

emm_df <- as.data.frame(emm)[, c("Day_Order",  "emmean", "SE", "df", "lower.CL", "upper.CL")]


alpha <- 0.05
 
contrasts <- as.data.frame(pairs(emm, adjust = "tukey"))

contrasts <- contrasts %>%
  mutate(
    Group1 = case_when(
      str_detect(contrast, "^\\(") ~ str_match(contrast, "^\\((.*?)\\)")[,2],
      TRUE ~ str_match(contrast, "^(.*?) - ")[,2]
    ),
    Group2 = case_when(
      str_detect(contrast, "\\) - \\(") ~ str_match(contrast, "- \\((.*?)\\)$")[,2],
      str_detect(contrast, "^\\(") ~ str_replace(contrast, "^\\(.*?\\) - ", ""),
      TRUE ~ str_match(contrast, "- (.*)$")[,2]
    ),
    Group1 = str_remove_all(str_trim(Group1), "[()]"),
    Group2 = str_remove_all(str_trim(Group2), "[()]")
  )

pairwise_tidy <- contrasts %>%
  filter(!is.na(p.value)) %>% 
  mutate(
    Significant = ifelse(p.value <= alpha, "Yes", "No"),
    Stars = case_when(
      p.value <= 0.001 ~ "***",
      p.value <= 0.01  ~ "**",
      p.value <= 0.05  ~ "*",
      p.value <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p.value = round(p.value, 4) ) %>%
  arrange(p.value) 

# Filter out insignificant pairwise results
sig_df <- pairwise_tidy %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,    
    label = Stars
  )
x_levels <- levels(factor(chlorhex_vs_all_same_site$Day_Order))

sig_df <- sig_df %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p.value) %>%
  mutate(
    y.position = max(chlorhex_vs_all_same_site$Bray_Distance, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )
 
 

p_control <- ggplot(chlorhex_vs_all_same_site, aes(x = Day_Order, y = Bray_Distance)) +
  coord_cartesian(ylim = c(0, 1.7)) +
  geom_boxplot(fill = "#CC99FF") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +

  geom_hline(yintercept = emm_df$emmean[1], linetype = "dashed", color="red") +
  geom_point(data = emm_df, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = emm_df, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +

  labs(
    title = "Control-Site: BC Dissimilarity to Chlorhex",
    subtitle = "Trend line indicating estimated marginal means per timepoint, horizontal red line indicates emmean of Admission",
    x = "Timepoint",
    y = "Bray-Curtis Dissimilarity to Chlorhex Sample"
  ) +
  theme_minimal() +
 # stat_pvalue_manual(
   # sig_df_surg,
    #label = "label",
    #xmin = "xmin",
  #  xmax = "xmax",
   # y.position = "y.position",
   # tip.length = 0.01,
  #  bracket.size = 0.5,
   # size = 5
  #) +
    theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  ) +
  coord_cartesian(ylim = c(0.3, 1.6)) 

p_control
```



### Plot taxa 
```{r, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
genus_rel_df_sig <- genus_rel_df %>% 
  filter(Sample_Site %in% c("Control", "Surgical"))%>%
 filter(Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0b - Surgery", "Day 0c - Post Surgery", "Day 1"))%>%
  filter(Genus %in% c("Lactococcus")) %>%
  group_by(Genus, Day_Order, Sample_Site, Dog_ID) %>%
  summarise(
    avg_rel_abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )
 

ggplot(
  genus_rel_df_sig,
  aes(
    x = Day_Order,
    y = avg_rel_abundance,
    fill = Sample_Site,
    group = interaction(Day_Order, Sample_Site)
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.65,
    outlier.alpha = 0.6
  ) +
  geom_jitter(
    aes(color = Sample_Site),
    position = position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.75
    ),
    size = 1.8,
    alpha = 0.7
  ) +
  facet_wrap(~ Genus, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Control" = "purple",
      "Surgical" = "blue"
    )
  ) +
  scale_color_manual(
    values = c(
      "Control" = "purple",
      "Surgical" = "blue"
    ),
    guide = "none"
  ) +
  labs(
    x = "Time group",
    y = "Relative abundance per dog",
    fill = "Sample site",
    title = "Genus abundance by time group and sample site"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )+ stat_summary(
    fun = median,
    geom = "point",
    aes(group = Sample_Site),
    position = position_dodge(width = 0.75),
    shape = 21,
    size = 2,
    stroke = 0.8,
    fill = "white",
    color = "black"
  )
  
```
 
```{r, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
genus_rel_df_sig <- genus_rel_df %>% 
  filter(Sample_Site %in% c("Control", "Surgical"))%>%
 filter(Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0b - Surgery", "Day 0c - Post Surgery", "Day 1"))%>%
  filter(Genus %in% c("Erysipelotrichaceae UCG-003")) %>%
  group_by(Genus, Day_Order, Sample_Site, Dog_ID) %>%
  summarise(
    avg_rel_abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )
 

ggplot(
  genus_rel_df_sig,
  aes(
    x = Day_Order,
    y = avg_rel_abundance,
    fill = Sample_Site,
    group = interaction(Day_Order, Sample_Site)
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.65,
    outlier.alpha = 0.6
  ) +
  geom_jitter(
    aes(color = Sample_Site),
    position = position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.75
    ),
    size = 1.8,
    alpha = 0.7
  ) +
  facet_wrap(~ Genus, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Control" = "purple",
      "Surgical" = "blue"
    )
  ) +
  scale_color_manual(
    values = c(
      "Control" = "purple",
      "Surgical" = "blue"
    ),
    guide = "none"
  ) +
  labs(
    x = "Time group",
    y = "Relative abundance per dog",
    fill = "Sample site",
    title = "Genus abundance by time group and sample site"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )+ stat_summary(
    fun = median,
    geom = "point",
    aes(group = Sample_Site),
    position = position_dodge(width = 0.75),
    shape = 21,
    size = 2,
    stroke = 0.8,
    fill = "white",
    color = "black"
  )
  
```
 
```{r, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
genus_rel_df_sig <- genus_rel_df %>% 
  filter(Sample_Site %in% c("Control", "Surgical"))%>%
 filter(Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0b - Surgery", "Day 0c - Post Surgery", "Day 1"))%>%
  filter(Genus %in% c("Faecalibacterium")) %>%
  group_by(Genus, Day_Order, Sample_Site, Dog_ID) %>%
  summarise(
    avg_rel_abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )
 

ggplot(
  genus_rel_df_sig,
  aes(
    x = Day_Order,
    y = avg_rel_abundance,
    fill = Sample_Site,
    group = interaction(Day_Order, Sample_Site)
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.65,
    outlier.alpha = 0.6
  ) +
  geom_jitter(
    aes(color = Sample_Site),
    position = position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.75
    ),
    size = 1.8,
    alpha = 0.7
  ) +
  facet_wrap(~ Genus, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Control" = "purple",
      "Surgical" = "blue"
    )
  ) +
  scale_color_manual(
    values = c(
      "Control" = "purple",
      "Surgical" = "blue"
    ),
    guide = "none"
  ) +
  labs(
    x = "Time group",
    y = "Relative abundance per dog",
    fill = "Sample site",
    title = "Genus abundance by time group and sample site"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )+ stat_summary(
    fun = median,
    geom = "point",
    aes(group = Sample_Site),
    position = position_dodge(width = 0.75),
    shape = 21,
    size = 2,
    stroke = 0.8,
    fill = "white",
    color = "black"
  )
  
```



### Plot taxa 
## Top 10 Genera
```{r, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=TRUE}
# Aggregate ASV counts for Genus-Level Analysis + remove ASVs with unknown Genus
ps_genus <- tax_glom(ps_1, taxrank = "Genus", NArm = TRUE)
# bUnique genera
ntaxa(ps_genus)

ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))


# turn phyloseq object into a dataframe 
genus_rel_df <- psmelt(ps_genus_rel) %>%
arrange(Dog_ID, Genus, Sample_Site, Day_Order)

paired_data <- genus_rel_df %>%
  filter(Sample_Site %in% c("Control", "Surgical")) %>%
  group_by(Dog_ID, Day_Order) %>%
  summarize(n_sites = n_distinct(Sample_Site), .groups = "drop") %>%
  filter(n_sites == 2) %>%
  dplyr::select(Dog_ID, Day_Order)


genus_rel_df_filtered <- genus_rel_df %>% 
  filter(Sample_Site %in% c("Chlorhex", "Surgical", "Control")) %>%
  filter(Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0b - Surgery", "Day 0c - Post Surgery", "Day 1")) %>%
  filter(Genus %in% c("Lactococcus", "Erysipelotrichaceae UCG-003", "Faecalibacterium")) %>%
  group_by(Genus, Day_Order, Sample_Site, Dog_ID) %>%
  summarise(
    avg_rel_abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Paired control and surgical samples
control_surgical_paired <- genus_rel_df_filtered %>%
  filter(Sample_Site %in% c("Control", "Surgical")) %>%
  semi_join(paired_data, by = c("Dog_ID", "Day_Order"))

# Chlorhex  samples
chlorhex_all <- genus_rel_df_filtered %>%
  filter(Sample_Site == "Chlorhex")

 
genus_rel_df_sig <- bind_rows(control_surgical_paired, chlorhex_all)
 

ggplot(
  genus_rel_df_sig,
  aes(
    x = Day_Order,
    y = avg_rel_abundance,
    fill = Sample_Site,
    group = interaction(Genus, Sample_Site, Day_Order)
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.65,
    outlier.alpha = 0.6
  ) +
  geom_jitter(
    aes(color = Sample_Site),
    position = position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.75
    ),
    size = 1.8,
    alpha = 0.7
  ) +
 facet_wrap(~ Genus, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Chlorhex" = "orange",
    "Kit" = "green",
    "Air" = "red",
    "Control" = "purple",
    "Surgical" = "blue"
    )
  ) +
  scale_color_manual(
    values = c(
      "Chlorhex" = "orange",
    "Kit" = "green",
    "Air" = "red",
    "Control" = "purple",
    "Surgical" = "blue"
    ),
    guide = "none"
  ) +
  labs(
    x = "Sample_Site",
    y = "Relative abundance per dog",
    fill = "Sample site",
    title = "Genera abundance with significantly higher abundances in surgical groups compared to control skin samples post-surgery."
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )+ stat_summary(
    fun = median,
    geom = "point",
    aes(group = Sample_Site),
    position = position_dodge(width = 0.75),
    shape = 21,
    size = 2,
    stroke = 0.8,
    fill = "white",
    color = "black"
  )
  
```


 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 