---
title: "Canine Skin Microbiome: Microbial Diversity Analysis "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---
**Purpose:**
Identify a change-point in the skin microbial diversity of dogs after surgery over the course of 6 weeks. 

**Description:**
Each dog has its own control prior to surgery as well as a control site that did not undergo surgery (a before-after-control-impact, BACI design). We will fit a change-point model of the dissimilarity (using Bray-Curtis) between the surgical site and control skin site on each of the dogs in the study. We will test the null hypothesis of no change-point in this dissimilarity measure before surgery. 

**References:**
Cutaneous Microbiome Profiles Following Chlorhexidine Treatment in a 72-Hour Daily Follow-Up Paired Design: a Pilot Study (ASM - June 2022) <https://journals.asm.org/doi/full/10.1128/spectrum.01753-21>
Longitudinal profiling of the microbiome at four body sites reveals core stability and individualized dynamics during health and disease (Cell Host Microbe - April 2025) <https://pmc.ncbi.nlm.nih.gov/articles/PMC11022754/>

### Load files
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(knitr)
library(FSA)
library(tidyverse)
library(vegan)
library(phyloseq)
library(ggplot2)
library(ggtree)
library(scales)  
library(RColorBrewer)
library(segmented)
library(stringr)
library(pairwiseAdonis)
library(ggpubr)
library(changepoint)
library(dplyr)
library(lme4)
library(emmeans)
library(nortest)
library(mcp)
```

### Create phyloseq objects

1. Load each output file for each sequencing run processed in decontam 
2. Filter out negative-controls (only Skin-Control and Skin-Surgical sites)
3. Merge all phyloseq objects for downstream analysis
```{r, message=FALSE, warning=FALSE, echo=TRUE}
load("../02_output_files/01_filtered_ASV_tables/CH1_filtered_canine_skin_microbiome.rda")

ps1 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps1 <- prune_samples(
  grepl("Skin", sample_data(ps1)$Location_Type),
  ps1
)
load("../02_output_files/01_filtered_ASV_tables/CH2_filtered_canine_skin_microbiome.rda")

ps2 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps2 <- prune_samples(
  grepl("Skin", sample_data(ps2)$Location_Type),
  ps2
)
load("../02_output_files/01_filtered_ASV_tables/CH3_filtered_canine_skin_microbiome.rda")

ps3 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
 ps3 <- prune_samples(
  grepl("Skin", sample_data(ps3)$Location_Type),
  ps3
)
load("../02_output_files/01_filtered_ASV_tables/CH4_filtered_canine_skin_microbiome.rda")

ps4 <- phyloseq(otu_table(ASV_feature_table_decontam, taxa_are_rows=FALSE), 
               sample_data(sample_metadata_decontam), 
               tax_table(taxa_annotations_decontam))
  ps4 <- prune_samples(
  grepl("Skin", sample_data(ps4)$Location_Type),
  ps4
)
ps <- merge_phyloseq(ps1, ps2, ps3, ps4)
ps <- prune_samples(
  grepl("Skin", sample_data(ps)$Location_Type),
  ps
)
```

# Bioinformatics Analysis

### Remove uncharacterized ASVs
```{r, message=FALSE, warning=FALSE, echo=TRUE}
otu_matrix <- as(otu_table(ps), "matrix")

# Get total ASV count per Sample_Plate
# Sample_Plate indicates sequencing run 
total_asvs_per_plate <- data.frame(
  Sample_Plate = sample_data(ps)$Sample_Plate,
  as.data.frame(otu_matrix)
) %>%
  group_by(Sample_Plate) %>%
  summarise(
    Total_ASVs = sum(colSums(across(-where(is.character))) > 0),
    .groups = "drop"
  )

unknown_asvs <- rownames(tax_table(ps))[
  is.na(tax_table(ps)[, "Phylum"]) | tax_table(ps)[, "Phylum"] %in% c("", "uncharacterized")
]


otu_unknown <- otu_matrix[, unknown_asvs, drop = FALSE]

# Provide summary for uncharacterized ASV count per seq. run
out <- data.frame(
  Sample_Plate = sample_data(ps)$Sample_Plate,
  as.data.frame(otu_unknown)
) %>%
  group_by(Sample_Plate) %>%
  summarise(
    Unknown_ASVs_Count = sum(colSums(across(-where(is.character))) > 0),
    .groups = "drop"
  ) %>%
  left_join(total_asvs_per_plate, by = "Sample_Plate") %>%
  mutate(
    Percent_Removed = 100 * Unknown_ASVs_Count / Total_ASVs
  )
print(out)
 
# Subset phyloseq object for downstream analysis
ps_1 <- subset_taxa(ps, !is.na(Phylum) & !Phylum %in% c("", "uncharacterized"))
```

### Check for low-prevalence taxa
```{r, fig.width=11, fig.height=8}
# Compute prevalence for each ASV 
prevdf = apply(X = otu_table(ps_1),
               MARGIN = ifelse(taxa_are_rows(ps_1), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})
 
# Add taxa and total read counts to prevalence df
prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(ps_1),
                    tax_table(ps_1))

# Are there phyla that are comprised of mostly low-prevalence features? 
# Compute the total and average prevalences of the features in each phylum.
phylum_summary <- plyr::ddply(prevdf, "Phylum", function(df1){
  cbind(mean(df1$Prevalence), sum(df1$Prevalence))
})
colnames(phylum_summary) <- c("Phylum", "Mean_Prevalence", "Sum_Prevalence")

# Phyla by Total ASV Prevalence 
phylum_summary_sorted <- phylum_summary[order(-phylum_summary$Sum_Prevalence), ]
print(phylum_summary_sorted, row.names = FALSE)
ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(ps_1),color=Phylum)) +
  ggtitle("Prevalence of Phyla Across Skin-Swab Samples", subtitle="dashed line = 5% prevalence") +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) +  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```

### Total Reads + ASVs per Individual and Sequencing Run
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE}
out_dir <- "../02_output_files/02_image_files"

# Get reads and metadata per sample
reads_per_sample <- sample_sums(ps_1)
md <- as(sample_data(ps_1), "data.frame")
 
sample_names_ps <- sample_names(ps_1)

# Ensure sample order matches
if (!all(sample_names_ps == rownames(md))) {
  md <- md[sample_names_ps, ]
}

# Create dataframe with sample-level stats
df <- data.frame(
  sample = sample_names_ps,
  total_reads = reads_per_sample,
  Sample_Plate = md$Sample_Plate,
  Individual = md$Individual,
  stringsAsFactors = FALSE
)

# Calculate total reads and ASVs per sample and add to metadata
otu <- otu_table(ps_1)
if (taxa_are_rows(otu)) {
  total_asvs <- colSums(otu > 0)
} else {
  total_asvs <- rowSums(otu > 0)
}

total_reads <- sample_sums(ps_1)
 
md$total_asv <- total_asvs
md$total_reads <- total_reads

sample_data(ps_1) <- sample_data(md)

md <- md %>%
  arrange(Surgery_Type, Individual) %>%
  mutate(Individual = factor(Individual, levels = unique(Individual)))

my_cols <- colorRampPalette(brewer.pal(8, "Dark2"))(20)

# Boxplot for total reads by Individual (each point = one sample)
p_reads <- ggplot(md, aes(x = Individual, y = total_reads, fill = Surgery_Type)) +
  geom_boxplot()  +
  scale_fill_manual(values = c(
    "Abdomen" = "orange",
    "Stifle" = "pink"
  )) +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "K"),
    name = "Total Reads (per Sample, in thousands)"
  ) +
  labs(
    title = "Total Reads per Sample by Dog",
    x = "Dog_ID"
  ) +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

p_reads

out_reads <- paste0(out_dir, "/total_reads_by_individual.png")
ggplot2::ggsave(out_reads, p_reads, width = 8, height = 6)
p_reads

# Boxplot for total ASVs by Individual (each point = one sample)
p_asv <- ggplot(by_ind, aes(x = Individual, y = total_asv, fill = Individual)) +
  geom_boxplot() +
  scale_fill_manual(values = c(
    "Abdomen" = "orange",
    "Stifle" = "pink"
  )) +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "k"),
    name = "Total ASVs (per Individual, in thousands)"
  ) +
  labs(
    title = "Total ASVs per Sample by Dog",
    x = "Dog_ID"
  ) +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
p_asv

out_asv <- paste0(out_dir, "/total_asvs_by_individual.png")
ggplot2::ggsave(out_asv, p_asv, width = 8, height = 6)

print(by_ind)

# Aggregate reads and ASVs per Individual (summing across all samples from same dog)
by_ind <- aggregate(
  cbind(total_reads, total_asv) ~ Individual + Sample_Plate,
  data = md,
  FUN = sum
)

# Boxplot for total reads by Sample_Plate (each point = one Individual)
p_reads <- ggplot(by_ind, aes(x = Sample_Plate, y = total_reads, fill = Sample_Plate)) +
  geom_boxplot() +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "K"),
    name = "Total Reads (per Individual, in thousands)"
  ) +
  labs(
    title = "Total Reads per Individual by Sample Plate",
    x = "Sample Plate"
  ) +
  theme_minimal()

out_reads <- paste0(out_dir, "/total_reads_by_sample_plate.png")
ggplot2::ggsave(out_reads, p_reads, width = 8, height = 6)
p_reads

# Boxplot for total ASVs by Sample_Plate (each point = one Individual)
p_asv <- ggplot(by_ind, aes(x = Sample_Plate, y = total_asv, fill = Sample_Plate)) +
  geom_boxplot() +
  scale_y_continuous(
    labels = function(x) paste0(x / 1000, "k"),
    name = "Total ASVs (per Individual, in thousands)"
  ) +
  labs(
    title = "Total ASVs per Individual by Sample Plate",
    x = "Sample Plate"
  ) +
  theme_minimal()
p_asv

out_asv <- paste0(out_dir, "/total_asvs_by_sample_plate.png")
ggplot2::ggsave(out_asv, p_asv, width = 8, height = 6)

print(by_ind)

```

### Normalize Read Counts  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Normalization using relative abundances
ps_rel <- transform_sample_counts(ps_1, function(x) x / sum(x))
individual_list <- unique(sample_data(ps_rel)$Individual)
``` 

### Plot Relative Abundances per Individual 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# create color palette so phyla is the same color across all individual dog plots 
all_phyla <- tax_table(ps_rel)[, "Phylum"] %>% as.vector() %>% unique() %>% na.omit() %>% sort()
n_phyla <- length(all_phyla)
base_cols <- brewer.pal(min(n_phyla, 12), "Paired")
phylum_cols_vec <- colorRampPalette(base_cols)(n_phyla)
phylum_colors <- setNames(phylum_cols_vec, all_phyla)

# Generate per dog plots showing (top 10) phyla diversity across timepoints
for (ind in individual_list) {
  ps_ind <- subset_samples(ps_rel, Individual == ind)
    plot_df <- psmelt(ps_ind) %>% 
    filter(Sample_Site %in% c("Control", "Surgical")) %>% 
    mutate(
      Surgical_Location = unique(Sample_Location[Sample_Site == "Surgical" & !is.na(Sample_Location)])[1]
    ) %>% 
    mutate(Surgical_Location = factor(Surgical_Location))
  
  # Identify top 10 phyla specific to this dog's profile
  top20_ind <- plot_df %>% 
    group_by(Phylum) %>% 
    summarise(Total_Abundance = sum(Abundance), .groups = "drop") %>% 
    arrange(desc(Total_Abundance)) %>% 
    slice_head(n = 10) %>% 
    pull(Phylum)
  
  phylum_levels_ind <- c(top20_ind, "Other")
   
  plot_df <- plot_df %>% mutate(
    Phylum = ifelse(Phylum %in% top20_ind, Phylum, "Other"),
    Phylum = factor(Phylum, levels = phylum_levels_ind),
    Sample_Site = factor(Sample_Site, levels = c("Control", "Surgical")),
    Surgical_Location = factor(Surgical_Location, levels = c("Abdomen", "Stifle")),
    Day_Order = factor(Day_Order)
  )
  
  local_cols <- phylum_colors[top20_ind]
  missing_indices <- is.na(local_cols)
  if(any(missing_indices)) {
    fallback_pool <- brewer.pal(12, "Paired")
    local_cols[missing_indices] <- sample(fallback_pool, sum(missing_indices), replace = TRUE)
  }
  values_ind <- c(local_cols, "Other" = "grey70")
  
  p <- ggplot(plot_df, aes(x = Day_Order, y = Abundance, fill = Phylum)) + 
    geom_col(position = "stack", width = 0.7) +  
    facet_wrap(~Sample_Site, nrow = 1) +          
    scale_fill_manual(
      values = values_ind,
      breaks = phylum_levels_ind,
      na.value = "black"
    ) + 
    theme_minimal() + 
    theme(
      axis.text.x = element_text(angle = 75, hjust = 1),
      panel.spacing = unit(1.5, "lines")        
    ) + 
    labs(
      title = paste0("Relative Abundance Across Timepoints by Phylum"),
      subtitle = paste0("Dog: ", ind, " Surgical Location: ", unique(plot_df$Surgical_Location)),
      x = "Timepoint",
      y = "Relative Abundance"
    )
  print(p)
}

 
# create color palette so genus is the same color across all individual dog plots 
all_phyla <- tax_table(ps_rel)[, "Genus"] %>% as.vector() %>% unique() %>% na.omit() %>% sort()
n_phyla <- length(all_phyla)
base_cols <- brewer.pal(min(n_phyla, 12), "Paired")
phylum_cols_vec <- colorRampPalette(base_cols)(n_phyla)
phylum_colors <- setNames(phylum_cols_vec, all_phyla)

# Generate per dog plots showing (top 10) phyla diversity across timepoints
for (ind in individual_list) {
  ps_ind <- subset_samples(ps_rel, Individual == ind)
    plot_df <- psmelt(ps_ind) %>% 
    filter(Sample_Site %in% c("Control", "Surgical")) %>% 
    mutate(
      Surgical_Location = unique(Sample_Location[Sample_Site == "Surgical" & !is.na(Sample_Location)])[1]
    ) %>% 
    mutate(Surgical_Location = factor(Surgical_Location))
  
  # Identify top 10 genus specific to this dog's profile
  top20_ind <- plot_df %>% 
    group_by(Genus) %>% 
    summarise(Total_Abundance = sum(Abundance), .groups = "drop") %>% 
    arrange(desc(Total_Abundance)) %>% 
    slice_head(n = 10) %>% 
    pull(Genus)
  
  phylum_levels_ind <- c(top20_ind, "Other")
   
  plot_df <- plot_df %>% mutate(
    Genus = ifelse(Genus %in% top20_ind, Genus, "Other"),
    Genus = factor(Genus, levels = phylum_levels_ind),
    Sample_Site = factor(Sample_Site, levels = c("Control", "Surgical")),
    Surgical_Location = factor(Surgical_Location, levels = c("Abdomen", "Stifle")),
    Day_Order = factor(Day_Order)
  )
  
  local_cols <- phylum_colors[top20_ind]
  missing_indices <- is.na(local_cols)
  if(any(missing_indices)) {
    fallback_pool <- brewer.pal(12, "Paired")
    local_cols[missing_indices] <- sample(fallback_pool, sum(missing_indices), replace = TRUE)
  }
  values_ind <- c(local_cols, "Other" = "grey70")
  
  p <- ggplot(plot_df, aes(x = Day_Order, y = Abundance, fill = Genus)) + 
    geom_col(position = "stack", width = 0.7) +  
    facet_wrap(~Sample_Site, nrow = 1) +          
    scale_fill_manual(
      values = values_ind,
      breaks = phylum_levels_ind,
      na.value = "black"
    ) + 
    theme_minimal() + 
    theme(
      axis.text.x = element_text(angle = 75, hjust = 1),
      panel.spacing = unit(1.5, "lines")        
    ) + 
    labs(
      title = paste0("Relative Abundance Across Timepoints by Genus"),
      subtitle = paste0("Dog: ", ind, " Surgical Location: ", unique(plot_df$Surgical_Location)),
      x = "Timepoint",
      y = "Relative Abundance"
    )
  print(p)
}
```

# Statistical Analysis
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALE, eval=FALSE}
# Calculate alpha-diversity using Shannon Diversity Index for each skin sample 
alphadiv_plot_data <- estimate_richness(ps_1, measures = "Shannon") %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(as.data.frame(sample_data(ps_1)), by = "SampleID") %>% 
  group_by(Individual) %>% 
  mutate(Surgical_Location = Sample_Location[Sample_Site == "Surgical"][1]) %>% 
  ungroup() %>% 
  mutate(Surgical_Location = factor(Surgical_Location)) %>% 
  arrange(Surgical_Location, Individual)
ggplot(alphadiv_plot_data, aes(x = factor(Day_Order), y = Shannon, color = Sample_Site, group = Sample_Site)) + 
  geom_point(size = 1, position = position_dodge(width = 0.3)) + 
  geom_line(position = position_dodge(width = 0.3)) + 
  facet_wrap(
    Surgical_Location ~ Individual, 
    labeller = label_both
  ) + 
  scale_color_manual(values = c("Control" = "purple", "Surgical" = "blue")) + 
  theme_bw() + 
  labs(
    title = "Shannon Diversity Across Timepoints per Dog",
    x = "Timepoint",
    y = "Shannon Index",
    color = "Skin Site"
  ) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```
 
### Remove unpaired timepoint samples   
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Extract sample data
metadata <- as(sample_data(ps_rel), "data.frame")
 
# Set timepoint order 
day_levels <- c(
  "Admission",
  "Day 0a - Pre Surgery",
  "Day 0c - Post Surgery",
  "Day 1",
  "Day 3",
  "Week 1",
  "Week 2",
  "Week 3",
  "Week 4",
  "Week 5",
  "Week 6"
)
metadata$Day_Order <- factor(metadata$Day_Order, levels = day_levels)
metadata$Sample_Site <- factor(metadata$Sample_Site, levels = c("Control", "Surgical"))

#Count skin samples across timepoints
count_df <- metadata %>% 
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>% 
  distinct() %>% 
  group_by(Day_Order, Sample_Site) %>% 
  summarise(n = n(), .groups = "drop")
 

count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)
count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Control", "Surgical"))

p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Control" = "blue", "Surgical" = "purple")) +
  labs(
    x = "Timepoint",
    y = "Number of Samples",
    fill = "Skin Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)

# Check which dogs are missing samples (either control or surgical) at each timepoint
missing_sample_unique <- metadata %>% 
  group_by(Individual, Day_Order) %>% 
  summarise(
    site_absent = if (all(c("Control", "Surgical") %in% Sample_Site)) NA_character_ 
                  else paste(setdiff(c("Control", "Surgical"), unique(Sample_Site)), collapse = ", "), 
    Sample_Unique = paste(sort(unique(Sample_Unique)), collapse = ", "), 
    n_sites = n_distinct(Sample_Site), 
    .groups = "drop"
  ) %>% 
  filter(n_sites < 2) %>% 
  dplyr::select(Individual, Day_Order, site_absent, Sample_Unique) %>%
  # Arrange by Individual and chronological factor timepoint
  arrange(Individual, Day_Order)

print(missing_sample_unique)

# Keep only samples that have control and surgical at each Day_Order
paired_groups <- metadata %>%
  group_by(Individual, Day_Order) %>%
  summarize(n_sites = n_distinct(Sample_Site), .groups = "drop") %>%
  filter(n_sites == 2) %>%
  transmute(Individual, Day_Order)
filtered_sample_data <- metadata %>%
  inner_join(paired_groups, by = c("Individual", "Day_Order"))

 
# Prune phyloseq object so it only contains these remaining samples 
ps_2 <- prune_samples(filtered_sample_data$SampleID, ps_rel)
 

# Count unique samples by Day_Order and Sample_Site
count_df <- filtered_sample_data %>%
  dplyr::select(Day_Order, Sample_Site, Sample_Unique) %>%
  distinct() %>%
  group_by(Day_Order, Sample_Site) %>%
  summarise(n = n(), .groups = "drop")


count_df$Day_Order <- factor(count_df$Day_Order, levels = day_levels)


count_df$Sample_Site <- factor(count_df$Sample_Site, levels = c("Control", "Surgical"))

# Updated after removing dog-timepoint samples with only one site
p <- ggplot(count_df, aes(x = Day_Order, y = n, fill = Sample_Site)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = c("Control" = "blue", "Surgical" = "purple")) +
  labs(
    x = "Timepoint",
    y = "Number of Samples",
    fill = "Skin Site"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p)

filtered_sample_data$Sample_Site <- factor(filtered_sample_data$Sample_Site)
filtered_sample_data$Dog_ID <- factor(filtered_sample_data$Dog_ID)
filtered_sample_data$Sample_Plate <- factor(filtered_sample_data$Sample_Plate)
filtered_sample_data$Collection_Day <- as.numeric(filtered_sample_data$Collection_Day)

```

### Bray-Curtis (BC) Dissimilarity  
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# Generate bray-curtis distance matrix using ASV relative abundances
bray_dist <- phyloseq::distance(ps_2, method = "bray")
bray_dist <- as.matrix(bray_dist)

keep_ids <- filtered_sample_data$SampleID
bray_matrix <- bray_dist[keep_ids, keep_ids]

# get vector for histogram
bray_vec <- as.vector(bray_matrix)

# Plot Distribution of Bray-Curtis Distances
histogram_distances <- ggplot(data = data.frame(bray_vec), aes(x = bray_vec)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of Bray-Curtis Distances",
    x = "Bray-Curtis Distance",
    y = "Count"
  )

print(histogram_distances)
```

## Sequencing Run Batch Effect

### PERMANOVA Test 

*Notes:*

* Batch effects were estimated using PERMANOVA analysis when both batch and Dog_ID were included
* Total variance explained by sequencing run is 10.67% 
* However batch effect is confounded with dogs since they are not spread across the runs but are restricted to one plate 
 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
bray_matrix <- as.dist(bray_matrix)

set.seed(100)
bray_pcoa <- ordinate(ps_2, method = "PCoA", distance =  bray_matrix)
p1 <- plot_ordination(ps_2, bray_pcoa, color = "Sample_Plate")


p1 + ggtitle("PCoA of Bray-Curtis Dissimilarities", subtitle = "Ellipses encircle and color samples belonging to the same sequencing run.") +
  geom_point(size = 1.5) + 
  theme_bw() +
  stat_ellipse(aes(fill = Sample_Plate, group = Sample_Plate), alpha = 0.2, geom = "polygon")


p1 + ggtitle("PCoA of Bray-Curtis Dissimilarities", subtitle = "Ellipses encircle subject replicates and colored according sequencing run.") +
  geom_point(size = 1.5) + 
  theme_bw() +
  stat_ellipse(aes(fill = Sample_Plate, group = Dog_ID), alpha = 0.2, geom = "polygon")


# PERMANOVA test
set.seed(100)
adonis2(bray_matrix ~ Sample_Plate + Dog_ID, data = filtered_sample_data, permutations = 999, by = "terms")  
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
# PERMDISP2 test  
bray_matrix <- as.dist(bray_matrix)
betadisper_result <- betadisper(bray_matrix, group = filtered_sample_data$Sample_Plate)

average_by_plate <- data.frame(
  Plate = filtered_sample_data$Sample_Plate,
  Dist_to_Centroid = betadisper_result$distances
)

average_by_plate_table <- average_by_plate %>%
  group_by(Plate) %>%
  summarize(
    Mean_Dist = mean(Dist_to_Centroid, na.rm = TRUE),
    SD_Dist   = sd(Dist_to_Centroid, na.rm = TRUE),
    CV_Percent = (SD_Dist / Mean_Dist *100) 
  )

print(average_by_plate_table)

# Extract the coordinates and centroids
df_points <- as.data.frame(scores(betadisper_result, display = "sites"))
df_points$Sample_Plate <- filtered_sample_data$Sample_Plate

df_centroids <- as.data.frame(scores(betadisper_result, display = "centroids"))
df_centroids$Sample_Plate <- rownames(df_centroids)

set.seed(100)
print(permutest(betadisper_result, perm=999))


# Pairwise post-hoc comparisons between sequencing runs
print(TukeyHSD(betadisper_result))

 
# PERMDISP2 test on dogs
betadisper_result <- betadisper(bray_matrix, group = filtered_sample_data$Dog_ID)

average_by_plate <- data.frame(
  Plate = filtered_sample_data$Dog_ID,
  Dist_to_Centroid = betadisper_result$distances
)

average_by_plate_table <- average_by_plate %>%
  group_by(Plate) %>%
  summarize(
    Mean_Dist = mean(Dist_to_Centroid, na.rm = TRUE),
    SD_Dist   = sd(Dist_to_Centroid, na.rm = TRUE),
    CV_Percent = (SD_Dist / Mean_Dist *100) 
  )

print(average_by_plate_table)

# Extract the coordinates and centroids
df_points <- as.data.frame(scores(betadisper_result, display = "sites"))
df_points$Dog_ID <- filtered_sample_data$Dog_ID

df_centroids <- as.data.frame(scores(betadisper_result, display = "centroids"))
df_centroids$Dog_ID <- rownames(df_centroids)

set.seed(100)
print(permutest(betadisper_result, perm=999))


# Pairwise post-hoc comparisons between dogs
print(TukeyHSD(betadisper_result))

bray_matrix <- as.matrix(bray_matrix)
```

## Comparison between timepoints  

*Notes:*

* A three-factor fixed design consisting of “Sample_Plate” (sequencing runs), “Sample_Site” (Surgery versus No Sugery ie Control), and “Day_Order” (categorical timepoint), with permutations within “Dog_ID” (subjects)  
* Day Order and Sample_Plate had significant effects on global BC dissimilarity, 2.93% and 10.7%, however Sample_Site did not. 
* Running pairwise permanova comparisons between all timepoints, after BH-adjusting p-values:
  * Day 0c - Post Surgery vs Week 5: adjusted p = 0.0138.
  * Day 1 vs Week 5: adjusted p = 0.0138.
  * Day 1 vs Week 4: adjusted p = 0.0183.
  * Day 0c - Post Surgery vs Week 4: adjusted p = 0.0344.

### Pairwise PERMANOVA Test 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# PERMANOVA test  
set.seed(100)
adonis2(bray_matrix ~ Sample_Plate + Sample_Site + Day_Order, data = filtered_sample_data, permutations = 999, strata = filtered_sample_data$Dog_ID, by = "terms")

# pairwise PERMANOVA between timepoints (Day_Order)  
set.seed(100)
pairwise_results <- pairwise.adonis2(
  bray_matrix ~ Day_Order,
  strata = "Dog_ID",
  data = filtered_sample_data,
  nperm = 9999)

p_values <- sapply(names(pairwise_results)[-1], function(x) {
  pairwise_results[[x]][["Pr(>F)"]][1]
})

# Benjamini-Hochberg (BH) adjustment since we are testing multiple comparisons
p_adjusted <- p.adjust(p_values, method = "BH")

pairwise_df <- data.frame(
  Comparison = names(p_values),
  p_value    = p_values,
  p_adj      = p_adjusted,
  row.names  = NULL
)

groups <- str_split_fixed(pairwise_df$Comparison, "_vs_", 2)

pairwise_df$Group1 <- groups[, 1]
pairwise_df$Group2 <- groups[, 2]

pairwise_df <- pairwise_df[, c("Group1", "Group2", "p_value", "p_adj")]

alpha <- 0.05
pairwise_tidy <- pairwise_df %>%
  filter(!is.na(p_adj)) %>% 
  mutate(
    Significant = ifelse(p_adj <= alpha, "Yes", "No"),
    Stars = case_when(
      p_adj <= 0.001 ~ "***",
      p_adj <= 0.01  ~ "**",
      p_adj <= 0.05  ~ "*",
      p_adj <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p_adj = round(p_adj, 4) ) %>%
  arrange(p_adj) 

 pairwise_tidy
  
```

## Within-dog control–surgical BC dissimilarity  

### Bray-curtis matrix transformation

The pairwise Bray-Curtis distances were transformed to improve normality before LMM analysis. An arcsine transformation was ultimately selected, after evaluating eight potential transformations, including log, square-root, arcsine, and ratio-based variants of the original distance values. Normality of the transformed data was assessed using the Anderson–Darling test, and the transformation with the lowest Anderson–Darling statistic was chosen. 

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
### Transformation BC
eps <- 1e-8

x <- as.vector(bray_matrix)

transforms <- list(
  log1p     = log1p(x),
  log_eps   = log(x + eps),
  sqrt      = sqrt(x),
  sqrt1p    = sqrt(x + 1),
  ratio_1px = x / (1 + x),
  ratio_max = x / max(x),
  arcsin_s  = asin(sqrt(x / max(x))),
  arcsin_p  = asin(sqrt(pmin(pmax(x, 0), 1)))
)

res <- data.frame(
  transformation = names(transforms),
  A = sapply(transforms, function(v) ad.test(v)$statistic),
  p = sapply(transforms, function(v) ad.test(v)$p.value)
)

res <- res[order(res$A), ]
res
best_transform <- res$transformation[1]
best_transform


bray_arcsin_s <- asin(sqrt(bray_matrix))
 
# get vector for histogram
bray_vec <- as.vector(bray_arcsin_s)
# Plot Distribution of Transformed Bray-Curtis Distances
histogram_distances <- ggplot(data = data.frame(bray_vec), aes(x = bray_vec)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of Transformed Bray-Curtis Distances",
    x = "(arcsine-transformed) Bray-Curtis Distance",
    y = "Count"
  )

print(histogram_distances)



```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
### Linear Mixed Model: Time between samples
Reference: Longitudinal profiling of the microbiome at four body sites reveals core stability and individualized dynamics during health and disease (Cell Host Microbe - April 2025) <https://pmc.ncbi.nlm.nih.gov/articles/PMC11022754/>

Using subset of the BC matrix that includes only pairwise distances between samples from the same dog and Sample_Site (control or surgical) because LMM requires BC as a vector not a full matrix so by pre-blocking the comparisons we can fit the model to the data specific to our question. To account for pairwise comparisons from different days (collection_day) we use diff_days which is the number of days in-between two samples. 

bray_same_dog_site <- as.data.frame(bray_arcsin_s) %>%
  tibble::rownames_to_column("SampleID_1") %>%
  tidyr::pivot_longer(-SampleID_1, names_to = "SampleID_2", values_to = "Bray_Distance") %>%
  dplyr::filter(!is.na(Bray_Distance))

# remove duplicate comparisons
bray_long <- bray_same_dog_site %>%
  filter(SampleID_1 < SampleID_2)  

paired_distances <- bray_long %>%
  left_join(
    filtered_sample_data %>%
      dplyr::select(Dog_ID, SampleID, Sample_Plate, Collection_Day, Sample_Site),
    by = c("SampleID_1" = "SampleID")
  ) %>%
  dplyr::rename_with(~ paste0(.x, "_1"), c(Collection_Day, Sample_Site)) %>%
  
  left_join(
    filtered_sample_data %>%
      dplyr::select(SampleID, Collection_Day, Sample_Site),
    by = c("SampleID_2" = "SampleID")
  ) %>%
  dplyr::rename_with(~ paste0(.x, "_2"), c(Collection_Day, Sample_Site)) %>%
  # Only looking at within-sample site comparisons
  dplyr::filter(Sample_Site_1 == Sample_Site_2) %>%
  dplyr::mutate(Sample_Site = Sample_Site_1) %>%
  dplyr::mutate(Diff_Days = abs(Collection_Day_1 - Collection_Day_2))


# Full model with Sample_Plate is significant 
lmer_bray_red <- lmer(data = paired_distances, Bray_Distance ~ Diff_Days + (1|Dog_ID))

lmer_bray_full <- lmer(data = paired_distances, Bray_Distance ~ Diff_Days + Sample_Plate + (1|Dog_ID))

anova(lmer_bray_red, lmer_bray_full, test = "LRT")

# Full model with Sample_Site is significant 
lmer_bray_red <- lmer(data = paired_distances, Bray_Distance ~ Diff_Days + Sample_Plate + (1|Dog_ID))

lmer_bray_full <- lmer(data = paired_distances, Bray_Distance ~ Diff_Days + Sample_Site + Sample_Plate + (1|Dog_ID))

anova(lmer_bray_red, lmer_bray_full, test = "LRT")
 
anova(lmer_bray_full)
summary(lmer_bray_full)
confint(lmer_bray_full)

#Notes: 
#* Diff_Days: For every day that passes between SampleID_1 and SampleID_2 at the Control site (since this is our reference), the BC dissimilarity shifts by an average of 3.625e-04 units 
#* Sample_SiteSurgical: On any given day (Diff_Days = 0), the Surgical site has a BC dissimilarity that is 0.00258 lower than the Control site 
 
```

### Linear Mixed Model

We analyzed Bray-Curtis distances between matched control and surgical samples from the same dog and timepoint. This allowed us to model how within-dog control–surgical dissimilarity changed with number of days since surgery while excluding between-dog variation from the response. Additionally lmms expect numeric vector not a matrix as a response. 

*Results:*

* Fit LMM with subsetted-BC matrix as response, Day_Order (categorical timepoint) as fixed effect, and Dog_ID as random effect.
  * Using LRT, this reduced model was more significant compared to full model where sequencing run was added as a random effect
* Contrasts between estimated marginal means of timepoints from model, using Tukey-adjusting p-values:
  * Day 0c - Post Surgery vs Week 5: estimate = 0.1737, p = 0.0287
  * Day 0a - Pre Surgery vs Day 0c - Post Surgery: estimate = -0.1749, p = 0.0308
  * Day 0c - Post Surgery vs Week 3: estimate = 0.1709, p = 0.0341
  * Admission vs Day 0c - Post Surgery: estimate = -0.1668, p = 0.0436
  
```{r,fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# generate paired-BC matrix where control and surgical samples within dogs are measured
paired_distances <- filtered_sample_data %>%
 group_by(Dog_ID, Day_Order) %>%
  summarise(
    Control_Sample = SampleID[Sample_Site == "Control"][1],
    Surgical_Sample = SampleID[Sample_Site == "Surgical"][1],
    Bray_Distance = bray_arcsin_s[Control_Sample, Surgical_Sample],
    Sample_Plate = Sample_Plate[1],
    Day_Order = Day_Order[1],   
    Collection_Day = Collection_Day[1],
   .groups = "drop"
  )
 

print("Group means by Day_Order:")
print(paired_distances %>%
  group_by(Day_Order) %>%
  summarise(
    Mean_Bray = mean(Bray_Distance),
    Median_Bray = median(Bray_Distance),
    SD_Bray = sd(Bray_Distance),
    N = n(),
    .groups = "drop"  ) %>% 
    arrange(Day_Order)
)

# Don't include sequencing run as random effect 
lmer_bray_red <- lmer(data = paired_distances, Bray_Distance ~ Day_Order + (1 | Dog_ID)  )
lmer_bray_full <- lmer(data = paired_distances, Bray_Distance ~ Day_Order + (1 | Dog_ID) + (1 | Sample_Plate))
anova(lmer_bray_red, lmer_bray_full, test = "LRT")

anova(lmer_bray_red)
summary(lmer_bray_red)
confint(lmer_bray_red)


# estimated marginal means 
emm <- emmeans(lmer_bray_red, ~ Day_Order)
contrasts <- data.frame(pairs(emm, Day_Order = "tukey"))

emm_df <- as.data.frame(emm)[, c("Day_Order", "emmean", "SE", "df", "lower.CL", "upper.CL")]

alpha <- 0.05
 

contrasts <- contrasts %>%
  mutate(
    Group1 = case_when(
      str_detect(contrast, "^\\(") ~ str_match(contrast, "^\\((.*?)\\)")[,2],
      TRUE ~ str_match(contrast, "^(.*?) - ")[,2]
    ),
    Group2 = case_when(
      str_detect(contrast, "\\) - \\(") ~ str_match(contrast, "- \\((.*?)\\)$")[,2],
      str_detect(contrast, "^\\(") ~ str_replace(contrast, "^\\(.*?\\) - ", ""),
      TRUE ~ str_match(contrast, "- (.*)$")[,2]
    ),
    Group1 = str_remove_all(str_trim(Group1), "[()]"),
    Group2 = str_remove_all(str_trim(Group2), "[()]")
  )

pairwise_tidy <- contrasts %>%
  filter(!is.na(p.value)) %>% 
  mutate(
    Significant = ifelse(p.value <= alpha, "Yes", "No"),
    Stars = case_when(
      p.value <= 0.001 ~ "***",
      p.value <= 0.01  ~ "**",
      p.value <= 0.05  ~ "*",
      p.value <= 0.10 ~ ".",  
      TRUE ~ ""
    ),
    p.value = round(p.value, 4) ) %>%
  arrange(p.value) 

# Filter out insignificant pairwise results
sig_df <- pairwise_tidy %>%
  filter(Significant == "Yes") %>%
  mutate(
    group1 = Group1,
    group2 = Group2,
    y.position = NA_real_,    
    label = Stars
  )
x_levels <- levels(factor(paired_distances$Day_Order))

sig_df <- sig_df %>%
  mutate(
    xmin = match(group1, x_levels),
    xmax = match(group2, x_levels)
  ) %>%
  arrange(p.value) %>%
  mutate(
    y.position = max(paired_distances$Bray_Distance, na.rm = TRUE) +
      seq(0.05, by = 0.05, length.out = n())
  )
print(sig_df)

 

ggplot(paired_distances, aes(x = Day_Order, y = Bray_Distance)) +
  geom_boxplot(fill = "lightblue") +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black") +
  geom_point(data = emm_df, aes(x = Day_Order, y = emmean),
             inherit.aes = FALSE, color = "darkblue", size = 3) +
  geom_line(data = emm_df, aes(x = Day_Order, y = emmean, group = 1),
            inherit.aes = FALSE, color = "darkblue", linewidth = 1) +
  stat_pvalue_manual(
    sig_df,
    label = "label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.5,
    size = 5
  ) +
  labs(
    title = "Control-Surgical Bray-Curtis Distances Across Timepoints",
    subtitle = "Trend line indicating estimated marginal means per timepoint",
    x = "Timepoint",
    y = "Bray-Curtis Distances"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  )
```


## Change-point analysis
### Change-point modeling
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

```

### Segmented regression
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

```

### Pruned Exact Linear Time (PELT) method
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
# Don't include sequencing run as random effect 
lmer_bray_red <- lmer(data = paired_distances, Bray_Distance ~ Collection_Day + (1 | Dog_ID))
lmer_bray_full <- lmer(data = paired_distances, Bray_Distance ~ Collection_Day + (1 | Dog_ID) + (1 | Sample_Plate))
anova(lmer_bray_red, lmer_bray_full, test = "LRT")

anova(lmer_bray_red)
summary(lmer_bray_red)
confint(lmer_bray_red)


# estimated marginal means 
emm <- emmeans(lmer_bray_red, ~ Collection_Day)
emm_df <- as.data.frame(emm)[, c("Collection_Day", "emmean", "SE", "df", "lower.CL", "upper.CL")]


x <- emm_df$emmean

fit_mean <- cpt.mean(x, method = "PELT", penalty = "MBIC")
fit_var  <- cpt.var(x,  method = "PELT", penalty = "MBIC")
fit_both <- cpt.meanvar(x, method = "PELT", penalty = "MBIC")

cpts(fit_mean)
cpts(fit_var)
cpts(fit_both)

plot(fit_both, main = "Changepoint on median Bray-Curtis by timepoint")
```

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```
