---
title: "Canine Skin Microbiome: 16S Differential Abundance Analysis "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

**Purpose:**
Identify specific taxa differences between post-surgery timepoint and other stages in the study. 

# Load files 
## R packages
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(knitr)
library(FSA)
library(tidyverse)
library(vegan)
library(phyloseq)
library(ggplot2)
library(ggtree)
library(scales)  
library(RColorBrewer)
library(segmented)
library(stringr)
library(pairwiseAdonis)
library(ggpubr)
library(changepoint)
library(dplyr)
library(lme4)
library(emmeans)
library(nortest)
library(mcp)
library(purrr)
library(kableExtra)
library(forcats)
library(microbiome)
library(lmerTest)       
library(broom.mixed)
library(ComplexHeatmap)
library(circlize)
```
 
## Import paired-dataset
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Paired-only ASV feature table
ps <- readRDS("../01_input_files/phyloseq/ps_2.rds")
 
sd <- as(sample_data(ps), "data.frame")

# define healing stages
sd <- sd %>%
  mutate(
    Time_Group = case_when(
      Day_Order %in% c("Admission", "Day 0a - Pre Surgery") ~ "Before_Surgery",
      Day_Order %in% c("Day 0c - Post Surgery")           ~ "Post_Surgery",
      Day_Order %in% c("Day 1", "Day 3")                  ~ "Inflammation",
      Day_Order %in% c("Week 1", "Week 2")                ~ "Proliferation",
      Day_Order %in% c("Week 3", "Week 4", "Week 5", "Week 6")  ~ "Maturation I",
      TRUE                                                 ~ NA_character_
    ),
    Time_Category = case_when(
      Day_Order %in% c("Admission", "Day 0a - Pre Surgery") ~ "Before",
      TRUE                                                  ~ "After"
    )
  )


sd$Time_Group <- factor(sd$Time_Group, levels = c("Before_Surgery", "Post_Surgery", "Inflammation", "Proliferation", "Maturation I", "Maturation II"))
  
sample_data(ps) <- sample_data(sd)

range(sample_sums(ps)) 
```

# Paired Wilcoxen Signed-Rank Test on Surgical Samples  
 
### Relative abundances of ASV + Genus
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Normalization of ASV counts 
ps_ASV_rel <- transform_sample_counts(ps, function(x) x / sum(x))
 
# generate df
ASV_rel_df <- psmelt(ps_ASV_rel) %>%
arrange(Dog_ID, Genus, Day_Order)

ASV_rel_df_surg <- ASV_rel_df%>%
   filter(Sample_Site == "Surgical")

# Unique ASVs
ntaxa(ps)
tax <- tax_table(ps)

# % ASVs removed due to unknown Genus 
no_genus <- is.na(tax[, "Genus"]) | tax[, "Genus"] == ""
n_no_genus <- sum(no_genus, na.rm = TRUE)
n_total_asv <- ntaxa(ps)
pct_no_genus <- 100 * n_no_genus / n_total_asv

n_no_genus
pct_no_genus

cat("Excluded", n_no_genus, "ASVs without genus assignment (", round(pct_no_genus, 1), "% of total ASVs).\n")


# Aggregate ASV counts for Genus-Level Analysis + remove ASVs with unknown Genus
ps_genus <- tax_glom(ps, taxrank = "Genus", NArm = TRUE)
#Unique genus
ntaxa(ps_genus)

ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))
 
# ntaxa should be smaller since we aggregated at Genus level 
ntaxa(ps_genus_rel)
 

# turn phyloseq object into a dataframe 
genus_rel_df <- psmelt(ps_genus_rel) %>%
arrange(Dog_ID, Genus, Day_Order)

# Check for genus-level (each sample should have one row for each Genus, if multiple appear then it is still at ASV-level)
check <- genus_rel_df %>%
  count(Sample, Genus) %>%
  filter(n > 1)
 check
 
genus_rel_df_surg <- genus_rel_df%>%
   filter(Sample_Site == "Surgical")
```


## Genus-level
  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE} 
run_paired_wilcoxon_by_genus <- function(
    data,
    time_group_1,
    time_group_2,
    exact = TRUE
) {
  
  required_columns <- c("Genus", time_group_1, time_group_2)
  
  missing_columns <- setdiff(required_columns, names(data))
  
  if (length(missing_columns) > 0) {
    stop(
      "The following columns are missing: ",
      paste(missing_columns, collapse = ", ")
    )
  }
  
  results <- data %>%
    group_by(Genus) %>%
    group_modify(~ {
      
      dat <- .x %>%
        filter(
          !is.na(.data[[time_group_1]]),
          !is.na(.data[[time_group_2]])
        )
      
      group_1_values <- dat[[time_group_1]]
      group_2_values <- dat[[time_group_2]]
      
      differences <- group_1_values - group_2_values
      
      n_pairs <- length(differences)
      n_zero_diff <- sum(differences == 0)
      n_nonzero_diff <- sum(differences != 0)
      
      # No signed-rank test can be performed when all differences are zero
      if (n_nonzero_diff == 0) {
        return(tibble(
          n_pairs = n_pairs,
          n_zero_diff = n_zero_diff,
          n_nonzero_diff = n_nonzero_diff,
          stat = NA_real_,
          p = 1,
          estimate = 0,
          conf_low = 0,
          conf_high = 0
        ))
      }
      
      test <- wilcox.test(
        group_1_values,
        group_2_values,
        paired = TRUE,
        exact = exact,
        conf.int = TRUE
      )
      
      tibble(
        n_pairs = n_pairs,
        n_zero_diff = n_zero_diff,
        n_nonzero_diff = n_nonzero_diff,
        stat = unname(test$statistic),
        p = test$p.value,
        estimate = unname(test$estimate),
        conf_low = test$conf.int[1],
        conf_high = test$conf.int[2]
      )
    }) %>%
    ungroup() %>%
    mutate(
      p_adj = p.adjust(p, method = "BH")
    ) %>%
    arrange(p_adj, p, Genus)
  
  return(results)
}

```

```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
run_paired_wilcoxon_by_genus <- function(
    data,
    time_group_1,
    time_group_2,
    exact = TRUE
) {
  
  required_columns <- c("Genus", time_group_1, time_group_2)
  
  missing_columns <- setdiff(required_columns, names(data))
  
  if (length(missing_columns) > 0) {
    stop(
      "The following columns are missing: ",
      paste(missing_columns, collapse = ", ")
    )
  }
  
  results <- data %>%
    group_by(Genus) %>%
    group_modify(~ {
      
      dat <- .x %>%
        filter(
          !is.na(.data[[time_group_1]]),
          !is.na(.data[[time_group_2]])
        )
      
      group_1_values <- dat[[time_group_1]]
      group_2_values <- dat[[time_group_2]]
      
      differences <- group_1_values - group_2_values
      
      n_pairs <- length(differences)
      n_zero_diff <- sum(differences == 0)
      n_nonzero_diff <- sum(differences != 0)
      
      test <- wilcox.test(
        group_1_values,
        group_2_values,
        paired = TRUE,
        exact = exact,
        conf.int = TRUE
      )
      
      tibble(
        n_pairs = n_pairs,
        n_zero_diff = n_zero_diff,
        n_nonzero_diff = n_nonzero_diff,
        stat = unname(test$statistic),
        p = test$p.value,
        estimate = unname(test$estimate),
        conf_low = test$conf.int[1],
        conf_high = test$conf.int[2]
      )
    }) %>%
    ungroup() %>%
    mutate(
      p_adj = p.adjust(p, method = "BH")
    ) %>%
    arrange(p_adj, p, Genus)
  
  return(results)
}

```

### Before_Surgery (Admission + Pre-Surgery) vs Post-Surgery 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify genera present in at least 10% of samples
taxa_present_10pct <- genus_rel_df_surg_2 %>%
  group_by(Genus) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.10) %>%
  dplyr::select(Genus)

# Number of retained genera
filtered_genus_count <- n_distinct(taxa_present_10pct$Genus)

cat(
  "Total genera present in at least 10% of samples:",
  filtered_genus_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "Genus")


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = median(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2 
)

tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_sig <- results_wt_annotated %>%
  filter(results_wt_annotated$p_adj <= 0.1)

results_wt_annotated_sig

nrow(results_wt_annotated)

significant_genera_1 <- unique(results_wt_annotated_sig$Genus) 
```
 
### Post-Surgery vs Stage: Inflammation (Days 1 + 3) 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Post_Surgery"
Time_Group_2 <- "Inflammation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify genera present in at least 10% of samples
taxa_present_10pct <- genus_rel_df_surg_2 %>%
  group_by(Genus) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.10) %>%
  dplyr::select(Genus)

# Number of retained genera
filtered_genus_count <- n_distinct(taxa_present_10pct$Genus)

cat(
  "Total genera present in at least 10% of samples:",
  filtered_genus_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "Genus")


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = median(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2 
)

tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_sig <- results_wt_annotated %>%
  filter(results_wt_annotated$p_adj <= 0.1)

results_wt_annotated_sig

nrow(results_wt_annotated)

significant_genera_2 <- unique(results_wt_annotated_sig$Genus) 
```
 
### Post-Surgery vs Stage: Proliferation (Weeks 1 + 2) 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Post_Surgery"
Time_Group_2 <- "Proliferation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify genera present in at least 10% of samples
taxa_present_10pct <- genus_rel_df_surg_2 %>%
  group_by(Genus) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.10) %>%
  dplyr::select(Genus)

# Number of retained genera
filtered_genus_count <- n_distinct(taxa_present_10pct$Genus)

cat(
  "Total genera present in at least 10% of samples:",
  filtered_genus_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "Genus")


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = median(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2 
)

tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_sig <- results_wt_annotated %>%
  filter(results_wt_annotated$p_adj <= 0.1)

results_wt_annotated_sig

nrow(results_wt_annotated)

significant_genera_3 <- unique(results_wt_annotated_sig$Genus) 
```
 
### Post-Surgery vs Stage: Maturation I (Weeks 3 + 4) 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Post_Surgery"
Time_Group_2 <- "Maturation I"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify genera present in at least 10% of samples
taxa_present_10pct <- genus_rel_df_surg_2 %>%
  group_by(Genus) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.10) %>%
  dplyr::select(Genus)

# Number of retained genera
filtered_genus_count <- n_distinct(taxa_present_10pct$Genus)

cat(
  "Total genera present in at least 10% of samples:",
  filtered_genus_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "Genus")


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = median(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2 
)

tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_sig <- results_wt_annotated %>%
  filter(results_wt_annotated$p_adj <= 0.1)

results_wt_annotated_sig

nrow(results_wt_annotated)

significant_genera_4 <- unique(results_wt_annotated_sig$Genus) 
```
 
### Post-Surgery vs Stage: Maturation II (Weeks 5 + 6) 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Post_Surgery"
Time_Group_2 <- "Maturation II"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
genus_rel_df_surg_2 <- genus_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify genera present in at least 10% of samples
taxa_present_10pct <- genus_rel_df_surg_2 %>%
  group_by(Genus) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.10) %>%
  dplyr::select(Genus)

# Number of retained genera
filtered_genus_count <- n_distinct(taxa_present_10pct$Genus)

cat(
  "Total genera present in at least 10% of samples:",
  filtered_genus_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
genus_rel_df_surg_2_filt<- genus_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "Genus")


# Summarise abundance per dog, genus, and time group
genus_rel_df_median <- genus_rel_df_surg_2_filt %>%
  group_by(Dog_ID, Genus, Time_Group) %>%
  summarise(
    median_abun = median(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
genus_rel_df_median_formatted <- genus_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = median_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_genus(
  data = genus_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2 
)

tax_df <- as.data.frame(tax_table(ps_genus))
 

results_wt_annotated <- results_wt %>%
  left_join(tax_df, by = "Genus")

results_wt_annotated_sig <- results_wt_annotated %>%
  filter(results_wt_annotated$p_adj <= 0.1)

results_wt_annotated_sig

nrow(results_wt_annotated)

significant_genera_5 <- unique(results_wt_annotated_sig$Genus) 
```

Get significant genus list 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
 significant_genera_list <- unique(c(significant_genera_1, significant_genera_2, significant_genera_3, significant_genera_4, significant_genera_5))
 
length(significant_genera_list)
```
 
### Sig. Genus Plot: Surgical vs. Control

Surgical Samples
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 

# turn phyloseq object into a dataframe 
ps_genus_df <- psmelt(ps_genus) %>%
arrange(Dog_ID, Genus, Day_Order)
 
ps_genus_df_surg <- ps_genus_df %>%
  filter(Sample_Site == "Surgical")   


# Plot only the top genuses
ps_genus_df_surg_filt <- ps_genus_df_surg %>%
  group_by(Genus, Day_Order) %>%
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Day_Order) %>%
  mutate(
    relative_abundance = total_abundance / sum(total_abundance)
  ) %>%
  ungroup()%>%
  filter(Genus %in% significant_genera_list)  

range(ps_genus_df_surg_filt$relative_abundance)
 
n_taxa <- length(significant_genera_list)
n_taxa
base_cols <- brewer.pal(12, "Paired")
taxa_cols_vec <- colorRampPalette(base_cols)(n_taxa)

idx <- c(rbind(1:ceiling(n_taxa/2), n_taxa:ceiling(n_taxa/2 + 1)))
idx <- idx[!is.na(idx)]

taxa_cols_vec_reordered <- taxa_cols_vec[idx]

before_order <- ps_genus_df_surg_filt %>%
  filter(Day_Order == "Admission") %>%
  group_by(Genus) %>%
  summarise(mean_before = mean(relative_abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_before)) %>%
  pull(Genus)

ps_genus_df_surg_filt <- ps_genus_df_surg_filt %>%
  mutate(Genus = factor(Genus, levels = before_order))

surgical_taxa_colors <- setNames(taxa_cols_vec, before_order)
 
plot <- ps_genus_df_surg_filt %>% 
  ggplot(aes(x = Day_Order, y = relative_abundance, fill = Genus)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = surgical_taxa_colors, drop = FALSE) +
  scale_y_continuous(
    limits = c(0, 0.2),
    breaks = seq(0, 0.2, by = 0.01)
  ) +
  labs(
    title = "Genus Abundance Across Surgical Groups",
    x = "Collection day",
    y = "Relative abundance",
    fill = "Genus"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot
```
 
Control Samples
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 

# Use counts to calculate per Day relative abundance 

ps_genus_df_ctrl <- ps_genus_df %>%
  filter(Sample_Site == "Control")   


# Plot only the top genuses
ps_genus_df_ctrl_filt <- ps_genus_df_ctrl %>%
  group_by(Genus, Day_Order) %>%
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Day_Order) %>%
  mutate(
    relative_abundance = total_abundance / sum(total_abundance)
  ) %>%
  ungroup()%>%
  filter(Genus %in% significant_genera_list)  

range(ps_genus_df_ctrl_filt$relative_abundance)
 
n_taxa <- length(significant_genera_list)

base_cols <- brewer.pal(12, "Paired")
taxa_cols_vec <- colorRampPalette(base_cols)(n_taxa)

idx <- c(rbind(1:ceiling(n_taxa/2), n_taxa:ceiling(n_taxa/2 + 1)))
idx <- idx[!is.na(idx)]

taxa_cols_vec_reordered <- taxa_cols_vec[idx]

before_order <- ps_genus_df_ctrl_filt %>%
  filter(Day_Order == "Admission") %>%
  group_by(Genus) %>%
  summarise(mean_before = mean(relative_abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_before)) %>%
  pull(Genus)

ps_genus_df_ctrl_filt <- ps_genus_df_ctrl_filt %>%
  mutate(Genus = factor(Genus, levels = before_order))

surgical_taxa_colors <- setNames(taxa_cols_vec, before_order)
 
plot <- ps_genus_df_ctrl_filt %>% 
  ggplot(aes(x = Day_Order, y = relative_abundance, fill = Genus)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = surgical_taxa_colors, drop = FALSE) +
  scale_y_continuous(
    limits = c(0, 0.2),
    breaks = seq(0, 0.2, by = 0.01)
  ) +
  labs(
    title = "Genus Abundance Across Control Groups",
    x = "Collection day",
    y = "Relative abundance",
    fill = "Genus"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot
```
 
 
 
## ASV-level
 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
run_paired_wilcoxon_by_ASV <- function(
    data,
    time_group_1,
    time_group_2,
    exact = TRUE
) {
  
  required_columns <- c("OTU", time_group_1, time_group_2)
  
  missing_columns <- setdiff(required_columns, names(data))
  
  if (length(missing_columns) > 0) {
    stop(
      "The following columns are missing: ",
      paste(missing_columns, collapse = ", ")
    )
  }
  
  results <- data %>%
    group_by(OTU) %>%
    group_modify(~ {
      
      dat <- .x %>%
        filter(
          !is.na(.data[[time_group_1]]),
          !is.na(.data[[time_group_2]])
        )
      
      group_1_values <- dat[[time_group_1]]
      group_2_values <- dat[[time_group_2]]
      
      differences <- group_1_values - group_2_values
      
      n_pairs <- length(differences)
      n_zero_diff <- sum(differences == 0)
      n_nonzero_diff <- sum(differences != 0)
 
      
      test <- wilcox.test(
        group_1_values,
        group_2_values,
        paired = TRUE,
        exact = exact,
        conf.int = TRUE
      )
      
      tibble(
        n_pairs = n_pairs,
        n_zero_diff = n_zero_diff,
        n_nonzero_diff = n_nonzero_diff,
        stat = unname(test$statistic),
        p = test$p.value,
        estimate = unname(test$estimate),
        conf_low = test$conf.int[1],
        conf_high = test$conf.int[2]
      )
    }) %>%
    ungroup() %>%
    mutate(
      p_adj = p.adjust(p, method = "BH")
    ) %>%
    arrange(p_adj, p, OTU)
  
  return(results)
}

```

### Before_Surgery (Admission + Pre-Surgery) vs Post-Surgery 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Before_Surgery"
Time_Group_2 <- "Post_Surgery"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
ASV_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify ASVs present in at least 10% of samples
taxa_present_10pct <- ASV_rel_df_surg_2 %>%
  group_by(OTU) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.10) %>%
  dplyr::select(OTU)

# Number of retained genera
filtered_ASV_count <- n_distinct(taxa_present_10pct$OTU)

cat(
  "Total ASV present in at least 10% of samples:",
  filtered_ASV_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
ASV_rel_df_surg_2_filt <- ASV_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "OTU")
 

# Summarise abundance per dog, genus, and time group
ASV_rel_df_median <- ASV_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_rel_abun = median(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
ASV_rel_df_median_formatted <- ASV_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_rel_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = ASV_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2 
)
 

tax_df <- tax_table(ps) %>%
  as("matrix") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>%
  tibble::as_tibble()

results_wt_annotated <- results_wt %>%
  mutate(OTU = as.character(OTU)) %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_sig <- results_wt_annotated %>%
  filter(results_wt_annotated$p_adj <= 0.1)

results_wt_annotated_sig

nrow(results_wt_annotated)

significant_genera_1 <- unique(results_wt_annotated_sig$OTU) 
```
 
### Post-Surgery vs Stage: Inflammation (Days 1 + 3) 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Post_Surgery"
Time_Group_2 <- "Inflammation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
ASV_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify ASVs present in at least 10% of samples
taxa_present_10pct <- ASV_rel_df_surg_2 %>%
  group_by(OTU) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.10) %>%
  dplyr::select(OTU)

# Number of retained genera
filtered_ASV_count <- n_distinct(taxa_present_10pct$OTU)

cat(
  "Total ASV present in at least 10% of samples:",
  filtered_ASV_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
ASV_rel_df_surg_2_filt <- ASV_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "OTU")
 

# Summarise abundance per dog, genus, and time group
ASV_rel_df_median <- ASV_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_rel_abun = median(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
ASV_rel_df_median_formatted <- ASV_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_rel_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = ASV_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2 
)
 

tax_df <- tax_table(ps) %>%
  as("matrix") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>%
  tibble::as_tibble()

results_wt_annotated <- results_wt %>%
  mutate(OTU = as.character(OTU)) %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_sig <- results_wt_annotated %>%
  filter(results_wt_annotated$p_adj <= 0.1)

results_wt_annotated_sig

nrow(results_wt_annotated)

significant_genera_2 <- unique(results_wt_annotated_sig$OTU) 
```
 
### Post-Surgery vs Stage: Proliferation (Weeks 1 + 2) 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Post_Surgery"
Time_Group_2 <- "Proliferation"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
ASV_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify ASVs present in at least 10% of samples
taxa_present_10pct <- ASV_rel_df_surg_2 %>%
  group_by(OTU) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.10) %>%
  dplyr::select(OTU)

# Number of retained genera
filtered_ASV_count <- n_distinct(taxa_present_10pct$OTU)

cat(
  "Total ASV present in at least 10% of samples:",
  filtered_ASV_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
ASV_rel_df_surg_2_filt <- ASV_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "OTU")
 

# Summarise abundance per dog, genus, and time group
ASV_rel_df_median <- ASV_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_rel_abun = median(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
ASV_rel_df_median_formatted <- ASV_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_rel_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = ASV_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2 
)
 

tax_df <- tax_table(ps) %>%
  as("matrix") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>%
  tibble::as_tibble()

results_wt_annotated <- results_wt %>%
  mutate(OTU = as.character(OTU)) %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_sig <- results_wt_annotated %>%
  filter(results_wt_annotated$p_adj <= 0.1)

results_wt_annotated_sig

nrow(results_wt_annotated)

significant_genera_3 <- unique(results_wt_annotated_sig$OTU) 
```
 
### Post-Surgery vs Stage: Maturation I (Weeks 3 + 4) 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Post_Surgery"
Time_Group_2 <- "Maturation I"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
ASV_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify ASVs present in at least 10% of samples
taxa_present_10pct <- ASV_rel_df_surg_2 %>%
  group_by(OTU) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.10) %>%
  dplyr::select(OTU)

# Number of retained genera
filtered_ASV_count <- n_distinct(taxa_present_10pct$OTU)

cat(
  "Total ASV present in at least 10% of samples:",
  filtered_ASV_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
ASV_rel_df_surg_2_filt <- ASV_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "OTU")
 

# Summarise abundance per dog, genus, and time group
ASV_rel_df_median <- ASV_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_rel_abun = median(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
ASV_rel_df_median_formatted <- ASV_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_rel_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = ASV_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2 
)
 

tax_df <- tax_table(ps) %>%
  as("matrix") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>%
  tibble::as_tibble()

results_wt_annotated <- results_wt %>%
  mutate(OTU = as.character(OTU)) %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_sig <- results_wt_annotated %>%
  filter(results_wt_annotated$p_adj <= 0.1)

results_wt_annotated_sig

nrow(results_wt_annotated)

significant_genera_4 <- unique(results_wt_annotated_sig$OTU) 
```
 
### Post-Surgery vs Stage: Maturation II (Weeks 5 + 6) 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
# Define the two paired time groups
Time_Group_1 <- "Post_Surgery"
Time_Group_2 <- "Maturation II"

time_groups <- c(Time_Group_1, Time_Group_2)

# Keep only the two surgery time groups
ASV_rel_df_surg_2 <- ASV_rel_df_surg %>%
  filter(Time_Group %in% time_groups)

# Identify ASVs present in at least 10% of samples
taxa_present_10pct <- ASV_rel_df_surg_2 %>%
  group_by(OTU) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.10) %>%
  dplyr::select(OTU)

# Number of retained genera
filtered_ASV_count <- n_distinct(taxa_present_10pct$OTU)

cat(
  "Total ASV present in at least 10% of samples:",
  filtered_ASV_count,
  "\n"
)

# Keep genera satisfying the prevalence criterion
ASV_rel_df_surg_2_filt <- ASV_rel_df_surg_2 %>%
  semi_join(taxa_present_10pct, by = "OTU")
 

# Summarise abundance per dog, genus, and time group
ASV_rel_df_median <- ASV_rel_df_surg_2_filt %>%
  group_by(Dog_ID, OTU, Time_Group) %>%
  summarise(
    median_rel_abun = median(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Reformat so each time group has its own column
ASV_rel_df_median_formatted <- ASV_rel_df_median %>%
  pivot_wider(
    id_cols = c(Dog_ID, OTU),
    names_from = Time_Group,
    values_from = median_rel_abun,
    values_fill = 0
  )

# Conduct paired Wilcoxon signed-rank tests by genus
results_wt <- run_paired_wilcoxon_by_ASV(
  data = ASV_rel_df_median_formatted,
  time_group_1 = Time_Group_1,
  time_group_2 = Time_Group_2 
)
 

tax_df <- tax_table(ps) %>%
  as("matrix") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>%
  tibble::as_tibble()

results_wt_annotated <- results_wt %>%
  mutate(OTU = as.character(OTU)) %>%
  left_join(tax_df, by = "OTU")

results_wt_annotated_sig <- results_wt_annotated %>%
  filter(results_wt_annotated$p_adj <= 0.1)

results_wt_annotated_sig

nrow(results_wt_annotated)

significant_genera_5 <- unique(results_wt_annotated_sig$OTU) 
```
 
 
# Taxon-specific linear mixed model

### CLR transformation
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Aggregate ASV counts for Genus-Level Analysis + remove ASVs with unknown Genus
ps_genus <- tax_glom(ps, taxrank = "Genus", NArm = TRUE)
#Unique genus
ntaxa(ps_genus)
ps_genus_df <- psmelt(ps_genus) %>%
arrange(Dog_ID, Genus, Day_Order)
range(sample_sums(ps_genus)) 

 
ps_genus_pseudo <- transform_sample_counts(ps_genus, function(x) x + 1)

# Convert to relative abundances (TSS) then CLR
ps_genus_clr <- microbiome::transform(ps_genus_pseudo, "clr")
 
 
# ntaxa should be smaller since we aggregated at Genus level 
ntaxa(ps_genus_clr)
 

# turn phyloseq object into a dataframe 
genus_clr_df <- psmelt(ps_genus_clr) %>%
arrange(Dog_ID, Genus, Day_Order)

# Check for genus-level (each sample should have one row for each Genus, if multiple appear then it is still at ASV-level)
check <- genus_clr_df %>%
  count(Sample, Genus) %>%
  filter(n > 1)
 check
 
genus_clr_df_surg <- genus_clr_df%>%
   filter(Sample_Site == "Surgical")

genus_clr_df_ctrl <- genus_clr_df%>%
   filter(Sample_Site == "Control")
```

```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE} 
 # other method: get genus in at least 10% of all samples 
taxa_present_10pct <- genus_rel_df_surg %>%
  group_by(Genus) %>%
  summarise(
    prevalence = n_distinct(Sample[Abundance > 0]) /
                 n_distinct(Sample),
    .groups = "drop"
  ) %>%
  filter(prevalence >= 0.10) %>%
  pull(Genus)
```

## Surgical Samples
```{r, fig.width=8, fig.height=11, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
 # model the genus identified in Wilcoxen test as significant
species_df_sub <- genus_clr_df_surg %>%
  filter(Genus %in% significant_genera_list)
  
species_df_sub <- species_df_sub %>%
  mutate(
    Time_Group = relevel(Time_Group, ref = "Before_Surgery")   
  )
 

genus_models <- species_df_sub %>%
  group_by(Genus) %>%
  group_nest() %>%
  mutate(
    model = map(
      data,
      ~ lmerTest::lmer(Abundance ~ Time_Group + (1 | Dog_ID), data = .x)
    ),
    sum_tbl = map(model, ~ summary(.x))
  )
 

# Extract coefficient tables with p-values AND the fixed effect (Time_Group)
results <- genus_models %>%
  mutate(
    coefs = map(
      model,
      ~ as.data.frame(summary(.x)$coefficients) %>%
        tibble::rownames_to_column("Term")
    )
  ) %>%
  dplyr::select(Genus, coefs) %>%
  unnest(coefs) %>%
  filter(grepl("^Time_Group", Term)) %>%
  mutate(
    Time_Group = sub("^Time_Group", "", Term)
  )

# Rename columns if needed (lmerTest gives: Estimate, Std. Error, df, t value, Pr(>|t|))
results <- results %>%
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    df = df,
    statistic = `t value`,
    p.value = `Pr(>|t|)`
  )%>%
  # BH correction
  mutate(
    p.adj = p.adjust(p.value, method = "BH")
  ) 

 results_sig <- results %>%
   filter(p.adj <= 0.1)
 
 results_sig
 
 results <- results %>%
  mutate(
    p_value = pmax(p.value, .Machine$double.xmin),
    signed_log10_p = sign(estimate) * -log10(p.value)
  )
 
 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation I",
  "Maturation II"
)

results <- results %>%
  mutate(
    Time_Group = factor(
      Time_Group,
      levels = timepoint_order
    )
  )

 score_matrix <- results %>%
  dplyr::select(Genus, Time_Group, signed_log10_p) %>%
  pivot_wider(
    names_from = Time_Group,
    values_from = signed_log10_p
  ) %>%
  column_to_rownames("Genus") %>%
  as.matrix()
 
 
 timepoint_pairs <- combn(
  colnames(score_matrix),
  2,
  simplify = FALSE
)

cor_results <- map_dfr(
  timepoint_pairs,
  function(pair) {
    keep <- complete.cases(score_matrix[, pair])

    if (sum(keep) < 3) {
      return(tibble(
        Timepoint_1 = pair[1],
        Timepoint_2 = pair[2],
        N_genera = sum(keep),
        rho = NA_real_,
        p_value = NA_real_
      ))
    }

    test <- cor.test(
      score_matrix[keep, pair[1]],
      score_matrix[keep, pair[2]],
      method = "spearman",
      exact = FALSE
    )

    tibble(
      Timepoint_1 = pair[1],
      Timepoint_2 = pair[2],
      N_genera = sum(keep),
      rho = unname(test$estimate),
      p_value = test$p.value
    )
  }
) %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    FDR_significant = p_adj < 0.1
  )

cor_results

heatmap_matrix <- score_matrix[
  complete.cases(score_matrix),
  ,
  drop = FALSE
]
 

signed_max <- max(abs(heatmap_matrix), na.rm = TRUE)

col_fun <- circlize::colorRamp2(
  c(-signed_max, 0, signed_max),
  c("#2166AC", "white", "#B2182B")
)
 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation I",
  "Maturation II"
)

heatmap_matrix_ordered <- heatmap_matrix[
  ,
  timepoint_order[timepoint_order %in% colnames(heatmap_matrix)],
  drop = FALSE
]

ht <- ComplexHeatmap::Heatmap(
  heatmap_matrix_ordered,
  name = "Signed\n-log10(p)",
  col = col_fun,

  cluster_rows = TRUE,
  cluster_columns = FALSE,
  column_order = seq_len(ncol(heatmap_matrix_ordered)),

  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",

  na_col = "grey90",
  column_title = "Surgical Sites",
  column_title_gp = grid::gpar(fontsize = 12),
  row_title = "Genus",
  row_title_gp = grid::gpar(fontsize = 10),
  row_names_side = "right",
  row_names_gp = grid::gpar(fontsize = 7)
)

plot(ht)
```

## Control Samples
```{r, fig.width=8, fig.height=11, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE} 
 # model the genus identified in Wilcoxen test as significant
species_df_sub <- genus_clr_df_ctrl %>%
  filter(Genus %in% significant_genera_list)
  
species_df_sub <- species_df_sub %>%
  mutate(
    Time_Group = relevel(Time_Group, ref = "Before_Surgery")   
  )
 

genus_models <- species_df_sub %>%
  group_by(Genus) %>%
  group_nest() %>%
  mutate(
    model = map(
      data,
      ~ lmerTest::lmer(Abundance ~ Time_Group + (1 | Dog_ID), data = .x)
    ),
    sum_tbl = map(model, ~ summary(.x))
  )
 

# Extract coefficient tables with p-values AND the fixed effect (Time_Group)
results <- genus_models %>%
  mutate(
    coefs = map(
      model,
      ~ as.data.frame(summary(.x)$coefficients) %>%
        tibble::rownames_to_column("Term")
    )
  ) %>%
  dplyr::select(Genus, coefs) %>%
  unnest(coefs) %>%
  filter(grepl("^Time_Group", Term)) %>%
  mutate(
    Time_Group = sub("^Time_Group", "", Term)
  )

# Rename columns if needed (lmerTest gives: Estimate, Std. Error, df, t value, Pr(>|t|))
results <- results %>%
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    df = df,
    statistic = `t value`,
    p.value = `Pr(>|t|)`
  )%>%
  # BH correction
  mutate(
    p.adj = p.adjust(p.value, method = "BH")
  ) 

 results_sig <- results %>%
   filter(p.adj <= 0.1)
 
 results_sig
 
 results <- results %>%
  mutate(
    p_value = pmax(p.value, .Machine$double.xmin),
    signed_log10_p = sign(estimate) * -log10(p.value)
  )
 
 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation I",
  "Maturation II"
)

results <- results %>%
  mutate(
    Time_Group = factor(
      Time_Group,
      levels = timepoint_order
    )
  )

 score_matrix <- results %>%
  dplyr::select(Genus, Time_Group, signed_log10_p) %>%
  pivot_wider(
    names_from = Time_Group,
    values_from = signed_log10_p
  ) %>%
  column_to_rownames("Genus") %>%
  as.matrix()
 
 
 timepoint_pairs <- combn(
  colnames(score_matrix),
  2,
  simplify = FALSE
)

cor_results <- map_dfr(
  timepoint_pairs,
  function(pair) {
    keep <- complete.cases(score_matrix[, pair])

    if (sum(keep) < 3) {
      return(tibble(
        Timepoint_1 = pair[1],
        Timepoint_2 = pair[2],
        N_genera = sum(keep),
        rho = NA_real_,
        p_value = NA_real_
      ))
    }

    test <- cor.test(
      score_matrix[keep, pair[1]],
      score_matrix[keep, pair[2]],
      method = "spearman",
      exact = FALSE
    )

    tibble(
      Timepoint_1 = pair[1],
      Timepoint_2 = pair[2],
      N_genera = sum(keep),
      rho = unname(test$estimate),
      p_value = test$p.value
    )
  }
) %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    FDR_significant = p_adj < 0.1
  )

cor_results

heatmap_matrix <- score_matrix[
  complete.cases(score_matrix),
  ,
  drop = FALSE
]
 

signed_max <- max(abs(heatmap_matrix), na.rm = TRUE)

col_fun <- circlize::colorRamp2(
  c(-signed_max, 0, signed_max),
  c("#2166AC", "white", "#B2182B")
)
 timepoint_order <- c(
  "Post_Surgery",
  "Inflammation",
  "Proliferation",
  "Maturation I",
  "Maturation II"
)

heatmap_matrix_ordered <- heatmap_matrix[
  ,
  timepoint_order[timepoint_order %in% colnames(heatmap_matrix)],
  drop = FALSE
]

ht <- ComplexHeatmap::Heatmap(
  heatmap_matrix_ordered,
  name = "Signed\n-log10(p)",
  col = col_fun,

  cluster_rows = TRUE,
  cluster_columns = FALSE,
  column_order = seq_len(ncol(heatmap_matrix_ordered)),

  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",

  na_col = "grey90",
  column_title = "Control Sites",
  column_title_gp = grid::gpar(fontsize = 12),
  row_title = "Genus",
  row_title_gp = grid::gpar(fontsize = 10),
  row_names_side = "right",
  row_names_gp = grid::gpar(fontsize = 7)
)

plot(ht)
```

  
 
```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
sessionInfo()
```
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
  
 
 
 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE} 

# Differential Abundance
# get mean relative abundance for each phylum
phy_phyla_df <- psmelt(ps_rel_2) %>%
  filter(Sample_Location == "Stifle") %>%
  filter(!is.na(Phylum)) %>%
  group_by(Phylum) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rel_abundance))

top_phyla <- phy_phyla_df$Phylum[1:min(10, nrow(phy_phyla_df))]

# Recode non-top phyla as Other
phy_phyla_df <- ps_rel_2 %>%
  subset_samples(Surgery_Type == "Stifle") %>%
  psmelt() %>%
  mutate(
    Phylum = if_else(Phylum %in% top_phyla, as.character(Phylum), "Other"),
    Sample_Site = factor(Sample_Site, levels = c("Control", "Surgical")),
    Day_Order = factor(Day_Order, levels = sort(unique(Day_Order)))
  ) %>%
  group_by(Sample_Site, Day_Order, Phylum) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")%>%
  arrange(desc(mean_rel_abundance))
 

# Colors for top 10 + Other
all_phyla <- c(top_phyla, "Other")
n_phyla <- length(all_phyla)
base_cols <- brewer.pal(min(n_phyla, 12), "Paired")
phylum_cols_vec <- colorRampPalette(base_cols)(n_phyla)
phylum_colors <- setNames(phylum_cols_vec, all_phyla)

plot_site <- function(site_name) {
  ggplot(
    phy_phyla_df %>% filter(Sample_Site == site_name),
    aes(x = Day_Order, y = mean_rel_abundance, fill = factor(Phylum, levels = all_phyla))
  ) +
    geom_col(position = "stack") +
    scale_fill_manual(values = phylum_colors, drop = FALSE) +
    labs(
      title = paste("Sample-Site:", site_name),
      x = "Timepoint",
      y = "Average Relative abundance",
      fill = "Phylum"
    ) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
 

p_control  <- plot_site("Control")
p_surgical <- plot_site("Surgical")
  
print(p_control)
print(p_surgical)

 

# get mean relative abundance for each family
family_df <- psmelt(ps_rel_2) %>%
  filter(Sample_Location == "Stifle") %>%
  filter(!is.na(Family)) %>%
  group_by(Phylum, Family) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rel_abundance)) 

top_10_family <- family_df$Family[1:min(10, nrow(family_df))]

# Recode non-top genus as Other
family_df <- ps_rel_2 %>%
  subset_samples(Surgery_Type == "Stifle") %>%
  psmelt() %>%
  mutate(
    Family = if_else(Family %in% top_10_family, as.character(Family), "Other"),
    Sample_Site = factor(Sample_Site, levels = c("Control", "Surgical")),
    Day_Order = factor(Day_Order, levels = sort(unique(Day_Order)))
  ) %>%
  group_by(Dog_ID, Sample_Site, Day_Order, Family) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")%>%
  arrange(desc(mean_rel_abundance))
 

# Colors for top 10 + Other
all_family <- c(top_10_family, "Other")
n_family <- length(all_family)
base_cols <- brewer.pal(min(n_family, 12), "Paired")
cols_vec <- colorRampPalette(base_cols)(n_family)
colors <- setNames(cols_vec, all_family)

plot_site <- function(site_name) {
  ggplot(
    family_df %>% filter(Sample_Site == site_name),
    aes(x = Day_Order, y = mean_rel_abundance, fill = factor(Family, levels = all_family))
  ) +
    geom_col(position = "stack") +
    scale_fill_manual(values = colors, drop = FALSE) +
    labs(
      title = paste("Sample-Site:", site_name),
      x = "Timepoint",
      y = "Average Relative abundance",
      fill = "Family"
    ) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
 

p_control  <- plot_site("Control")
p_surgical <- plot_site("Surgical")
 
p_control
p_surgical


# get mean relative abundance for each genus
phy_genus_df <- psmelt(ps_rel_2) %>%
  filter(Sample_Location == "Stifle") %>%
  filter(!is.na(Genus)) %>%
  group_by(Phylum, Genus) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rel_abundance)) 

top_genus <- phy_genus_df$Genus[1:min(10, nrow(phy_genus_df))]

# Recode non-top genus as Other
phy_genus_df <- ps_rel_2 %>%
  subset_samples(Surgery_Type == "Stifle") %>%
  psmelt() %>%
  mutate(
    Genus = if_else(Genus %in% top_genus, as.character(Genus), "Other"),
    Sample_Site = factor(Sample_Site, levels = c("Control", "Surgical")),
    Day_Order = factor(Day_Order, levels = sort(unique(Day_Order)))
  ) %>%
  group_by(Sample_Site, Day_Order, Genus) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")%>%
  arrange(desc(mean_rel_abundance))
 

# Colors for top 10 + Other
all_genus <- c(top_genus, "Other")
n_genus <- length(all_genus)
base_cols <- brewer.pal(min(n_genus, 12), "Paired")
genus_cols_vec <- colorRampPalette(base_cols)(n_genus)
genus_colors <- setNames(genus_cols_vec, all_genus)

plot_site <- function(site_name) {
  ggplot(
    phy_genus_df %>% filter(Sample_Site == site_name),
    aes(x = Day_Order, y = mean_rel_abundance, fill = factor(Genus, levels = all_genus))
  ) +
    geom_col(position = "stack") +
    scale_fill_manual(values = genus_colors, drop = FALSE) +
    labs(
      title = paste("Sample-Site:", site_name),
      x = "Timepoint",
      y = "Average Relative abundance",
      fill = "Genus"
    ) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
 

p_control  <- plot_site("Control")
p_surgical <- plot_site("Surgical")
 
p_control
p_surgical



species_df <- psmelt(ps_rel_2) %>%
  filter(Sample_Location == "Stifle") %>%
  filter(!is.na(Species)) %>%
  group_by(Species) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rel_abundance))

top_species <- species_df$Species[1:min(10, nrow(species_df))]

# Recode non-top phyla as Other
species_df <- ps_rel_2 %>%
  subset_samples(Surgery_Type == "Stifle") %>%
  psmelt() %>%
  mutate(
    Species = if_else(Species %in% top_phyla, as.character(Species), "Other"),
    Sample_Site = factor(Sample_Site, levels = c("Control", "Surgical")),
    Day_Order = factor(Day_Order, levels = sort(unique(Day_Order)))
  ) %>%
  group_by(Sample_Site, Day_Order, Species) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")%>%
  arrange(desc(mean_rel_abundance))
 

# Colors for top 10 + Other
all_phyla <- c(top_phyla, "Other")
n_phyla <- length(all_phyla)
base_cols <- brewer.pal(min(n_phyla, 12), "Paired")
phylum_cols_vec <- colorRampPalette(base_cols)(n_phyla)
phylum_colors <- setNames(phylum_cols_vec, all_phyla)

plot_site <- function(site_name) {
  ggplot(
    species_df %>% filter(Sample_Site == site_name),
    aes(x = Day_Order, y = mean_rel_abundance, fill = factor(Species, levels = all_phyla))
  ) +
    geom_col(position = "stack") +
    scale_fill_manual(values = phylum_colors, drop = FALSE) +
    labs(
      title = paste("Sample-Site:", site_name),
      x = "Timepoint",
      y = "Average Relative abundance",
      fill = "Species"
    ) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
 

p_control  <- plot_site("Control")
p_surgical <- plot_site("Surgical")
  
print(p_control)
print(p_surgical)

```

 
```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
sd <- as(sample_data(ps_rel_2), "data.frame")

sd$Time_Group <- ifelse(
  sd$Day_Order %in% c("Admission", "Day 0a - Pre Surgery"),
  "Before",
  "After"
)

sd$Time_Group <- factor(sd$Time_Group, levels = c("Before", "After"))
sample_data(ps_rel_2) <- sample_data(sd)

# get mean relative abundance for each phylum
phy_phyla_df <- psmelt(ps_rel_2) %>%
  filter(Sample_Location == "Stifle") %>%
  filter(!is.na(Phylum)) %>%
  group_by(Phylum) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rel_abundance))
 

surg_df_phyla <- psmelt(ps_rel_2)%>%
  filter(
    Sample_Location == "Stifle",
    #Sample_Site == "Surgical",
    Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery", "Day 1"),
    !is.na(Phylum)
  ) %>%
  group_by(Dog_ID, Time_Group, Sample_Site, Phylum) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

results_phyla <- surg_df_phyla %>%
  tidyr::pivot_wider(
    id_cols = c(Dog_ID, Phylum, Sample_Site),
    names_from = Time_Group,
    values_from = mean_rel_abundance
  ) %>%
  group_by(Sample_Site) %>%
  filter((Before > 0) | (After > 0))  %>%
  summarise(
    n_pairs = n(),
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    estimate = test[[1]]$estimate,
    conf_low = test[[1]]$conf.int[1],
    conf_high = test[[1]]$conf.int[2],
    .groups = "drop"
  ) %>%
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)

print(results_phyla)

# Post-Hoc 
results_phyla <- surg_df_phyla %>%
  filter(mean_rel_abundance > 0.0001) %>% 
  tidyr::pivot_wider(
    id_cols = c(Dog_ID, Phylum, Sample_Site),
    names_from = Time_Group,
    values_from = mean_rel_abundance
  ) %>%
  group_by(Sample_Site) %>%
  filter(!is.na(Before), !is.na(After)) %>%
  summarise(
    n_pairs = n(),
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)

print(results_phyla)
 
 
## Genus 
 
surg_df_phyla <- psmelt(ps_rel_2)%>%
  filter(
    Sample_Location == "Stifle",
    #Sample_Site == "Surgical",
    Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery", "Day 1"),
    !is.na(Genus)
  ) %>%
  group_by(Dog_ID, Time_Group, Sample_Site, Genus) %>%
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")


results_phyla <- surg_df_phyla %>%
  tidyr::pivot_wider(
    id_cols = c(Dog_ID, Genus, Sample_Site),
    names_from = Time_Group,
    values_from = mean_rel_abundance
  ) %>%
  group_by(Sample_Site) %>%
  filter((Before > 0.001) | (After > 0.001))  %>%
  summarise(
    n_pairs = n(),
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)

print(results_phyla)

results_phyla <- surg_df_phyla %>%
  tidyr::pivot_wider(
    id_cols = c(Dog_ID, Genus),
    names_from = Time_Group,
    values_from = mean_rel_abundance
  ) %>%
  group_by(Genus) %>%
  filter((Before > 0), (After > 0)) %>% 
  summarise(
    n_pairs = n(),
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  arrange(p_adj)

print(results_phyla)


## Species 

surg_df_species <- psmelt(ps_rel_2) %>% 
  filter(
 #  Sample_Location == "Stifle", 
    Sample_Site == "Surgical", 
    Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery", "Day 1"), 
    !is.na(Species) 
  ) %>%
  group_by(Dog_ID, Time_Group, Genus, Species) %>% 
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

results_phyla <- surg_df_species %>% 
  pivot_wider(
    id_cols = c(Dog_ID, Genus, Species), 
    names_from = Time_Group, 
    values_from = mean_rel_abundance
  ) %>% 
  filter(!is.na(Before), !is.na(After)) %>%
  filter((Before > 0.001) | (After > 0.001))  %>%
  # 4. Group strictly by the taxonomic level you are testing
  group_by(Genus, Species) %>% 
  
  # 5. Run the statistical analysis safely
  summarise(
    n_pairs = n(),
    # Explicitly calculate the test statistics
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>% 
  
  # 6. Apply Benjamini-Hochberg false discovery rate correction
  mutate(p_adj = p.adjust(p, method = "BH")) %>% 
  arrange(p)

print(results_phyla)

```

 
```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
library(lefser)
library(SummarizedExperiment) 
ps_rel_2_surg <- ps_rel_2 %>%
  subset_samples(
    Sample_Site == "Surgical")%>% 
  subset_samples(Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery")) %>% 
  subset_taxa(!is.na(Species)) 

# Total ASVs across surgical samples from pre- and post-surgery timepoints 
phyloseq::ntaxa(ps_rel_2_surg)
 


sd <- as(sample_data(ps_rel_2_surg), "data.frame")

sd$Time_Group <- ifelse(
  sd$Day_Order %in% c("Admission", "Day 0a - Pre Surgery"),
  "Before",
  "After"
)
sd$Dog_ID <- factor(sd$Dog_ID)
sd$Time_Group <- factor(sd$Time_Group, levels = c("Before", "After"))
sample_data(ps_rel_2_surg) <- sample_data(sd)



keep <- phyloseq::filter_taxa(
  ps_rel_2_surg,
  function(x) {
    sd <- phyloseq::sample_data(ps_rel_2_surg_filt)
    adm  <- sd$Day_Order == "Admission"
    pre  <- sd$Day_Order == "Day 0a - Pre Surgery"
    post <- sd$Day_Order == "Day 0c - Post Surgery"

    # Require presence in all three timepoints
    present_all <- any(x[adm] > 0) && any(x[pre] > 0) && any(x[post] > 0)

    prevalence_all <- mean(x > 0) >= 0.05

    present_all && prevalence_all
  },
  prune = FALSE
)

ps_rel_2_surg_filt <- phyloseq::prune_taxa(keep, ps_rel_2_surg)
phyloseq::ntaxa(ps_rel_2_surg_filt)


ps_lefse <- ps_rel_2_surg_filt  
 

ps_lefse <- tax_glom(ps_lefse, taxrank = "Species", NArm = TRUE)
 

ps_surgical <- subset_samples(ps_lefse, Sample_Site == "Surgical")
ps_surgical <- prune_taxa(taxa_sums(ps_surgical) > 0, ps_surgical)
 
keep_dogs <- sample_data(ps_surgical) |>
  data.frame() |>
  dplyr::group_by(Dog_ID) |>
  dplyr::filter(all(c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery") %in% Day_Order)) |>
  dplyr::distinct(Dog_ID) |>
  dplyr::pull(Dog_ID)

ps_surgical <- prune_samples(sample_data(ps_surgical)$Dog_ID %in% keep_dogs, ps_surgical)

## Split data tables
counts <- as(otu_table(ps_surgical), "matrix")
if (!taxa_are_rows(otu_table(ps_surgical))) {
  counts <- t(counts)
}
coldata <- as(sample_data(ps_surgical), "data.frame")


## Create a SummarizedExperiment object
se <- SummarizedExperiment(assays = list(counts = counts), colData = coldata)

 
table(colData(se)$Time_Group, colData(se)$Dog_ID)

set.seed(1234)
res <- lefser(
  se,
  kruskal.threshold = .1, wilcox.threshold = .1,
  classCol = "Time_Group" 
) 

lefserPlot(res)
```
 
 
 
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE} 

# Differential Abundance
# get mean relative abundance for each phylum
  

species_df_summary <- species_df %>% 
  mutate(genus_species = paste(Genus, Species, sep = " ")) %>%
  group_by(Day_Order, Genus, Species, genus_species) %>% 
  summarise(
    mean_rel_abundance = mean(mean_rel_abundance, na.rm = TRUE),
    .groups = "drop"
  )

species_list <- unique(species_df_summary$genus_species)

# Colors for top 10 + Other
all_species <- c(species_list)
n_taxa <- length(all_species)
base_cols <- brewer.pal(min(n_taxa, 12), "Paired")
taxa_cols_vec <- colorRampPalette(base_cols)(n_taxa)
taxa_colors <- setNames(taxa_cols_vec, all_species)

species_df_summary %>%
  mutate(genus_species = factor(genus_species, levels = all_species)) %>%
  ggplot(aes(x = Day_Order, y = mean_rel_abundance, fill = genus_species)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = taxa_colors, drop = FALSE) +
  labs(
    title = paste("Taxa in Pre-Survery vs. Post-Surgery Samples"),
    x = "Timepoint",
    y = "Average Relative abundance",
    fill = "Species"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
 
 
 

```
 

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
 ## Species 

ctrl_df_species <- psmelt(ps_rel_2) %>% 
  filter(
 #  Sample_Location == "Stifle", 
    Sample_Site == "Control", 
    Day_Order %in% c("Admission", "Day 0a - Pre Surgery", "Day 0c - Post Surgery", "Day 1"), 
    !is.na(Species) 
  ) %>%
  group_by(Dog_ID, Time_Group, Genus, Species) %>% 
  summarise(mean_rel_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

 

results_phyla <- ctrl_df_species %>% 
  pivot_wider(
    id_cols = c(Dog_ID, Genus, Species), 
    names_from = Time_Group, 
    values_from = mean_rel_abundance
  ) %>% 
  filter(!is.na(Before), !is.na(After)) %>%
  filter((Before > 0.001) | (After > 0.001))  %>%
  # 4. Group strictly by the taxonomic level you are testing
  group_by(Genus, Species) %>% 
  
  # 5. Run the statistical analysis safely
  summarise(
    n_pairs = n(),
    # Explicitly calculate the test statistics
    stat = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$statistic,
    p = wilcox.test(Before, After, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>% 
  
  # 6. Apply Benjamini-Hochberg false discovery rate correction
  mutate(p_adj = p.adjust(p, method = "BH")) %>% 
  arrange(p)

print(results_phyla)
```
 
  
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE} 
results_species_annotated_notsig <- results_species_annotated %>%
  filter(results_species_annotated$p > 0.05)

results_species_annotated_notsig
shared_seqs <- intersect(results_species_annotated_sig$OTU, asvs_negative_samples$ASV)

subset_not_in_negatives <- results_species_annotated_sig[
  !results_species_annotated_sig$OTU %in% asvs_negative_samples$ASV,
]
subset_not_in_negatives


 

species_df_formatted_sub <- species_df_formatted %>%
  filter(OTU == "ASV_114")
species_long <- species_df_formatted_sub %>%
  pivot_longer(
    cols = c(Post_Surgery, Maturation II),
    names_to = "Timepoint",
    values_to = "RelAbundance"
  )
ggplot(species_long,
       aes(x = RelAbundance + 1e-6, fill = Timepoint)) +
  ggtitle("Rel Abundance of Signficant ASV_114 (p.adj <0.1)") +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  coord_cartesian(ylim = c(0, 2.5)) +
  theme_bw() +
  labs(x = "Relative abundance (log10 scale)",
       y = "Density")


species_df_formatted_sub <- species_df_formatted %>%
  filter(OTU == "ASV_1488")
species_long <- species_df_formatted_sub %>%
  pivot_longer(
    cols = c(Post_Surgery, Maturation),
    names_to = "Timepoint",
    values_to = "RelAbundance"
  )
ggplot(species_long,
       aes(x = RelAbundance + 1e-6, fill = Timepoint)) +
  ggtitle("Rel Abundance of Non-signficant ASV_1488 (p.adj > 0.1)") +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  coord_cartesian(ylim = c(0, 2.5)) +
  theme_bw() +
  labs(x = "Relative abundance (log10 scale)",
       y = "Density")



 

phylum_colors <- c(
  "Bacillota" = "orange",
  "Pseudomonadota" = "#e7298a",
  "Actinomycetota" = "#1f78b4",
  "Bacteroidota" = "#1b9e77",
  "Cyanobacteriota" = "#a6761d",
  "Thermodesulfobacteriota" = "red",
  "Fusobacteriota" = "yellow",
  "Campylobacterota" = "#66a61e",
  "Acidobacteriota" = "green",
  "Chloroflexota" = "pink",
  "Other" = "gray"
)

phylum_counts <- results_species_annotated_sig %>%
  mutate(Phylum = ifelse(is.na(Phylum) | Phylum == "", "Other", Phylum),
         Phylum = ifelse(Phylum %in% names(phylum_colors), Phylum, "Other")) %>%
  count(Phylum, name = "Count") %>%
  mutate(Phylum = factor(Phylum, levels = Phylum[order(Count, decreasing = TRUE)]))

ggplot(phylum_counts, aes(x = Phylum, y = Count, fill = Phylum)) +
  ggtitle("Phyla of the 45 ASVs found to have significantly higher relative abundances (p_adj =<= 0.1)") +
  geom_col() +
  scale_fill_manual(values = phylum_colors, drop = FALSE) +
  labs(x = "Phylum", y = "Count") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```
 



 