---
title: "Canine Skin Microbiome: Curate Read and ASV Samples "
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

**Purpose:**
  Analyze read count and ASVs across samples, specifically in comparison to controls (Chlor and Kit) with "zero" microbial signatures. 
  
**Overview**
 1. Curation of reads and ASVs https://benjjneb.github.io/FelineBileManuscript/Analyses/FelineBile_Curate.html  2. Perform decontam diagnostics to investigate core microbiome. 
https://benjjneb.github.io/FelineBileManuscript/Analyses/FelineBile_Core.html  

# Load R Packages
```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(decontam); packageVersion("decontam")
library(ggplot2); packageVersion("ggplot2")
library(data.table)
library(dplyr)
```


# Input files 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
ASV_feature_table <- fread("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/input_csv_files/canine_skin_microbiome_filtered_nochim_asvs_names.csv", header=TRUE)

sample_metadata <- read.delim("/Users/spencertoth/Desktop/Research/canine_skin_project/R_visualizations/input_csv_files/metadata/CH-All_16S_Metadata_20220124.tsv", header=TRUE)

rownames(ASV_feature_table) <- ASV_feature_table$V1
rownames(sample_metadata) <- sample_metadata$SampleID

# Confirm identical samples 
identical(sort(rownames(sample_metadata)), sort(rownames(ASV_feature_table)))

table(sample_metadata$Sample_Site, useNA="ifany")
```
# Curate Data
## Define a generic column with TRUE for all negative controls, and FALSE for real (positive) samples
```{r, message=FALSE, warning=FALSE, echo=TRUE}
sample_metadata$NegativeControl <- grepl("Chlorhex|Air|Kit", sample_metadata$Sample_Site)
table(sample_metadata$Sample_Site, sample_metadata$NegativeControl)
```
## Inspect the DNA concentrations.
```{r, message=FALSE, warning=FALSE, echo=TRUE}
class(sample_metadata$DNA_Conc)
summary(sample_metadata$DNA_Conc)

table(sample_metadata$Sample_Site, sample_metadata$DNA_Conc>3.635, useNA="ifany")
```
# Curate Read Samples
### Check the distribution of library sizes (read depths) for samples  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
if(!identical(sort(rownames(sample_metadata)), sort(rownames(ASV_feature_table)))) stop("Mismatched samples in df and st.")
ASV_feature_table_subset1 <- ASV_feature_table %>% 
  filter(V1 %in% rownames(sample_metadata))
sample_metadata$LibrarySize <- rowSums(ASV_feature_table_subset1[, -1])
ggplot(data=sample_metadata, aes(x=rank(LibrarySize), y=LibrarySize, color=Sample_Site)) +   geom_point(size = 1) +  # Standard scatter plot points
  scale_color_manual(values = c(
    "Air"       = "red",
    "Chlorhex"  = "red",
    "Kit"       = "red",
    "Surgical"  = "blue",
    "Control"   = "blue"
  ))

sort(sample_metadata$LibrarySize)[1:50]
table(sample_metadata$Sample_Site, sample_metadata$NegativeControl)
```


### Remove the low-depth samples  
```{r, message=FALSE, warning=FALSE, echo=TRUE}

sam <- rownames(ASV_feature_table_subset1)[sample_metadata$LibrarySize > 9000]

sample_metadata_subset <- sample_metadata[sam,]

ASV_feature_table_subset <- ASV_feature_table_subset1 %>% 
  filter(V1 %in% sam)

cat("original")
table(sample_metadata$Sample_Site, sample_metadata$NegativeControl)
cat("\nRemoved samples < 9000 reads")
table(sample_metadata_subset$Sample_Site, sample_metadata_subset$NegativeControl)
```